---
title: "Module 6: Organizing Tabular Data with tidyr"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
---

## Welcome to Data Reshaping and Tidying! 📊🔄

Data rarely comes in the format you need for analysis. Sometimes it's too wide, sometimes too long, sometimes messy with multiple values in single cells. The `tidyr` package provides powerful tools to reshape and tidy your data, transforming it into the clean, analysis-ready format that other tidyverse packages expect.

::: {.callout-note title="Shape follows purpose"}
“Wide” and “long” are not measures of data quality. Long data often suits analysis and visualisation; wide data can suit a management report or modelling matrix. First define the observational unit and intended consumer, then choose a shape. Always retain stable identifiers so the transformation can be validated or reversed.
:::

### A safe reshaping workflow

1. State what one row represents before and after the transformation.
2. Check that identifier columns uniquely locate each observation.
3. Reshape with explicit `cols`, `names_to`, and `values_to` arguments.
4. Compare row counts, totals, missingness, and duplicates.
5. Keep analysis-ready data separate from presentation-ready tables.

## What You'll Learn

In this module, we'll master the art of data reshaping and tidying with tidyr:

### 1. Tidy Data Principles
- Understanding the three fundamental rules of tidy data
- Identifying common "messy" data patterns
- Diagnosing data structure problems
- Planning transformation strategies

### 2. Pivoting Data
- `pivot_longer()`: Converting wide data to long format
- `pivot_wider()`: Converting long data to wide format
- Handling multiple value columns and complex pivots
- Real-world reshaping scenarios

### 3. Separating and Uniting
- `separate_wider_delim()` and related functions: splitting one column into multiple columns
- `unite()`: Combining multiple columns into one
- Working with delimiters and patterns
- Handling missing values in separation

### 4. Nested Data and Rectangling
- Working with list-columns and nested data
- `unnest()`: Expanding nested data structures
- `hoist()`: Extracting specific elements from lists
- Rectangling JSON-like data

### 5. Advanced Tidying Techniques
- Handling multiple types of messiness simultaneously
- Custom separation and parsing functions
- Performance considerations for large datasets
- Integration with other tidyverse packages

## Learning Objectives

By the end of this module, you will be able to:

- ✅ Identify and apply tidy data principles to any dataset
- ✅ Reshape data between wide and long formats efficiently
- ✅ Split and combine columns using various delimiters
- ✅ Work with complex nested data structures
- ✅ Handle real-world messy data transformation challenges
- ✅ Choose the right tidyr function for any data structure problem
- ✅ Build data cleaning pipelines that integrate with dplyr workflows

## Why Master tidyr?

Data reshaping is essential because:

1. **Analysis requirements**: Different analyses need different data shapes
2. **Visualization needs**: ggplot2 works best with long-format data
3. **Modeling prep**: Machine learning models need specific data structures
4. **Real-world messiness**: Raw data is rarely in the format you need
5. **Reproducible workflows**: Systematic tidying ensures consistent results

## The Tidy Data Philosophy

```{r}
#| label: tidy-philosophy
#| message: false

library(tidyverse)

cat("The three fundamental rules of tidy data:\n")
cat("1. Each variable forms a column\n")
cat("2. Each observation forms a row\n")
cat("3. Each type of observational unit forms a table\n\n")

cat("Common tidyr functions:\n")
cat("- pivot_longer(): Wide to long (gather)\n")
cat("- pivot_wider(): Long to wide (spread)\n")
cat("- separate(): Split one column into many\n")
cat("- unite(): Combine many columns into one\n")
cat("- unnest(): Expand list-columns\n")
cat("- fill(): Fill missing values\n")
```

## Real-World Messy Data Examples

Let's start with realistic examples of the messy data you'll encounter and need to clean:

```{r}
#| label: messy-data-examples

# Example 1: Wide sales data (common in spreadsheets)
sales_wide <- tibble(
  sales_rep = c("Alice", "Bob", "Charlie", "Diana"),
  region = c("North", "South", "East", "West"),
  Q1_2023 = c(45000, 52000, 48000, 51000),
  Q2_2023 = c(47000, 54000, 50000, 53000),
  Q3_2023 = c(49000, 56000, 52000, 55000),
  Q4_2023 = c(51000, 58000, 54000, 57000),
  Q1_2024 = c(53000, 60000, 56000, 59000),
  Q2_2024 = c(55000, 62000, 58000, 61000)
)

cat("Example 1: Wide sales data (quarters as columns)\n")
print(sales_wide)

# Example 2: Multiple variables in column names
survey_wide <- tibble(
  respondent_id = 1:4,
  age = c(25, 34, 28, 45),
  satisfaction_pre = c(3, 4, 2, 5),
  satisfaction_post = c(4, 5, 4, 5),
  recommendation_pre = c(2, 3, 2, 4),
  recommendation_post = c(4, 4, 3, 5)
)

cat("\nExample 2: Pre/post measurements in column names\n")
print(survey_wide)

# Example 3: Multiple values in single cells
contact_messy <- tibble(
  id = 1:4,
  name = c("John Doe", "Jane Smith", "Bob Johnson", "Alice Brown"),
  contact = c("john@email.com; 555-1234",
              "jane@company.org; 555-5678; LinkedIn: jsmith",
              "bob.j@domain.com",
              "alice@startup.io; 555-9999; Twitter: @alice_b"),
  skills = c("R, Python, SQL", "Excel, Tableau, SQL", "Python, Java", "R, Statistics, Machine Learning")
)

cat("\nExample 3: Multiple values in single cells\n")
print(contact_messy)

# Example 4: Missing structure (common in real data)
sales_messy <- tibble(
  entry = 1:8,
  data = c("Product: Widget A", "Q1: 1000", "Q2: 1200", "Q3: 1100",
           "Product: Widget B", "Q1: 800", "Q2: 950", "Q3: 900")
)

cat("\nExample 4: Unstructured data that needs parsing\n")
print(sales_messy)
```

These examples represent real challenges you'll face. By the end of this module, you'll be able to transform all of these into clean, tidy formats!

## The Power of Tidy Data

Let's see the dramatic difference between messy and tidy data for analysis:

```{r}
#| label: tidy-transformation-preview

# Transform the wide sales data to tidy format
sales_tidy <- sales_wide %>%
  pivot_longer(
    cols = starts_with("Q"),
    names_to = "quarter_year",
    values_to = "sales"
  ) %>%
  separate(quarter_year, into = c("quarter", "year"), sep = "_") %>%
  mutate(
    year = as.numeric(year),
    date = ymd(paste(year, case_when(
      quarter == "Q1" ~ "03-31",
      quarter == "Q2" ~ "06-30",
      quarter == "Q3" ~ "09-30",
      quarter == "Q4" ~ "12-31"
    )))
  )

cat("Transformed to tidy format:\n")
print(head(sales_tidy, 10))

# Now analysis becomes straightforward
cat("\nWith tidy data, analysis is simple:\n")

# Growth analysis
growth_analysis <- sales_tidy %>%
  arrange(sales_rep, date) %>%
  group_by(sales_rep) %>%
  mutate(
    previous_sales = lag(sales),
    growth_rate = (sales - previous_sales) / previous_sales * 100
  ) %>%
  filter(!is.na(growth_rate))

cat("Average quarterly growth by rep:\n")
growth_analysis %>%
  group_by(sales_rep) %>%
  summarise(avg_growth = round(mean(growth_rate, na.rm = TRUE), 1), .groups = "drop") %>%
  arrange(desc(avg_growth)) %>%
  print()

# Regional performance
cat("\nBest performing region by year:\n")
sales_tidy %>%
  group_by(region, year) %>%
  summarise(total_sales = sum(sales), .groups = "drop") %>%
  group_by(year) %>%
  slice_max(total_sales, n = 1) %>%
  print()
```

This transformation from wide to long format makes complex analyses straightforward!

## Module Structure

This module is organized into four comprehensive sections:

1. **[Tidy Data Principles](tidy-principles.qmd)**: Master the fundamentals of tidy data
2. **[Pivoting Data](pivoting-data.qmd)**: Reshape between wide and long formats
3. **[Separating and Uniting](separating-uniting.qmd)**: Split and combine columns
4. **[Nested Data and Rectangling](nested-rectangling.qmd)**: Work with complex structures

Each section includes:
- Core concepts with clear explanations
- Before-and-after transformation examples
- Real-world data scenarios and solutions
- Integration with dplyr for complete workflows
- Performance tips and best practices

## Common Data Shapes and When to Use Them

Understanding when to use different data shapes is crucial:

```{r}
#| label: data-shapes

# Wide format: Good for human reading, reporting
cat("WIDE FORMAT - Good for:\n")
cat("- Spreadsheet-style reports\n")
cat("- Side-by-side comparisons\n")
cat("- Correlation matrices\n")
cat("- Human-readable summaries\n\n")

# Long format: Good for analysis, visualization
cat("LONG FORMAT - Good for:\n")
cat("- ggplot2 visualizations\n")
cat("- Statistical modeling\n")
cat("- Group-by operations\n")
cat("- Time series analysis\n\n")

# Demonstrate with a simple example
demo_wide <- tibble(
  person = c("Alice", "Bob"),
  height = c(165, 180),
  weight = c(60, 75)
)

demo_long <- demo_wide %>%
  pivot_longer(cols = c(height, weight), names_to = "measurement", values_to = "value")

cat("Wide format (2 rows, 3 columns):\n")
print(demo_wide)

cat("\nLong format (4 rows, 3 columns):\n")
print(demo_long)

cat("\nUse wide for comparison, long for analysis!\n")
```

## Integration with the Tidyverse Ecosystem

tidyr works seamlessly with other tidyverse packages:

```{r}
#| label: tidyverse-integration

# Complete data cleaning pipeline
messy_sales <- tibble(
  rep_region = c("Alice_North", "Bob_South", "Charlie_East"),
  jan_feb = c("1000,1200", "800,900", "1100,1050"),
  mar_apr = c("1300,1400", "950,1000", "1200,1250"),
  contact = c("alice@co.com", "bob@co.com", "charlie@co.com")
)

cat("Starting with messy data:\n")
print(messy_sales)

# Complete tidying pipeline
clean_sales <- messy_sales %>%
  # Step 1: Separate rep and region
  separate(rep_region, into = c("sales_rep", "region"), sep = "_") %>%

  # Step 2: Pivot longer to get month pairs in rows
  pivot_longer(
    cols = c(jan_feb, mar_apr),
    names_to = "month_pair",
    values_to = "sales_pair"
  ) %>%

  # Step 3: Separate the sales pairs
  separate(sales_pair, into = c("month1_sales", "month2_sales"), sep = ",") %>%

  # Step 4: Convert to proper structure
  mutate(
    month1_sales = as.numeric(month1_sales),
    month2_sales = as.numeric(month2_sales)
  ) %>%

  # Step 5: Pivot longer again to get one row per month
  pivot_longer(
    cols = c(month1_sales, month2_sales),
    names_to = "month_type",
    values_to = "sales"
  ) %>%

  # Step 6: Clean up month names with dplyr
  mutate(
    month = case_when(
      month_pair == "jan_feb" & month_type == "month1_sales" ~ "January",
      month_pair == "jan_feb" & month_type == "month2_sales" ~ "February",
      month_pair == "mar_apr" & month_type == "month1_sales" ~ "March",
      month_pair == "mar_apr" & month_type == "month2_sales" ~ "April"
    )
  ) %>%

  # Step 7: Select final columns
  select(sales_rep, region, month, sales, contact) %>%
  arrange(sales_rep, match(month, c("January", "February", "March", "April")))

cat("\nAfter complete tidying pipeline:\n")
print(clean_sales)

cat("\nNow ready for analysis with dplyr, ggplot2, etc.!\n")
```

This shows the power of combining tidyr with dplyr for complete data transformation workflows.

## Prerequisites

Before starting this module, make sure you have:

- Completed Module 3 (Tidyverse Introduction) and Module 5 (Data Wrangling with dplyr)
- Understanding of the pipe operator and dplyr verbs
- Basic knowledge of data types and tibbles
- Familiarity with the concept of "tidy data"

## Getting Started

Ready to master data reshaping and tidying? Let's begin with the fundamental principles: **[Tidy Data Principles](tidy-principles.qmd)**

## Summary

Data tidying with tidyr is both an art and a science. The key principles to remember:

- **Think tidy first**: Always consider what the final tidy structure should look like
- **One step at a time**: Complex transformations are easier when broken into steps
- **Combine tools**: tidyr + dplyr creates powerful data cleaning pipelines
- **Document your logic**: Reshaping operations can be complex - comment your code
- **Test incrementally**: Verify each transformation step before moving to the next

tidyr transforms chaotic, real-world data into the clean, consistent format that enables powerful analysis. With these skills, no data structure challenge will be too complex to handle!

Let's begin your journey to data tidying mastery! 🧹📊
