---
title: "Nested Data and Rectangling: Advanced Data Structures"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Understanding Nested Data Structures 🗂️

In the real world, data often comes in complex, nested formats - think JSON from APIs, lists within data frames, or hierarchical structures from databases. tidyr's "rectangling" functions help you transform these complex structures into the rectangular, tidy format that R analysis tools expect.

## What is Nested Data?

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(jsonlite)

cat("NESTED DATA STRUCTURES - Common Forms:\n\n")

# Example 1: List-columns in data frames
survey_nested <- tibble(
  survey_id = 1:4,
  participant = c("Alice", "Bob", "Carol", "David"),
  responses = list(
    c(5, 4, 3, 5, 4),
    c(3, 3, 4, 2, 5),
    c(4, 5, 5, 4, 4),
    c(2, 3, 3, 3, 2)
  ),
  demographics = list(
    list(age = 25, education = "Bachelor", income = 50000),
    list(age = 34, education = "Master", income = 75000),
    list(age = 28, education = "PhD", income = 85000),
    list(age = 42, education = "Bachelor", income = 60000)
  )
)

cat("Example 1: Survey data with list-columns\n")
print(survey_nested)
cat("Column types:", sapply(survey_nested, class), "\n")

# Example 2: JSON-like nested structure
api_response <- tibble(
  endpoint = c("/users", "/products", "/orders"),
  data = list(
    list(
      users = list(
        list(id = 1, name = "John", email = "john@example.com", purchases = c(100, 250, 150)),
        list(id = 2, name = "Jane", email = "jane@example.com", purchases = c(300, 180))
      )
    ),
    list(
      products = list(
        list(id = "A1", name = "Widget", price = 25.99, categories = c("electronics", "gadgets")),
        list(id = "B2", name = "Gadget", price = 45.50, categories = c("electronics"))
      )
    ),
    list(
      orders = list(
        list(order_id = 1001, customer_id = 1, items = c("A1", "B2"), total = 71.49),
        list(order_id = 1002, customer_id = 2, items = c("A1"), total = 25.99)
      )
    )
  )
)

cat("\nExample 2: API response with deeply nested structure\n")
str(api_response, max.level = 2)
```

## `unnest()`: Expanding List-Columns

### Basic Unnesting Operations

```{r}
#| label: unnest-basics

# Unnest simple vector lists
responses_unnested <- survey_nested %>%
  select(survey_id, participant, responses) %>%
  unnest(responses) %>%
  group_by(survey_id, participant) %>%
  mutate(question = paste0("Q", row_number())) %>%
  ungroup()

cat("Unnested survey responses:\n")
print(responses_unnested)

# Create analysis-ready format
responses_wide <- responses_unnested %>%
  pivot_wider(names_from = question, values_from = responses)

cat("\nAnalysis-ready wide format:\n")
print(responses_wide)

# Unnest named lists (demographics)
demographics_unnested <- survey_nested %>%
  select(survey_id, participant, demographics) %>%
  unnest_wider(demographics)

cat("\nUnnested demographics (wide format):\n")
print(demographics_unnested)
```

### Complex Unnesting Scenarios

```{r}
#| label: complex-unnesting

# E-commerce order data with multiple levels
ecommerce_data <- tibble(
  order_id = 1:3,
  customer = c("Alice Johnson", "Bob Smith", "Carol Davis"),
  order_date = as.Date(c("2023-01-15", "2023-01-18", "2023-01-22")),
  items = list(
    list(
      list(product_id = "WIDGET_A", name = "Premium Widget", quantity = 2, unit_price = 25.99),
      list(product_id = "GADGET_B", name = "Smart Gadget", quantity = 1, unit_price = 45.50)
    ),
    list(
      list(product_id = "WIDGET_A", name = "Premium Widget", quantity = 3, unit_price = 25.99),
      list(product_id = "DEVICE_C", name = "IoT Device", quantity = 1, unit_price = 89.99)
    ),
    list(
      list(product_id = "GADGET_B", name = "Smart Gadget", quantity = 2, unit_price = 45.50)
    )
  ),
  shipping = list(
    list(method = "Standard", cost = 5.99, estimated_days = 5),
    list(method = "Express", cost = 12.99, estimated_days = 2),
    list(method = "Standard", cost = 5.99, estimated_days = 5)
  )
)

cat("Complex e-commerce nested data:\n")
str(ecommerce_data, max.level = 2)

# Unnest items to get order details
order_items <- ecommerce_data %>%
  select(order_id, customer, order_date, items) %>%
  unnest(items) %>%
  unnest_wider(items) %>%
  mutate(
    line_total = quantity * unit_price,
    item_number = row_number()
  )

cat("\nOrder items unnested:\n")
print(order_items)

# Unnest shipping information
shipping_info <- ecommerce_data %>%
  select(order_id, shipping) %>%
  unnest_wider(shipping)

cat("\nShipping information unnested:\n")
print(shipping_info)

# Combine for complete order analysis
complete_orders <- order_items %>%
  group_by(order_id, customer, order_date) %>%
  summarise(
    total_items = sum(quantity),
    order_total = sum(line_total),
    unique_products = n(),
    .groups = "drop"
  ) %>%
  left_join(shipping_info, by = "order_id") %>%
  mutate(
    grand_total = order_total + cost,
    total_delivery_days = estimated_days
  )

cat("\nComplete order analysis:\n")
print(complete_orders)
```

### Working with Varied List Structures

```{r}
#| label: varied-structures

# Customer data with inconsistent nested structures
customer_profiles <- tibble(
  customer_id = 1:5,
  profile = list(
    list(name = "Alice", age = 25, preferences = c("electronics", "books"),
         contact = list(email = "alice@email.com", phone = "555-0123")),
    list(name = "Bob", age = 34, preferences = c("sports", "electronics"),
         contact = list(email = "bob@email.com")),  # No phone
    list(name = "Carol", age = 28, preferences = c("fashion", "beauty", "books"),
         contact = list(email = "carol@email.com", phone = "555-0156", address = "123 Main St")),
    list(name = "David", preferences = c("music"),  # No age
         contact = list(phone = "555-0189")),       # No email
    list(name = "Emma", age = 31, preferences = c("travel", "photography"))  # No contact
  )
)

cat("Customer profiles with inconsistent structures:\n")
str(customer_profiles, max.level = 2)

# Safe unnesting with error handling
basic_info <- customer_profiles %>%
  unnest_wider(profile) %>%
  select(customer_id, name, age) %>%
  mutate(age = as.numeric(age))  # Handle missing ages

cat("\nBasic customer information:\n")
print(basic_info)

# Handle preferences (vector lists)
preferences_unnested <- customer_profiles %>%
  unnest_wider(profile) %>%
  select(customer_id, name, preferences) %>%
  unnest(preferences)

cat("\nCustomer preferences (long format):\n")
print(preferences_unnested)

# Handle contact information (nested lists with varying structures)
contact_info <- customer_profiles %>%
  unnest_wider(profile) %>%
  select(customer_id, name, contact) %>%
  # Handle missing contact info
  filter(!is.null(contact)) %>%
  unnest_wider(contact)

cat("\nContact information:\n")
print(contact_info)

# Comprehensive customer analysis
customer_summary <- basic_info %>%
  left_join(
    preferences_unnested %>%
      group_by(customer_id) %>%
      summarise(
        preference_count = n(),
        top_preferences = paste(preferences, collapse = ", "),
        .groups = "drop"
      ),
    by = "customer_id"
  ) %>%
  left_join(
    contact_info %>%
      select(customer_id, email, phone),
    by = "customer_id"
  ) %>%
  mutate(
    age_group = case_when(
      age < 25 ~ "Under 25",
      age < 35 ~ "25-34",
      age < 45 ~ "35-44",
      TRUE ~ "45+"
    ),
    contact_completeness = case_when(
      !is.na(email) & !is.na(phone) ~ "Complete",
      !is.na(email) | !is.na(phone) ~ "Partial",
      TRUE ~ "Missing"
    )
  )

cat("\nComprehensive customer summary:\n")
print(customer_summary)
```

## `hoist()`: Extracting Specific Elements

### Selective Extraction from Nested Lists

```{r}
#| label: hoist-basics

# Product catalog with nested specifications
product_catalog <- tibble(
  product_id = c("LAPTOP_001", "PHONE_002", "TABLET_003"),
  category = c("Electronics", "Electronics", "Electronics"),
  specifications = list(
    list(
      brand = "TechCorp",
      model = "UltraBook Pro",
      specs = list(cpu = "Intel i7", ram = "16GB", storage = "512GB SSD"),
      features = c("Touchscreen", "Backlit Keyboard", "Fingerprint Reader"),
      price = list(base = 1299.99, sale = 1199.99, currency = "USD")
    ),
    list(
      brand = "MobileTech",
      model = "SmartPhone X",
      specs = list(cpu = "Snapdragon 888", ram = "8GB", storage = "128GB"),
      features = c("5G", "Wireless Charging", "Water Resistant"),
      price = list(base = 899.99, sale = 799.99, currency = "USD")
    ),
    list(
      brand = "TabletCorp",
      model = "Tablet Pro",
      specs = list(cpu = "A14 Bionic", ram = "6GB", storage = "256GB"),
      features = c("Apple Pencil Support", "Face ID", "Retina Display"),
      price = list(base = 649.99, currency = "USD")  # No sale price
    )
  )
)

cat("Product catalog with complex nested specifications:\n")
str(product_catalog, max.level = 2)

# Use hoist() to extract specific elements
products_hoisted <- product_catalog %>%
  hoist(specifications,
    brand = "brand",
    model = "model",
    base_price = c("price", "base"),
    sale_price = c("price", "sale"),
    cpu = c("specs", "cpu"),
    ram = c("specs", "ram")
  ) %>%
  mutate(
    has_sale = !is.na(sale_price),
    discount_percent = round((base_price - sale_price) / base_price * 100, 1)
  )

cat("\nProducts with hoisted key information:\n")
print(products_hoisted)

# Extract features using unnest after hoist
product_features <- product_catalog %>%
  hoist(specifications, features = "features") %>%
  select(product_id, category, features) %>%
  unnest(features)

cat("\nProduct features extracted:\n")
print(product_features)
```

### Advanced Hoisting Techniques

```{r}
#| label: advanced-hoisting

# Financial data with nested time series and metadata
financial_data <- tibble(
  symbol = c("AAPL", "GOOGL", "MSFT"),
  data = list(
    list(
      company = "Apple Inc.",
      sector = "Technology",
      prices = list(
        "2023-01-01" = 150.25,
        "2023-01-02" = 152.10,
        "2023-01-03" = 149.80
      ),
      metrics = list(
        pe_ratio = 25.4,
        market_cap = "2.8T",
        dividend_yield = 0.52
      ),
      analysts = list(
        rating = "Buy",
        target_price = 180.00,
        consensus = list(buy = 15, hold = 3, sell = 1)
      )
    ),
    list(
      company = "Alphabet Inc.",
      sector = "Technology",
      prices = list(
        "2023-01-01" = 88.73,
        "2023-01-02" = 89.45,
        "2023-01-03" = 87.92
      ),
      metrics = list(
        pe_ratio = 18.7,
        market_cap = "1.1T",
        dividend_yield = 0.0
      ),
      analysts = list(
        rating = "Buy",
        target_price = 120.00,
        consensus = list(buy = 18, hold = 2, sell = 0)
      )
    ),
    list(
      company = "Microsoft Corporation",
      sector = "Technology",
      prices = list(
        "2023-01-01" = 240.22,
        "2023-01-02" = 242.15,
        "2023-01-03" = 238.90
      ),
      metrics = list(
        pe_ratio = 28.1,
        market_cap = "1.8T",
        dividend_yield = 0.73
      ),
      analysts = list(
        rating = "Strong Buy",
        target_price = 280.00,
        consensus = list(buy = 20, hold = 1, sell = 0)
      )
    )
  )
)

cat("Financial data with deeply nested structure:\n")
str(financial_data, max.level = 2)

# Extract key financial metrics
financial_metrics <- financial_data %>%
  hoist(data,
    company = "company",
    sector = "sector",
    pe_ratio = c("metrics", "pe_ratio"),
    market_cap = c("metrics", "market_cap"),
    dividend_yield = c("metrics", "dividend_yield"),
    analyst_rating = c("analysts", "rating"),
    target_price = c("analysts", "target_price"),
    buy_ratings = c("analysts", "consensus", "buy"),
    hold_ratings = c("analysts", "consensus", "hold")
  ) %>%
  mutate(
    total_ratings = buy_ratings + hold_ratings,
    buy_percentage = round(buy_ratings / total_ratings * 100, 1)
  )

cat("\nExtracted financial metrics:\n")
print(financial_metrics)

# Extract and process price data
price_data <- financial_data %>%
  hoist(data, prices = "prices") %>%
  select(symbol, prices) %>%
  unnest_longer(prices, indices_to = "date") %>%
  mutate(
    date = as.Date(date),
    price = as.numeric(prices)
  ) %>%
  select(-prices) %>%
  group_by(symbol) %>%
  mutate(
    price_change = price - lag(price),
    price_change_pct = round((price / lag(price) - 1) * 100, 2)
  ) %>%
  ungroup()

cat("\nProcessed price data:\n")
print(price_data)

# Combine for investment analysis
investment_analysis <- financial_metrics %>%
  left_join(
    price_data %>%
      group_by(symbol) %>%
      summarise(
        avg_price = round(mean(price), 2),
        price_volatility = round(sd(price_change, na.rm = TRUE), 2),
        latest_price = last(price),
        .groups = "drop"
      ),
    by = "symbol"
  ) %>%
  mutate(
    upside_potential = round((target_price - latest_price) / latest_price * 100, 1),
    investment_score = round((buy_percentage / 100) * (upside_potential / 10) + (dividend_yield * 10), 1)
  ) %>%
  arrange(desc(investment_score))

cat("\nInvestment analysis summary:\n")
print(investment_analysis)
```

## `unnest_longer()` and `unnest_wider()`: Specialized Unnesting

### Understanding the Difference

```{r}
#| label: unnest-comparison

# Sample data for comparison
sample_nested <- tibble(
  id = 1:3,
  measurements = list(
    list(height = 175, weight = 70, bmi = 22.9),
    list(height = 180, weight = 75, bmi = 23.1),
    list(height = 165, weight = 60, bmi = 22.0)
  ),
  tags = list(
    c("healthy", "active", "young"),
    c("athletic", "healthy"),
    c("young", "active", "healthy", "student")
  )
)

cat("Sample nested data:\n")
print(sample_nested)

# unnest_wider: Spreads list elements into columns
measurements_wide <- sample_nested %>%
  select(id, measurements) %>%
  unnest_wider(measurements)

cat("\nunnest_wider result (measurements as columns):\n")
print(measurements_wide)

# unnest_longer: Stacks list elements into rows
tags_long <- sample_nested %>%
  select(id, tags) %>%
  unnest_longer(tags)

cat("\nunnest_longer result (tags as rows):\n")
print(tags_long)

# Combined approach for comprehensive analysis
complete_analysis <- sample_nested %>%
  unnest_wider(measurements) %>%
  left_join(
    tags_long %>%
      group_by(id) %>%
      summarise(tag_list = paste(tags, collapse = ", "), .groups = "drop"),
    by = "id"
  )

cat("\nCombined analysis:\n")
print(complete_analysis)
```

### Real-World JSON Processing

```{r}
#| label: json-processing

# Simulate API response data (like from a social media platform)
social_media_data <- tibble(
  user_id = 1:4,
  profile = list(
    list(
      username = "tech_alice",
      followers = 1250,
      following = 340,
      posts = list(
        list(id = 1, text = "Great day at the conference!", likes = 45, comments = 8, hashtags = c("tech", "networking")),
        list(id = 2, text = "New project launching soon", likes = 32, comments = 12, hashtags = c("startup", "innovation"))
      ),
      interests = c("technology", "startups", "design")
    ),
    list(
      username = "fitness_bob",
      followers = 890,
      following = 180,
      posts = list(
        list(id = 3, text = "Morning workout complete!", likes = 67, comments = 4, hashtags = c("fitness", "morning")),
        list(id = 4, text = "Healthy meal prep tips", likes = 89, comments = 15, hashtags = c("nutrition", "health"))
      ),
      interests = c("fitness", "nutrition", "wellness")
    ),
    list(
      username = "travel_carol",
      followers = 2100,
      following = 450,
      posts = list(
        list(id = 5, text = "Amazing sunset in Bali", likes = 156, comments = 23, hashtags = c("travel", "sunset", "bali")),
        list(id = 6, text = "Travel tips for budget backpacking", likes = 98, comments = 31, hashtags = c("travel", "budget", "backpacking"))
      ),
      interests = c("travel", "photography", "culture")
    ),
    list(
      username = "foodie_david",
      followers = 670,
      following = 290,
      posts = list(
        list(id = 7, text = "Homemade pasta recipe", likes = 78, comments = 19, hashtags = c("cooking", "pasta", "homemade"))
      ),
      interests = c("cooking", "food", "recipes")
    )
  )
)

cat("Social media nested data structure:\n")
str(social_media_data, max.level = 2)

# Extract user profile information
user_profiles <- social_media_data %>%
  hoist(profile,
    username = "username",
    followers = "followers",
    following = "following",
    interests = "interests"
  ) %>%
  select(-profile) %>%
  unnest_longer(interests) %>%
  group_by(user_id, username, followers, following) %>%
  summarise(
    interest_count = n(),
    interest_list = paste(interests, collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(
    follower_ratio = round(followers / following, 2),
    popularity_tier = case_when(
      followers >= 2000 ~ "High",
      followers >= 1000 ~ "Medium",
      TRUE ~ "Low"
    )
  )

cat("\nUser profile analysis:\n")
print(user_profiles)

# Extract and analyze posts
posts_analysis <- social_media_data %>%
  hoist(profile, username = "username", posts = "posts") %>%
  select(user_id, username, posts) %>%
  unnest_longer(posts) %>%
  unnest_wider(posts) %>%
  unnest_longer(hashtags) %>%
  group_by(user_id, username, id, text, likes, comments) %>%
  summarise(
    hashtag_count = n(),
    hashtag_list = paste(hashtags, collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(
    engagement_rate = round((likes + comments) / likes * 100, 1),
    content_category = case_when(
      str_detect(text, "workout|fitness|health") ~ "Fitness",
      str_detect(text, "travel|trip|vacation") ~ "Travel",
      str_detect(text, "food|recipe|cooking") ~ "Food",
      str_detect(text, "tech|project|conference") ~ "Technology",
      TRUE ~ "Other"
    )
  )

cat("\nPost analysis:\n")
print(posts_analysis)

# Content performance by category
category_performance <- posts_analysis %>%
  group_by(content_category) %>%
  summarise(
    post_count = n(),
    avg_likes = round(mean(likes), 1),
    avg_comments = round(mean(comments), 1),
    avg_hashtags = round(mean(hashtag_count), 1),
    avg_engagement = round(mean(engagement_rate), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_engagement))

cat("\nContent performance by category:\n")
print(category_performance)
```

## Advanced Rectangling Patterns

### Handling Deeply Nested JSON-like Structures

```{r}
#| label: advanced-rectangling

# Complex enterprise data structure
enterprise_data <- tibble(
  department_id = 1:3,
  department_info = list(
    list(
      name = "Engineering",
      budget = 2500000,
      teams = list(
        list(
          team_name = "Backend",
          members = list(
            list(name = "Alice", role = "Senior", salary = 120000, skills = c("Python", "AWS", "Docker")),
            list(name = "Bob", role = "Junior", salary = 80000, skills = c("Python", "MySQL"))
          ),
          projects = list(
            list(name = "API Redesign", status = "Active", budget = 150000),
            list(name = "Database Migration", status = "Planning", budget = 200000)
          )
        ),
        list(
          team_name = "Frontend",
          members = list(
            list(name = "Carol", role = "Senior", salary = 115000, skills = c("React", "TypeScript", "CSS")),
            list(name = "David", role = "Mid", salary = 95000, skills = c("Vue", "JavaScript"))
          ),
          projects = list(
            list(name = "UI Refresh", status = "Active", budget = 100000)
          )
        )
      )
    ),
    list(
      name = "Marketing",
      budget = 800000,
      teams = list(
        list(
          team_name = "Digital",
          members = list(
            list(name = "Emma", role = "Manager", salary = 105000, skills = c("SEO", "Analytics", "AdWords")),
            list(name = "Frank", role = "Specialist", salary = 70000, skills = c("Social Media", "Content"))
          ),
          projects = list(
            list(name = "Brand Campaign", status = "Active", budget = 250000),
            list(name = "Website Optimization", status = "Completed", budget = 75000)
          )
        )
      )
    ),
    list(
      name = "Sales",
      budget = 1200000,
      teams = list(
        list(
          team_name = "Enterprise",
          members = list(
            list(name = "Grace", role = "Director", salary = 140000, skills = c("B2B Sales", "Negotiation")),
            list(name = "Henry", role = "Rep", salary = 85000, skills = c("Cold Calling", "CRM"))
          ),
          projects = list(
            list(name = "Q1 Targets", status = "Active", budget = 50000)
          )
        )
      )
    )
  )
)

cat("Complex enterprise data structure:\n")
str(enterprise_data, max.level = 3)

# Extract department overview
departments <- enterprise_data %>%
  hoist(department_info,
    dept_name = "name",
    budget = "budget",
    teams = "teams"
  )

cat("\nDepartment overview:\n")
print(departments)

# Extract team and member information
team_members <- departments %>%
  unnest_longer(teams) %>%
  hoist(teams,
    team_name = "team_name",
    members = "members"
  ) %>%
  unnest_longer(members) %>%
  unnest_wider(members) %>%
  unnest_longer(skills) %>%
  group_by(department_id, dept_name, budget, team_name, name, role, salary) %>%
  summarise(
    skill_count = n(),
    skill_list = paste(skills, collapse = ", "),
    .groups = "drop"
  )

cat("\nTeam members with skills:\n")
print(team_members)

# Department analytics
dept_analytics <- team_members %>%
  group_by(department_id, dept_name, budget) %>%
  summarise(
    total_employees = n(),
    avg_salary = round(mean(salary), 0),
    total_payroll = sum(salary),
    skill_diversity = length(unique(unlist(str_split(skill_list, ", ")))),
    .groups = "drop"
  ) %>%
  mutate(
    payroll_ratio = round(total_payroll / budget * 100, 1),
    budget_per_employee = round(budget / total_employees, 0)
  ) %>%
  arrange(desc(budget))

cat("\nDepartment analytics:\n")
print(dept_analytics)

# Skills analysis across organization
skills_analysis <- team_members %>%
  select(dept_name, team_name, name, skill_list) %>%
  separate_rows(skill_list, sep = ", ") %>%
  rename(skill = skill_list) %>%
  group_by(skill) %>%
  summarise(
    employee_count = n(),
    departments = length(unique(dept_name)),
    teams = length(unique(paste(dept_name, team_name))),
    .groups = "drop"
  ) %>%
  arrange(desc(employee_count))

cat("\nSkills analysis across organization:\n")
print(skills_analysis)
```

## Performance Considerations and Best Practices

```{r}
#| label: performance-best-practices

cat("🚀 PERFORMANCE OPTIMIZATION FOR NESTED DATA:\n\n")

cat("1. MEMORY MANAGEMENT:\n")
cat("   - Process nested data in chunks for large datasets\n")
cat("   - Remove unnecessary nesting levels early\n")
cat("   - Use specific hoist() extractions instead of full unnesting\n")
cat("   - Consider data.table::rbindlist() for very large list processing\n\n")

cat("2. STRATEGIC APPROACH:\n")
cat("   - Plan your rectangling strategy before starting\n")
cat("   - Extract what you need, not everything\n")
cat("   - Combine unnest operations efficiently\n")
cat("   - Validate structure at each step\n\n")

# Demonstrate efficient vs inefficient approaches
cat("EFFICIENT APPROACH:\n")
cat("data %>%\n")
cat("  hoist(column, key1 = 'key1', key2 = c('nested', 'key2')) %>%\n")
cat("  select(needed_columns) %>%\n")
cat("  process_further()\n\n")

cat("LESS EFFICIENT APPROACH:\n")
cat("data %>%\n")
cat("  unnest_wider(column) %>%          # Expands everything\n")
cat("  unnest_wider(nested_column) %>%   # More expansion\n")
cat("  select(needed_columns) %>%        # Selection after expansion\n")
cat("  process_further()\n\n")

cat("3. ERROR HANDLING:\n")
error_safe_unnest <- function(data, col) {
  tryCatch({
    data %>% unnest_wider(!!sym(col))
  }, error = function(e) {
    warning(paste("Unnesting failed for column", col, ":", e$message))
    return(data)
  })
}

cat("   - Use tryCatch() for robust unnesting\n")
cat("   - Check for NULL values before unnesting\n")
cat("   - Validate list structures before processing\n")
cat("   - Provide default values for missing elements\n\n")

cat("4. VALIDATION HELPERS:\n")
check_list_structure <- function(data, col) {
  col_data <- data[[col]]
  if (is.list(col_data)) {
    list(
      is_list_column = TRUE,
      length = length(col_data),
      first_element_type = class(col_data[[1]]),
      has_names = !is.null(names(col_data[[1]]))
    )
  } else {
    list(is_list_column = FALSE)
  }
}

cat("   - Create helper functions to inspect list structures\n")
cat("   - Use str() and glimpse() liberally during development\n")
cat("   - Test with small samples before processing large datasets\n")
```

## Integration with Other tidyr Functions

```{r}
#| label: integration-workflow

# Complete data processing pipeline combining all tidyr functions
complex_workflow_data <- tibble(
  survey_id = 1:3,
  metadata = c("Study_A|2023|Online", "Study_B|2023|Phone", "Study_C|2024|Online"),
  responses = list(
    list(
      demographics = list(age = "25-34", income = "50-75K", education = "Bachelor"),
      ratings = list(q1 = 4, q2 = 5, q3 = 3, q4 = 4),
      comments = list(positive = c("Great service", "Easy to use"),
                     negative = c("Slow response"))
    ),
    list(
      demographics = list(age = "35-44", income = "75-100K", education = "Master"),
      ratings = list(q1 = 3, q2 = 4, q3 = 4, q4 = 5),
      comments = list(positive = c("Professional staff"),
                     negative = c("Website issues", "Long wait"))
    ),
    list(
      demographics = list(age = "45-54", income = "100K+", education = "PhD"),
      ratings = list(q1 = 5, q2 = 5, q3 = 4, q4 = 4),
      comments = list(positive = c("Excellent quality", "Quick delivery", "Good value"))
    )
  )
)

cat("Complex survey data requiring multiple tidyr operations:\n")
str(complex_workflow_data, max.level = 2)

# Complete processing pipeline
processed_survey <- complex_workflow_data %>%
  # Step 1: Separate metadata
  separate(metadata, into = c("study", "year", "method"), sep = "\\|") %>%

  # Step 2: Extract demographics using hoist
  hoist(responses,
    age_group = c("demographics", "age"),
    income_bracket = c("demographics", "income"),
    education_level = c("demographics", "education"),
    ratings = "ratings",
    comments = "comments"
  ) %>%

  # Step 3: Process ratings (unnest_wider)
  unnest_wider(ratings) %>%
  rename_with(~ paste0("rating_", .x), .cols = c(q1, q2, q3, q4)) %>%

  # Step 4: Calculate rating metrics
  rowwise() %>%
  mutate(
    avg_rating = mean(c(rating_q1, rating_q2, rating_q3, rating_q4)),
    rating_variance = var(c(rating_q1, rating_q2, rating_q3, rating_q4))
  ) %>%
  ungroup() %>%

  # Step 5: Process comments separately
  select(-comments) %>%
  left_join(
    complex_workflow_data %>%
      select(survey_id, responses) %>%
      hoist(responses, comments = "comments") %>%
      unnest_wider(comments) %>%
      mutate(
        positive_count = map_dbl(positive, length),
        negative_count = map_dbl(negative, length),
        sentiment_ratio = positive_count / (positive_count + negative_count)
      ) %>%
      select(survey_id, positive_count, negative_count, sentiment_ratio),
    by = "survey_id"
  ) %>%

  # Step 6: Create final analysis metrics
  mutate(
    satisfaction_score = round((avg_rating * 25) + (sentiment_ratio * 25), 1),
    response_completeness = ifelse(is.na(rating_variance), "Incomplete", "Complete")
  )

cat("\nFinal processed survey data:\n")
print(processed_survey)

# Summary analysis
cat("\nSurvey analysis summary:\n")
processed_survey %>%
  group_by(study, method) %>%
  summarise(
    responses = n(),
    avg_satisfaction = round(mean(satisfaction_score, na.rm = TRUE), 1),
    avg_rating = round(mean(avg_rating, na.rm = TRUE), 2),
    avg_sentiment = round(mean(sentiment_ratio, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_satisfaction)) %>%
  print()
```

## Summary

Mastering nested data and rectangling enables you to:

- **🗂️ Handle complex data structures**: Transform JSON, APIs, and hierarchical data into analysis-ready formats
- **🎯 Extract precisely what you need**: Use `hoist()` for selective extraction instead of full unnesting
- **📊 Process real-world data**: Handle inconsistent structures and missing elements gracefully
- **🔄 Integrate workflows**: Combine rectangling with other tidyr functions for complete data pipelines
- **⚡ Optimize performance**: Choose the right unnesting strategy for your data size and structure

Key principles to remember:

- **Plan your approach**: Understand the nested structure before choosing your rectangling strategy
- **Extract strategically**: Use `hoist()` for specific elements, unnesting for comprehensive expansion
- **Handle variation**: Real-world nested data is often inconsistent - prepare for missing elements
- **Validate continuously**: Check structure and results at each transformation step
- **Combine techniques**: Rectangling works best when integrated with pivoting, separating, and filtering

These advanced data structure skills unlock the ability to work with modern, complex data sources that are increasingly common in data science! 🎯