---
title: "Pivoting Data: Wide ↔ Long Transformations"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## The Art of Data Reshaping 🔄

Data pivoting is the cornerstone of data tidying - the ability to transform data between wide and long formats. Think of it as rotating your data table to change its perspective. Wide format spreads variables across columns (good for human reading), while long format stacks data into rows (ideal for analysis and visualization).

## Understanding Wide vs. Long Format

Let's start with a clear comparison:

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(lubridate)

# Wide format: Values spread across columns
sales_wide <- tibble(
  quarter = c("Q1", "Q2", "Q3", "Q4"),
  north = c(150000, 162000, 158000, 175000),
  south = c(142000, 155000, 168000, 180000),
  east = c(138000, 148000, 152000, 165000),
  west = c(145000, 158000, 163000, 172000)
)

cat("WIDE FORMAT: Each region is a separate column\n")
print(sales_wide)

# Long format: Values stacked in rows
sales_long <- sales_wide %>%
  pivot_longer(
    cols = c(north, south, east, west),
    names_to = "region",
    values_to = "sales"
  )

cat("\nLONG FORMAT: All sales values in one column\n")
print(sales_long)

cat("\nComparison:\n")
cat("Wide: ", nrow(sales_wide), " rows x ", ncol(sales_wide), " columns\n")
cat("Long: ", nrow(sales_long), " rows x ", ncol(sales_long), " columns\n")
```

## When to Use Each Format

```{r}
#| label: format-comparison

cat("📊 WIDE FORMAT - Best for:\n")
cat("- Human-readable reports and dashboards\n")
cat("- Side-by-side comparisons\n")
cat("- Spreadsheet-style presentations\n")
cat("- Correlation matrices\n")
cat("- Crosstabs and pivot tables\n\n")

cat("📈 LONG FORMAT - Best for:\n")
cat("- ggplot2 visualizations\n")
cat("- Statistical modeling and analysis\n")
cat("- Group-by operations with dplyr\n")
cat("- Time series analysis\n")
cat("- Machine learning datasets\n\n")

# Demonstrate why ggplot2 prefers long format
library(ggplot2)

# This works beautifully with long format
sales_long %>%
  ggplot(aes(x = quarter, y = sales, color = region, group = region)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "K")) +
  labs(
    title = "Regional Sales Performance by Quarter",
    subtitle = "Long format makes this visualization straightforward",
    x = "Quarter",
    y = "Sales ($)",
    color = "Region"
  ) +
  theme_minimal()
```

## `pivot_longer()`: Wide to Long

`pivot_longer()` transforms wide data to long format by "gathering" multiple columns into key-value pairs.

### Basic Syntax and Common Patterns

```{r}
#| label: pivot-longer-basics

# Real-world example: Employee performance ratings
performance_wide <- tibble(
  employee_id = 1:5,
  name = c("Alice Johnson", "Bob Smith", "Carol Davis", "David Wilson", "Emma Brown"),
  department = c("Sales", "Marketing", "IT", "Sales", "Marketing"),
  communication_q1 = c(4.2, 3.8, 4.5, 3.9, 4.1),
  communication_q2 = c(4.3, 4.0, 4.6, 4.1, 4.2),
  teamwork_q1 = c(4.0, 4.2, 4.3, 3.8, 4.4),
  teamwork_q2 = c(4.1, 4.4, 4.4, 4.0, 4.6),
  leadership_q1 = c(3.8, 3.5, 4.1, 3.6, 3.9),
  leadership_q2 = c(4.0, 3.7, 4.3, 3.8, 4.1)
)

cat("Original wide format (multiple rating columns):\n")
print(performance_wide)

# Method 1: Basic pivot_longer
performance_long_basic <- performance_wide %>%
  pivot_longer(
    cols = communication_q1:leadership_q2,  # Columns to pivot
    names_to = "skill_quarter",             # Name for the new key column
    values_to = "rating"                    # Name for the new value column
  )

cat("\nBasic pivot_longer result:\n")
print(head(performance_long_basic, 10))
```

### Advanced Column Selection

```{r}
#| label: pivot-longer-selection

# Method 2: Using column selection helpers
performance_long_helpers <- performance_wide %>%
  pivot_longer(
    cols = contains("_q"),               # All columns containing "_q"
    names_to = "skill_quarter",
    values_to = "rating"
  )

# Method 3: Excluding specific columns
performance_long_exclude <- performance_wide %>%
  pivot_longer(
    cols = -c(employee_id, name, department),  # Everything except these
    names_to = "skill_quarter",
    values_to = "rating"
  )

# Method 4: Using patterns with matches()
performance_long_pattern <- performance_wide %>%
  pivot_longer(
    cols = matches("(communication|teamwork|leadership)_q[12]"),  # Regex pattern
    names_to = "skill_quarter",
    values_to = "rating"
  )

cat("All methods produce identical results:\n")
cat("Basic: ", nrow(performance_long_basic), " rows\n")
cat("Helpers: ", nrow(performance_long_helpers), " rows\n")
cat("Exclude: ", nrow(performance_long_exclude), " rows\n")
cat("Pattern: ", nrow(performance_long_pattern), " rows\n")
```

### Separating Column Names

The real power comes when you separate the pivoted column names:

```{r}
#| label: pivot-longer-separation

# Separate skill and quarter into different columns
performance_tidy <- performance_wide %>%
  pivot_longer(
    cols = contains("_q"),
    names_to = "skill_quarter",
    values_to = "rating"
  ) %>%
  separate(
    skill_quarter,
    into = c("skill", "quarter"),
    sep = "_"
  ) %>%
  mutate(
    quarter = str_replace(quarter, "q", "Q"),  # Clean up quarter format
    quarter_num = as.numeric(str_extract(quarter, "\\d+")),
    skill = str_to_title(skill)
  )

cat("Cleaned and separated data:\n")
print(head(performance_tidy, 12))

# Now analysis becomes powerful
cat("\nAverage ratings by skill and quarter:\n")
performance_tidy %>%
  group_by(skill, quarter) %>%
  summarise(
    avg_rating = round(mean(rating), 2),
    employees = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = quarter, values_from = avg_rating) %>%
  print()
```

### Handling Multiple Value Columns

```{r}
#| label: multiple-values

# Complex example: Survey with both scores and confidence levels
survey_complex <- tibble(
  participant_id = 1:4,
  age_group = c("18-25", "26-35", "36-45", "46-55"),
  satisfaction_score_pre = c(3.2, 4.1, 2.8, 3.9),
  satisfaction_confidence_pre = c(0.8, 0.9, 0.7, 0.85),
  satisfaction_score_post = c(4.1, 4.5, 3.9, 4.2),
  satisfaction_confidence_post = c(0.9, 0.95, 0.85, 0.9),
  recommendation_score_pre = c(2.8, 3.5, 2.1, 3.2),
  recommendation_confidence_pre = c(0.7, 0.8, 0.6, 0.75),
  recommendation_score_post = c(3.9, 4.2, 3.4, 4.0),
  recommendation_confidence_post = c(0.85, 0.9, 0.8, 0.88)
)

cat("Complex survey data with scores and confidence:\n")
print(survey_complex)

# Pivot multiple value types simultaneously
survey_long <- survey_complex %>%
  pivot_longer(
    cols = -c(participant_id, age_group),
    names_to = c("metric", "measure", "period"),
    names_pattern = "(.+)_(score|confidence)_(pre|post)",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = measure,
    values_from = value
  ) %>%
  mutate(
    metric = str_to_title(metric),
    period = str_to_title(period)
  )

cat("\nTransformed to tidy format with separate score/confidence columns:\n")
print(head(survey_long, 10))

# Calculate improvement with confidence intervals
improvements <- survey_long %>%
  select(-age_group) %>%
  pivot_wider(names_from = period, values_from = c(score, confidence)) %>%
  mutate(
    score_improvement = score_Post - score_Pre,
    confidence_avg = (confidence_Pre + confidence_Post) / 2,
    improvement_category = case_when(
      score_improvement >= 1.0 ~ "Significant Improvement",
      score_improvement >= 0.5 ~ "Moderate Improvement",
      score_improvement >= 0.0 ~ "Slight Improvement",
      TRUE ~ "Decline"
    )
  )

cat("\nImprovement analysis:\n")
improvements %>%
  group_by(metric, improvement_category) %>%
  summarise(
    participants = n(),
    avg_improvement = round(mean(score_improvement), 2),
    avg_confidence = round(mean(confidence_avg), 2),
    .groups = "drop"
  ) %>%
  arrange(metric, desc(avg_improvement)) %>%
  print()
```

## `pivot_wider()`: Long to Wide

`pivot_wider()` spreads key-value pairs into separate columns, creating the inverse of `pivot_longer()`.

### Basic Wide Transformation

```{r}
#| label: pivot-wider-basics

# Start with our long performance data
cat("Starting with long format data:\n")
print(head(performance_tidy, 8))

# Convert back to wide format for reporting
performance_report <- performance_tidy %>%
  # Create quarter-skill combinations
  unite("quarter_skill", quarter, skill, sep = "_") %>%
  pivot_wider(
    names_from = quarter_skill,
    values_from = rating
  ) %>%
  arrange(department, name)

cat("\nConverted to wide format for reporting:\n")
print(performance_report)
```

### Business Intelligence Dashboard Format

```{r}
#| label: bi-dashboard

# Create dashboard-ready format
dashboard_data <- performance_tidy %>%
  group_by(employee_id, name, department, quarter) %>%
  summarise(
    avg_rating = round(mean(rating), 2),
    skills_evaluated = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = quarter,
    values_from = c(avg_rating, skills_evaluated),
    names_sep = "_"
  ) %>%
  # Calculate improvements
  mutate(
    improvement_q1_to_q2 = round(avg_rating_Q2 - avg_rating_Q1, 2),
    improvement_category = case_when(
      improvement_q1_to_q2 >= 0.3 ~ "Strong Growth",
      improvement_q1_to_q2 >= 0.1 ~ "Steady Growth",
      improvement_q1_to_q2 >= -0.1 ~ "Stable",
      TRUE ~ "Needs Attention"
    )
  )

cat("Dashboard-ready format:\n")
print(dashboard_data)

# Summary by department
cat("\nDepartment performance summary:\n")
dashboard_data %>%
  group_by(department) %>%
  summarise(
    employees = n(),
    avg_q1_rating = round(mean(avg_rating_Q1), 2),
    avg_q2_rating = round(mean(avg_rating_Q2), 2),
    avg_improvement = round(mean(improvement_q1_to_q2), 2),
    strong_growth_pct = round(100 * mean(improvement_category == "Strong Growth"), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_improvement)) %>%
  print()
```

### Handling Missing Values in Pivoting

```{r}
#| label: missing-values

# Realistic scenario with missing data
incomplete_sales <- tibble(
  month = rep(c("Jan", "Feb", "Mar"), each = 3),
  region = rep(c("North", "South", "West"), 3),
  sales = c(100, 150, NA, 120, NA, 180, 130, 160, 170)
)

cat("Data with missing values:\n")
print(incomplete_sales)

# Pivot wider with missing values
sales_matrix <- incomplete_sales %>%
  pivot_wider(
    names_from = month,
    values_from = sales,
    values_fill = 0  # Fill missing with 0 for this business case
  )

cat("\nPivoted wide with missing values filled as 0:\n")
print(sales_matrix)

# Alternative: Keep NAs and handle explicitly
sales_matrix_na <- incomplete_sales %>%
  pivot_wider(
    names_from = month,
    values_from = sales
    # No values_fill - keeps NAs
  ) %>%
  mutate(
    total_known = rowSums(select(., Jan:Mar), na.rm = TRUE),
    months_reported = rowSums(!is.na(select(., Jan:Mar))),
    avg_monthly = round(total_known / months_reported, 1)
  )

cat("\nWith explicit missing value handling:\n")
print(sales_matrix_na)
```

## Complex Real-World Scenarios

### Financial Data Transformation

```{r}
#| label: financial-example

# Quarterly financial statements across multiple companies
financials_raw <- tibble(
  company = rep(c("TechCorp", "RetailCo", "ManufactureInc"), each = 4),
  metric = rep(c("revenue", "costs", "profit", "employees"), 3),
  q1_2023 = c(
    # TechCorp
    5200, 3800, 1400, 450,
    # RetailCo
    8900, 7100, 1800, 1200,
    # ManufactureInc
    6700, 5200, 1500, 800
  ),
  q2_2023 = c(
    5800, 4100, 1700, 470,
    9200, 7300, 1900, 1250,
    7100, 5500, 1600, 820
  ),
  q3_2023 = c(
    6100, 4200, 1900, 485,
    9800, 7600, 2200, 1300,
    7400, 5800, 1600, 835
  ),
  q4_2023 = c(
    6500, 4400, 2100, 500,
    10200, 7900, 2300, 1320,
    7800, 6100, 1700, 850
  )
)

cat("Raw financial data (wide format):\n")
print(head(financials_raw, 12))

# Transform to analysis-ready format
financials_clean <- financials_raw %>%
  # First pivot to long format
  pivot_longer(
    cols = starts_with("q"),
    names_to = "quarter",
    values_to = "value"
  ) %>%
  # Then pivot wider to separate metrics
  pivot_wider(
    names_from = metric,
    values_from = value
  ) %>%
  # Clean and enhance
  mutate(
    quarter = str_replace(quarter, "q(\\d)_", "Q\\1 "),
    profit_margin = round(profit / revenue * 100, 1),
    revenue_per_employee = round(revenue / employees * 1000, 0),
    cost_efficiency = round(costs / revenue * 100, 1)
  ) %>%
  arrange(company, quarter)

cat("\nCleaned financial analysis format:\n")
print(head(financials_clean, 9))

# Performance analysis
cat("\nCompany performance comparison:\n")
financials_clean %>%
  group_by(company) %>%
  summarise(
    avg_profit_margin = round(mean(profit_margin), 1),
    revenue_growth = round((last(revenue) - first(revenue)) / first(revenue) * 100, 1),
    avg_revenue_per_employee = round(mean(revenue_per_employee), 0),
    cost_trend = round(last(cost_efficiency) - first(cost_efficiency), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_profit_margin)) %>%
  print()
```

### Customer Behavior Analysis

```{r}
#| label: customer-behavior

# Customer purchase behavior across channels
customer_behavior <- tibble(
  customer_id = 1:6,
  segment = c("Premium", "Standard", "Premium", "Standard", "Budget", "Budget"),
  online_purchases = c(12, 8, 15, 5, 3, 2),
  store_purchases = c(3, 6, 2, 8, 4, 5),
  phone_purchases = c(2, 1, 3, 2, 1, 0),
  online_revenue = c(2400, 960, 3200, 450, 180, 120),
  store_revenue = c(950, 1200, 580, 1800, 320, 400),
  phone_revenue = c(450, 180, 720, 380, 95, 0)
)

cat("Customer behavior data (wide format):\n")
print(customer_behavior)

# Transform to analyzable format
behavior_analysis <- customer_behavior %>%
  # Pivot purchases and revenue separately, then combine
  pivot_longer(
    cols = c(online_purchases:phone_purchases, online_revenue:phone_revenue),
    names_to = c("channel", "metric"),
    names_pattern = "(.+)_(purchases|revenue)",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = metric,
    values_from = value
  ) %>%
  # Calculate derived metrics
  mutate(
    avg_order_value = round(ifelse(purchases > 0, revenue / purchases, 0), 2),
    channel = str_to_title(channel)
  ) %>%
  arrange(customer_id, channel)

cat("\nTransformed behavior analysis:\n")
print(head(behavior_analysis, 12))

# Channel preference analysis
cat("\nChannel preferences by segment:\n")
behavior_analysis %>%
  filter(purchases > 0) %>%  # Only active channels
  group_by(segment, channel) %>%
  summarise(
    customers = n(),
    avg_purchases = round(mean(purchases), 1),
    avg_revenue = round(mean(revenue), 0),
    avg_order_value = round(mean(avg_order_value), 2),
    .groups = "drop"
  ) %>%
  arrange(segment, desc(avg_revenue)) %>%
  print()
```

## Performance Tips and Best Practices

```{r}
#| label: best-practices

cat("🚀 PERFORMANCE TIPS:\n\n")

cat("1. COLUMN SELECTION:\n")
cat("   - Use specific column ranges: pivot_longer(cols = col1:col5)\n")
cat("   - Avoid pivot_longer(cols = everything()) on large datasets\n")
cat("   - Use selection helpers: starts_with(), contains(), matches()\n\n")

cat("2. MEMORY MANAGEMENT:\n")
cat("   - Pivot smaller chunks of large datasets\n")
cat("   - Remove unnecessary columns before pivoting\n")
cat("   - Consider data.table::melt() for very large datasets\n\n")

cat("3. DATA VALIDATION:\n")
cat("   - Check for duplicate key combinations before pivot_wider()\n")
cat("   - Validate data types after pivoting\n")
cat("   - Handle missing values explicitly with values_fill\n\n")

# Demonstrate validation
cat("Example: Checking for duplicates before pivot_wider()\n")

# This would fail due to duplicates
problematic_data <- tibble(
  id = c(1, 1, 2, 2),
  metric = c("sales", "sales", "profit", "profit"),  # Duplicate keys!
  value = c(100, 150, 50, 75)
)

# Check for duplicates
duplicate_check <- problematic_data %>%
  count(id, metric) %>%
  filter(n > 1)

if(nrow(duplicate_check) > 0) {
  cat("⚠️  Duplicates found! Handle before pivot_wider():\n")
  print(duplicate_check)

  # Fix by aggregating
  fixed_data <- problematic_data %>%
    group_by(id, metric) %>%
    summarise(value = sum(value), .groups = "drop")

  cat("\nFixed by aggregating:\n")
  print(fixed_data)
}

cat("\n4. NAMING CONVENTIONS:\n")
cat("   - Use descriptive names_to and values_to parameters\n")
cat("   - Consider names_prefix to add context\n")
cat("   - Use names_sep for multiple name columns\n")
```

## Integration with dplyr Workflows

```{r}
#| label: dplyr-integration

# Complete data analysis pipeline combining pivot with dplyr
sales_pipeline <- tibble(
  date = seq(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "month"),
  product_a_units = sample(100:500, 12),
  product_a_revenue = sample(5000:25000, 12),
  product_b_units = sample(80:400, 12),
  product_b_revenue = sample(4000:20000, 12),
  product_c_units = sample(60:300, 12),
  product_c_revenue = sample(3000:15000, 12)
) %>%
  # Transform to long format
  pivot_longer(
    cols = -date,
    names_to = c("product", "metric"),
    names_pattern = "product_([abc])_(units|revenue)",
    values_to = "value"
  ) %>%
  # Spread metrics to columns
  pivot_wider(
    names_from = metric,
    values_from = value
  ) %>%
  # Add derived metrics with dplyr
  mutate(
    product = paste("Product", str_to_upper(product)),
    avg_price = round(revenue / units, 2),
    quarter = paste("Q", quarter(date), year(date)),
    month_name = month(date, label = TRUE, abbr = FALSE)
  ) %>%
  # Group analysis
  group_by(product) %>%
  mutate(
    revenue_growth = round((revenue / lag(revenue) - 1) * 100, 1),
    units_growth = round((units / lag(units) - 1) * 100, 1),
    price_change = round(avg_price - lag(avg_price), 2)
  ) %>%
  ungroup()

cat("Complete analysis pipeline result:\n")
print(head(sales_pipeline, 10))

# Quarterly summary
cat("\nQuarterly performance summary:\n")
sales_pipeline %>%
  group_by(product, quarter) %>%
  summarise(
    total_units = sum(units),
    total_revenue = sum(revenue),
    avg_price = round(mean(avg_price), 2),
    .groups = "drop"
  ) %>%
  arrange(product, quarter) %>%
  print()
```

## Common Pitfalls and Solutions

```{r}
#| label: common-pitfalls

cat("⚠️  COMMON PITFALLS AND SOLUTIONS:\n\n")

cat("1. PROBLEM: Mixed data types in value columns\n")
mixed_types <- tibble(
  id = 1:3,
  score_2023 = c("85", "92", "78"),    # Character instead of numeric
  score_2024 = c(88, 94, 82)          # Numeric
)

cat("   Solution: Fix data types before pivoting\n")
fixed_types <- mixed_types %>%
  mutate(score_2023 = as.numeric(score_2023)) %>%
  pivot_longer(cols = starts_with("score"), names_to = "year", values_to = "score")

cat("2. PROBLEM: Inconsistent column naming\n")
inconsistent_names <- tibble(
  region = c("North", "South"),
  Q1_sales = c(100, 120),
  q2_Sales = c(110, 125),              # Inconsistent case
  "Q3 sales" = c(115, 130)            # Space in name
)

cat("   Solution: Standardize names first\n")
standardized <- inconsistent_names %>%
  rename_with(~ str_to_lower(str_replace_all(.x, " ", "_"))) %>%
  pivot_longer(cols = contains("sales"), names_to = "quarter", values_to = "sales")

cat("3. PROBLEM: Unexpected pivoting results\n")
cat("   - Always check dimensions before and after\n")
cat("   - Use glimpse() to verify structure\n")
cat("   - Test with small data first\n\n")

cat("4. SOLUTION TEMPLATE for complex pivots:\n")
cat("   data %>%\n")
cat("     # 1. Clean column names\n")
cat("     rename_with(clean_names_function) %>%\n")
cat("     # 2. Fix data types\n")
cat("     mutate(across(where(is.character), type_convert)) %>%\n")
cat("     # 3. Pivot with explicit parameters\n")
cat("     pivot_longer(...) %>%\n")
cat("     # 4. Validate results\n")
cat("     glimpse()\n")
```

## Summary

Data pivoting with `pivot_longer()` and `pivot_wider()` is essential for:

- **Reshaping for analysis**: Convert wide data to long format for ggplot2 and statistical models
- **Creating reports**: Transform analytical results back to wide format for human consumption
- **Data integration**: Standardize datasets from different sources into consistent formats
- **Performance optimization**: Choose the right shape for your specific analysis needs

Key principles to remember:

- 📊 **Think about your end goal**: What format does your analysis or visualization need?
- 🔄 **One step at a time**: Complex reshaping is easier when broken into multiple pivots
- 🧹 **Clean first**: Standardize column names and data types before pivoting
- ✅ **Validate results**: Check dimensions and data types after each transformation
- 🚀 **Combine tools**: Pivoting works best when integrated with dplyr workflows

Master these pivoting techniques and you'll be able to reshape any dataset into the perfect format for your analysis! 🎯