---
title: "Separating and Uniting: Column Manipulation Mastery"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## The Art of Column Surgery 🔧

Real-world data often comes with information packed into single columns that needs to be split, or scattered across multiple columns that should be combined. The `separate()` and `unite()` functions in tidyr are your surgical tools for precise column manipulation.

### Prefer explicit modern separators

For new code, use `separate_wider_delim()`, `separate_wider_position()`, or `separate_wider_regex()`. They make the parsing strategy explicit and provide clearer handling of malformed records than superseded `separate()`:

```{r}
#| eval: false
products |>
  separate_wider_delim(
    sku,
    delim = "-",
    names = c("division", "family", "size", "colour", "variant"),
    too_few = "debug",
    too_many = "debug"
  )
```

The expected five fields form a contract. Debug mode retains diagnostic columns when a supplier sends too few or too many components, allowing analysts to quarantine exceptions instead of silently losing text. After parsing, validate allowed division codes and convert numeric fields explicitly.

::: {.callout-note title="Do not split opaque identifiers"}
Separate a code only when its components have documented business meaning. If an identifier is merely a key, keep it intact: inferred substrings can change when a source system revises its numbering scheme.
:::

## Understanding the Problem

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(lubridate)

# Common messy data scenarios
cat("TYPICAL MESSY DATA PATTERNS:\n\n")

# Example 1: Multiple values in one column
contact_data <- tibble(
  employee_id = 1:5,
  full_name = c("Sarah Johnson", "Michael Chen", "Emily Rodriguez-Smith",
                "David O'Connor", "Anna-Marie Dubois"),
  contact_info = c("sarah.j@company.com | 555-0123",
                   "m.chen@company.com | 555-0156 | ext:234",
                   "emily.rs@company.com",
                   "d.oconnor@company.com | 555-0189 | LinkedIn: davidoc",
                   "anna.dubois@company.com | 555-0167 | Skype: annadubois")
)

cat("Problem 1: Multiple contact methods in single column\n")
print(contact_data)

# Example 2: Date components scattered across columns
date_components <- tibble(
  transaction_id = 1:4,
  year = c(2023, 2023, 2024, 2024),
  month = c("Jan", "Dec", "Mar", "Jul"),
  day = c(15, 28, 8, 22),
  amount = c(1250.50, 899.99, 2100.00, 567.80)
)

cat("\nProblem 2: Date components in separate columns\n")
print(date_components)

# Example 3: Product codes with embedded information
product_data <- tibble(
  sku = c("ELC-LTOP-15-BLK-256", "HME-CHAI-STD-BRN-001", "OFF-DESK-ADJ-WHT-042"),
  description = c("Electronics Laptop 15inch Black 256GB",
                  "Home Chair Standard Brown",
                  "Office Desk Adjustable White"),
  price = c(899.99, 249.99, 399.99)
)

cat("\nProblem 3: Structured codes that need parsing\n")
print(product_data)
```

## `separate()`: Splitting One Column Into Many

### Basic Separation with Delimiters

```{r}
#| label: separate-basics

# Basic separation using delimiters
contact_separated <- contact_data %>%
  separate(
    col = contact_info,
    into = c("email", "phone", "other"),
    sep = " \\| ",  # Regex pattern for " | "
    fill = "right"  # Fill missing values on the right with NA
  )

cat("Basic separation result:\n")
print(contact_separated)

# Handle names with various formats
names_separated <- contact_data %>%
  separate(
    col = full_name,
    into = c("first_name", "last_name"),
    sep = " ",
    extra = "merge"  # If more than 2 parts, merge extras into last column
  )

cat("\nName separation (handling middle names/hyphenated names):\n")
print(names_separated)

# Better approach for complex names
names_improved <- contact_data %>%
  separate(
    col = full_name,
    into = c("first_name", "last_name"),
    sep = " (?=[^ ]*$)",  # Split on last space (regex lookahead)
    extra = "merge"
  )

cat("\nImproved name separation (preserves compound first names):\n")
print(names_improved)
```

### Advanced Separation Techniques

```{r}
#| label: separate-advanced

# Complex product code parsing
product_parsed <- product_data %>%
  separate(
    col = sku,
    into = c("category", "type", "size", "color", "variant"),
    sep = "-",
    remove = FALSE  # Keep original column
  ) %>%
  mutate(
    category = case_when(
      category == "ELC" ~ "Electronics",
      category == "HME" ~ "Home",
      category == "OFF" ~ "Office",
      TRUE ~ category
    ),
    type = case_when(
      type == "LTOP" ~ "Laptop",
      type == "CHAI" ~ "Chair",
      type == "DESK" ~ "Desk",
      TRUE ~ type
    ),
    size = case_when(
      size %in% c("15", "STD", "ADJ") ~ ifelse(size == "15", "15 inch",
                                               ifelse(size == "STD", "Standard", "Adjustable")),
      TRUE ~ size
    )
  )

cat("Complex product code parsing:\n")
print(product_parsed)

# Handling mixed delimiters in contact info
contact_advanced <- contact_data %>%
  separate(
    col = contact_info,
    into = c("email", "rest"),
    sep = " \\| ",
    extra = "merge",
    fill = "right"
  ) %>%
  # Further parse the "rest" column
  mutate(
    phone = str_extract(rest, "\\d{3}-\\d{4}"),
    extension = str_extract(rest, "ext:(\\d+)") %>% str_remove("ext:"),
    social_platform = str_extract(rest, "(LinkedIn|Skype)") ,
    social_handle = str_extract(rest, "(?:LinkedIn: |Skype: )([^|]+)") %>%
                   str_remove("^(LinkedIn: |Skype: )")
  ) %>%
  select(-rest)

cat("\nAdvanced contact parsing with mixed formats:\n")
print(contact_advanced)
```

### Pattern-Based Separation

```{r}
#| label: separate-patterns

# Email analysis - separate local and domain parts
email_data <- tibble(
  user_id = 1:6,
  email = c("john.doe@company.com", "sarah_123@university.edu",
            "mike+marketing@startup.io", "admin@gov.agency.uk",
            "customer.service@big-corp.co.uk", "test.user@sub.domain.org")
)

email_parsed <- email_data %>%
  separate(
    col = email,
    into = c("local_part", "domain"),
    sep = "@",
    remove = FALSE
  ) %>%
  # Further separate domain components
  separate(
    col = domain,
    into = c("domain_name", "tld"),
    sep = "\\.",
    extra = "merge",  # Handle multi-part TLDs like .co.uk
    remove = FALSE
  ) %>%
  mutate(
    # Extract username patterns
    username_base = str_remove(local_part, "[+.].*"),
    has_plus_addressing = str_detect(local_part, "\\+"),
    domain_type = case_when(
      str_detect(tld, "edu") ~ "Educational",
      str_detect(tld, "gov") ~ "Government",
      str_detect(tld, "org") ~ "Organization",
      str_detect(tld, "com|co\\.") ~ "Commercial",
      TRUE ~ "Other"
    )
  )

cat("Email parsing with pattern recognition:\n")
print(email_parsed)

# Analyze email patterns
cat("\nEmail pattern analysis:\n")
email_parsed %>%
  count(domain_type, has_plus_addressing) %>%
  arrange(desc(n)) %>%
  print()
```

### Handling Messy Real-World Data

```{r}
#| label: messy-data-separation

# Realistic messy customer data
messy_customers <- tibble(
  record_id = 1:8,
  customer_info = c(
    "John Smith, 42, Manager, New York, NY, 10001",
    "Sarah Johnson; 35; Developer; San Francisco; CA; 94102",
    "Mike Chen | 28 | Designer | Austin TX 78701",
    "Emily Rodriguez-Davis, 39, Director, Chicago, IL, 60601",
    "David Kim; 45; Analyst; Seattle WA 98101",
    "Anna Petrova, 33, Consultant, Boston, MA, 02101",
    "Carlos Mendez | 41 | Manager | Miami FL 33101",
    "Jennifer Wu, 37, Engineer, Portland, OR, 97201"
  )
)

cat("Messy customer data with mixed delimiters:\n")
print(messy_customers)

# Clean step by step
customers_cleaned <- messy_customers %>%
  # First, standardize delimiters
  mutate(
    customer_info = str_replace_all(customer_info, "[;|]", ","),
    customer_info = str_replace_all(customer_info, ", +", ",")
  ) %>%
  # Separate into main components
  separate(
    col = customer_info,
    into = c("name", "age", "title", "location"),
    sep = ",",
    extra = "merge"
  ) %>%
  # Parse location further
  separate(
    col = location,
    into = c("city", "state", "zip"),
    sep = ",",
    extra = "merge",
    fill = "right"
  ) %>%
  # Clean up the data
  mutate(
    across(everything(), str_trim),  # Remove extra whitespace
    age = as.numeric(age),
    # Handle state abbreviations with/without zip
    state_zip = ifelse(is.na(zip), state, paste(state, zip)),
    state = str_extract(state_zip, "^[A-Z]{2}"),
    zip = str_extract(state_zip, "\\d{5}$")
  ) %>%
  select(-state_zip)

cat("\nCleaned customer data:\n")
print(customers_cleaned)

# Validation and summary
cat("\nData quality check:\n")
customers_cleaned %>%
  summarise(
    complete_names = sum(!is.na(name)),
    valid_ages = sum(age >= 18 & age <= 100, na.rm = TRUE),
    valid_states = sum(str_length(state) == 2, na.rm = TRUE),
    valid_zips = sum(str_length(zip) == 5, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

## `unite()`: Combining Multiple Columns

### Basic Column Combination

```{r}
#| label: unite-basics

# Start with separated date components
cat("Combining date components:\n")
print(date_components)

# Basic unite operation
dates_united <- date_components %>%
  unite(
    col = "date_string",
    year, month, day,
    sep = "-",
    remove = FALSE  # Keep original columns
  ) %>%
  # Convert to proper date format
  mutate(
    date_proper = dmy(paste(day, month, year)),
    # Alternative approach for US format
    date_string_clean = str_replace_all(date_string, "-(Jan|Feb|Mar|Apr|May|Jun|Jul|Aug|Sep|Oct|Nov|Dec)-", function(x) {
      month_num <- case_when(
        str_detect(x, "Jan") ~ "-01-",
        str_detect(x, "Feb") ~ "-02-",
        str_detect(x, "Mar") ~ "-03-",
        str_detect(x, "Apr") ~ "-04-",
        str_detect(x, "May") ~ "-05-",
        str_detect(x, "Jun") ~ "-06-",
        str_detect(x, "Jul") ~ "-07-",
        str_detect(x, "Aug") ~ "-08-",
        str_detect(x, "Sep") ~ "-09-",
        str_detect(x, "Oct") ~ "-10-",
        str_detect(x, "Nov") ~ "-11-",
        str_detect(x, "Dec") ~ "-12-"
      )
      return(month_num)
    }),
    date_final = ymd(date_string_clean)
  )

cat("\nUnited date columns:\n")
print(dates_united)
```

### Creating Composite Keys and Identifiers

```{r}
#| label: composite-keys

# Employee data that needs composite identifiers
employee_data <- tibble(
  department = c("IT", "HR", "IT", "Finance", "HR", "IT"),
  location = c("NYC", "LA", "CHI", "NYC", "CHI", "LA"),
  employee_num = c(1001, 2001, 1002, 3001, 2002, 1003),
  first_name = c("John", "Sarah", "Mike", "Emily", "David", "Anna"),
  last_name = c("Smith", "Johnson", "Chen", "Rodriguez", "Kim", "Petrova")
)

# Create various composite identifiers
employee_ids <- employee_data %>%
  unite(
    col = "dept_location",
    department, location,
    sep = "_",
    remove = FALSE
  ) %>%
  unite(
    col = "full_name",
    first_name, last_name,
    sep = " ",
    remove = FALSE
  ) %>%
  unite(
    col = "employee_id",
    department, location, employee_num,
    sep = "-"
  ) %>%
  unite(
    col = "display_name",
    last_name, first_name,
    sep = ", "
  )

cat("Composite identifiers created:\n")
print(employee_ids)

# Create user-friendly codes
user_codes <- employee_data %>%
  mutate(
    dept_code = case_when(
      department == "IT" ~ "IT",
      department == "HR" ~ "HR",
      department == "Finance" ~ "FIN"
    ),
    location_code = case_when(
      location == "NYC" ~ "NY",
      location == "LA" ~ "LA",
      location == "CHI" ~ "CH"
    ),
    name_initials = paste0(str_sub(first_name, 1, 1), str_sub(last_name, 1, 1))
  ) %>%
  unite(
    col = "user_code",
    dept_code, location_code, name_initials,
    sep = ""
  )

cat("\nUser-friendly codes:\n")
print(user_codes)
```

### Business Applications

```{r}
#| label: business-applications

# Product catalog management
product_components <- tibble(
  brand = c("TechCorp", "TechCorp", "HomeLife", "HomeLife", "OfficeMax"),
  category = c("Electronics", "Electronics", "Furniture", "Furniture", "Supplies"),
  subcategory = c("Laptops", "Tablets", "Chairs", "Desks", "Storage"),
  model_year = c(2024, 2024, 2023, 2023, 2024),
  color = c("Silver", "Black", "Brown", "White", "Gray"),
  size = c("15in", "10in", "Standard", "Large", "Medium")
)

# Create comprehensive product identifiers
product_catalog <- product_components %>%
  # Create hierarchical category
  unite(
    col = "category_hierarchy",
    category, subcategory,
    sep = " > ",
    remove = FALSE
  ) %>%
  # Create product descriptor
  unite(
    col = "product_name",
    brand, subcategory, size, color,
    sep = " ",
    remove = FALSE
  ) %>%
  # Create SKU-style code
  mutate(
    brand_code = str_sub(brand, 1, 3) %>% str_to_upper(),
    cat_code = str_sub(category, 1, 3) %>% str_to_upper(),
    subcat_code = str_sub(subcategory, 1, 3) %>% str_to_upper(),
    color_code = str_sub(color, 1, 2) %>% str_to_upper()
  ) %>%
  unite(
    col = "sku",
    brand_code, cat_code, subcat_code, model_year, color_code,
    sep = "-"
  ) %>%
  select(sku, product_name, category_hierarchy, everything())

cat("Product catalog with unified identifiers:\n")
print(product_catalog)

# Customer segmentation
customer_segments <- tibble(
  age_group = c("25-34", "35-44", "25-34", "45-54", "35-44", "25-34"),
  income_bracket = c("50-75K", "75-100K", "25-50K", "100K+", "50-75K", "75-100K"),
  location_type = c("Urban", "Suburban", "Urban", "Rural", "Suburban", "Urban"),
  purchase_frequency = c("Monthly", "Quarterly", "Weekly", "Annually", "Monthly", "Bi-weekly")
)

# Create segment profiles
segment_profiles <- customer_segments %>%
  unite(
    col = "demographic_profile",
    age_group, income_bracket,
    sep = " | $",
    remove = FALSE
  ) %>%
  unite(
    col = "behavior_profile",
    location_type, purchase_frequency,
    sep = " ",
    remove = FALSE
  ) %>%
  unite(
    col = "segment_id",
    age_group, income_bracket, location_type,
    sep = "_"
  ) %>%
  # Create readable segment names
  mutate(
    segment_name = case_when(
      str_detect(segment_id, "25-34.*Urban") ~ "Young Urban Professionals",
      str_detect(segment_id, "35-44.*Suburban") ~ "Suburban Families",
      str_detect(segment_id, "45-54.*Rural") ~ "Rural Established",
      TRUE ~ "Other Segment"
    )
  )

cat("\nCustomer segment profiles:\n")
print(segment_profiles)
```

## Advanced Separation and Union Workflows

### Multi-Step Data Cleaning Pipeline

```{r}
#| label: advanced-pipeline

# Complex sales data requiring multiple operations
complex_sales <- tibble(
  record = 1:6,
  sales_rep_info = c("Alice Johnson (North) - alice.j@company.com",
                     "Bob Smith (South) - bob.smith@company.com",
                     "Carol Davis (East) - c.davis@company.com",
                     "David Wilson (West) - d.wilson@company.com",
                     "Emma Brown (North) - emma.b@company.com",
                     "Frank Miller (South) - f.miller@company.com"),
  transaction_details = c("Product:Widget A|Qty:100|Price:$25.99|Date:2023-01-15",
                         "Product:Widget B|Qty:75|Price:$45.50|Date:2023-01-18",
                         "Product:Widget C|Qty:150|Price:$15.75|Date:2023-01-22",
                         "Product:Widget A|Qty:200|Price:$25.99|Date:2023-01-25",
                         "Product:Widget B|Qty:125|Price:$45.50|Date:2023-01-28",
                         "Product:Widget C|Qty:80|Price:$15.75|Date:2023-02-01")
)

cat("Complex sales data to clean:\n")
print(complex_sales)

# Multi-step cleaning pipeline
sales_cleaned <- complex_sales %>%
  # Step 1: Separate rep info
  separate(
    col = sales_rep_info,
    into = c("rep_name_region", "email"),
    sep = " - "
  ) %>%
  # Step 2: Extract name and region
  separate(
    col = rep_name_region,
    into = c("rep_name", "region"),
    sep = " \\("
  ) %>%
  mutate(region = str_remove(region, "\\)")) %>%

  # Step 3: Parse transaction details
  separate(
    col = transaction_details,
    into = c("product_raw", "qty_raw", "price_raw", "date_raw"),
    sep = "\\|"
  ) %>%

  # Step 4: Clean individual transaction components
  mutate(
    product = str_remove(product_raw, "Product:"),
    quantity = as.numeric(str_remove(qty_raw, "Qty:")),
    unit_price = as.numeric(str_remove_all(price_raw, "Price:\\$")),
    transaction_date = ymd(str_remove(date_raw, "Date:")),
    total_amount = quantity * unit_price
  ) %>%

  # Step 5: Create final identifiers
  unite(
    col = "rep_id",
    rep_name, region,
    sep = "_",
    remove = FALSE
  ) %>%
  mutate(
    rep_id = str_replace_all(rep_id, " ", "_"),
    month_year = format(transaction_date, "%Y-%m")
  ) %>%

  # Step 6: Clean up
  select(record, rep_id, rep_name, region, email, product, quantity,
         unit_price, total_amount, transaction_date, month_year)

cat("\nCleaned sales data:\n")
print(sales_cleaned)

# Analysis ready data
cat("\nSales summary by rep and month:\n")
sales_cleaned %>%
  group_by(rep_name, region, month_year) %>%
  summarise(
    transactions = n(),
    total_quantity = sum(quantity),
    total_revenue = sum(total_amount),
    avg_transaction = round(mean(total_amount), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(total_revenue)) %>%
  print()
```

### Dynamic Column Operations

```{r}
#| label: dynamic-operations

# Survey data with varying response structures
survey_responses <- tibble(
  respondent_id = 1:5,
  demographics = c("Age:25|Gender:F|Education:Bachelor",
                   "Age:34|Gender:M|Education:Master|Income:75000",
                   "Age:42|Gender:F|Education:PhD|Income:95000|Years_Experience:15",
                   "Age:28|Gender:M|Education:Bachelor|Income:55000",
                   "Age:51|Gender:F|Education:Master|Income:85000|Years_Experience:20"),
  responses = c("Q1:5|Q2:4|Q3:3",
                "Q1:4|Q2:5|Q3:4|Q4:3",
                "Q1:3|Q2:3|Q3:5|Q4:4|Q5:5",
                "Q1:5|Q2:4|Q3:4",
                "Q1:4|Q2:3|Q3:3|Q4:5|Q5:4")
)

cat("Variable structure survey data:\n")
print(survey_responses)

# Function to parse key-value pairs
parse_key_value <- function(data, col_name) {
  data %>%
    separate_rows(!!sym(col_name), sep = "\\|") %>%
    separate(!!sym(col_name), into = c("key", "value"), sep = ":") %>%
    pivot_wider(names_from = key, values_from = value)
}

# Parse demographics
demographics_parsed <- survey_responses %>%
  select(respondent_id, demographics) %>%
  parse_key_value("demographics") %>%
  mutate(
    Age = as.numeric(Age),
    Income = as.numeric(Income),
    Years_Experience = as.numeric(Years_Experience)
  )

cat("\nParsed demographics:\n")
print(demographics_parsed)

# Parse responses
responses_parsed <- survey_responses %>%
  select(respondent_id, responses) %>%
  parse_key_value("responses") %>%
  mutate(across(starts_with("Q"), as.numeric))

cat("\nParsed survey responses:\n")
print(responses_parsed)

# Combine for analysis
complete_survey <- demographics_parsed %>%
  left_join(responses_parsed, by = "respondent_id") %>%
  mutate(
    avg_response = round(rowMeans(select(., starts_with("Q")), na.rm = TRUE), 2),
    response_count = rowSums(!is.na(select(., starts_with("Q"))))
  )

cat("\nComplete survey analysis dataset:\n")
print(complete_survey)
```

## Performance Optimization and Best Practices

```{r}
#| label: performance-tips

cat("🚀 PERFORMANCE AND BEST PRACTICES:\n\n")

cat("1. PREPARATION STRATEGIES:\n")
cat("   - Clean delimiters before separation\n")
cat("   - Standardize formats across rows\n")
cat("   - Handle missing values explicitly\n")
cat("   - Validate data types after operations\n\n")

cat("2. MEMORY EFFICIENCY:\n")
cat("   - Use remove = TRUE to drop source columns when not needed\n")
cat("   - Process large datasets in chunks\n")
cat("   - Consider data.table for very large datasets\n\n")

cat("3. ERROR HANDLING:\n")
# Demonstrate robust separation
robust_separation <- function(data, col, into_cols, sep_pattern) {
  tryCatch({
    data %>%
      separate(
        col = !!sym(col),
        into = into_cols,
        sep = sep_pattern,
        fill = "right",
        extra = "merge"
      )
  }, error = function(e) {
    warning(paste("Separation failed:", e$message))
    return(data)
  })
}

cat("   - Use tryCatch() for error handling\n")
cat("   - Set fill and extra parameters appropriately\n")
cat("   - Validate results after each step\n\n")

cat("4. TESTING APPROACH:\n")
test_data <- tibble(
  id = 1:3,
  messy_col = c("A|B|C", "X|Y", "P|Q|R|S|T")
)

# Test with small sample first
test_result <- robust_separation(test_data, "messy_col", c("col1", "col2", "col3"), "\\|")
cat("   - Always test with small samples first\n")
cat("   - Check edge cases (missing values, extra delimiters)\n")
cat("   - Verify column counts and types\n")

print(test_result)

cat("\n5. REGEX BEST PRACTICES:\n")
cat("   - Escape special characters: \\\\| for literal pipe\n")
cat("   - Use anchors: ^ for start, $ for end\n")
cat("   - Test patterns: str_detect() before separate()\n")
cat("   - Consider str_split() for complex patterns\n")
```

## Integration with Other tidyr Functions

```{r}
#| label: integration-example

# Complete data transformation workflow
messy_financial <- tibble(
  company_quarter = c("TechCorp_Q1_2023", "TechCorp_Q2_2023", "RetailCo_Q1_2023", "RetailCo_Q2_2023"),
  metrics = c("Revenue:5200|Costs:3800|Employees:450",
              "Revenue:5800|Costs:4100|Employees:470",
              "Revenue:8900|Costs:7100|Employees:1200",
              "Revenue:9200|Costs:7300|Employees:1250"),
  growth_data = c("YoY_Revenue:12.5%|QoQ_Costs:8.1%",
                  "YoY_Revenue:15.2%|QoQ_Costs:7.9%",
                  "YoY_Revenue:8.7%|QoQ_Costs:2.8%",
                  "YoY_Revenue:9.1%|QoQ_Costs:2.8%")
)

# Complete transformation pipeline
financial_clean <- messy_financial %>%
  # Step 1: Separate company info
  separate(company_quarter, into = c("company", "quarter", "year"), sep = "_") %>%

  # Step 2: Parse metrics using separate_rows and pivot
  separate_rows(metrics, sep = "\\|") %>%
  separate(metrics, into = c("metric", "value"), sep = ":") %>%
  mutate(value = as.numeric(value)) %>%
  pivot_wider(names_from = metric, values_from = value) %>%

  # Step 3: Parse growth data
  separate_rows(growth_data, sep = "\\|") %>%
  separate(growth_data, into = c("growth_type", "growth_rate"), sep = ":") %>%
  mutate(growth_rate = as.numeric(str_remove(growth_rate, "%"))) %>%
  pivot_wider(names_from = growth_type, values_from = growth_rate) %>%

  # Step 4: Create derived metrics
  mutate(
    profit = Revenue - Costs,
    profit_margin = round(profit / Revenue * 100, 1),
    revenue_per_employee = round(Revenue / Employees, 0)
  ) %>%

  # Step 5: Create period identifier
  unite(period, quarter, year, sep = " ") %>%
  arrange(company, period)

cat("Complete financial transformation:\n")
print(financial_clean)

cat("\nFinancial performance summary:\n")
financial_clean %>%
  group_by(company) %>%
  summarise(
    avg_profit_margin = round(mean(profit_margin), 1),
    revenue_growth = round(mean(YoY_Revenue, na.rm = TRUE), 1),
    cost_efficiency = round(mean(QoQ_Costs, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  print()
```

## Common Use Cases and Solutions

```{r}
#| label: common-use-cases

cat("📋 COMMON USE CASES AND SOLUTIONS:\n\n")

cat("1. ADDRESS PARSING:\n")
address_data <- tibble(
  id = 1:3,
  full_address = c("123 Main St, Apt 4B, New York, NY 10001",
                   "456 Oak Avenue, San Francisco, CA 94102",
                   "789 Pine Road, Unit 12, Chicago, IL 60601")
)

addresses_parsed <- address_data %>%
  separate(full_address, into = c("street", "city_state_zip"), sep = ", (?=[^,]*,[^,]*$)") %>%
  separate(city_state_zip, into = c("city", "state_zip"), sep = ", ") %>%
  separate(state_zip, into = c("state", "zip"), sep = " ")

print(addresses_parsed)

cat("\n2. NAME STANDARDIZATION:\n")
name_variations <- tibble(
  id = 1:4,
  name_input = c("Dr. John Smith Jr.", "Ms. Sarah Johnson-Brown",
                 "Michael Chen", "Prof. Emily Rodriguez PhD")
)

names_standardized <- name_variations %>%
  mutate(
    # Remove titles and suffixes for core name
    clean_name = str_remove_all(name_input, "\\b(Dr\\.|Ms\\.|Prof\\.|Jr\\.|PhD)\\b"),
    clean_name = str_trim(clean_name)
  ) %>%
  separate(clean_name, into = c("first", "last"), sep = " ", extra = "merge")

print(names_standardized)

cat("\n3. IDENTIFIER CREATION:\n")
cat("   unite(col = 'new_id', col1, col2, sep = '_')\n")
cat("   - Use consistent separators\n")
cat("   - Include entity type prefixes\n")
cat("   - Consider sort order in combinations\n")

cat("\n4. DATA DENORMALIZATION:\n")
cat("   - Combine related fields for reporting\n")
cat("   - Create display-friendly formats\n")
cat("   - Unite for export to external systems\n")
```

## Summary

Mastering `separate()` and `unite()` enables you to:

- **🔧 Parse complex data structures**: Break down compound fields into analyzable components
- **🎯 Create meaningful identifiers**: Combine data elements into useful composite keys
- **🧹 Clean messy real-world data**: Handle inconsistent formats and delimiters
- **📊 Prepare data for analysis**: Transform between storage and analysis formats
- **🔗 Integrate workflows**: Combine with other tidyr functions for complete data pipelines

Key principles to remember:

- **Test patterns first**: Use `str_detect()` and small samples to validate separation logic
- **Handle edge cases**: Plan for missing values, extra delimiters, and format variations
- **Validate results**: Check data types and completeness after operations
- **Think about reversibility**: Document transformations for potential recreation
- **Optimize for performance**: Remove unnecessary columns and process in logical chunks

These column manipulation skills are essential for transforming real-world messy data into analysis-ready datasets! 🎯
