---
title: "Tidy Data Principles"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to Tidy Data

Tidy data is a consistent way to organize data that makes analysis easier and more intuitive. The concept, formalized by Hadley Wickham, provides a standard way to structure datasets that works seamlessly with the tidyverse ecosystem.

```{r}
#| label: setup
#| message: false

library(tidyverse)

cat("The three fundamental rules of tidy data:\n")
cat("1. Each variable forms a column\n")
cat("2. Each observation forms a row\n")
cat("3. Each type of observational unit forms a table\n")
```

## The Three Rules of Tidy Data

### Rule 1: Each Variable Forms a Column

A variable is something you can measure, count, or describe. Each variable should have its own column.

```{r}
#| label: rule1-examples

# TIDY: Each variable (name, age, height) has its own column
tidy_people <- tibble(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  height = c(165, 180, 175)
)

cat("TIDY - Each variable in its own column:\n")
print(tidy_people)

# MESSY: Multiple variables squeezed into one column
messy_people <- tibble(
  person = c("Alice", "Bob", "Charlie"),
  age_height = c("25_165", "30_180", "35_175")
)

cat("\nMESSY - Multiple variables in one column:\n")
print(messy_people)

# MESSY: Variable names as data
messy_measurements <- tibble(
  person = c("Alice", "Alice", "Bob", "Bob"),
  measurement = c("age", "height", "age", "height"),
  value = c(25, 165, 30, 180)
)

cat("\nMESSY - Variable names stored as data:\n")
print(messy_measurements)
```

### Rule 2: Each Observation Forms a Row

An observation is all the values measured on the same unit (person, company, country) at the same time.

```{r}
#| label: rule2-examples

# TIDY: Each person (observation) gets one row
tidy_sales <- tibble(
  sales_rep = c("Alice", "Bob", "Charlie"),
  region = c("North", "South", "East"),
  Q1_sales = c(10000, 12000, 11000),
  Q2_sales = c(11000, 13000, 12000)
)

cat("TIDY - Each sales rep (observation) in one row:\n")
print(tidy_sales)

# MESSY: Observations split across multiple rows
messy_sales_split <- tibble(
  rep_quarter = c("Alice_Q1", "Alice_Q2", "Bob_Q1", "Bob_Q2"),
  region = c("North", "North", "South", "South"),
  sales = c(10000, 11000, 12000, 13000)
)

cat("\nMESSY - Observations split across rows:\n")
print(messy_sales_split)

# MESSY: Multiple observations in one row
messy_sales_combined <- tibble(
  quarter = c("Q1", "Q2"),
  alice_north = c(10000, 11000),
  bob_south = c(12000, 13000),
  charlie_east = c(11000, 12000)
)

cat("\nMESSY - Multiple observations in one row:\n")
print(messy_sales_combined)
```

### Rule 3: Each Type of Observational Unit Forms a Table

Different types of data should be stored in separate tables. For example, customer data and order data are different types of observational units.

```{r}
#| label: rule3-examples

# TIDY: Separate tables for different observational units
customers <- tibble(
  customer_id = c("C001", "C002", "C003"),
  customer_name = c("Alice Corp", "Bob Ltd", "Charlie Inc"),
  industry = c("Tech", "Finance", "Healthcare")
)

orders <- tibble(
  order_id = c("O001", "O002", "O003", "O004"),
  customer_id = c("C001", "C002", "C001", "C003"),
  order_date = ymd(c("2024-01-15", "2024-01-16", "2024-01-17", "2024-01-18")),
  amount = c(1000, 1500, 800, 2000)
)

cat("TIDY - Customers table:\n")
print(customers)
cat("\nTIDY - Orders table:\n")
print(orders)

# MESSY: Everything mashed together
messy_combined <- tibble(
  order_id = c("O001", "O002", "O003", "O004"),
  customer_id = c("C001", "C002", "C001", "C003"),
  customer_name = c("Alice Corp", "Bob Ltd", "Alice Corp", "Charlie Inc"),
  industry = c("Tech", "Finance", "Tech", "Healthcare"),
  order_date = ymd(c("2024-01-15", "2024-01-16", "2024-01-17", "2024-01-18")),
  amount = c(1000, 1500, 800, 2000)
)

cat("\nMESSY - Everything in one table (redundant customer info):\n")
print(messy_combined)
```

## Common Messy Data Patterns

### Pattern 1: Column Headers are Values

This is one of the most common problems in real-world data.

```{r}
#| label: pattern1-headers-as-values

# Common in spreadsheets: Years as column headers
messy_population <- tibble(
  country = c("USA", "Canada", "Mexico"),
  `2020` = c(331, 38, 129),
  `2021` = c(332, 38, 130),
  `2022` = c(333, 39, 131),
  `2023` = c(334, 39, 132)
)

cat("MESSY - Years as column headers:\n")
print(messy_population)

# How to identify this pattern:
cat("\nHow to identify this pattern:\n")
cat("- Column names represent values of a variable (years, quarters, treatments)\n")
cat("- You see repeated patterns in column names\n")
cat("- Data is 'wider' than it should be\n")

# What the tidy version should look like:
tidy_population <- messy_population %>%
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "population_millions"
  ) %>%
  mutate(year = as.numeric(year))

cat("\nTIDY version:\n")
print(tidy_population)
```

### Pattern 2: Multiple Variables in Column Names

Column names often encode multiple pieces of information.

```{r}
#| label: pattern2-multiple-variables

# Treatment and measurement type encoded in column names
messy_experiment <- tibble(
  subject_id = 1:3,
  treatment_A_weight = c(70, 75, 68),
  treatment_A_height = c(170, 180, 165),
  treatment_B_weight = c(71, 76, 69),
  treatment_B_height = c(171, 181, 166)
)

cat("MESSY - Treatment and measurement in column names:\n")
print(messy_experiment)

cat("\nHow to identify this pattern:\n")
cat("- Column names contain underscores or separators\n")
cat("- Multiple pieces of info encoded in names\n")
cat("- Similar patterns repeated across columns\n")

# Tidy approach: separate the information
tidy_experiment <- messy_experiment %>%
  pivot_longer(
    cols = -subject_id,
    names_to = "measurement_info",
    values_to = "value"
  ) %>%
  separate(measurement_info,
           into = c("treatment_label", "treatment", "measurement"),
           sep = "_") %>%
  select(-treatment_label) %>%  # Remove redundant column
  pivot_wider(names_from = measurement, values_from = value)

cat("\nTIDY version:\n")
print(tidy_experiment)
```

### Pattern 3: Variables Stored in Rows and Columns

Sometimes data is stored in a matrix-like format where both rows and columns contain variable information.

```{r}
#| label: pattern3-matrix-format

# Sales data with products as rows and quarters as columns
messy_sales_matrix <- tibble(
  product = c("Widget A", "Widget B", "Widget C"),
  Q1_North = c(100, 150, 120),
  Q1_South = c(90, 140, 110),
  Q2_North = c(110, 160, 130),
  Q2_South = c(95, 145, 115)
)

cat("MESSY - Products in rows, quarters and regions in columns:\n")
print(messy_sales_matrix)

# Tidy version needs multiple steps
tidy_sales_matrix <- messy_sales_matrix %>%
  pivot_longer(
    cols = -product,
    names_to = "quarter_region",
    values_to = "sales"
  ) %>%
  separate(quarter_region, into = c("quarter", "region"), sep = "_")

cat("\nTIDY version:\n")
print(tidy_sales_matrix)
```

### Pattern 4: Multiple Observational Units in One Table

Different types of information mixed together.

```{r}
#| label: pattern4-mixed-units

# Student and course information mixed
messy_grades <- tibble(
  record_id = 1:4,
  student_name = c("Alice", "Alice", "Bob", "Bob"),
  student_age = c(20, 20, 22, 22),
  student_major = c("Math", "Math", "Physics", "Physics"),
  course_name = c("Calculus", "Statistics", "Calculus", "Physics"),
  course_credits = c(3, 3, 3, 4),
  grade = c("A", "B", "A-", "B+")
)

cat("MESSY - Student and course info mixed:\n")
print(messy_grades)

# Should be separate tables
students <- messy_grades %>%
  distinct(student_name, student_age, student_major)

courses <- messy_grades %>%
  distinct(course_name, course_credits)

enrollments <- messy_grades %>%
  select(student_name, course_name, grade)

cat("\nTIDY - Students table:\n")
print(students)
cat("\nTIDY - Courses table:\n")
print(courses)
cat("\nTIDY - Enrollments table:\n")
print(enrollments)
```

## Diagnosing Data Structure Problems

### Quick Diagnostic Questions

```{r}
#| label: diagnostic-questions

diagnostic_checklist <- function(data) {
  cat("TIDY DATA DIAGNOSTIC CHECKLIST\n")
  cat("==============================\n\n")

  cat("1. COLUMNS (Variables):\n")
  cat("   □ Does each column represent exactly one variable?\n")
  cat("   □ Are column names variable names (not values)?\n")
  cat("   □ Do you see patterns like '2020', '2021' in column names?\n")
  cat("   □ Do column names contain multiple pieces of info?\n\n")

  cat("2. ROWS (Observations):\n")
  cat("   □ Does each row represent exactly one observation?\n")
  cat("   □ Are observations split across multiple rows?\n")
  cat("   □ Are multiple observations squeezed into one row?\n\n")

  cat("3. CELLS (Values):\n")
  cat("   □ Does each cell contain exactly one value?\n")
  cat("   □ Are multiple values separated by commas/semicolons?\n")
  cat("   □ Are values and variable names mixed in cells?\n\n")

  cat("4. TABLES (Observational Units):\n")
  cat("   □ Does this table mix different types of entities?\n")
  cat("   □ Is information duplicated across rows?\n")
  cat("   □ Should this be split into multiple tables?\n\n")

  # Basic analysis of the provided data
  cat("QUICK ANALYSIS OF YOUR DATA:\n")
  cat("============================\n")
  cat("Dimensions:", nrow(data), "rows ×", ncol(data), "columns\n")
  cat("Column names:", paste(names(data), collapse = ", "), "\n")

  # Check for common patterns
  col_names <- names(data)
  if (any(str_detect(col_names, "\\d{4}"))) {
    cat("⚠️  WARNING: Found years in column names - might need pivot_longer()\n")
  }
  if (any(str_detect(col_names, "_"))) {
    cat("⚠️  WARNING: Found underscores in column names - might need separate()\n")
  }
  if (length(unique(apply(data, 1, paste, collapse = ""))) < nrow(data)) {
    cat("⚠️  WARNING: Found duplicate rows - check for repeated observations\n")
  }
}

# Example usage
sample_messy <- tibble(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie"),
  `2023_sales` = c(1000, 1200, 1100),
  `2024_sales` = c(1100, 1300, 1200)
)

diagnostic_checklist(sample_messy)
```

### Visual Pattern Recognition

```{r}
#| label: visual-patterns

# Function to visualize data structure issues
show_data_problems <- function() {
  cat("COMMON MESSY DATA PATTERNS - VISUAL GUIDE\n")
  cat("=========================================\n\n")

  # Pattern 1: Too wide
  cat("PATTERN 1: TOO WIDE (headers are values)\n")
  cat("┌─────────┬─────┬─────┬─────┬─────┐\n")
  cat("│ country │2020 │2021 │2022 │2023 │\n")
  cat("├─────────┼─────┼─────┼─────┼─────┤\n")
  cat("│ USA     │ 331 │ 332 │ 333 │ 334 │\n")
  cat("│ Canada  │  38 │  38 │  39 │  39 │\n")
  cat("└─────────┴─────┴─────┴─────┴─────┘\n")
  cat("SHOULD BE: pivot_longer() to make it tall\n\n")

  # Pattern 2: Multiple variables in names
  cat("PATTERN 2: MULTIPLE VARIABLES IN COLUMN NAMES\n")
  cat("┌─────┬──────────┬──────────┬──────────┐\n")
  cat("│ id  │treat_A_pre│treat_A_post│treat_B_pre│\n")
  cat("├─────┼──────────┼──────────┼──────────┤\n")
  cat("│  1  │    70    │    75    │    72    │\n")
  cat("└─────┴──────────┴──────────┴──────────┘\n")
  cat("SHOULD BE: pivot_longer() then separate()\n\n")

  # Pattern 3: Multiple values in cells
  cat("PATTERN 3: MULTIPLE VALUES IN CELLS\n")
  cat("┌─────────┬─────────────┬─────────────┐\n")
  cat("│ person  │   contact   │   skills    │\n")
  cat("├─────────┼─────────────┼─────────────┤\n")
  cat("│ Alice   │email;phone  │ R,Python    │\n")
  cat("│ Bob     │email;phone  │Excel,SQL    │\n")
  cat("└─────────┴─────────────┴─────────────┘\n")
  cat("SHOULD BE: separate() into multiple columns\n\n")

  # Pattern 4: Variables as rows
  cat("PATTERN 4: VARIABLES STORED AS ROWS\n")
  cat("┌─────────┬─────────┬───────┐\n")
  cat("│ person  │variable │ value │\n")
  cat("├─────────┼─────────┼───────┤\n")
  cat("│ Alice   │ height  │  165  │\n")
  cat("│ Alice   │ weight  │   60  │\n")
  cat("│ Bob     │ height  │  180  │\n")
  cat("│ Bob     │ weight  │   75  │\n")
  cat("└─────────┴─────────┴───────┘\n")
  cat("SHOULD BE: pivot_wider() to make columns\n\n")
}

show_data_problems()
```

## Planning Your Tidying Strategy

### Step-by-Step Planning Process

```{r}
#| label: planning-strategy

# Function to help plan tidying strategy
plan_tidying <- function(data, target_description = NULL) {
  cat("TIDYING STRATEGY PLANNER\n")
  cat("========================\n\n")

  cat("STEP 1: UNDERSTAND YOUR CURRENT DATA\n")
  cat("Current dimensions:", nrow(data), "rows ×", ncol(data), "columns\n")
  cat("Current columns:", paste(names(data), collapse = ", "), "\n\n")

  cat("STEP 2: ENVISION YOUR TARGET STRUCTURE\n")
  if (!is.null(target_description)) {
    cat("Target:", target_description, "\n")
  } else {
    cat("Ask yourself:\n")
    cat("- What should each row represent?\n")
    cat("- What variables do I need as columns?\n")
    cat("- What is the observational unit?\n")
  }
  cat("\n")

  cat("STEP 3: IDENTIFY REQUIRED TRANSFORMATIONS\n")

  # Check for common patterns
  col_names <- names(data)

  if (any(str_detect(col_names, "\\d{4}"))) {
    cat("📋 NEEDS: pivot_longer() - Found years/dates in column names\n")
  }

  if (any(str_detect(col_names, "_.*_"))) {
    cat("📋 NEEDS: separate() - Found multiple variables in column names\n")
  }

  # Check for values that look like they contain multiple items
  char_cols <- data %>% select(where(is.character))
  if (ncol(char_cols) > 0) {
    multi_value_check <- char_cols %>%
      summarise(across(everything(), ~ any(str_detect(.x, "[,;|]"), na.rm = TRUE)))

    if (any(multi_value_check)) {
      cat("📋 NEEDS: separate() - Found multiple values in cells\n")
    }
  }

  cat("\n")

  cat("STEP 4: PLAN THE SEQUENCE\n")
  cat("1. Start with the most fundamental transformation\n")
  cat("2. Work from general to specific\n")
  cat("3. Clean up column names and types at the end\n\n")

  cat("STEP 5: VERIFY EACH STEP\n")
  cat("- Check dimensions after each transformation\n")
  cat("- Verify that you haven't lost or duplicated data\n")
  cat("- Test with a small sample first\n")
}

# Example usage
messy_survey <- tibble(
  respondent = 1:3,
  Q1_pre = c(3, 4, 2),
  Q1_post = c(4, 5, 4),
  Q2_pre = c(2, 3, 2),
  Q2_post = c(3, 4, 3),
  demographics = c("25_Male", "34_Female", "28_Male")
)

plan_tidying(messy_survey, "One row per respondent-question-time combination")
```

## Real-World Tidying Examples

### Example 1: Financial Data

```{r}
#| label: financial-example

# Typical financial data from spreadsheets
financial_messy <- tibble(
  company = c("Apple", "Google", "Microsoft"),
  `2021_Revenue` = c(365.8, 257.6, 168.1),
  `2021_Profit` = c(94.7, 76.0, 61.3),
  `2022_Revenue` = c(394.3, 282.8, 198.3),
  `2022_Profit` = c(99.8, 82.5, 72.7),
  `2023_Revenue` = c(383.3, 307.4, 211.9),
  `2023_Profit` = c(97.0, 73.8, 72.4)
)

cat("MESSY: Financial data with years and metrics in column names\n")
print(financial_messy)

# Step 1: Identify the structure
cat("\nANALYSIS:\n")
cat("- Years (2021, 2022, 2023) are in column names → need pivot_longer()\n")
cat("- Metrics (Revenue, Profit) are in column names → need separate()\n")
cat("- Each company-year should be one observation\n")

# Step 2: Tidy transformation
financial_tidy <- financial_messy %>%
  # First, pivot longer to get year_metric combinations
  pivot_longer(
    cols = -company,
    names_to = "year_metric",
    values_to = "value"
  ) %>%
  # Then separate the year and metric
  separate(year_metric, into = c("year", "metric"), sep = "_") %>%
  # Convert year to numeric
  mutate(year = as.numeric(year)) %>%
  # Pivot wider to get metrics as columns
  pivot_wider(names_from = metric, values_from = value) %>%
  # Calculate derived metrics
  mutate(
    profit_margin = round(Profit / Revenue * 100, 1),
    profit_growth = round((Profit / lag(Profit) - 1) * 100, 1)
  ) %>%
  arrange(company, year)

cat("\nTIDY: Clean financial data ready for analysis\n")
print(financial_tidy)

# Now analysis is straightforward
cat("\nAnalysis example - Average profit margin by company:\n")
financial_tidy %>%
  group_by(company) %>%
  summarise(avg_profit_margin = round(mean(profit_margin), 1), .groups = "drop") %>%
  arrange(desc(avg_profit_margin)) %>%
  print()
```

### Example 2: Survey Data

```{r}
#| label: survey-example

# Complex survey data with multiple issues
survey_messy <- tibble(
  id = 1:4,
  name_age = c("Alice_25", "Bob_30", "Charlie_35", "Diana_28"),
  pre_satisfaction = c(3, 4, 2, 5),
  post_satisfaction = c(4, 5, 4, 5),
  pre_recommendation = c(2, 3, 2, 4),
  post_recommendation = c(4, 4, 3, 5),
  contact_info = c("alice@email.com;555-1234", "bob@company.org;555-5678",
                   "charlie@domain.com", "diana@startup.io;555-9999")
)

cat("MESSY: Survey data with multiple structural issues\n")
print(survey_messy)

# Multi-step tidying process
survey_tidy <- survey_messy %>%
  # Step 1: Separate name and age
  separate(name_age, into = c("name", "age"), sep = "_") %>%
  mutate(age = as.numeric(age)) %>%

  # Step 2: Pivot longer to get measurement types in rows
  pivot_longer(
    cols = matches("(satisfaction|recommendation)"),
    names_to = "measurement_info",
    values_to = "score"
  ) %>%

  # Step 3: Separate measurement info
  separate(measurement_info, into = c("time", "metric"), sep = "_") %>%

  # Step 4: Handle contact info (separate into multiple rows for multiple contacts)
  separate_rows(contact_info, sep = ";") %>%

  # Step 5: Clean up and organize
  mutate(
    contact_type = case_when(
      str_detect(contact_info, "@") ~ "email",
      str_detect(contact_info, "^\\d{3}-\\d{4}$") ~ "phone",
      TRUE ~ "other"
    )
  ) %>%
  arrange(id, time, metric)

cat("\nTIDY: Survey data in analysis-ready format\n")
print(head(survey_tidy, 12))

# Now we can easily analyze pre/post changes
cat("\nAnalysis example - Average improvement by metric:\n")
survey_improvement <- survey_tidy %>%
  filter(contact_type == "email") %>%  # One record per person
  select(id, name, time, metric, score) %>%
  pivot_wider(names_from = time, values_from = score) %>%
  mutate(improvement = post - pre) %>%
  group_by(metric) %>%
  summarise(avg_improvement = round(mean(improvement, na.rm = TRUE), 2), .groups = "drop")

print(survey_improvement)
```

## Best Practices for Tidy Data

### Documentation and Naming

```{r}
#| label: best-practices

# Good practices for maintaining tidy data
cat("BEST PRACTICES FOR TIDY DATA\n")
cat("============================\n\n")

cat("1. CONSISTENT NAMING CONVENTIONS:\n")
cat("   ✓ Use snake_case for column names\n")
cat("   ✓ Use descriptive, unambiguous names\n")
cat("   ✓ Avoid spaces, special characters\n")
cat("   ✓ Be consistent with abbreviations\n\n")

# Example of good naming
good_names_example <- tibble(
  customer_id = c("C001", "C002"),
  order_date = ymd(c("2024-01-15", "2024-01-16")),
  product_category = c("Electronics", "Clothing"),
  unit_price_usd = c(299.99, 49.99),
  quantity_ordered = c(1, 2),
  discount_percent = c(0.1, 0.0)
)

cat("GOOD NAMING EXAMPLE:\n")
print(good_names_example)

cat("\n2. DATA TYPES:\n")
cat("   ✓ Use appropriate data types (dates as Date, numbers as numeric)\n")
cat("   ✓ Use factors for categorical data with known levels\n")
cat("   ✓ Use logical (TRUE/FALSE) for binary variables\n")
cat("   ✓ Document any special coding (NA, missing values)\n\n")

cat("3. DOCUMENTATION:\n")
cat("   ✓ Document your tidying process\n")
cat("   ✓ Keep original data unchanged\n")
cat("   ✓ Comment complex transformations\n")
cat("   ✓ Include data dictionaries\n\n")

cat("4. VALIDATION:\n")
cat("   ✓ Check row counts before/after transformations\n")
cat("   ✓ Verify no data was lost or duplicated\n")
cat("   ✓ Test with edge cases and missing data\n")
cat("   ✓ Compare aggregated totals before/after\n")
```

### Common Pitfalls and How to Avoid Them

```{r}
#| label: common-pitfalls

cat("COMMON PITFALLS IN DATA TIDYING\n")
cat("===============================\n\n")

# Pitfall 1: Losing data during transformation
cat("PITFALL 1: LOSING DATA DURING TRANSFORMATION\n")
example_data <- tibble(
  id = 1:3,
  value1 = c(10, 20, 30),
  value2 = c(15, 25, 35)
)

# Bad: Forgetting to check row counts
before_rows <- nrow(example_data)
transformed <- example_data %>% pivot_longer(cols = -id, names_to = "variable", values_to = "value")
after_rows <- nrow(transformed)

cat("Original rows:", before_rows, "\n")
cat("After transformation:", after_rows, "\n")
cat("✓ GOOD: Always verify row counts make sense\n\n")

# Pitfall 2: Mixing data types
cat("PITFALL 2: MIXING DATA TYPES\n")
mixed_types <- tibble(
  metric = c("revenue", "count", "rate"),
  value = c("1000", "50", "0.05")  # All stored as character!
)

cat("✗ BAD: All values stored as character\n")
print(mixed_types)

# Better approach
proper_types <- tibble(
  metric = c("revenue", "count", "rate"),
  value = c(1000, 50, 0.05),
  unit = c("USD", "items", "proportion")
)

cat("✓ GOOD: Proper data types with units documented\n")
print(proper_types)

cat("\nPITFALL 3: OVER-TIDYING\n")
cat("✗ BAD: Making data so long it's hard to understand\n")
cat("✓ GOOD: Keep related variables together when it makes sense\n")
cat("✓ GOOD: Consider your analysis needs when choosing structure\n\n")

cat("PITFALL 4: NOT HANDLING MISSING VALUES\n")
cat("✗ BAD: Ignoring how NAs will behave in transformations\n")
cat("✓ GOOD: Explicitly handle missing values with coalesce(), replace_na()\n")
cat("✓ GOOD: Document what missing values mean\n")
```

## Exercises

### Exercise 1: Diagnostic Practice
Given these messy datasets, identify what makes them untidy:
1. A table with years as column headers
2. A survey with pre_post measurements combined
3. A contact list with multiple phone numbers in one cell
4. A gradebook mixing student and course information

### Exercise 2: Planning Transformations
For each messy dataset, write out:
1. What the tidy version should look like
2. Which tidyr functions you'll need
3. The order of operations
4. How to verify the transformation worked

### Exercise 3: Multi-Step Tidying
Practice with complex datasets that require:
1. Multiple pivot operations
2. Separation and recombination
3. Handling missing values
4. Type conversions

### Exercise 4: Real-World Scenarios
Work with realistic messy data:
1. Financial statements from spreadsheets
2. Survey data with encoded responses
3. Log files with mixed formats
4. Scientific data with experimental conditions in names

## Summary

Understanding tidy data principles is fundamental to effective data analysis:

### The Three Rules:
1. **Each variable forms a column**
2. **Each observation forms a row**
3. **Each type of observational unit forms a table**

### Common Problems:
- **Headers as values**: Years, treatments, conditions in column names
- **Multiple variables in names**: Combined information that should be separated
- **Multiple values in cells**: Lists or combined data in single cells
- **Mixed observational units**: Different types of entities in one table

### Diagnostic Process:
1. **Examine the structure**: Look at dimensions and column names
2. **Identify patterns**: Spot repeated elements and encoding
3. **Plan transformations**: Map current state to desired tidy state
4. **Execute step-by-step**: Transform incrementally and verify
5. **Validate results**: Check row counts and data integrity

### Best Practices:
- **Plan before you pivot**: Understand your target structure
- **Document your process**: Comment complex transformations
- **Validate each step**: Check that data is preserved
- **Use consistent naming**: Follow naming conventions
- **Handle missing data explicitly**: Don't let NAs surprise you

Mastering tidy data principles provides the foundation for all other data analysis work. When your data is tidy, analysis becomes straightforward and intuitive!

Next: **[Pivoting Data](pivoting-data.qmd)**