---
title: "Advanced dplyr Techniques"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Beyond the Basics: Advanced dplyr

Once you've mastered the core dplyr verbs, it's time to explore advanced techniques that will make your code more efficient, reusable, and elegant. This section covers programming with dplyr, performance optimization, and sophisticated data manipulation patterns.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(lubridate)

# Create a comprehensive dataset for demonstrating advanced techniques
set.seed(123)
n_transactions <- 5000

transactions <- tibble(
  transaction_id = 1:n_transactions,
  customer_id = sample(paste0("CUST-", str_pad(1:500, 3, pad = "0")), n_transactions, replace = TRUE),
  product_id = sample(paste0("PROD-", str_pad(1:100, 3, pad = "0")), n_transactions, replace = TRUE),
  store_id = sample(paste0("STORE-", 1:20), n_transactions, replace = TRUE),
  timestamp = sample(seq(ymd_hms("2024-01-01 00:00:00"), ymd_hms("2024-12-31 23:59:59"), by = "hour"), n_transactions),
  quantity = sample(1:10, n_transactions, replace = TRUE, prob = c(0.4, 0.3, 0.15, 0.08, 0.04, 0.02, 0.005, 0.005, 0.005, 0.005)),
  unit_price = round(rlnorm(n_transactions, meanlog = 3, sdlog = 1), 2),
  discount_rate = sample(c(0, 0.05, 0.10, 0.15, 0.20, 0.25), n_transactions, replace = TRUE, prob = c(0.6, 0.2, 0.1, 0.05, 0.03, 0.02)),
  payment_method = sample(c("Credit Card", "Cash", "Debit Card", "Mobile Payment"), n_transactions, replace = TRUE, prob = c(0.4, 0.25, 0.2, 0.15)),
  customer_segment = sample(c("Regular", "Premium", "VIP"), n_transactions, replace = TRUE, prob = c(0.7, 0.25, 0.05))
) %>%
  mutate(
    date = as_date(timestamp),
    hour = hour(timestamp),
    day_of_week = wday(date, label = TRUE),
    month = month(date, label = TRUE),
    quarter = quarter(date),
    gross_amount = quantity * unit_price,
    discount_amount = gross_amount * discount_rate,
    net_amount = gross_amount - discount_amount
  )

cat("Created dataset with", format(nrow(transactions), big.mark = ","), "transactions\n")
cat("Date range:", as.character(min(transactions$date)), "to", as.character(max(transactions$date)), "\n")
cat("Total revenue: $", format(sum(transactions$net_amount), big.mark = ",", nsmall = 2), "\n")
```

## The across() Function

`across()` allows you to apply operations to multiple columns simultaneously:

```{r}
#| label: across-basics

# Summarize multiple columns at once
summary_stats <- transactions %>%
  summarise(
    across(
      c(quantity, unit_price, gross_amount, discount_rate, net_amount),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE),
        min = ~min(.x, na.rm = TRUE),
        max = ~max(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(
    everything(),
    names_to = c("variable", "stat"),
    names_sep = "_(?=mean|median|sd|min|max)",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = stat, values_from = value)

print(summary_stats)

# Apply transformations to multiple columns
transactions_scaled <- transactions %>%
  mutate(
    across(
      c(quantity, gross_amount, net_amount),
      list(
        log = ~log1p(.x),
        sqrt = ~sqrt(.x),
        z_score = ~(.x - mean(.x)) / sd(.x)
      ),
      .names = "{.col}_{.fn}"
    )
  )

cat("\nCreated", ncol(transactions_scaled) - ncol(transactions), "new transformed variables\n")
```

## Column Selection Helpers

dplyr provides powerful helpers for selecting columns:

```{r}
#| label: selection-helpers

# Various selection methods
selection_examples <- transactions %>%
  select(
    # Specific columns
    transaction_id, customer_id,
    
    # Range of columns
    quantity:discount_rate,
    
    # Columns matching patterns
    starts_with("customer"),
    ends_with("amount"),
    contains("price"),
    
    # Columns by type
    where(is.numeric),
    
    # Exclude columns
    -matches("^(timestamp|date)$")
  )

# Using where() with custom functions
high_variance_cols <- transactions %>%
  select(where(~is.numeric(.x) && var(.x, na.rm = TRUE) > 100))

cat("Selected columns with high variance:\n")
names(high_variance_cols)

# Complex selections with multiple conditions
financial_cols <- transactions %>%
  select(
    transaction_id,
    where(~is.numeric(.x) && any(str_detect(names(.), "amount|price|discount")))
  )

cat("\n\nFinancial columns:", paste(names(financial_cols), collapse = ", "))
```

## Programming with dplyr

Creating reusable functions with dplyr requires understanding tidy evaluation:

```{r}
#| label: programming-basics

# Function to analyze any grouping variable
analyze_by_category <- function(data, group_col, value_col) {
  data %>%
    group_by({{ group_col }}) %>%
    summarise(
      count = n(),
      total = sum({{ value_col }}, na.rm = TRUE),
      avg = mean({{ value_col }}, na.rm = TRUE),
      median = median({{ value_col }}, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      percentage = total / sum(total) * 100,
      cumulative_pct = cumsum(percentage)
    ) %>%
    arrange(desc(total))
}

# Use the function with different variables
by_segment <- analyze_by_category(transactions, customer_segment, net_amount)
by_payment <- analyze_by_category(transactions, payment_method, net_amount)
by_store <- analyze_by_category(transactions, store_id, net_amount) %>% head(10)

cat("Analysis by customer segment:\n")
print(by_segment)

cat("\n\nAnalysis by payment method:\n")
print(by_payment)
```

## Advanced Programming Patterns

```{r}
#| label: advanced-programming

# Function with dynamic column names
create_kpi_summary <- function(data, ..., prefix = "kpi") {
  group_vars <- enquos(...)
  
  data %>%
    group_by(!!!group_vars) %>%
    summarise(
      "{prefix}_transactions" := n(),
      "{prefix}_revenue" := sum(net_amount),
      "{prefix}_units" := sum(quantity),
      "{prefix}_avg_basket" := mean(net_amount),
      "{prefix}_unique_customers" := n_distinct(customer_id),
      "{prefix}_unique_products" := n_distinct(product_id),
      .groups = "drop"
    )
}

# Generate KPIs for different time periods
monthly_kpis <- transactions %>%
  create_kpi_summary(month, prefix = "monthly")

quarterly_kpis <- transactions %>%
  create_kpi_summary(quarter, customer_segment, prefix = "q")

cat("Monthly KPIs (first 6 months):\n")
print(head(monthly_kpis))

# Function that accepts string inputs
summarize_columns <- function(data, cols, funs) {
  data %>%
    summarise(
      across(all_of(cols), funs, .names = "{.col}_{.fn}")
    )
}

numeric_summary <- transactions %>%
  summarize_columns(
    c("quantity", "net_amount"), 
    list(total = sum, avg = mean, sd = sd)
  )

print(numeric_summary)
```

## Row-wise Operations

Sometimes you need to perform operations across rows:

```{r}
#| label: rowwise-operations

# Create a dataset with multiple score columns
performance_data <- tibble(
  employee_id = paste0("EMP-", str_pad(1:100, 3, pad = "0")),
  q1_score = runif(100, 60, 100),
  q2_score = runif(100, 65, 100),
  q3_score = runif(100, 70, 100),
  q4_score = runif(100, 68, 100),
  bonus_points = sample(0:10, 100, replace = TRUE)
)

# Row-wise calculations
performance_summary <- performance_data %>%
  rowwise() %>%
  mutate(
    avg_score = mean(c(q1_score, q2_score, q3_score, q4_score)),
    min_score = min(c(q1_score, q2_score, q3_score, q4_score)),
    max_score = max(c(q1_score, q2_score, q3_score, q4_score)),
    total_score = sum(c(q1_score, q2_score, q3_score, q4_score, bonus_points)),
    score_variance = var(c(q1_score, q2_score, q3_score, q4_score))
  ) %>%
  ungroup() %>%
  mutate(
    performance_category = case_when(
      avg_score >= 90 ~ "Outstanding",
      avg_score >= 80 ~ "Excellent",
      avg_score >= 70 ~ "Good",
      TRUE ~ "Needs Improvement"
    )
  )

# Summary by performance category
performance_summary %>%
  count(performance_category) %>%
  mutate(percentage = n / sum(n) * 100)

# Alternative using c_across()
performance_alt <- performance_data %>%
  rowwise() %>%
  mutate(
    avg_score = mean(c_across(q1_score:q4_score)),
    score_range = max(c_across(q1_score:q4_score)) - min(c_across(q1_score:q4_score))
  ) %>%
  ungroup()
```

## Complex Conditional Logic

Advanced use of case_when() and if_else():

```{r}
#| label: complex-conditionals

# Multi-condition categorization
transactions_categorized <- transactions %>%
  mutate(
    # Complex time-based categorization
    time_category = case_when(
      hour >= 6 & hour < 12 ~ "Morning",
      hour >= 12 & hour < 17 ~ "Afternoon", 
      hour >= 17 & hour < 21 ~ "Evening",
      TRUE ~ "Night"
    ),
    
    # Nested conditions
    purchase_type = case_when(
      quantity == 1 & net_amount < 20 ~ "Small Single Item",
      quantity == 1 & net_amount >= 20 ~ "Large Single Item",
      quantity > 1 & quantity <= 5 & net_amount < 50 ~ "Small Basket",
      quantity > 1 & quantity <= 5 & net_amount >= 50 ~ "Medium Basket",
      quantity > 5 ~ "Bulk Purchase",
      TRUE ~ "Other"
    ),
    
    # Complex business rules
    customer_value = case_when(
      customer_segment == "VIP" ~ "High Value",
      customer_segment == "Premium" & net_amount > 100 ~ "High Value",
      customer_segment == "Premium" ~ "Medium Value",
      net_amount > 200 ~ "Potential High Value",
      TRUE ~ "Standard"
    ),
    
    # Seasonal pricing strategy
    pricing_season = case_when(
      month %in% c("Nov", "Dec") ~ "Holiday Premium",
      month %in% c("Jun", "Jul", "Aug") ~ "Summer Sale",
      month %in% c("Jan", "Feb") & discount_rate > 0.15 ~ "New Year Clearance",
      day_of_week %in% c("Sat", "Sun") & time_category == "Morning" ~ "Weekend Special",
      TRUE ~ "Regular Pricing"
    )
  )

# Analyze the categorizations
cat("Purchase type distribution:\n")
transactions_categorized %>%
  count(purchase_type, sort = TRUE) %>%
  mutate(percentage = n / sum(n) * 100)
```

## Performance Optimization

Tips for making dplyr operations faster:

```{r}
#| label: performance-optimization

# 1. Use data.table backend for large datasets
# library(dtplyr)
# transactions_dt <- lazy_dt(transactions)

# 2. Minimize grouping operations
# Inefficient: multiple group_by operations
start_time <- Sys.time()
result_slow <- transactions %>%
  group_by(store_id) %>%
  mutate(store_total = sum(net_amount)) %>%
  ungroup() %>%
  group_by(customer_id) %>%
  mutate(customer_total = sum(net_amount)) %>%
  ungroup()
time_slow <- Sys.time() - start_time

# Efficient: calculate summaries separately and join
start_time <- Sys.time()
store_totals <- transactions %>%
  group_by(store_id) %>%
  summarise(store_total = sum(net_amount), .groups = "drop")

customer_totals <- transactions %>%
  group_by(customer_id) %>%
  summarise(customer_total = sum(net_amount), .groups = "drop")

result_fast <- transactions %>%
  left_join(store_totals, by = "store_id") %>%
  left_join(customer_totals, by = "customer_id")
time_fast <- Sys.time() - start_time

cat("Performance comparison:\n")
cat("Multiple group_by:", format(time_slow), "\n")
cat("Separate summaries:", format(time_fast), "\n")
cat("Speedup:", round(as.numeric(time_slow) / as.numeric(time_fast), 2), "x\n")

# 3. Use vectorized operations
# Slow: row-wise string operations
slow_categorize <- function(df) {
  df %>%
    rowwise() %>%
    mutate(
      description = paste(customer_segment, "customer bought", quantity, 
                         "items via", payment_method)
    ) %>%
    ungroup()
}

# Fast: vectorized operations
fast_categorize <- function(df) {
  df %>%
    mutate(
      description = str_c(customer_segment, " customer bought ", quantity, 
                         " items via ", payment_method)
    )
}

# Compare a subset
small_trans <- head(transactions, 1000)
time_slow <- system.time(slow_result <- slow_categorize(small_trans))
time_fast <- system.time(fast_result <- fast_categorize(small_trans))

cat("\nString operation performance (1000 rows):\n")
cat("Row-wise:", time_slow["elapsed"], "seconds\n")
cat("Vectorized:", time_fast["elapsed"], "seconds\n")
```

## Advanced Joining Techniques

```{r}
#| label: advanced-joins

# Rolling joins (finding the nearest match)
price_changes <- tibble(
  product_id = rep(paste0("PROD-", str_pad(1:10, 3, pad = "0")), each = 4),
  effective_date = rep(c(ymd("2024-01-01"), ymd("2024-04-01"), 
                        ymd("2024-07-01"), ymd("2024-10-01")), 10),
  price = round(runif(40, 10, 100), 2)
)

# Function to find applicable price for each transaction
find_applicable_price <- function(trans_data, price_data) {
  trans_data %>%
    left_join(
      price_data %>%
        rename(price_date = effective_date),
      by = "product_id"
    ) %>%
    filter(date >= price_date) %>%
    group_by(transaction_id) %>%
    filter(price_date == max(price_date)) %>%
    ungroup() %>%
    select(-price_date)
}

# Apply to a subset of transactions
sample_trans <- transactions %>%
  filter(product_id %in% paste0("PROD-", str_pad(1:10, 3, pad = "0"))) %>%
  slice_sample(n = 100)

trans_with_prices <- find_applicable_price(sample_trans, price_changes)

cat("Sample of transactions with historical prices:\n")
trans_with_prices %>%
  select(transaction_id, date, product_id, unit_price, price) %>%
  rename(transaction_price = unit_price, catalog_price = price) %>%
  slice_head(n = 5)
```

## Creating Custom Verb Functions

```{r}
#| label: custom-verbs

# Create a custom verb for common operations
add_time_features <- function(data, date_col) {
  date_col <- enquo(date_col)
  
  data %>%
    mutate(
      year = year(!!date_col),
      quarter = quarter(!!date_col),
      month = month(!!date_col),
      week = week(!!date_col),
      day_of_month = day(!!date_col),
      day_of_week = wday(!!date_col, label = TRUE),
      is_weekend = wday(!!date_col) %in% c(1, 7),
      is_month_start = day(!!date_col) <= 7,
      is_month_end = day(!!date_col) >= day(ceiling_date(!!date_col, "month") - 1) - 6
    )
}

# Create business metric calculations
calculate_business_metrics <- function(data) {
  data %>%
    mutate(
      margin = (net_amount - (net_amount * 0.7)) / net_amount,  # Assume 30% margin
      margin_amount = net_amount * margin,
      revenue_per_unit = net_amount / quantity,
      discount_impact = discount_amount / gross_amount * 100
    )
}

# Chain custom functions
enhanced_transactions <- transactions %>%
  slice_sample(n = 1000) %>%
  add_time_features(date) %>%
  calculate_business_metrics() %>%
  select(transaction_id, date, is_weekend, margin, revenue_per_unit)

cat("Enhanced transactions with custom features:\n")
print(head(enhanced_transactions))
```

## Advanced Reshaping Patterns

```{r}
#| label: advanced-reshaping

# Complex pivot with multiple value columns
monthly_metrics <- transactions %>%
  mutate(year_month = floor_date(date, "month")) %>%
  group_by(year_month, customer_segment) %>%
  summarise(
    transactions = n(),
    revenue = sum(net_amount),
    units = sum(quantity),
    avg_basket = mean(net_amount),
    .groups = "drop"
  )

# Pivot with custom names
wide_metrics <- monthly_metrics %>%
  pivot_wider(
    names_from = customer_segment,
    values_from = c(transactions, revenue, units, avg_basket),
    names_glue = "{customer_segment}_{.value}"
  )

cat("Wide format with custom column names:\n")
names(wide_metrics)

# Nested data structures
nested_analysis <- transactions %>%
  group_by(store_id) %>%
  nest() %>%
  mutate(
    # Perform analysis on each nested dataset
    summary = map(data, ~{
      .x %>%
        summarise(
          n_trans = n(),
          revenue = sum(net_amount),
          top_product = names(sort(table(.x$product_id), decreasing = TRUE)[1]),
          peak_hour = as.numeric(names(sort(table(.x$hour), decreasing = TRUE)[1]))
        )
    }),
    # Extract specific metrics
    total_revenue = map_dbl(summary, ~.x$revenue),
    transaction_count = map_int(summary, ~.x$n_trans)
  ) %>%
  arrange(desc(total_revenue))

cat("\nTop 5 stores by revenue:\n")
nested_analysis %>%
  select(store_id, total_revenue, transaction_count) %>%
  head(5)
```

## Error Handling and Validation

```{r}
#| label: error-handling

# Safe summary function with validation
safe_summarize <- function(data, group_var, value_var) {
  group_var <- enquo(group_var)
  value_var <- enquo(value_var)
  
  # Validate inputs
  if (!quo_name(value_var) %in% names(data)) {
    stop("Value variable not found in data")
  }
  
  if (!is.numeric(data %>% pull(!!value_var))) {
    warning("Value variable is not numeric, converting...")
    data <- data %>%
      mutate(!!value_var := as.numeric(!!value_var))
  }
  
  # Perform analysis with error handling
  tryCatch({
    data %>%
      group_by(!!group_var) %>%
      summarise(
        n = n(),
        mean = mean(!!value_var, na.rm = TRUE),
        median = median(!!value_var, na.rm = TRUE),
        sd = sd(!!value_var, na.rm = TRUE),
        missing = sum(is.na(!!value_var)),
        .groups = "drop"
      ) %>%
      mutate(
        cv = sd / mean,  # Coefficient of variation
        data_quality = case_when(
          missing / n > 0.1 ~ "Poor",
          cv > 1 ~ "High Variability", 
          TRUE ~ "Good"
        )
      )
  }, error = function(e) {
    message("Error in summarization: ", e$message)
    return(NULL)
  })
}

# Test the function
result <- safe_summarize(transactions, store_id, net_amount)
cat("Safe summarization completed successfully\n")
```

## Best Practices Summary

1. **Use across() for multiple columns**: More concise than multiple mutate calls
2. **Leverage selection helpers**: starts_with(), ends_with(), contains(), where()
3. **Program with {{}}**: For creating reusable functions
4. **Optimize performance**: Minimize groups, use vectorized operations
5. **Handle errors gracefully**: Validate inputs and use tryCatch()
6. **Document your functions**: Clear names and comments
7. **Test with small data first**: Develop on subsets before scaling up

## Practice Exercises

1. **Custom Analysis Function**: Create a function that performs a complete analysis pipeline
2. **Performance Comparison**: Compare different approaches to the same problem
3. **Complex Reshaping**: Transform nested transaction data into a report format
4. **Error-Robust Pipeline**: Build a data processing pipeline that handles various data quality issues

## Conclusion

Advanced dplyr techniques enable you to write more efficient, maintainable, and powerful data manipulation code. The key is to:

- Understand tidy evaluation for programming
- Use vectorized operations whenever possible
- Create reusable functions for common patterns
- Optimize for performance with large datasets
- Handle edge cases and errors gracefully

These advanced skills will make you a more effective data analyst and allow you to tackle complex real-world data challenges with confidence.