---
title: "Filtering and Selecting Data"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to Data Subsetting

Filtering and selecting are fundamental operations in data analysis. The `filter()` function chooses which rows to keep based on conditions, while `select()` chooses which columns to keep. These are often the first steps in any data analysis pipeline.

```{r}
#| label: setup
#| message: false

library(tidyverse)

# Load the sample dataset from Module 5 index
set.seed(123)
sales_raw <- tibble(
  order_id = paste0("ORD-", str_pad(1:1000, 4, pad = "0")),
  customer_id = sample(paste0("CUST-", 1:200), 1000, replace = TRUE),
  product_id = sample(paste0("PROD-", 1:50), 1000, replace = TRUE),
  order_date = sample(seq(ymd("2023-01-01"), ymd("2024-12-31"), by = "day"), 1000, replace = TRUE),
  quantity = sample(1:10, 1000, replace = TRUE),
  unit_price = round(runif(1000, 5, 200), 2),
  discount_pct = sample(c(0, 0.05, 0.10, 0.15, 0.20), 1000, replace = TRUE, prob = c(0.6, 0.2, 0.1, 0.05, 0.05)),
  sales_rep = sample(c("Alice Johnson", "Bob Smith", "Charlie Brown", "Diana Prince", "Eve Wilson"), 1000, replace = TRUE),
  region = sample(c("North", "South", "East", "West"), 1000, replace = TRUE),
  customer_type = sample(c("Individual", "Small Business", "Enterprise"), 1000, replace = TRUE, prob = c(0.5, 0.3, 0.2))
) %>%
  mutate(
    sales_rep = ifelse(row_number() %in% sample(1:1000, 50), NA, sales_rep),
    discount_pct = ifelse(row_number() %in% sample(1:1000, 30), NA, discount_pct),
    gross_amount = quantity * unit_price,
    discount_amount = gross_amount * coalesce(discount_pct, 0),
    net_amount = gross_amount - discount_amount,
    unit_price = ifelse(row_number() %in% sample(1:1000, 5), unit_price * 100, unit_price),
    quantity = ifelse(row_number() %in% sample(1:1000, 3), 0, quantity)
  )

cat("Working with", nrow(sales_raw), "sales transactions\n")
cat("Columns:", ncol(sales_raw), "variables\n")
```

## Filtering Rows with filter()

### Basic Filtering

```{r}
#| label: basic-filtering

# Single condition
large_orders <- sales_raw %>%
  filter(net_amount > 1000)

cat("Large orders (>$1000):", nrow(large_orders), "transactions\n")

# Multiple conditions with AND (comma-separated or &)
premium_orders <- sales_raw %>%
  filter(
    net_amount > 500,
    customer_type == "Enterprise",
    quantity >= 5
  )

cat("Premium enterprise orders:", nrow(premium_orders), "transactions\n")

# OR conditions
discount_or_large <- sales_raw %>%
  filter(discount_pct > 0.1 | net_amount > 800)

cat("High discount or large orders:", nrow(discount_or_large), "transactions\n")

# NOT conditions
no_discount <- sales_raw %>%
  filter(discount_pct == 0 | is.na(discount_pct))

cat("Orders without discount:", nrow(no_discount), "transactions\n")
```

### Working with Dates

```{r}
#| label: date-filtering

# Recent orders (last 6 months)
recent_cutoff <- ymd("2024-06-01")
recent_orders <- sales_raw %>%
  filter(order_date >= recent_cutoff)

cat("Recent orders (since June 2024):", nrow(recent_orders), "transactions\n")

# Specific date ranges
q4_2024 <- sales_raw %>%
  filter(
    order_date >= ymd("2024-10-01"),
    order_date <= ymd("2024-12-31")
  )

cat("Q4 2024 orders:", nrow(q4_2024), "transactions\n")

# Using between() for ranges
summer_2024 <- sales_raw %>%
  filter(between(order_date, ymd("2024-06-01"), ymd("2024-08-31")))

cat("Summer 2024 orders:", nrow(summer_2024), "transactions\n")

# Filtering by day of week or month
weekend_orders <- sales_raw %>%
  filter(wday(order_date, label = TRUE) %in% c("Sat", "Sun"))

december_orders <- sales_raw %>%
  filter(month(order_date) == 12)

cat("Weekend orders:", nrow(weekend_orders), "\n")
cat("December orders:", nrow(december_orders), "\n")
```

### String Filtering

```{r}
#| label: string-filtering

# Exact matches
alice_orders <- sales_raw %>%
  filter(sales_rep == "Alice Johnson")

# Pattern matching with str_detect()
johnson_reps <- sales_raw %>%
  filter(str_detect(sales_rep, "Johnson"))

# Case-insensitive patterns
alice_any_case <- sales_raw %>%
  filter(str_detect(str_to_lower(sales_rep), "alice"))

# Multiple string patterns
senior_reps <- sales_raw %>%
  filter(sales_rep %in% c("Alice Johnson", "Diana Prince"))

# String starts with or ends with
north_products <- sales_raw %>%
  filter(str_starts(product_id, "PROD-1"))

cat("Alice's orders:", nrow(alice_orders), "\n")
cat("Johnson family reps:", nrow(johnson_reps), "\n")
cat("Senior reps orders:", nrow(senior_reps), "\n")
```

### Handling Missing Values

```{r}
#| label: missing-value-filtering

# Remove rows with missing sales rep
complete_sales_rep <- sales_raw %>%
  filter(!is.na(sales_rep))

# Only rows with missing values
missing_sales_rep <- sales_raw %>%
  filter(is.na(sales_rep))

# Complete cases only (no missing values in any column)
complete_cases <- sales_raw %>%
  filter(complete.cases(.))

# Rows with any missing values
any_missing <- sales_raw %>%
  filter(!complete.cases(.))

cat("Orders with sales rep assigned:", nrow(complete_sales_rep), "\n")
cat("Orders missing sales rep:", nrow(missing_sales_rep), "\n")
cat("Complete cases (no missing data):", nrow(complete_cases), "\n")
cat("Cases with any missing data:", nrow(any_missing), "\n")
```

### Advanced Filtering Techniques

```{r}
#| label: advanced-filtering

# Using %in% for multiple values
target_regions <- c("North", "East")
target_region_orders <- sales_raw %>%
  filter(region %in% target_regions)

# Complex logical conditions
complex_filter <- sales_raw %>%
  filter(
    (customer_type == "Enterprise" & net_amount > 500) |
    (customer_type == "Small Business" & net_amount > 200) |
    (customer_type == "Individual" & net_amount > 100)
  )

# Filtering based on ranking
top_10_percent <- sales_raw %>%
  filter(net_amount >= quantile(net_amount, 0.9, na.rm = TRUE))

# Outlier detection and removal
q1 <- quantile(sales_raw$net_amount, 0.25, na.rm = TRUE)
q3 <- quantile(sales_raw$net_amount, 0.75, na.rm = TRUE)
iqr <- q3 - q1
outlier_bounds <- c(q1 - 1.5 * iqr, q3 + 1.5 * iqr)

no_outliers <- sales_raw %>%
  filter(between(net_amount, outlier_bounds[1], outlier_bounds[2]))

cat("Target regions:", nrow(target_region_orders), "orders\n")
cat("Complex filter:", nrow(complex_filter), "orders\n")
cat("Top 10% by value:", nrow(top_10_percent), "orders\n")
cat("After removing outliers:", nrow(no_outliers), "orders\n")
```

## Selecting Columns with select()

### Basic Column Selection

```{r}
#| label: basic-selecting

# Select specific columns
basic_info <- sales_raw %>%
  select(order_id, customer_id, order_date, net_amount)

# Select columns by position
first_five <- sales_raw %>%
  select(1:5)

# Select all except certain columns
no_internals <- sales_raw %>%
  select(-gross_amount, -discount_amount)

# Reorder columns
reordered <- sales_raw %>%
  select(order_id, order_date, customer_id, region, everything())

cat("Basic info columns:", ncol(basic_info), "\n")
cat("First five columns:", ncol(first_five), "\n")
cat("Without internal calculations:", ncol(no_internals), "\n")
cat("Reordered columns:", paste(names(reordered)[1:6], collapse = ", "), "...\n")
```

### Helper Functions for Selection

```{r}
#| label: selection-helpers

# Select columns by name pattern
id_columns <- sales_raw %>%
  select(ends_with("_id"))

# Select by data type
numeric_columns <- sales_raw %>%
  select(where(is.numeric))

character_columns <- sales_raw %>%
  select(where(is.character))

# Select columns containing certain text
amount_columns <- sales_raw %>%
  select(contains("amount"))

# Select columns matching regex
date_and_id <- sales_raw %>%
  select(matches("(date|_id)"))

# Multiple helper functions
key_metrics <- sales_raw %>%
  select(starts_with("order"), contains("amount"), region, customer_type)

cat("ID columns:", ncol(id_columns), "variables\n")
cat("Numeric columns:", ncol(numeric_columns), "variables\n")
cat("Character columns:", ncol(character_columns), "variables\n")
cat("Amount columns:", ncol(amount_columns), "variables\n")
cat("Key metrics:", ncol(key_metrics), "variables\n")

# Show the helper function results
cat("\nID columns:", paste(names(id_columns), collapse = ", "), "\n")
cat("Amount columns:", paste(names(amount_columns), collapse = ", "), "\n")
```

### Renaming During Selection

```{r}
#| label: rename-select

# Rename while selecting
renamed_selection <- sales_raw %>%
  select(
    order = order_id,
    customer = customer_id,
    date = order_date,
    value = net_amount,
    rep = sales_rep
  )

# Rename with everything()
cleaner_names <- sales_raw %>%
  select(
    order_number = order_id,
    customer_number = customer_id,
    everything()
  )

cat("Renamed selection columns:\n")
print(names(renamed_selection))

cat("\nCleaner names (first 8 columns):\n")
print(names(cleaner_names)[1:8])
```

## Combining filter() and select()

### Common Patterns

```{r}
#| label: filter-select-combo

# Filter then select (most common)
recent_summary <- sales_raw %>%
  filter(
    order_date >= ymd("2024-01-01"),
    !is.na(sales_rep)
  ) %>%
  select(order_id, order_date, sales_rep, region, net_amount, customer_type)

# Select then filter (less common, but sometimes useful)
high_value_basics <- sales_raw %>%
  select(order_id, customer_id, net_amount, customer_type) %>%
  filter(net_amount > 500)

# Complex combination with multiple steps
analysis_ready <- sales_raw %>%
  # Remove problematic data
  filter(
    quantity > 0,                    # No zero quantities
    unit_price <= 300,               # Remove pricing errors
    !is.na(sales_rep)                # Must have assigned rep
  ) %>%
  # Select relevant columns
  select(
    order_id,
    customer_id,
    order_date,
    quantity,
    unit_price,
    discount_pct,
    net_amount,
    sales_rep,
    region,
    customer_type
  ) %>%
  # Add derived columns for analysis
  mutate(
    order_month = floor_date(order_date, "month"),
    is_discounted = !is.na(discount_pct) & discount_pct > 0,
    order_size = case_when(
      net_amount >= 1000 ~ "Large",
      net_amount >= 300 ~ "Medium",
      TRUE ~ "Small"
    )
  )

cat("Recent summary:", nrow(recent_summary), "rows,", ncol(recent_summary), "columns\n")
cat("Analysis ready:", nrow(analysis_ready), "rows,", ncol(analysis_ready), "columns\n")

# Show the first few rows of analysis-ready data
print(head(analysis_ready, 5))
```

### Working with Grouped Data

```{r}
#| label: grouped-filtering

# Filter within groups
top_orders_by_region <- sales_raw %>%
  group_by(region) %>%
  filter(net_amount >= quantile(net_amount, 0.9, na.rm = TRUE)) %>%
  ungroup()

# Select and filter with grouping
regional_summaries <- sales_raw %>%
  filter(!is.na(sales_rep)) %>%
  group_by(region, sales_rep) %>%
  select(region, sales_rep, net_amount, quantity) %>%
  summarise(
    total_sales = sum(net_amount),
    total_orders = n(),
    avg_order_value = mean(net_amount),
    .groups = "drop"
  )

cat("Top orders by region:", nrow(top_orders_by_region), "orders\n")
cat("Regional summaries:", nrow(regional_summaries), "rep-region combinations\n")

print(head(regional_summaries))
```

## Advanced Filtering Techniques

### Using filter() with Window Functions

```{r}
#| label: window-filtering

# Top N orders per customer
top_orders_per_customer <- sales_raw %>%
  group_by(customer_id) %>%
  filter(rank(desc(net_amount)) <= 3) %>%
  arrange(customer_id, desc(net_amount)) %>%
  ungroup()

# Orders above regional average
above_regional_avg <- sales_raw %>%
  group_by(region) %>%
  filter(net_amount > mean(net_amount, na.rm = TRUE)) %>%
  ungroup()

# Recent orders within each customer
recent_per_customer <- sales_raw %>%
  group_by(customer_id) %>%
  filter(order_date >= max(order_date) - days(30)) %>%
  ungroup()

cat("Top 3 orders per customer:", nrow(top_orders_per_customer), "orders\n")
cat("Above regional average:", nrow(above_regional_avg), "orders\n")
cat("Recent per customer:", nrow(recent_per_customer), "orders\n")
```

### Conditional Filtering

```{r}
#| label: conditional-filtering

# Filter based on conditions in other columns
inconsistent_discounts <- sales_raw %>%
  filter(
    case_when(
      customer_type == "Enterprise" ~ discount_pct <= 0.15,
      customer_type == "Small Business" ~ discount_pct <= 0.10,
      customer_type == "Individual" ~ discount_pct <= 0.05,
      TRUE ~ TRUE
    )
  )

# Filter using if_else logic
weekend_enterprise <- sales_raw %>%
  filter(
    if_else(
      wday(order_date) %in% c(1, 7),  # Weekend
      customer_type == "Enterprise",   # Only enterprise on weekends
      TRUE                            # Any customer type on weekdays
    )
  )

cat("Consistent discount policies:", nrow(inconsistent_discounts), "orders\n")
cat("Weekend enterprise filter:", nrow(weekend_enterprise), "orders\n")
```

## Performance and Best Practices

### Efficient Filtering

```{r}
#| label: efficient-filtering

# Early filtering reduces computation
# Good: Filter early
efficient_pipeline <- sales_raw %>%
  filter(
    order_date >= ymd("2024-01-01"),
    !is.na(sales_rep),
    net_amount > 100
  ) %>%
  select(order_id, customer_id, order_date, net_amount, region) %>%
  mutate(
    quarter = quarter(order_date),
    high_value = net_amount > 500
  )

# Less efficient: Filter after expensive operations
# (We won't run this, but it shows the pattern to avoid)
# inefficient_pipeline <- sales_raw %>%
#   mutate(complex_calculation = net_amount * some_expensive_function()) %>%
#   left_join(large_table) %>%
#   filter(order_date >= ymd("2024-01-01"))

# Use specific conditions instead of complex ones
# Good: Specific numeric ranges
specific_range <- sales_raw %>%
  filter(between(net_amount, 100, 1000))

# Less good: Complex calculations in filter
# avoid_complex <- sales_raw %>%
#   filter(log(net_amount) > some_threshold)

cat("Efficient pipeline result:", nrow(efficient_pipeline), "rows\n")
cat("Specific range filter:", nrow(specific_range), "rows\n")
```

### Readable Filter Conditions

```{r}
#| label: readable-filters

# Create intermediate variables for complex conditions
high_value_threshold <- 1000
recent_date_cutoff <- ymd("2024-06-01")
target_customer_types <- c("Enterprise", "Small Business")

readable_filter <- sales_raw %>%
  filter(
    net_amount >= high_value_threshold,
    order_date >= recent_date_cutoff,
    customer_type %in% target_customer_types,
    !is.na(sales_rep)
  )

# Use meaningful helper functions
is_valid_order <- function(data) {
  data %>%
    filter(
      quantity > 0,
      unit_price > 0,
      unit_price <= 500,  # Remove pricing errors
      !is.na(sales_rep)
    )
}

valid_orders <- sales_raw %>%
  is_valid_order()

cat("Readable filter:", nrow(readable_filter), "orders\n")
cat("Valid orders:", nrow(valid_orders), "orders\n")
```

## Real-World Examples

### Example 1: Customer Segmentation Analysis

```{r}
#| label: customer-segmentation

# Identify different customer segments for targeted analysis
customer_segments <- sales_raw %>%
  filter(
    !is.na(sales_rep),
    quantity > 0,
    order_date >= ymd("2024-01-01")
  ) %>%
  group_by(customer_id) %>%
  summarise(
    total_orders = n(),
    total_spent = sum(net_amount),
    avg_order_value = mean(net_amount),
    first_order = min(order_date),
    last_order = max(order_date),
    customer_type = first(customer_type),
    .groups = "drop"
  ) %>%
  mutate(
    recency_days = as.numeric(Sys.Date() - last_order),
    customer_lifetime = as.numeric(last_order - first_order),
    segment = case_when(
      total_spent >= 5000 & recency_days <= 30 ~ "VIP Active",
      total_spent >= 5000 & recency_days > 30 ~ "VIP At Risk",
      total_spent >= 1000 & recency_days <= 60 ~ "Regular Active",
      total_spent >= 1000 & recency_days > 60 ~ "Regular At Risk",
      recency_days <= 30 ~ "New Active",
      TRUE ~ "Low Value"
    )
  )

# Filter for each segment
vip_customers <- customer_segments %>%
  filter(str_starts(segment, "VIP"))

at_risk_customers <- customer_segments %>%
  filter(str_ends(segment, "At Risk"))

# Segment summary
segment_summary <- customer_segments %>%
  count(segment, sort = TRUE) %>%
  mutate(percentage = round(n / sum(n) * 100, 1))

cat("Customer segmentation results:\n")
print(segment_summary)

cat("\nVIP customers:", nrow(vip_customers), "\n")
cat("At-risk customers:", nrow(at_risk_customers), "\n")
```

### Example 2: Sales Performance Analysis

```{r}
#| label: sales-performance

# Analyze sales rep performance with filters
rep_performance <- sales_raw %>%
  filter(
    !is.na(sales_rep),
    order_date >= ymd("2024-01-01"),
    quantity > 0,
    unit_price <= 300  # Remove pricing errors
  ) %>%
  group_by(sales_rep, region) %>%
  summarise(
    total_sales = sum(net_amount),
    order_count = n(),
    avg_order_value = mean(net_amount),
    unique_customers = n_distinct(customer_id),
    .groups = "drop"
  ) %>%
  mutate(
    performance_score = scale(total_sales)[,1] + scale(unique_customers)[,1],
    performance_tier = case_when(
      performance_score >= 1 ~ "Top Performer",
      performance_score >= 0 ~ "Good Performer",
      performance_score >= -1 ~ "Average Performer",
      TRUE ~ "Needs Improvement"
    )
  )

# Filter for different performance tiers
top_performers <- rep_performance %>%
  filter(performance_tier == "Top Performer")

underperformers <- rep_performance %>%
  filter(performance_tier == "Needs Improvement")

cat("Sales rep performance analysis:\n")
print(rep_performance %>% arrange(desc(performance_score)))

cat("\nTop performers:", nrow(top_performers), "\n")
cat("Need improvement:", nrow(underperformers), "\n")
```

### Example 3: Product Analysis

```{r}
#| label: product-analysis

# Analyze product performance across different dimensions
product_analysis <- sales_raw %>%
  filter(
    !is.na(sales_rep),
    quantity > 0,
    order_date >= ymd("2024-01-01")
  ) %>%
  group_by(product_id, region, customer_type) %>%
  summarise(
    units_sold = sum(quantity),
    revenue = sum(net_amount),
    orders = n(),
    avg_price = mean(unit_price),
    .groups = "drop"
  ) %>%
  group_by(product_id) %>%
  mutate(
    product_total_revenue = sum(revenue),
    regional_share = revenue / product_total_revenue
  ) %>%
  ungroup()

# Filter for top products
top_products <- product_analysis %>%
  group_by(product_id) %>%
  summarise(total_revenue = sum(revenue), .groups = "drop") %>%
  filter(total_revenue >= quantile(total_revenue, 0.8)) %>%
  pull(product_id)

top_product_details <- product_analysis %>%
  filter(product_id %in% top_products) %>%
  arrange(desc(revenue))

# Products that perform well in specific segments
enterprise_favorites <- product_analysis %>%
  filter(customer_type == "Enterprise") %>%
  group_by(product_id) %>%
  summarise(enterprise_revenue = sum(revenue), .groups = "drop") %>%
  filter(enterprise_revenue >= quantile(enterprise_revenue, 0.9))

cat("Product analysis results:\n")
cat("Top products (top 20%):", length(top_products), "products\n")
cat("Enterprise favorites (top 10%):", nrow(enterprise_favorites), "products\n")

print(head(top_product_details))
```

## Common Pitfalls and Solutions

### Missing Data Handling

```{r}
#| label: missing-data-pitfalls

# Pitfall: Forgetting about NA values in conditions
# This might not work as expected
# problematic_filter <- sales_raw %>%
#   filter(discount_pct > 0.1)  # Excludes NA values!

# Better: Explicitly handle NA values
explicit_na_handling <- sales_raw %>%
  filter(!is.na(discount_pct) & discount_pct > 0.1)

# Or include NA values if appropriate
include_na <- sales_raw %>%
  filter(is.na(discount_pct) | discount_pct > 0.1)

cat("Explicit NA handling:", nrow(explicit_na_handling), "rows\n")
cat("Including NA values:", nrow(include_na), "rows\n")
```

### Logical Operator Precedence

```{r}
#| label: logical-precedence

# Pitfall: Operator precedence can cause unexpected results
# This might not work as intended:
# ambiguous <- sales_raw %>%
#   filter(customer_type == "Enterprise" | customer_type == "Small Business" & net_amount > 500)

# Better: Use parentheses for clarity
clear_logic <- sales_raw %>%
  filter(
    (customer_type == "Enterprise" | customer_type == "Small Business") &
    net_amount > 500
  )

# Or use %in% for multiple values
cleaner_multiple <- sales_raw %>%
  filter(
    customer_type %in% c("Enterprise", "Small Business"),
    net_amount > 500
  )

cat("Clear logic with parentheses:", nrow(clear_logic), "rows\n")
cat("Cleaner multiple value filter:", nrow(cleaner_multiple), "rows\n")
```

## Exercises

### Exercise 1: Data Quality Assessment
Using the sales dataset:
1. Filter for orders with potential data quality issues
2. Identify orders with suspicious pricing (very high or very low)
3. Find orders with missing critical information
4. Create a "clean" dataset by removing problematic records

### Exercise 2: Time-Based Analysis
Create filters for:
1. Orders from the last quarter
2. Weekend vs weekday patterns
3. Month-end effects (last 3 days of each month)
4. Seasonal patterns (summer vs winter orders)

### Exercise 3: Customer Analysis
Filter and analyze:
1. First-time customers vs repeat customers
2. High-value customers (top 10% by total spending)
3. Customers with recent activity
4. Customers who haven't ordered in 90+ days

### Exercise 4: Advanced Selection
Practice column selection:
1. Create different views of the data for different stakeholders
2. Select columns based on data types and patterns
3. Rename columns for better readability
4. Create summary datasets with only key metrics

## Summary

Filtering and selecting are the foundation of data manipulation in dplyr:

### Key Functions:
- **filter()**: Choose rows based on conditions
- **select()**: Choose columns by name, pattern, or type
- **Helper functions**: `starts_with()`, `ends_with()`, `contains()`, `where()`

### Best Practices:
- **Filter early** in your pipeline to improve performance
- **Handle missing values explicitly** in filter conditions
- **Use meaningful variable names** for complex conditions
- **Combine multiple simple filters** rather than one complex filter
- **Test filters incrementally** to ensure they work as expected

### Common Patterns:
- Filter then select (most common workflow)
- Use `%in%` for multiple value matching
- Combine with date functions for time-based analysis
- Use window functions for relative filtering

### Remember:
- Logical operators: `&` (AND), `|` (OR), `!` (NOT)
- Missing values require special handling
- Use parentheses to clarify complex logical conditions
- Performance matters: filter early and select what you need

Mastering filtering and selecting sets you up for success with all other dplyr operations. These skills form the foundation for every data analysis pipeline!

Next: **[Mutating and Arranging Data](mutating-arranging.qmd)**