---
title: "Grouping and Summarizing Data"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## The Power of Group-Wise Operations

One of the most powerful features of dplyr is the ability to perform operations on groups of data. The combination of `group_by()` and `summarise()` allows you to calculate statistics for each group in your data, implementing the split-apply-combine strategy that's fundamental to data analysis.

```{r}
#| label: setup
#| message: false

library(tidyverse)

# Load the sample dataset from Module 5 index
set.seed(123)
sales_raw <- tibble(
  order_id = paste0("ORD-", str_pad(1:1000, 4, pad = "0")),
  customer_id = sample(paste0("CUST-", 1:200), 1000, replace = TRUE),
  product_id = sample(paste0("PROD-", 1:50), 1000, replace = TRUE),
  order_date = sample(seq(ymd("2023-01-01"), ymd("2024-12-31"), by = "day"), 1000, replace = TRUE),
  quantity = sample(1:10, 1000, replace = TRUE),
  unit_price = round(runif(1000, 5, 200), 2),
  discount_pct = sample(c(0, 0.05, 0.10, 0.15, 0.20), 1000, replace = TRUE, prob = c(0.6, 0.2, 0.1, 0.05, 0.05)),
  sales_rep = sample(c("Alice Johnson", "Bob Smith", "Charlie Brown", "Diana Prince", "Eve Wilson"), 1000, replace = TRUE),
  region = sample(c("North", "South", "East", "West"), 1000, replace = TRUE),
  customer_type = sample(c("Individual", "Small Business", "Enterprise"), 1000, replace = TRUE, prob = c(0.5, 0.3, 0.2))
) %>%
  mutate(
    sales_rep = ifelse(row_number() %in% sample(1:1000, 50), NA, sales_rep),
    discount_pct = ifelse(row_number() %in% sample(1:1000, 30), NA, discount_pct),
    gross_amount = quantity * unit_price,
    discount_amount = gross_amount * coalesce(discount_pct, 0),
    net_amount = gross_amount - discount_amount,
    unit_price = ifelse(row_number() %in% sample(1:1000, 5), unit_price * 100, unit_price),
    quantity = ifelse(row_number() %in% sample(1:1000, 3), 0, quantity)
  )

cat("Working with", nrow(sales_raw), "sales transactions\n")
```

## Understanding group_by()

The `group_by()` function doesn't change the data itself - it adds metadata that tells dplyr to perform subsequent operations by group.

```{r}
#| label: basic-grouping

# Group by a single variable
sales_by_region <- sales_raw %>%
  group_by(region)

# The data looks the same, but it's now grouped
cat("Grouped data has", n_groups(sales_by_region), "groups\n")
cat("Group sizes:\n")
count(sales_by_region, region)

# Group by multiple variables
sales_by_region_type <- sales_raw %>%
  group_by(region, customer_type)

cat("\nGrouping by region and customer type creates", n_groups(sales_by_region_type), "groups\n")
```

## Basic Summarization

`summarise()` (or `summarize()`) reduces grouped data to summary statistics:

```{r}
#| label: basic-summarize

# Calculate summary statistics by region
regional_summary <- sales_raw %>%
  group_by(region) %>%
  summarise(
    total_sales = sum(net_amount),
    avg_sale = mean(net_amount),
    median_sale = median(net_amount),
    total_orders = n(),
    unique_customers = n_distinct(customer_id),
    .groups = "drop"  # Good practice to specify group handling
  )

print(regional_summary)

# Multiple grouping variables
detailed_summary <- sales_raw %>%
  filter(quantity > 0) %>%  # Exclude invalid orders
  group_by(region, customer_type) %>%
  summarise(
    revenue = sum(net_amount),
    orders = n(),
    avg_order_value = mean(net_amount),
    avg_discount = mean(discount_pct, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(revenue))

cat("\nTop 5 segments by revenue:\n")
print(head(detailed_summary, 5))
```

## Common Summary Functions

dplyr provides many useful functions for summarization:

```{r}
#| label: summary-functions

# Comprehensive summary statistics
sales_raw %>%
  filter(!is.na(sales_rep), quantity > 0) %>%
  group_by(sales_rep) %>%
  summarise(
    # Count functions
    total_orders = n(),
    distinct_customers = n_distinct(customer_id),
    distinct_products = n_distinct(product_id),
    
    # Central tendency
    mean_sale = mean(net_amount),
    median_sale = median(net_amount),
    
    # Spread
    sd_sale = sd(net_amount),
    min_sale = min(net_amount),
    max_sale = max(net_amount),
    range = max_sale - min_sale,
    
    # Position
    q25 = quantile(net_amount, 0.25),
    q75 = quantile(net_amount, 0.75),
    iqr = IQR(net_amount),
    
    # Totals
    total_revenue = sum(net_amount),
    total_units = sum(quantity),
    
    .groups = "drop"
  ) %>%
  arrange(desc(total_revenue))
```

## Grouped Mutations

You can also use `mutate()` with grouped data to add group-wise calculations:

```{r}
#| label: grouped-mutate

# Add group-level statistics to each row
sales_with_stats <- sales_raw %>%
  filter(quantity > 0) %>%
  group_by(region) %>%
  mutate(
    regional_avg = mean(net_amount),
    pct_of_regional_avg = net_amount / regional_avg * 100,
    regional_rank = row_number(desc(net_amount)),
    regional_percentile = percent_rank(net_amount)
  ) %>%
  ungroup()

# Show top performers in each region
sales_with_stats %>%
  filter(regional_rank <= 3) %>%
  select(order_id, region, net_amount, regional_avg, pct_of_regional_avg, regional_rank) %>%
  arrange(region, regional_rank)
```

## Window Functions

Window functions are particularly useful with grouped data:

```{r}
#| label: window-functions

# Various window functions
monthly_sales <- sales_raw %>%
  filter(quantity > 0) %>%
  mutate(
    year_month = floor_date(order_date, "month")
  ) %>%
  group_by(year_month, region) %>%
  summarise(
    total_sales = sum(net_amount),
    .groups = "drop"
  ) %>%
  group_by(region) %>%
  arrange(year_month) %>%
  mutate(
    # Cumulative functions
    cumulative_sales = cumsum(total_sales),
    
    # Lag and lead
    prev_month_sales = lag(total_sales, 1),
    next_month_sales = lead(total_sales, 1),
    
    # Month-over-month growth
    mom_growth = (total_sales - prev_month_sales) / prev_month_sales * 100,
    
    # Rolling calculations (3-month average)
    rolling_avg_3m = (lag(total_sales, 2) + lag(total_sales, 1) + total_sales) / 3
  ) %>%
  ungroup()

# Show recent trends
monthly_sales %>%
  filter(year_month >= "2024-01-01") %>%
  select(year_month, region, total_sales, mom_growth, rolling_avg_3m) %>%
  arrange(region, year_month)
```

## Multiple Summary Statistics at Once

You can calculate multiple statistics for multiple variables efficiently:

```{r}
#| label: multiple-summaries

# Summary statistics for multiple numeric variables
sales_raw %>%
  filter(quantity > 0, unit_price <= 500) %>%
  group_by(customer_type) %>%
  summarise(
    across(
      c(quantity, gross_amount, discount_pct, net_amount),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    ),
    n_orders = n(),
    .groups = "drop"
  ) %>%
  mutate(
    across(where(is.numeric), ~round(.x, 2))
  )
```

## Handling Missing Values in Groups

Proper handling of missing values is crucial in grouped operations:

```{r}
#| label: missing-values-groups

# Compare different approaches to missing values
missing_comparison <- sales_raw %>%
  group_by(region) %>%
  summarise(
    # Different ways to handle NAs
    total_orders = n(),
    orders_with_rep = sum(!is.na(sales_rep)),
    orders_missing_rep = sum(is.na(sales_rep)),
    
    # Average discount - different approaches
    avg_discount_na_rm = mean(discount_pct, na.rm = TRUE) * 100,
    avg_discount_zeros = mean(coalesce(discount_pct, 0)) * 100,
    
    # Count of discounted orders
    discounted_orders = sum(discount_pct > 0, na.rm = TRUE),
    
    .groups = "drop"
  )

print(missing_comparison)
```

## Advanced Grouping Techniques

### Dynamic Grouping

```{r}
#| label: dynamic-grouping

# Function to analyze sales by any grouping variable
analyze_by_group <- function(data, group_var) {
  data %>%
    group_by({{ group_var }}) %>%
    summarise(
      revenue = sum(net_amount),
      orders = n(),
      avg_order = mean(net_amount),
      customers = n_distinct(customer_id),
      .groups = "drop"
    ) %>%
    mutate(
      revenue_pct = revenue / sum(revenue) * 100
    ) %>%
    arrange(desc(revenue))
}

# Apply to different grouping variables
cat("Analysis by sales rep:\n")
sales_raw %>%
  filter(!is.na(sales_rep)) %>%
  analyze_by_group(sales_rep)

cat("\nAnalysis by customer type:\n")
sales_raw %>%
  analyze_by_group(customer_type)
```

### Conditional Grouping

```{r}
#| label: conditional-grouping

# Group by calculated categories
sales_categorized <- sales_raw %>%
  filter(quantity > 0) %>%
  mutate(
    order_size = case_when(
      net_amount < 100 ~ "Small",
      net_amount < 500 ~ "Medium",
      net_amount < 1000 ~ "Large",
      TRUE ~ "Enterprise"
    ),
    discount_category = case_when(
      is.na(discount_pct) | discount_pct == 0 ~ "No discount",
      discount_pct <= 0.05 ~ "Low discount",
      discount_pct <= 0.10 ~ "Medium discount",
      TRUE ~ "High discount"
    )
  ) %>%
  group_by(order_size, discount_category) %>%
  summarise(
    orders = n(),
    revenue = sum(net_amount),
    avg_margin = mean((net_amount - gross_amount * 0.6) / net_amount) * 100,  # Assuming 60% cost
    .groups = "drop"
  ) %>%
  arrange(order_size, desc(revenue))

print(sales_categorized)
```

## Performance Tips

When working with large grouped datasets:

```{r}
#| label: performance-tips

# Efficient grouping - calculate once, use multiple times
efficient_summary <- sales_raw %>%
  filter(quantity > 0) %>%
  group_by(region, customer_type) %>%
  summarise(
    n = n(),
    revenue = sum(net_amount),
    .groups = "drop"
  )

# Add multiple calculated fields after summarizing
efficient_summary <- efficient_summary %>%
  mutate(
    avg_order = revenue / n,
    revenue_share = revenue / sum(revenue) * 100,
    cumulative_share = cumsum(revenue_share)
  ) %>%
  arrange(desc(revenue))

cat("Most efficient to summarize first, then calculate derived metrics\n")
print(head(efficient_summary, 10))
```

## Common Patterns and Use Cases

### Top N per Group

```{r}
#| label: top-n-per-group

# Find top 3 customers in each region
top_customers <- sales_raw %>%
  filter(quantity > 0) %>%
  group_by(region, customer_id) %>%
  summarise(
    total_spent = sum(net_amount),
    order_count = n(),
    .groups = "drop"
  ) %>%
  group_by(region) %>%
  slice_max(total_spent, n = 3) %>%
  ungroup()

cat("Top 3 customers by region:\n")
print(top_customers)
```

### Year-over-Year Comparisons

```{r}
#| label: yoy-comparison

# Year-over-year analysis
yoy_analysis <- sales_raw %>%
  filter(quantity > 0) %>%
  mutate(
    year = year(order_date),
    month = month(order_date)
  ) %>%
  group_by(year, month, region) %>%
  summarise(
    revenue = sum(net_amount),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = year,
    values_from = revenue,
    names_prefix = "year_"
  ) %>%
  mutate(
    yoy_change = (year_2024 - year_2023) / year_2023 * 100
  ) %>%
  filter(!is.na(yoy_change))

cat("Year-over-year revenue change by region and month:\n")
print(yoy_analysis)
```

## Practice Exercises

1. **Customer Segmentation**: Group customers by their total purchase value and analyze their behavior
2. **Time-Based Analysis**: Create weekly summaries showing trends and patterns
3. **Product Performance**: Identify best and worst performing products by region
4. **Sales Team Metrics**: Create a comprehensive dashboard for sales rep performance

## Summary

Grouping and summarizing are essential skills for data analysis. Key takeaways:

- `group_by()` creates groups for subsequent operations
- `summarise()` reduces groups to single rows with summary statistics
- Use `.groups = "drop"` to avoid unexpected behavior
- Grouped `mutate()` adds group-level calculations to each row
- Window functions provide powerful tools for within-group calculations
- Always consider how to handle missing values in grouped operations
- Think about performance when working with large datasets

Next, we'll explore how to combine multiple datasets with [Joining Data](joining-data.qmd).