---
title: "Module 5: Data Wrangling with dplyr"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
---

## Welcome to Advanced Data Wrangling! 🔧📊

Now that you understand tidyverse fundamentals and can work with different data types, it's time to master the art of data wrangling. The `dplyr` package is the Swiss Army knife of data manipulation, providing intuitive functions that correspond to the most common data analysis tasks.

## What You'll Learn

In this module, we'll dive deep into the powerful world of dplyr, covering:

### 1. Filtering and Selecting
- Advanced row filtering with complex conditions
- Column selection with helper functions
- Combining and chaining operations
- Working with missing data in filters

### 2. Mutating and Arranging
- Creating and modifying variables
- Window functions and ranking
- Conditional mutations with case_when()
- Sorting data with arrange()

### 3. Grouping and Summarizing
- Group-wise operations and split-apply-combine
- Multiple summary statistics
- Grouped mutations and filtering
- Working with grouped data frames

### 4. Joining Data
- Inner, left, right, and full joins
- Anti-joins and semi-joins
- Handling key conflicts and duplicates
- Joining multiple tables

### 5. Advanced Techniques
- Across() for operating on multiple columns
- Programming with dplyr using {{ and !!
- Custom functions and non-standard evaluation
- Performance optimization and best practices

## Learning Objectives

By the end of this module, you will be able to:

- ✅ Filter and select data using complex conditions and patterns
- ✅ Create new variables with sophisticated transformations
- ✅ Perform group-wise operations and aggregations
- ✅ Join multiple data sources effectively
- ✅ Use advanced dplyr techniques for complex data manipulation
- ✅ Write reusable data wrangling functions
- ✅ Apply best practices for readable and efficient data pipelines

## Why Master dplyr?

dplyr is essential for data analysis because:

1. **Intuitive verbs**: Functions named like actions (filter, select, mutate)
2. **Consistent syntax**: All functions work the same way
3. **Pipe-friendly**: Designed for chaining operations together
4. **Performance**: Optimized for speed with large datasets
5. **Versatility**: Handles 90% of data manipulation tasks

## The dplyr Philosophy: Grammar of Data Manipulation

```{r}
#| label: dplyr-philosophy
#| message: false

library(tidyverse)

# dplyr provides a consistent set of verbs for data manipulation
cat("The main dplyr verbs:\n")
cat("- filter(): choose rows based on conditions\n")
cat("- select(): choose columns by name or pattern\n")
cat("- mutate(): create or modify columns\n")
cat("- arrange(): reorder rows\n")
cat("- summarise(): reduce to summary statistics\n")
cat("- group_by(): perform operations by groups\n")
cat("\nJoining verbs:\n")
cat("- left_join(), inner_join(), full_join(): combine tables\n")
cat("- anti_join(), semi_join(): filtering joins\n")
```

## Real-World Data Challenges

Let's start with a realistic, messy dataset that demonstrates why we need powerful data wrangling tools:

```{r}
#| label: realistic-dataset

# Create a realistic sales dataset with common data problems
set.seed(123)
sales_raw <- tibble(
  order_id = paste0("ORD-", str_pad(1:1000, 4, pad = "0")),
  customer_id = sample(paste0("CUST-", 1:200), 1000, replace = TRUE),
  product_id = sample(paste0("PROD-", 1:50), 1000, replace = TRUE),
  order_date = sample(seq(ymd("2023-01-01"), ymd("2024-12-31"), by = "day"), 1000, replace = TRUE),
  quantity = sample(1:10, 1000, replace = TRUE),
  unit_price = round(runif(1000, 5, 200), 2),
  discount_pct = sample(c(0, 0.05, 0.10, 0.15, 0.20), 1000, replace = TRUE, prob = c(0.6, 0.2, 0.1, 0.05, 0.05)),
  sales_rep = sample(c("Alice Johnson", "Bob Smith", "Charlie Brown", "Diana Prince", "Eve Wilson"), 1000, replace = TRUE),
  region = sample(c("North", "South", "East", "West"), 1000, replace = TRUE),
  customer_type = sample(c("Individual", "Small Business", "Enterprise"), 1000, replace = TRUE, prob = c(0.5, 0.3, 0.2))
)

# Add some realistic data quality issues
sales_raw <- sales_raw %>%
  mutate(
    # Some missing values
    sales_rep = ifelse(row_number() %in% sample(1:1000, 50), NA, sales_rep),
    discount_pct = ifelse(row_number() %in% sample(1:1000, 30), NA, discount_pct),

    # Calculate derived fields
    gross_amount = quantity * unit_price,
    discount_amount = gross_amount * coalesce(discount_pct, 0),
    net_amount = gross_amount - discount_amount,

    # Add some data entry errors
    unit_price = ifelse(row_number() %in% sample(1:1000, 5), unit_price * 100, unit_price),  # Pricing errors
    quantity = ifelse(row_number() %in% sample(1:1000, 3), 0, quantity)  # Zero quantities
  )

# Add customer information
customers <- tibble(
  customer_id = paste0("CUST-", 1:200),
  customer_name = paste("Customer", 1:200),
  industry = sample(c("Technology", "Healthcare", "Finance", "Retail", "Manufacturing"), 200, replace = TRUE),
  signup_date = sample(seq(ymd("2020-01-01"), ymd("2023-12-31"), by = "day"), 200),
  credit_limit = sample(c(1000, 5000, 10000, 25000, 50000), 200, replace = TRUE)
)

# Add product information
products <- tibble(
  product_id = paste0("PROD-", 1:50),
  product_name = paste("Product", LETTERS[1:50]),
  category = sample(c("Electronics", "Software", "Services", "Hardware"), 50, replace = TRUE),
  cost = round(runif(50, 2, 100), 2),
  launch_date = sample(seq(ymd("2020-01-01"), ymd("2024-01-01"), by = "day"), 50)
)

cat("Created realistic business dataset:\n")
cat("Sales transactions:", nrow(sales_raw), "rows\n")
cat("Unique customers:", n_distinct(sales_raw$customer_id), "\n")
cat("Date range:", as.character(min(sales_raw$order_date)), "to", as.character(max(sales_raw$order_date)), "\n")
cat("Total revenue: $", format(sum(sales_raw$net_amount), big.mark = ","), "\n")

# Show data quality issues
cat("\nData quality issues to address:\n")
cat("Missing sales reps:", sum(is.na(sales_raw$sales_rep)), "\n")
cat("Missing discounts:", sum(is.na(sales_raw$discount_pct)), "\n")
cat("Suspicious high prices (>$500):", sum(sales_raw$unit_price > 500), "\n")
cat("Zero quantities:", sum(sales_raw$quantity == 0), "\n")
```

This dataset represents the kind of messy, real-world data you'll encounter in business. Throughout this module, we'll learn how to clean, transform, and analyze it using dplyr's powerful tools.

## Module Structure

This module is organized into five comprehensive sections:

1. **[Filtering and Selecting](filtering-selecting.qmd)**: Master row and column selection
2. **[Mutating and Arranging](mutating-arranging.qmd)**: Create variables and sort data
3. **[Grouping and Summarizing](grouping-summarizing.qmd)**: Aggregate data by groups
4. **[Joining Data](joining-data.qmd)**: Combine multiple tables
5. **[Advanced Techniques](advanced-techniques.qmd)**: Programming and optimization

Each section includes:
- Core concepts with clear explanations
- Practical examples using real business scenarios
- Common patterns and best practices
- Performance tips and troubleshooting
- Hands-on exercises

## The Power of Pipes

Before diving in, let's appreciate the elegance of dplyr's pipe-friendly design:

```{r}
#| label: pipe-power

# Without pipes (nested and hard to read)
result_nested <- arrange(
  summarise(
    group_by(
      filter(sales_raw, !is.na(sales_rep)),
      region
    ),
    total_sales = sum(net_amount),
    avg_order = mean(net_amount),
    order_count = n()
  ),
  desc(total_sales)
)

# With pipes (readable and intuitive)
result_piped <- sales_raw %>%
  filter(!is.na(sales_rep)) %>%
  group_by(region) %>%
  summarise(
    total_sales = sum(net_amount),
    avg_order = mean(net_amount),
    order_count = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(total_sales))

# Both produce the same result
identical(result_nested, result_piped)

# Show the clean, readable result
cat("Regional sales summary:\n")
print(result_piped)
```

The pipe makes data analysis feel like describing what you want to do in plain English!

## A Motivating Example

Here's a taste of what you'll be able to do by the end of this module:

```{r}
#| label: motivating-example

# Complex business analysis in a single, readable pipeline
quarterly_performance <- sales_raw %>%
  # Data cleaning
  filter(
    quantity > 0,                    # Remove invalid orders
    unit_price <= 500,               # Remove pricing errors
    !is.na(sales_rep)                # Only orders with assigned reps
  ) %>%

  # Join with customer and product data
  left_join(customers, by = "customer_id") %>%
  left_join(products, by = "product_id") %>%

  # Create time-based variables
  mutate(
    quarter = paste0(year(order_date), "-Q", quarter(order_date)),
    profit = net_amount - (quantity * cost),
    profit_margin = profit / net_amount
  ) %>%

  # Group by multiple dimensions
  group_by(quarter, region, customer_type) %>%

  # Calculate comprehensive metrics
  summarise(
    # Revenue metrics
    total_revenue = sum(net_amount),
    total_profit = sum(profit),
    avg_profit_margin = mean(profit_margin, na.rm = TRUE),

    # Volume metrics
    total_orders = n(),
    total_units = sum(quantity),

    # Customer metrics
    unique_customers = n_distinct(customer_id),
    avg_order_value = mean(net_amount),

    .groups = "drop"
  ) %>%

  # Add rankings and comparisons
  group_by(quarter) %>%
  mutate(
    revenue_rank = row_number(desc(total_revenue)),
    revenue_pct_of_total = total_revenue / sum(total_revenue) * 100
  ) %>%
  ungroup() %>%

  # Final sorting
  arrange(quarter, revenue_rank)

cat("Quarterly performance analysis:\n")
quarterly_performance %>%
  filter(quarter %in% c("2024-Q1", "2024-Q2")) %>%
  select(quarter, region, customer_type, total_revenue, revenue_rank, revenue_pct_of_total) %>%
  print()
```

This analysis combines filtering, joining, grouping, summarizing, and ranking - all in a single, readable pipeline. By the end of this module, this kind of complex analysis will feel natural and intuitive.

## Prerequisites

Before starting this module, make sure you have:

- Completed Module 3 (Tidyverse Introduction) and Module 4 (Data Types)
- Understanding of the pipe operator (`%>%`)
- Basic knowledge of tibbles and data frames
- Familiarity with tidyverse loading and basic functions

## Getting Started

Ready to become a data wrangling expert? Let's start with the foundation: **[Filtering and Selecting](filtering-selecting.qmd)**

## Summary

Data wrangling with dplyr is both an art and a science. The key principles to remember:

- **Think in verbs**: Each dplyr function does one thing well
- **Chain operations**: Use pipes to create readable data pipelines
- **Be explicit**: Clear, descriptive code is better than clever shortcuts
- **Test incrementally**: Build complex operations step by step
- **Handle missing data**: Always consider how NAs affect your analysis

dplyr transforms messy, real-world data into clean, analysis-ready datasets. With these skills, you'll be able to tackle any data challenge with confidence and clarity.

Let's begin your journey to dplyr mastery! 🚀