---
title: "Joining Data"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Combining Data from Multiple Sources

In real-world data analysis, information is rarely contained in a single table. The ability to combine data from multiple sources is essential. dplyr provides a comprehensive set of join functions that mirror SQL operations while maintaining the tidyverse philosophy.

```{r}
#| label: setup
#| message: false

library(tidyverse)

# Create the sales dataset
set.seed(123)
sales <- tibble(
  order_id = paste0("ORD-", str_pad(1:500, 4, pad = "0")),
  customer_id = sample(paste0("CUST-", 1:100), 500, replace = TRUE),
  product_id = sample(paste0("PROD-", 1:30), 500, replace = TRUE),
  order_date = sample(seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "day"), 500, replace = TRUE),
  quantity = sample(1:10, 500, replace = TRUE),
  unit_price = round(runif(500, 10, 200), 2)
) %>%
  mutate(
    total_amount = quantity * unit_price
  )

# Create customer information
customers <- tibble(
  customer_id = paste0("CUST-", 1:100),
  customer_name = paste("Customer", 1:100),
  customer_type = sample(c("Individual", "Small Business", "Enterprise"), 100, replace = TRUE, prob = c(0.6, 0.3, 0.1)),
  join_date = sample(seq(ymd("2020-01-01"), ymd("2023-12-31"), by = "day"), 100),
  credit_limit = sample(c(1000, 5000, 10000, 25000, 50000), 100, replace = TRUE),
  city = sample(c("New York", "Los Angeles", "Chicago", "Houston", "Phoenix"), 100, replace = TRUE),
  state = c(rep("NY", 20), rep("CA", 20), rep("IL", 20), rep("TX", 20), rep("AZ", 20))
)

# Create product information
products <- tibble(
  product_id = paste0("PROD-", 1:30),
  product_name = paste("Product", LETTERS[1:30]),
  category = rep(c("Electronics", "Clothing", "Food", "Home", "Sports"), 6),
  cost = round(runif(30, 5, 100), 2),
  supplier_id = sample(paste0("SUPP-", 1:10), 30, replace = TRUE)
)

# Create some additional data for demonstrating different join scenarios
# Some products that don't have sales
new_products <- tibble(
  product_id = paste0("PROD-", 31:35),
  product_name = paste("New Product", 1:5),
  category = "New Items",
  cost = round(runif(5, 20, 150), 2),
  supplier_id = "SUPP-11"
)

# Some customers who haven't made purchases
inactive_customers <- tibble(
  customer_id = paste0("CUST-", 101:110),
  customer_name = paste("Inactive Customer", 1:10),
  customer_type = "Individual",
  join_date = ymd("2024-01-01"),
  credit_limit = 1000,
  city = "Boston",
  state = "MA"
)

cat("Created sample datasets:\n")
cat("- Sales:", nrow(sales), "transactions\n")
cat("- Customers:", nrow(customers), "records\n")
cat("- Products:", nrow(products), "items\n")
```

## Types of Joins

dplyr provides several types of joins, each serving different purposes:

### Inner Join

Keeps only rows that have matching keys in both tables:

```{r}
#| label: inner-join

# Join sales with customer information
sales_with_customers <- sales %>%
  inner_join(customers, by = "customer_id")

cat("Inner join result:\n")
cat("- Original sales:", nrow(sales), "rows\n")
cat("- After inner join:", nrow(sales_with_customers), "rows\n")
cat("- Columns added:", setdiff(names(sales_with_customers), names(sales)), "\n\n")

# Show a sample
sales_with_customers %>%
  select(order_id, customer_name, customer_type, order_date, total_amount) %>%
  slice_head(n = 5)
```

### Left Join

Keeps all rows from the left table, adding NA for non-matching right table rows:

```{r}
#| label: left-join

# Combine all customers and products
all_products <- bind_rows(products, new_products)

# Left join to keep all sales, even if product info is missing
sales_complete <- sales %>%
  left_join(customers, by = "customer_id") %>%
  left_join(all_products, by = "product_id")

cat("Left join preserves all sales records:\n")
cat("- Sales records:", nrow(sales_complete), "\n")
cat("- Products without info:", sum(is.na(sales_complete$product_name)), "\n\n")

# Check for missing product information
sales_complete %>%
  filter(is.na(product_name)) %>%
  select(order_id, product_id, product_name, category) %>%
  slice_head(n = 5)
```

### Right Join

Keeps all rows from the right table:

```{r}
#| label: right-join

# Find all products and their sales (if any)
product_sales <- sales %>%
  group_by(product_id) %>%
  summarise(
    total_sold = sum(quantity),
    revenue = sum(total_amount),
    .groups = "drop"
  ) %>%
  right_join(all_products, by = "product_id")

cat("Right join shows all products, including those without sales:\n")
product_sales %>%
  filter(is.na(total_sold)) %>%
  select(product_id, product_name, category, total_sold, revenue)
```

### Full Join

Keeps all rows from both tables:

```{r}
#| label: full-join

# Combine all customers (including inactive ones)
all_customers <- bind_rows(customers, inactive_customers)

# Full join to see all customers and all orders
customer_order_summary <- sales %>%
  group_by(customer_id) %>%
  summarise(
    order_count = n(),
    total_spent = sum(total_amount),
    .groups = "drop"
  ) %>%
  full_join(all_customers, by = "customer_id")

cat("Full join shows both customers with orders and those without:\n")
cat("- Total customers:", nrow(customer_order_summary), "\n")
cat("- Customers with orders:", sum(!is.na(customer_order_summary$order_count)), "\n")
cat("- Customers without orders:", sum(is.na(customer_order_summary$order_count)), "\n\n")

# Show customers without orders
customer_order_summary %>%
  filter(is.na(order_count)) %>%
  select(customer_id, customer_name, join_date, order_count) %>%
  slice_head(n = 5)
```

## Filtering Joins

These joins filter rows based on matches, without adding columns:

### Semi Join

Keeps rows in the first table that have a match in the second table:

```{r}
#| label: semi-join

# Find customers who bought electronics
electronics_products <- products %>%
  filter(category == "Electronics")

electronics_buyers <- sales %>%
  semi_join(electronics_products, by = "product_id") %>%
  distinct(customer_id) %>%
  semi_join(customers, by = "customer_id")

cat("Semi join to find electronics buyers:\n")
cat("- Total unique customers:", n_distinct(sales$customer_id), "\n")
cat("- Electronics buyers:", nrow(electronics_buyers), "\n")

# Get their details
electronics_buyers %>%
  left_join(customers, by = "customer_id") %>%
  select(customer_id, customer_name, customer_type) %>%
  slice_head(n = 5)
```

### Anti Join

Keeps rows in the first table that do NOT have a match in the second table:

```{r}
#| label: anti-join

# Find customers who have never bought electronics
non_electronics_buyers <- customers %>%
  anti_join(electronics_buyers, by = "customer_id")

cat("Anti join to find non-electronics buyers:\n")
cat("- Customers who never bought electronics:", nrow(non_electronics_buyers), "\n\n")

# Find products that have never been sold
unsold_products <- all_products %>%
  anti_join(sales, by = "product_id")

cat("Products with no sales:\n")
unsold_products %>%
  select(product_id, product_name, category)
```

## Joining on Multiple Keys

Sometimes you need to join on multiple columns:

```{r}
#| label: multiple-keys

# Create order details with multiple keys
order_details <- tibble(
  order_id = sample(sales$order_id, 300),
  product_id = sample(products$product_id, 300, replace = TRUE),
  discount_pct = sample(c(0, 5, 10, 15, 20), 300, replace = TRUE, prob = c(0.5, 0.2, 0.15, 0.1, 0.05))
)

# Join using both order_id and product_id
sales_with_discounts <- sales %>%
  left_join(order_details, by = c("order_id", "product_id")) %>%
  mutate(
    discount_amount = if_else(is.na(discount_pct), 0, total_amount * discount_pct / 100),
    final_amount = total_amount - discount_amount
  )

cat("Joining on multiple keys:\n")
cat("- Orders with discounts:", sum(!is.na(sales_with_discounts$discount_pct)), "\n")
cat("- Total discount amount: $", round(sum(sales_with_discounts$discount_amount), 2), "\n")
```

## Handling Key Conflicts

When column names don't match, specify the relationship:

```{r}
#| label: key-conflicts

# Create supplier data with different column name
suppliers <- tibble(
  supplier_code = paste0("SUPP-", 1:11),  # Different name than supplier_id
  supplier_name = paste("Supplier", LETTERS[1:11]),
  country = sample(c("USA", "Canada", "Mexico"), 11, replace = TRUE),
  rating = round(runif(11, 3, 5), 1)
)

# Join with name mapping
products_with_suppliers <- all_products %>%
  left_join(suppliers, by = c("supplier_id" = "supplier_code"))

cat("Joining with different column names:\n")
products_with_suppliers %>%
  select(product_name, category, supplier_id, supplier_name, country, rating) %>%
  slice_head(n = 5)
```

## Dealing with Duplicate Keys

Understanding how joins handle duplicates is crucial:

```{r}
#| label: duplicate-keys

# Create a scenario with duplicate keys
# Multiple promotions for the same product
promotions <- tibble(
  product_id = rep(paste0("PROD-", 1:5), each = 2),
  promo_name = c("Summer Sale", "Back to School", "Black Friday", "Cyber Monday", 
                 "Holiday Special", "New Year", "Spring Sale", "Flash Deal", 
                 "Weekend Special", "Member Exclusive"),
  discount = c(10, 15, 20, 25, 30, 10, 15, 20, 25, 30)
)

# Join will create multiple rows for each match
sales_with_promos <- sales %>%
  filter(product_id %in% paste0("PROD-", 1:5)) %>%
  inner_join(promotions, by = "product_id")

cat("Effect of duplicate keys in joins:\n")
cat("- Sales records for PROD-1 to PROD-5:", 
    sum(sales$product_id %in% paste0("PROD-", 1:5)), "\n")
cat("- Records after join with promotions:", nrow(sales_with_promos), "\n")
cat("- Each sale is duplicated for each matching promotion\n\n")

# Show example
sales_with_promos %>%
  filter(order_id == first(order_id)) %>%
  select(order_id, product_id, promo_name, discount)
```

## Complex Multi-Table Joins

Real-world analysis often requires joining multiple tables:

```{r}
#| label: complex-joins

# Create a complete sales analysis dataset
sales_analysis <- sales %>%
  # Add customer information
  left_join(customers, by = "customer_id") %>%
  # Add product information
  left_join(products, by = "product_id") %>%
  # Add supplier information
  left_join(suppliers, by = c("supplier_id" = "supplier_code")) %>%
  # Calculate additional metrics
  mutate(
    profit_margin = (unit_price - cost) / unit_price,
    profit = total_amount * profit_margin,
    order_month = floor_date(order_date, "month")
  )

cat("Complete sales analysis dataset:\n")
cat("- Total columns:", ncol(sales_analysis), "\n")
cat("- Column categories:\n")
cat("  - Order info:", sum(str_detect(names(sales_analysis), "order")), "columns\n")
cat("  - Customer info:", sum(str_detect(names(sales_analysis), "customer")), "columns\n")
cat("  - Product info:", sum(str_detect(names(sales_analysis), "product")), "columns\n")
cat("  - Financial metrics:", sum(names(sales_analysis) %in% c("profit", "profit_margin", "total_amount")), "columns\n")

# Sample the complete dataset
sales_analysis %>%
  select(order_id, customer_name, product_name, supplier_name, quantity, profit) %>%
  slice_sample(n = 5)
```

## Performance Considerations

When working with large datasets, join order matters:

```{r}
#| label: join-performance

# Example of efficient join ordering
# Start with the smallest dataset and join progressively

# Inefficient: Starting with large table
start_time1 <- Sys.time()
result1 <- sales %>%
  left_join(sales_analysis, by = "order_id") %>%
  nrow()
time1 <- Sys.time() - start_time1

# Efficient: Filter first, then join
start_time2 <- Sys.time()
result2 <- sales %>%
  filter(order_date >= "2024-06-01") %>%
  left_join(customers, by = "customer_id") %>%
  left_join(products, by = "product_id") %>%
  nrow()
time2 <- Sys.time() - start_time2

cat("Join performance tips:\n")
cat("1. Filter data before joining when possible\n")
cat("2. Join smaller tables first\n")
cat("3. Use appropriate join types (inner join is fastest)\n")
cat("4. Index key columns in databases\n")
```

## Common Join Patterns

### Customer Lifetime Value Analysis

```{r}
#| label: customer-ltv

# Calculate customer lifetime value
customer_ltv <- sales %>%
  group_by(customer_id) %>%
  summarise(
    first_order = min(order_date),
    last_order = max(order_date),
    total_orders = n(),
    total_spent = sum(total_amount),
    avg_order_value = mean(total_amount),
    .groups = "drop"
  ) %>%
  inner_join(customers, by = "customer_id") %>%
  mutate(
    customer_age_days = as.numeric(today() - join_date),
    days_active = as.numeric(last_order - first_order) + 1,
    order_frequency = total_orders / days_active * 30  # Orders per 30 days
  ) %>%
  arrange(desc(total_spent))

cat("Top 5 customers by lifetime value:\n")
customer_ltv %>%
  select(customer_name, customer_type, total_orders, total_spent, avg_order_value, order_frequency) %>%
  slice_head(n = 5)
```

### Product Affinity Analysis

```{r}
#| label: product-affinity

# Find products frequently bought together
product_pairs <- sales %>%
  inner_join(sales, by = "customer_id", suffix = c("_1", "_2")) %>%
  filter(product_id_1 < product_id_2) %>%  # Avoid duplicates and self-joins
  count(product_id_1, product_id_2, name = "times_bought_together") %>%
  filter(times_bought_together >= 5) %>%
  left_join(products, by = c("product_id_1" = "product_id")) %>%
  rename(product_name_1 = product_name) %>%
  left_join(products, by = c("product_id_2" = "product_id")) %>%
  rename(product_name_2 = product_name) %>%
  arrange(desc(times_bought_together))

cat("Products frequently bought together:\n")
product_pairs %>%
  select(product_name_1, product_name_2, times_bought_together) %>%
  slice_head(n = 10)
```

## Practice Exercises

1. **Sales Dashboard**: Create a comprehensive sales report joining all tables
2. **Customer Segmentation**: Use joins to categorize customers based on purchase behavior
3. **Inventory Analysis**: Join sales and product data to identify slow-moving items
4. **Geographic Analysis**: Join customer location data with sales to analyze regional patterns

## Summary

Mastering joins is essential for real-world data analysis. Key points:

- Choose the right join type for your analysis needs
- Understand how joins handle missing values and duplicates
- Pay attention to column names and key relationships
- Consider performance when joining large datasets
- Filter before joining when possible
- Use filtering joins (semi/anti) for efficient subsetting
- Always verify join results match expectations

Next, explore [Advanced Techniques](advanced-techniques.qmd) for more sophisticated dplyr operations.