---
title: "Mutating and Arranging Data"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to Data Transformation

While `filter()` and `select()` help you subset data, `mutate()` and `arrange()` help you transform and organize it. The `mutate()` function creates new variables or modifies existing ones, while `arrange()` sorts your data. These operations are essential for data preparation and analysis.

```{r}
#| label: setup
#| message: false

library(tidyverse)

# Create our sample sales dataset
set.seed(123)
sales_data <- tibble(
  order_id = paste0("ORD-", str_pad(1:1000, 4, pad = "0")),
  customer_id = sample(paste0("CUST-", 1:200), 1000, replace = TRUE),
  product_id = sample(paste0("PROD-", 1:50), 1000, replace = TRUE),
  order_date = sample(seq(ymd("2023-01-01"), ymd("2024-12-31"), by = "day"), 1000, replace = TRUE),
  quantity = sample(1:10, 1000, replace = TRUE),
  unit_price = round(runif(1000, 5, 200), 2),
  discount_pct = sample(c(0, 0.05, 0.10, 0.15, 0.20), 1000, replace = TRUE, prob = c(0.6, 0.2, 0.1, 0.05, 0.05)),
  sales_rep = sample(c("Alice Johnson", "Bob Smith", "Charlie Brown", "Diana Prince", "Eve Wilson"), 1000, replace = TRUE),
  region = sample(c("North", "South", "East", "West"), 1000, replace = TRUE),
  customer_type = sample(c("Individual", "Small Business", "Enterprise"), 1000, replace = TRUE, prob = c(0.5, 0.3, 0.2))
) %>%
  mutate(
    sales_rep = ifelse(row_number() %in% sample(1:1000, 50), NA, sales_rep),
    discount_pct = ifelse(row_number() %in% sample(1:1000, 30), NA, discount_pct)
  )

cat("Working with", nrow(sales_data), "sales transactions\n")
```

## Creating New Variables with mutate()

### Basic Calculations

```{r}
#| label: basic-mutations

# Calculate derived financial metrics
sales_enhanced <- sales_data %>%
  mutate(
    # Basic calculations
    gross_amount = quantity * unit_price,
    discount_amount = gross_amount * coalesce(discount_pct, 0),
    net_amount = gross_amount - discount_amount,

    # Profit calculations (assuming 60% margin)
    cost_amount = gross_amount * 0.6,
    profit = net_amount - cost_amount,
    profit_margin = profit / net_amount,

    # Unit economics
    revenue_per_unit = net_amount / quantity,
    profit_per_unit = profit / quantity
  )

# Show the new variables
cat("New financial variables created:\n")
sales_enhanced %>%
  select(order_id, quantity, unit_price, gross_amount, discount_amount,
         net_amount, profit, profit_margin) %>%
  head(5) %>%
  print()
```

### Date and Time Manipulations

```{r}
#| label: date-mutations

sales_with_dates <- sales_enhanced %>%
  mutate(
    # Extract date components
    order_year = year(order_date),
    order_month = month(order_date, label = TRUE),
    order_quarter = paste0("Q", quarter(order_date)),
    order_weekday = wday(order_date, label = TRUE),

    # Date calculations
    days_since_epoch = as.numeric(order_date - ymd("2023-01-01")),
    is_weekend = wday(order_date) %in% c(1, 7),
    is_month_end = day(order_date) >= 25,

    # Relative date calculations
    days_ago = as.numeric(Sys.Date() - order_date),
    is_recent = days_ago <= 30,

    # Business calendar
    order_month_year = floor_date(order_date, "month"),
    fiscal_quarter = case_when(
      month(order_date) %in% c(1, 2, 3) ~ paste0("FY", year(order_date), "-Q4"),
      month(order_date) %in% c(4, 5, 6) ~ paste0("FY", year(order_date + years(1)), "-Q1"),
      month(order_date) %in% c(7, 8, 9) ~ paste0("FY", year(order_date + years(1)), "-Q2"),
      month(order_date) %in% c(10, 11, 12) ~ paste0("FY", year(order_date + years(1)), "-Q3")
    )
  )

cat("Date-related variables:\n")
sales_with_dates %>%
  select(order_date, order_year, order_month, order_quarter,
         order_weekday, is_weekend, days_ago, fiscal_quarter) %>%
  head(5) %>%
  print()
```

### Conditional Mutations with case_when()

```{r}
#| label: case-when-mutations

sales_categorized <- sales_with_dates %>%
  mutate(
    # Order size categories
    order_size = case_when(
      net_amount >= 1000 ~ "Large",
      net_amount >= 500 ~ "Medium",
      net_amount >= 100 ~ "Small",
      TRUE ~ "Micro"
    ),

    # Customer segment based on type and order value
    customer_segment = case_when(
      customer_type == "Enterprise" & net_amount >= 500 ~ "Enterprise High",
      customer_type == "Enterprise" & net_amount < 500 ~ "Enterprise Standard",
      customer_type == "Small Business" & net_amount >= 300 ~ "SMB High",
      customer_type == "Small Business" & net_amount < 300 ~ "SMB Standard",
      customer_type == "Individual" & net_amount >= 200 ~ "Individual Premium",
      customer_type == "Individual" & net_amount < 200 ~ "Individual Standard",
      TRUE ~ "Other"
    ),

    # Discount tier
    discount_tier = case_when(
      is.na(discount_pct) | discount_pct == 0 ~ "No Discount",
      discount_pct <= 0.05 ~ "Low Discount",
      discount_pct <= 0.15 ~ "Medium Discount",
      discount_pct > 0.15 ~ "High Discount",
      TRUE ~ "Unknown"
    ),

    # Sales performance indicator
    performance_flag = case_when(
      profit_margin >= 0.3 & net_amount >= 500 ~ "Excellent",
      profit_margin >= 0.2 & net_amount >= 300 ~ "Good",
      profit_margin >= 0.1 ~ "Fair",
      profit_margin >= 0 ~ "Poor",
      TRUE ~ "Loss"
    ),

    # Regional priority (business logic example)
    regional_priority = case_when(
      region == "North" & customer_type == "Enterprise" ~ "High",
      region == "East" & customer_type %in% c("Enterprise", "Small Business") ~ "High",
      region %in% c("South", "West") & customer_type == "Enterprise" ~ "Medium",
      TRUE ~ "Standard"
    )
  )

# Show the categorizations
cat("Conditional mutations results:\n")
sales_categorized %>%
  count(order_size, customer_segment) %>%
  head(10) %>%
  print()

cat("\nDiscount tier distribution:\n")
sales_categorized %>%
  count(discount_tier, sort = TRUE) %>%
  print()
```

### Working with Strings

```{r}
#| label: string-mutations

sales_with_strings <- sales_categorized %>%
  mutate(
    # Extract information from IDs
    customer_number = str_extract(customer_id, "\\d+"),
    product_number = str_extract(product_id, "\\d+"),
    order_number = str_extract(order_id, "\\d+"),

    # Sales rep processing
    rep_first_name = str_extract(sales_rep, "^\\w+"),
    rep_last_name = str_extract(sales_rep, "\\w+$"),
    rep_initials = paste0(
      str_sub(str_extract(sales_rep, "^\\w+"), 1, 1),
      str_sub(str_extract(sales_rep, "\\w+$"), 1, 1)
    ),

    # Create display names and codes
    order_display = paste0(order_quarter, "-", str_pad(order_number, 4, pad = "0")),
    customer_code = paste0(str_sub(region, 1, 1), "-", customer_number),

    # Text classifications
    has_discount = if_else(!is.na(discount_pct) & discount_pct > 0, "Yes", "No"),
    weekend_order = if_else(is_weekend, "Weekend", "Weekday"),

    # Complex string operations
    order_summary = paste0(
      customer_type, " customer (", region, ") - ",
      order_size, " order: $", round(net_amount, 0),
      if_else(has_discount == "Yes", paste0(" (", round(discount_pct * 100, 0), "% off)"), "")
    )
  )

cat("String manipulation results:\n")
sales_with_strings %>%
  select(sales_rep, rep_first_name, rep_last_name, rep_initials,
         order_display, customer_code, order_summary) %>%
  head(5) %>%
  print()
```

## Working with Window Functions

### Ranking and Row Numbers

```{r}
#| label: window-functions

sales_ranked <- sales_with_strings %>%
  # Add various ranking methods
  mutate(
    # Overall rankings
    revenue_rank = row_number(desc(net_amount)),
    revenue_rank_tie = rank(desc(net_amount)),
    revenue_percentile = percent_rank(net_amount),

    # Quantity rankings
    quantity_rank = row_number(desc(quantity)),

    # Date rankings
    chronological_order = row_number(order_date),
    reverse_chronological = row_number(desc(order_date))
  ) %>%
  # Add group-based rankings
  group_by(region) %>%
  mutate(
    regional_revenue_rank = row_number(desc(net_amount)),
    regional_revenue_percentile = percent_rank(net_amount)
  ) %>%
  ungroup() %>%
  # Add customer-based rankings
  group_by(customer_id) %>%
  mutate(
    customer_order_sequence = row_number(order_date),
    customer_largest_order_rank = row_number(desc(net_amount)),
    is_largest_customer_order = customer_largest_order_rank == 1
  ) %>%
  ungroup()

# Show ranking results
cat("Top 10 orders by revenue:\n")
sales_ranked %>%
  filter(revenue_rank <= 10) %>%
  select(order_id, region, net_amount, revenue_rank, regional_revenue_rank) %>%
  arrange(revenue_rank) %>%
  print()

cat("\nCustomer order patterns (first few customers):\n")
sales_ranked %>%
  filter(customer_id %in% c("CUST-1", "CUST-2", "CUST-3")) %>%
  select(customer_id, order_date, net_amount, customer_order_sequence,
         customer_largest_order_rank, is_largest_customer_order) %>%
  arrange(customer_id, order_date) %>%
  print()
```

### Lead and Lag Operations

```{r}
#| label: lead-lag-functions

# Calculate period-over-period changes
sales_with_trends <- sales_ranked %>%
  arrange(order_date) %>%
  mutate(
    # Overall trends
    prev_order_amount = lag(net_amount),
    next_order_amount = lead(net_amount),
    amount_change_from_prev = net_amount - prev_order_amount,

    # Moving calculations
    rolling_avg_3 = (lag(net_amount, 2) + lag(net_amount, 1) + net_amount) / 3,

    # Time between orders
    prev_order_date = lag(order_date),
    days_since_last_order = as.numeric(order_date - prev_order_date)
  ) %>%
  # Customer-specific trends
  group_by(customer_id) %>%
  arrange(order_date) %>%
  mutate(
    prev_customer_order = lag(net_amount),
    customer_order_growth = net_amount - prev_customer_order,
    customer_order_growth_pct = (net_amount / prev_customer_order - 1) * 100,

    # Customer ordering patterns
    days_between_customer_orders = as.numeric(order_date - lag(order_date)),
    avg_days_between_orders = mean(days_between_customer_orders, na.rm = TRUE),

    # Customer value trends
    customer_cumulative_value = cumsum(net_amount),
    customer_running_avg = cummean(net_amount)
  ) %>%
  ungroup()

cat("Customer ordering trends:\n")
sales_with_trends %>%
  filter(!is.na(customer_order_growth)) %>%
  select(customer_id, order_date, net_amount, prev_customer_order,
         customer_order_growth, customer_order_growth_pct) %>%
  head(10) %>%
  print()
```

### Cumulative Operations

```{r}
#| label: cumulative-operations

sales_cumulative <- sales_with_trends %>%
  arrange(order_date) %>%
  mutate(
    # Running totals
    cumulative_revenue = cumsum(net_amount),
    cumulative_orders = row_number(),
    running_average = cummean(net_amount),

    # Running business metrics
    cumulative_profit = cumsum(profit),
    running_profit_margin = cumulative_profit / cumulative_revenue
  ) %>%
  # Monthly cumulative metrics
  group_by(order_month_year) %>%
  arrange(order_date) %>%
  mutate(
    monthly_cumulative_revenue = cumsum(net_amount),
    monthly_order_count = row_number(),
    month_to_date_avg = cummean(net_amount)
  ) %>%
  ungroup() %>%
  # Regional cumulative metrics
  group_by(region) %>%
  arrange(order_date) %>%
  mutate(
    regional_cumulative_revenue = cumsum(net_amount),
    regional_market_share = regional_cumulative_revenue / sum(net_amount) * 100
  ) %>%
  ungroup()

cat("Cumulative business metrics:\n")
sales_cumulative %>%
  filter(order_date >= ymd("2024-12-01")) %>%
  select(order_date, net_amount, cumulative_revenue, running_average,
         monthly_cumulative_revenue, monthly_order_count) %>%
  head(10) %>%
  print()
```

## Sorting Data with arrange()

### Basic Sorting

```{r}
#| label: basic-arranging

# Single column sorting
by_amount_desc <- sales_cumulative %>%
  arrange(desc(net_amount))

by_date_asc <- sales_cumulative %>%
  arrange(order_date)

# Multiple column sorting
by_region_then_amount <- sales_cumulative %>%
  arrange(region, desc(net_amount))

by_customer_then_date <- sales_cumulative %>%
  arrange(customer_id, order_date)

cat("Top 5 orders by amount:\n")
by_amount_desc %>%
  select(order_id, customer_id, region, net_amount, order_date) %>%
  head(5) %>%
  print()

cat("\nRecent orders by date:\n")
by_date_asc %>%
  select(order_id, order_date, customer_id, net_amount) %>%
  tail(5) %>%
  print()
```

### Advanced Sorting

```{r}
#| label: advanced-arranging

# Sorting with custom logic
custom_sort <- sales_cumulative %>%
  arrange(
    desc(customer_type == "Enterprise"),  # Enterprise customers first
    desc(regional_priority == "High"),    # High priority regions next
    desc(net_amount),                     # Then by order value
    order_date                            # Finally by date
  )

# Sorting by calculated values
by_profit_margin <- sales_cumulative %>%
  arrange(desc(profit_margin), desc(net_amount))

# Sorting with string functions
by_customer_alpha <- sales_cumulative %>%
  arrange(str_extract(customer_id, "\\d+") %>% as.numeric())

# Complex sorting for analysis
analysis_sort <- sales_cumulative %>%
  arrange(
    order_year,
    match(order_month, month.name),  # Sort months in calendar order
    desc(net_amount)
  )

cat("Custom priority sorting (top 5):\n")
custom_sort %>%
  select(customer_type, regional_priority, region, net_amount, order_date) %>%
  head(5) %>%
  print()

cat("\nBest profit margins (top 5):\n")
by_profit_margin %>%
  select(order_id, net_amount, profit, profit_margin) %>%
  head(5) %>%
  print()
```

### Grouped Sorting

```{r}
#| label: grouped-arranging

# Sort within groups
top_orders_per_region <- sales_cumulative %>%
  group_by(region) %>%
  arrange(desc(net_amount), .by_group = TRUE) %>%
  slice_head(n = 3) %>%
  ungroup()

# Sort by group statistics
regions_by_performance <- sales_cumulative %>%
  group_by(region) %>%
  summarise(
    total_revenue = sum(net_amount),
    avg_order_value = mean(net_amount),
    order_count = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(total_revenue))

# Complex grouped sorting
customer_analysis <- sales_cumulative %>%
  group_by(customer_id, customer_type) %>%
  summarise(
    total_spent = sum(net_amount),
    order_count = n(),
    avg_order_value = mean(net_amount),
    first_order = min(order_date),
    last_order = max(order_date),
    .groups = "drop"
  ) %>%
  arrange(
    customer_type,
    desc(total_spent)
  )

cat("Top 3 orders per region:\n")
print(top_orders_per_region %>% select(region, order_id, net_amount))

cat("\nRegions by performance:\n")
print(regions_by_performance)
```

## Advanced Mutation Techniques

### Conditional Mutations with if_else()

```{r}
#| label: advanced-if-else

sales_conditional <- sales_cumulative %>%
  mutate(
    # Simple conditional
    high_value = if_else(net_amount > 500, "High Value", "Standard"),

    # Nested conditionals
    customer_status = if_else(
      customer_type == "Enterprise",
      if_else(net_amount > 1000, "Enterprise VIP", "Enterprise Standard"),
      if_else(net_amount > 300, "High Value Individual", "Standard Individual")
    ),

    # Conditional with different data types
    discount_description = if_else(
      !is.na(discount_pct) & discount_pct > 0,
      paste0(round(discount_pct * 100, 0), "% discount applied"),
      "No discount"
    ),

    # Complex business logic
    approval_required = if_else(
      (customer_type == "Individual" & net_amount > 1000) |
      (customer_type == "Small Business" & net_amount > 2000) |
      (customer_type == "Enterprise" & net_amount > 5000),
      "Manager Approval Required",
      "Auto Approved"
    ),

    # Performance-based conditionals
    rep_performance = if_else(
      !is.na(sales_rep),
      case_when(
        profit_margin > 0.25 ~ "Excellent",
        profit_margin > 0.15 ~ "Good",
        profit_margin > 0.05 ~ "Satisfactory",
        TRUE ~ "Needs Improvement"
      ),
      "No Rep Assigned"
    )
  )

cat("Conditional mutation results:\n")
sales_conditional %>%
  count(customer_status, approval_required) %>%
  print()
```

### Working with Multiple Variables

```{r}
#| label: multiple-variables

# Using across() for multiple column operations
sales_normalized <- sales_conditional %>%
  mutate(
    # Scale numeric variables
    across(c(quantity, unit_price, net_amount), ~ scale(.x)[,1], .names = "{.col}_scaled"),

    # Log transform financial variables
    across(c(gross_amount, net_amount, profit), ~ log1p(.x), .names = "{.col}_log"),

    # Create text versions of numeric variables
    across(c(net_amount, profit), ~ paste0("$", format(.x, big.mark = ",")), .names = "{.col}_formatted"),

    # Apply business rules across columns
    across(ends_with("_amount"), ~ if_else(.x < 0, 0, .x), .names = "{.col}_positive")
  )

cat("Multiple variable operations:\n")
sales_normalized %>%
  select(net_amount, net_amount_scaled, net_amount_log,
         net_amount_formatted, net_amount_positive) %>%
  head(5) %>%
  print()
```

## Real-World Examples

### Example 1: Customer Lifetime Value Calculation

```{r}
#| label: customer-ltv

customer_ltv <- sales_cumulative %>%
  group_by(customer_id) %>%
  arrange(order_date) %>%
  mutate(
    # Customer order sequence
    order_sequence = row_number(),

    # Time-based calculations
    customer_tenure_days = as.numeric(max(order_date) - min(order_date)),
    days_since_first_order = as.numeric(order_date - min(order_date)),

    # Value calculations
    customer_total_value = sum(net_amount),
    customer_avg_order_value = mean(net_amount),
    order_value_percentile = percent_rank(net_amount),

    # Frequency calculations
    customer_order_frequency = customer_total_value / pmax(customer_tenure_days, 1),

    # Recency calculations
    days_since_last_order = as.numeric(max(order_date) - order_date),

    # Customer lifecycle stage
    lifecycle_stage = case_when(
      order_sequence == 1 ~ "New",
      customer_tenure_days <= 90 ~ "Developing",
      days_since_last_order <= 30 ~ "Active",
      days_since_last_order <= 90 ~ "At Risk",
      TRUE ~ "Inactive"
    ),

    # Predicted lifetime value (simplified model)
    predicted_ltv = customer_avg_order_value *
                   (365 / pmax(customer_tenure_days / pmax(order_sequence - 1, 1), 30)) * 2,

    # Customer score (simplified)
    customer_score = scale(customer_total_value)[,1] +
                    scale(customer_order_frequency)[,1] +
                    scale(-days_since_last_order)[,1]
  ) %>%
  ungroup()

cat("Customer Lifetime Value Analysis:\n")
customer_ltv %>%
  group_by(customer_id) %>%
  slice_tail(n = 1) %>%  # Get latest record per customer
  select(customer_id, customer_total_value, customer_avg_order_value,
         lifecycle_stage, predicted_ltv, customer_score) %>%
  arrange(desc(customer_score)) %>%
  head(10) %>%
  print()
```

### Example 2: Sales Performance Dashboard Data

```{r}
#| label: sales-dashboard

dashboard_data <- sales_cumulative %>%
  mutate(
    # Time period classifications
    period_type = case_when(
      order_date >= Sys.Date() - days(7) ~ "This Week",
      order_date >= Sys.Date() - days(30) ~ "This Month",
      order_date >= Sys.Date() - days(90) ~ "This Quarter",
      TRUE ~ "Historical"
    ),

    # Performance metrics
    is_target_achieved = net_amount >= 500,  # Example target
    performance_vs_target = net_amount / 500,

    # Trend indicators
    order_month_number = as.numeric(format(order_date, "%Y%m")),
    is_current_month = order_month_number == as.numeric(format(Sys.Date(), "%Y%m")),

    # KPI classifications
    kpi_revenue_tier = case_when(
      net_amount >= quantile(net_amount, 0.9, na.rm = TRUE) ~ "Top 10%",
      net_amount >= quantile(net_amount, 0.75, na.rm = TRUE) ~ "Top 25%",
      net_amount >= quantile(net_amount, 0.5, na.rm = TRUE) ~ "Above Median",
      TRUE ~ "Below Median"
    ),

    # Alert flags
    alert_high_value = net_amount > 2000,
    alert_low_margin = profit_margin < 0.1,
    alert_large_discount = coalesce(discount_pct, 0) > 0.15,

    # Summary flags
    needs_attention = alert_low_margin | alert_large_discount,
    opportunity_flag = alert_high_value & !needs_attention
  ) %>%
  # Add period comparisons
  group_by(period_type) %>%
  mutate(
    period_avg_order = mean(net_amount, na.rm = TRUE),
    period_performance = net_amount / period_avg_order,
    period_rank = row_number(desc(net_amount))
  ) %>%
  ungroup()

# Create dashboard summary
dashboard_summary <- dashboard_data %>%
  group_by(period_type, kpi_revenue_tier) %>%
  summarise(
    order_count = n(),
    total_revenue = sum(net_amount),
    avg_order_value = mean(net_amount),
    alert_count = sum(needs_attention),
    opportunity_count = sum(opportunity_flag),
    .groups = "drop"
  ) %>%
  arrange(match(period_type, c("This Week", "This Month", "This Quarter", "Historical")))

cat("Sales Dashboard Summary:\n")
print(dashboard_summary)

cat("\nHigh Priority Alerts:\n")
dashboard_data %>%
  filter(needs_attention) %>%
  select(order_id, order_date, net_amount, profit_margin, discount_pct,
         alert_low_margin, alert_large_discount) %>%
  head(5) %>%
  print()
```

### Example 3: Inventory and Product Analysis

```{r}
#| label: inventory-analysis

product_analysis <- sales_cumulative %>%
  group_by(product_id) %>%
  arrange(order_date) %>%
  mutate(
    # Product performance over time
    product_total_sales = sum(net_amount),
    product_total_units = sum(quantity),
    product_avg_price = mean(unit_price),

    # Product lifecycle indicators
    product_first_sale = min(order_date),
    product_last_sale = max(order_date),
    product_days_active = as.numeric(product_last_sale - product_first_sale),

    # Sales velocity
    days_since_product_launch = as.numeric(order_date - product_first_sale),
    product_cumulative_sales = cumsum(net_amount),
    sales_velocity = product_cumulative_sales / pmax(days_since_product_launch, 1),

    # Product ranking within its timeline
    product_order_sequence = row_number(),

    # Seasonal patterns
    product_quarter_sales = quarter(order_date),

    # Product performance classification
    product_performance = case_when(
      product_total_sales >= quantile(product_total_sales, 0.8, na.rm = TRUE) ~ "Star",
      product_total_sales >= quantile(product_total_sales, 0.6, na.rm = TRUE) ~ "Rising",
      product_total_sales >= quantile(product_total_sales, 0.4, na.rm = TRUE) ~ "Steady",
      product_total_sales >= quantile(product_total_sales, 0.2, na.rm = TRUE) ~ "Declining",
      TRUE ~ "Lagging"
    )
  ) %>%
  ungroup() %>%
  # Add cross-product comparisons
  mutate(
    product_market_share = product_total_sales / sum(product_total_sales) * 100,
    product_price_position = case_when(
      product_avg_price >= quantile(product_avg_price, 0.8, na.rm = TRUE) ~ "Premium",
      product_avg_price >= quantile(product_avg_price, 0.6, na.rm = TRUE) ~ "High",
      product_avg_price >= quantile(product_avg_price, 0.4, na.rm = TRUE) ~ "Medium",
      product_avg_price >= quantile(product_avg_price, 0.2, na.rm = TRUE) ~ "Low",
      TRUE ~ "Budget"
    )
  )

# Product portfolio summary
product_portfolio <- product_analysis %>%
  group_by(product_id) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  select(product_id, product_performance, product_price_position,
         product_total_sales, product_market_share, sales_velocity) %>%
  arrange(desc(product_total_sales))

cat("Product Portfolio Analysis (Top 10):\n")
print(head(product_portfolio, 10))

cat("\nProduct Performance Distribution:\n")
product_portfolio %>%
  count(product_performance, product_price_position) %>%
  print()
```

## Best Practices and Performance Tips

### Efficient Mutations

```{r}
#| label: efficient-mutations

# Good practices for performance
efficient_mutations <- sales_data %>%
  mutate(
    # Calculate once, use multiple times
    gross = quantity * unit_price,
    discount_amt = gross * coalesce(discount_pct, 0),
    net = gross - discount_amt,

    # Use vectorized operations
    is_large_order = net > 500,
    order_category = if_else(is_large_order, "Large", "Standard"),

    # Avoid redundant calculations
    profit_simple = net * 0.4,  # Instead of complex profit calculations

    # Use case_when efficiently
    priority = case_when(
      customer_type == "Enterprise" & net > 1000 ~ 1L,
      customer_type == "Enterprise" ~ 2L,
      net > 500 ~ 3L,
      TRUE ~ 4L
    )
  )

cat("Efficient mutations completed successfully\n")
```

### Common Pitfalls to Avoid

```{r}
#| label: common-pitfalls

# Demonstrate common issues and solutions

# Issue 1: Order of operations in mutate()
# Bad: trying to use a variable before it's created
# good_order <- sales_data %>%
#   mutate(
#     net_amount = gross_amount - discount_amount,  # Error: gross_amount doesn't exist yet
#     gross_amount = quantity * unit_price
#   )

# Good: create variables in the right order
good_order <- sales_data %>%
  mutate(
    gross_amount = quantity * unit_price,
    discount_amount = gross_amount * coalesce(discount_pct, 0),
    net_amount = gross_amount - discount_amount
  )

# Issue 2: Division by zero
safe_calculations <- sales_data %>%
  mutate(
    # Bad: could divide by zero
    # ratio = net_amount / quantity  # Error if quantity is 0

    # Good: handle zero values
    ratio = if_else(quantity > 0,
                   (quantity * unit_price) / quantity,
                   NA_real_)
  )

# Issue 3: Missing value propagation
na_handling <- sales_data %>%
  mutate(
    # NA values propagate through calculations
    discount_amt = coalesce(discount_pct, 0) * quantity * unit_price,

    # Explicit NA handling
    has_discount = !is.na(discount_pct) & discount_pct > 0,

    # Safe calculations with NA
    adjusted_price = case_when(
      is.na(discount_pct) ~ unit_price,
      discount_pct == 0 ~ unit_price,
      TRUE ~ unit_price * (1 - discount_pct)
    )
  )

cat("Common pitfalls addressed successfully\n")
```

## Exercises

### Exercise 1: Financial Analysis
Create a comprehensive financial analysis using mutate():
1. Calculate profit margins, ROI, and markup percentages
2. Create financial performance categories
3. Add cumulative financial metrics
4. Build a profitability scoring system

### Exercise 2: Time Series Features
Build time-based features:
1. Extract all relevant date components
2. Calculate time-based lags and leads
3. Create seasonal indicators
4. Build time-based rankings and percentiles

### Exercise 3: Customer Segmentation
Develop a customer segmentation system:
1. Calculate RFM (Recency, Frequency, Monetary) scores
2. Create customer lifecycle stages
3. Build predictive features
4. Develop customer risk scores

### Exercise 4: Product Performance
Analyze product performance:
1. Calculate product velocity and trends
2. Create product lifecycle classifications
3. Build cross-selling opportunity indicators
4. Develop inventory optimization metrics

## Summary

Mutating and arranging are essential for data transformation and organization:

### Key Functions:
- **mutate()**: Create or modify variables
- **arrange()**: Sort data by one or more variables
- **case_when()**: Multi-condition logic
- **if_else()**: Binary conditional logic

### Window Functions:
- **Ranking**: `row_number()`, `rank()`, `percent_rank()`
- **Offset**: `lag()`, `lead()`
- **Cumulative**: `cumsum()`, `cummean()`, `cummax()`

### Best Practices:
- **Order matters** in mutate() - create variables in logical sequence
- **Handle missing values** explicitly with `coalesce()` and `if_else()`
- **Use case_when()** for complex multi-condition logic
- **Group operations** when you need relative calculations
- **Think about performance** - avoid redundant calculations

### Common Patterns:
- Financial calculations and derived metrics
- Date/time feature engineering
- Categorical variable creation
- Ranking and percentile calculations
- Cumulative and rolling metrics

### Remember:
- mutate() preserves all original columns unless overwritten
- arrange() changes row order but preserves all data
- Window functions work within groups when data is grouped
- Always test complex calculations with simple examples first

These transformation skills enable you to prepare raw data for analysis and create the features needed for insights and modeling!

Next: **[Grouping and Summarizing Data](grouping-summarizing.qmd)**