---
title: "Control Structures in R"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to Control Structures

Control structures allow you to control the flow of execution in your R programs. They enable you to:

- Make decisions based on conditions (if/else statements)
- Repeat operations multiple times (loops)
- Handle different cases efficiently (switch statements)

While R's vectorized operations often eliminate the need for explicit loops, understanding control structures is essential for more complex programming tasks.

## Conditional Statements

### The `if` Statement

The most basic control structure tests a condition and executes code if the condition is true:

```{r}
#| label: basic-if

# Basic if statement
score <- 85

if (score >= 80) {
  print("You passed!")
}

# The condition must evaluate to a single TRUE or FALSE
x <- 10
if (x > 5) {
  cat("x is greater than 5\n")
}
```

### The `if-else` Statement

Use `else` to specify what happens when the condition is false:

```{r}
#| label: if-else

# Simple if-else
temperature <- 75

if (temperature > 80) {
  print("It's hot outside!")
} else {
  print("It's not too hot.")
}

# Another example
age <- 17

if (age >= 18) {
  status <- "adult"
} else {
  status <- "minor"
}

cat("Status:", status, "\n")
```

### Multiple Conditions with `else if`

Chain multiple conditions together:

```{r}
#| label: else-if

# Grade assignment
score <- 87

if (score >= 90) {
  grade <- "A"
} else if (score >= 80) {
  grade <- "B"
} else if (score >= 70) {
  grade <- "C"
} else if (score >= 60) {
  grade <- "D"
} else {
  grade <- "F"
}

cat("Your grade is:", grade, "\n")
```

### Nested `if` Statements

You can nest if statements inside other if statements:

```{r}
#| label: nested-if

# Weather decision making
temperature <- 75
is_raining <- FALSE

if (temperature > 70) {
  if (is_raining) {
    print("Warm but rainy - bring an umbrella!")
  } else {
    print("Perfect weather for a walk!")
  }
} else {
  if (is_raining) {
    print("Cold and rainy - stay inside!")
  } else {
    print("Cold but dry - wear a jacket!")
  }
}
```

### Vectorized Conditional: `ifelse()`

For vectorized operations, use `ifelse()`:

```{r}
#| label: ifelse-function

# Apply condition to entire vector
scores <- c(85, 92, 78, 96, 73, 88)

# Traditional if would only work with single values
# This creates a vector of results
results <- ifelse(scores >= 80, "Pass", "Fail")
print(results)

# More complex example
temperatures <- c(68, 75, 82, 79, 71)
weather_description <- ifelse(temperatures > 80, "Hot",
                             ifelse(temperatures > 70, "Warm", "Cool"))
print(weather_description)

# Numeric results
bonus <- ifelse(scores >= 90, scores * 0.1, 0)
print(bonus)
```

## Logical Operators in Conditions

### Basic Logical Operators

```{r}
#| label: logical-operators

age <- 25
income <- 45000
has_degree <- TRUE

# AND operator (&)
if (age >= 18 & income > 30000) {
  print("Eligible for loan")
}

# OR operator (|)
if (age < 18 | age > 65) {
  print("Special rate applies")
}

# NOT operator (!)
if (!has_degree) {
  print("Degree required")
} else {
  print("Education requirement met")
}

# Complex conditions
if ((age >= 21 & age <= 65) & (income > 40000 | has_degree)) {
  print("Premium membership available")
}
```

### Comparison Operators

```{r}
#| label: comparison-operators

x <- 10
y <- 20

# All comparison operators
print(x == y)    # Equal to
print(x != y)    # Not equal to
print(x < y)     # Less than
print(x <= y)    # Less than or equal to
print(x > y)     # Greater than
print(x >= y)    # Greater than or equal to

# Using with strings
name1 <- "Alice"
name2 <- "Bob"
print(name1 == name2)
print(name1 < name2)  # Alphabetical comparison
```

## Loops

### The `for` Loop

`for` loops iterate over a sequence of values:

```{r}
#| label: for-loop-basic

# Basic for loop
for (i in 1:5) {
  print(paste("Iteration:", i))
}

# Loop over vector elements
fruits <- c("apple", "banana", "cherry")
for (fruit in fruits) {
  print(paste("I like", fruit))
}

# Loop with calculations
squares <- numeric(5)  # Pre-allocate vector
for (i in 1:5) {
  squares[i] <- i^2
}
print(squares)
```

### Practical `for` Loop Examples

```{r}
#| label: for-loop-practical

# Calculate cumulative sum
numbers <- c(5, 10, 15, 20, 25)
cumulative <- numeric(length(numbers))
cumulative[1] <- numbers[1]

for (i in 2:length(numbers)) {
  cumulative[i] <- cumulative[i-1] + numbers[i]
}
print(cumulative)

# Compare with built-in function
print(cumsum(numbers))  # R's built-in cumulative sum

# Process data frame rows
students <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  midterm = c(85, 78, 92),
  final = c(88, 82, 89)
)

for (i in 1:nrow(students)) {
  overall <- 0.4 * students$midterm[i] + 0.6 * students$final[i]
  cat(students$name[i], "overall grade:", round(overall, 1), "\n")
}
```

### The `while` Loop

`while` loops continue until a condition becomes false:

```{r}
#| label: while-loop

# Basic while loop
counter <- 1
while (counter <= 5) {
  print(paste("Count:", counter))
  counter <- counter + 1
}

# While loop with condition
balance <- 1000
interest_rate <- 0.05
year <- 0

while (balance < 2000) {
  balance <- balance * (1 + interest_rate)
  year <- year + 1
}

cat("It takes", year, "years for balance to exceed $2000\n")
cat("Final balance: $", round(balance, 2), "\n")
```

### Loop Control: `break` and `next`

```{r}
#| label: loop-control

# Using break to exit loop early
for (i in 1:10) {
  if (i == 6) {
    print("Breaking at 6")
    break
  }
  print(i)
}

# Using next to skip iterations
for (i in 1:10) {
  if (i %% 2 == 0) {  # Skip even numbers
    next
  }
  print(paste("Odd number:", i))
}
```

## Alternative to Loops: Vectorization

In R, vectorized operations are usually preferred over loops:

```{r}
#| label: vectorization-vs-loops

# Calculate squares using loop (slower)
n <- 1000
squares_loop <- numeric(n)
system.time({
  for (i in 1:n) {
    squares_loop[i] <- i^2
  }
})

# Calculate squares using vectorization (faster)
system.time({
  squares_vector <- (1:n)^2
})

# Both give same result
identical(squares_loop, squares_vector)

# More examples of vectorization
numbers <- 1:100

# Instead of loop for condition
# Loop approach:
# result <- numeric(100)
# for (i in 1:100) {
#   if (numbers[i] > 50) result[i] <- "high" else result[i] <- "low"
# }

# Vectorized approach:
result <- ifelse(numbers > 50, "high", "low")
```

## The `switch` Statement

For multiple discrete choices, `switch` can be cleaner than multiple `if-else`:

```{r}
#| label: switch-statement

# Basic switch with character
day_type <- function(day) {
  switch(day,
    "Monday" = "Start of work week",
    "Tuesday" = "Work day",
    "Wednesday" = "Hump day",
    "Thursday" = "Almost there",
    "Friday" = "TGIF!",
    "Saturday" = "Weekend!",
    "Sunday" = "Weekend!",
    "Unknown day"  # Default case
  )
}

print(day_type("Friday"))
print(day_type("Sunday"))
print(day_type("Holiday"))

# Switch with numeric values
grade_points <- function(letter_grade) {
  switch(letter_grade,
    "A" = 4.0,
    "B" = 3.0,
    "C" = 2.0,
    "D" = 1.0,
    "F" = 0.0,
    NA  # Default for invalid grades
  )
}

print(grade_points("A"))
print(grade_points("C"))
print(grade_points("X"))
```

## Practical Examples

### Example 1: Data Validation and Cleaning

```{r}
#| label: data-validation

# Simulate messy data
raw_data <- data.frame(
  age = c(25, -5, 150, 30, 22),
  income = c(50000, 75000, -1000, 85000, 45000),
  score = c(85, 92, 78, 150, 88)
)

# Clean the data using control structures
cleaned_data <- raw_data

for (i in 1:nrow(cleaned_data)) {
  # Validate age
  if (cleaned_data$age[i] < 0 | cleaned_data$age[i] > 120) {
    cleaned_data$age[i] <- NA
    cat("Invalid age in row", i, "- set to NA\n")
  }

  # Validate income
  if (cleaned_data$income[i] < 0) {
    cleaned_data$income[i] <- NA
    cat("Invalid income in row", i, "- set to NA\n")
  }

  # Validate score (0-100 range)
  if (cleaned_data$score[i] < 0 | cleaned_data$score[i] > 100) {
    cleaned_data$score[i] <- NA
    cat("Invalid score in row", i, "- set to NA\n")
  }
}

print("Original data:")
print(raw_data)
print("Cleaned data:")
print(cleaned_data)
```

### Example 2: Simulation and Analysis

```{r}
#| label: simulation-analysis

# Simulate rolling dice until we get a 6
set.seed(123)
simulate_dice <- function() {
  rolls <- 0
  result <- 0

  while (result != 6) {
    result <- sample(1:6, 1)
    rolls <- rolls + 1
    cat("Roll", rolls, ":", result, "\n")
  }

  return(rolls)
}

cat("Simulation 1:\n")
rolls_needed <- simulate_dice()
cat("Took", rolls_needed, "rolls to get a 6\n\n")

# Run multiple simulations
num_simulations <- 1000
all_rolls <- numeric(num_simulations)

for (i in 1:num_simulations) {
  rolls <- 0
  result <- 0

  while (result != 6) {
    result <- sample(1:6, 1)
    rolls <- rolls + 1
  }

  all_rolls[i] <- rolls
}

cat("Average rolls needed:", mean(all_rolls), "\n")
cat("Maximum rolls needed:", max(all_rolls), "\n")
cat("95% of games finished within", quantile(all_rolls, 0.95), "rolls\n")
```

### Example 3: Processing Survey Data

```{r}
#| label: survey-processing

# Survey responses
survey_data <- data.frame(
  respondent = 1:10,
  satisfaction = c(5, 4, 3, 5, 2, 4, 5, 3, 4, 5),
  recommend = c("Yes", "Yes", "No", "Yes", "No", "Maybe", "Yes", "No", "Yes", "Yes"),
  comments = c("Great!", "", "Could be better", "Excellent", "Poor service",
              "", "Amazing!", "Not satisfied", "", "Very good")
)

# Process and categorize responses
processed_data <- survey_data

for (i in 1:nrow(processed_data)) {
  # Categorize satisfaction scores
  if (processed_data$satisfaction[i] >= 4) {
    processed_data$satisfaction_category[i] <- "High"
  } else if (processed_data$satisfaction[i] >= 3) {
    processed_data$satisfaction_category[i] <- "Medium"
  } else {
    processed_data$satisfaction_category[i] <- "Low"
  }

  # Convert recommendation to numeric score
  rec_score <- switch(processed_data$recommend[i],
    "Yes" = 3,
    "Maybe" = 2,
    "No" = 1,
    0  # Default for unexpected values
  )
  processed_data$recommend_score[i] <- rec_score

  # Flag responses with comments
  processed_data$has_comment[i] <- processed_data$comments[i] != ""
}

# Summary analysis
cat("Satisfaction Distribution:\n")
print(table(processed_data$satisfaction_category))

cat("\nRecommendation Distribution:\n")
print(table(processed_data$recommend))

cat("\nPercent with comments:",
    round(mean(processed_data$has_comment) * 100, 1), "%\n")

# Find dissatisfied customers with comments
dissatisfied_with_comments <- processed_data[
  processed_data$satisfaction <= 2 & processed_data$has_comment,
  c("respondent", "satisfaction", "comments")
]

if (nrow(dissatisfied_with_comments) > 0) {
  cat("\nDissatisfied customers with comments:\n")
  print(dissatisfied_with_comments)
}
```

## When to Use Control Structures vs. Vectorization

### Use Control Structures When:

1. **Complex logic**: Multiple conditions that can't be easily vectorized
2. **Sequential dependencies**: Each iteration depends on the previous result
3. **Early termination**: Need to stop based on a condition
4. **Complex data structures**: Working with lists or nested structures

### Use Vectorization When:

1. **Simple operations**: Applying the same operation to all elements
2. **Mathematical calculations**: Arithmetic operations on vectors
3. **Logical filtering**: Subsetting based on conditions
4. **Performance matters**: Vectorized operations are typically faster

```{r}
#| label: when-to-use-what

# Control structure appropriate: Fibonacci sequence
fibonacci <- function(n) {
  if (n <= 2) return(1)

  fib <- numeric(n)
  fib[1:2] <- 1

  for (i in 3:n) {
    fib[i] <- fib[i-1] + fib[i-2]
  }

  return(fib)
}

print(fibonacci(10))

# Vectorization appropriate: Simple transformations
data <- 1:1000
squared <- data^2  # Much faster than loop
log_values <- log(data)  # Vectorized function
```

## Best Practices and Common Pitfalls

### 1. Pre-allocate Vectors in Loops

```{r}
#| label: pre-allocation

# Bad: Growing vector in loop (slow)
# result <- c()
# for (i in 1:1000) {
#   result <- c(result, i^2)
# }

# Good: Pre-allocate (fast)
result <- numeric(1000)
for (i in 1:1000) {
  result[i] <- i^2
}
```

### 2. Avoid Unnecessary Loops

```{r}
#| label: avoid-loops

# Instead of loop:
# result <- numeric(length(data))
# for (i in 1:length(data)) {
#   result[i] <- data[i] * 2
# }

# Use vectorization:
data <- 1:100
result <- data * 2
```

### 3. Be Careful with Conditions

```{r}
#| label: condition-safety

# Dangerous: vector in condition
scores <- c(85, 92, 78)
# if (scores > 80) { ... }  # This would cause an error!

# Safe: single logical value
if (any(scores > 80)) {
  print("At least one score above 80")
}

if (all(scores > 80)) {
  print("All scores above 80")
}
```

## Exercises

### Exercise 1: Grade Calculator
Write a function that takes a numeric score and returns the letter grade using if-else statements:
- A: 90-100
- B: 80-89
- C: 70-79
- D: 60-69
- F: Below 60

### Exercise 2: Number Guessing Game
Create a simple number guessing game where:
1. The computer picks a random number between 1-100
2. User has to guess (simulate with a loop)
3. Provide "higher" or "lower" hints
4. Count the number of guesses

### Exercise 3: Data Processing
Given a vector of temperatures, write code that:
1. Uses a loop to categorize each temperature as "Freezing", "Cold", "Mild", "Warm", or "Hot"
2. Counts how many days fall into each category
3. Compares this approach with a vectorized solution

### Exercise 4: Compound Interest Calculator
Write a function that calculates compound interest using a while loop. The function should:
1. Take initial amount, interest rate, and target amount
2. Calculate how many years to reach the target
3. Print yearly balances along the way

## Summary

Control structures are essential tools for programming logic:

### Key Concepts:
- **Conditional statements**: `if`, `else if`, `else` for decision making
- **Loops**: `for` and `while` for repetitive operations
- **Loop control**: `break` and `next` for flow control
- **Vectorized alternatives**: Often faster and more R-like

### Best Practices:
- **Prefer vectorization** when possible for performance
- **Pre-allocate vectors** in loops to avoid memory issues
- **Use appropriate structures** for the task at hand
- **Test conditions carefully** to avoid errors

### Remember:
- R is designed for vectorized operations
- Control structures are powerful but should be used judiciously
- Always consider if there's a vectorized alternative
- Complex logic often requires control structures

Next, we'll explore how to organize code into reusable functions!