---
title: "Data Types and Objects in R"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to R Data Types

In R, everything is an object, and every object has a **data type**. Understanding data types is fundamental to working effectively with R because it determines what operations you can perform and how R stores and processes your data.

R has four fundamental data types:

- **integer**: Whole numbers
- **double**: Real numbers (with decimal points)
- **character**: Text/strings
- **logical**: TRUE/FALSE values

Let's explore each of these in detail.

## Numeric Data Types

### Integers and Doubles

R distinguishes between two types of numbers:

#### Doubles (Real Numbers)

By default, R treats all numbers as **double** (also called "numeric"):

```{r}
#| label: doubles-basic

# These are all doubles by default
x <- 1.2
y <- 3.0
z <- 5

# Check their types
typeof(x)
typeof(y)
typeof(z)  # Even though 5 looks like an integer, it's a double!
```

#### Integers

To create an integer, you must explicitly specify it by adding `L` after the number:

```{r}
#| label: integers-basic

# Creating integers
a <- 2L
b <- -123456789L
c <- 0L

# Check their types
typeof(a)
typeof(b)
typeof(c)
```

#### Why the Distinction Matters

The difference between integers and doubles affects:

1. **Memory usage**: Integers use less memory
2. **Precision**: Doubles can represent fractional values
3. **Operations**: Some functions behave differently with different types

```{r}
#| label: type-implications

# Mathematical operations
double_result <- 3.5 + 1.2
typeof(double_result)

integer_result <- 3L + 4L
typeof(integer_result)  # Still integer!

mixed_result <- 3L + 1.3  # Integer + double = double
typeof(mixed_result)
```

### Working with Large Numbers

R can handle very large numbers, but there are limits:

```{r}
#| label: large-numbers

# Scientific notation
large_number <- 1.5e8
print(large_number)

# Very large numbers become Inf (infinity)
too_large <- 1e1000
print(too_large)

# Very small numbers become 0
too_small <- 1e-1000
print(too_small)
```

### Checking and Converting Numeric Types

```{r}
#| label: numeric-functions

# Check if something is numeric
is.numeric(3.14)
is.numeric("hello")

# Check specific types
is.integer(5L)
is.double(5.0)

# Convert between types
as.integer(3.7)    # Truncates, doesn't round!
as.double(5L)      # Convert integer to double
```

## Character Data (Strings)

Character data represents text and is created using quotation marks:

### Creating Strings

```{r}
#| label: strings-basic

# Using double quotes
text1 <- "Hello, world!"

# Using single quotes
text2 <- 'This is also a string'

# Mixing quotes (useful when string contains quotes)
text3 <- "She said 'Hello!'"
text4 <- 'He replied "Hi there!"'

print(text1)
print(text3)
```

### String Properties

```{r}
#| label: string-properties

my_string <- "Data Science is awesome!"

# Check type
typeof(my_string)

# Check if it's character
is.character(my_string)

# Get string length
nchar(my_string)
```

### Working with Strings

```{r}
#| label: string-operations

# String concatenation
first_name <- "Jane"
last_name <- "Doe"
full_name <- paste(first_name, last_name)
print(full_name)

# Alternative concatenation
full_name2 <- paste0(first_name, " ", last_name)  # No space separator
print(full_name2)

# Substrings
substr(full_name, 1, 4)  # Characters 1 through 4
```

### Converting Between Numbers and Strings

```{r}
#| label: string-conversions

# Numbers to strings
age <- 25
age_text <- as.character(age)
print(age_text)
typeof(age_text)

# Strings to numbers (if they represent numbers)
number_text <- "42"
number_value <- as.numeric(number_text)
print(number_value)
typeof(number_value)

# What happens with non-numeric strings?
invalid_number <- as.numeric("hello")
print(invalid_number)  # Returns NA (Not Available)
```

## Logical Data (Boolean)

Logical data represents TRUE/FALSE values and is fundamental for conditional operations:

### Creating Logical Values

```{r}
#| label: logical-basic

# Direct assignment
is_student <- TRUE
has_job <- FALSE

# From comparisons
x <- 10
y <- 5

is_greater <- x > y
is_equal <- x == y
is_not_equal <- x != y

print(is_greater)
print(is_equal)
print(is_not_equal)
```

### Comparison Operators

R provides six comparison operators:

```{r}
#| label: comparison-operators

a <- 6
b <- 3

# All comparison operators
a < b   # Less than
a > b   # Greater than
a <= b  # Less than or equal to
a >= b  # Greater than or equal to
a == b  # Equal to
a != b  # Not equal to
```

### Logical Operations

Combine logical values using logical operators:

```{r}
#| label: logical-operations

x <- TRUE
y <- FALSE

# AND operation
x & y   # FALSE (both must be TRUE)

# OR operation
x | y   # TRUE (at least one must be TRUE)

# NOT operation
!x      # FALSE (flips the value)
!y      # TRUE

# Complex logical expressions
age <- 25
has_license <- TRUE
can_drive <- (age >= 16) & has_license
print(can_drive)
```

### Logical Values in Arithmetic

```{r}
#| label: logical-arithmetic

# TRUE = 1, FALSE = 0 in arithmetic operations
TRUE + TRUE    # 2
FALSE + TRUE   # 1
TRUE * 5       # 5

# Counting TRUE values
scores <- c(85, 92, 78, 96, 88)
passing_grades <- scores >= 80
sum(passing_grades)  # Count how many passed
mean(passing_grades) # Proportion who passed
```

## Special Values

R has several special values you should know about:

### NA (Not Available)

Represents missing data:

```{r}
#| label: na-values

# Creating NA values
missing_value <- NA
ages <- c(25, 30, NA, 35, 28)

# Check for NA
is.na(missing_value)
is.na(ages)

# Operations with NA
mean(ages)                    # Returns NA
mean(ages, na.rm = TRUE)      # Remove NA values first
```

### NULL

Represents "nothing" or absence of a value:

```{r}
#| label: null-values

# NULL represents absence
empty_var <- NULL
length(empty_var)  # 0

# NULL vs NA
is.null(NULL)
is.null(NA)
is.na(NULL)
is.na(NA)
```

### Inf and -Inf

Represent positive and negative infinity:

```{r}
#| label: infinity

# Division by zero
positive_inf <- 1/0
negative_inf <- -1/0

print(positive_inf)
print(negative_inf)

# Check for infinity
is.infinite(positive_inf)
is.finite(positive_inf)
```

## Type Checking and Conversion

### Checking Data Types

```{r}
#| label: type-checking

# Create different types
int_val <- 42L
dbl_val <- 3.14
chr_val <- "hello"
log_val <- TRUE

# typeof() shows the exact type
typeof(int_val)
typeof(dbl_val)
typeof(chr_val)
typeof(log_val)

# class() shows the object class
class(int_val)
class(dbl_val)

# Specific type checks
is.numeric(int_val)    # TRUE (integers are numeric)
is.integer(int_val)    # TRUE
is.double(int_val)     # FALSE
is.character(chr_val)  # TRUE
is.logical(log_val)    # TRUE
```

### Type Conversion (Coercion)

R can convert between types:

```{r}
#| label: type-conversion

# Explicit conversion
x <- 3.7
as.integer(x)      # Truncates to 3
as.character(x)    # "3.7"
as.logical(x)      # TRUE (non-zero numbers are TRUE)

# Converting strings to numbers
text_numbers <- c("1", "2.5", "3")
numeric_values <- as.numeric(text_numbers)
print(numeric_values)

# What happens with invalid conversions?
as.numeric(c("1", "hello", "3"))  # Returns c(1, NA, 3)
```

### Automatic Type Conversion

R automatically converts types when needed:

```{r}
#| label: automatic-conversion

# Mixing types in operations
result1 <- 5L + 3.2      # integer + double = double
typeof(result1)

result2 <- TRUE + 5      # logical + numeric = numeric
print(result2)           # TRUE becomes 1

# result3 <- "Number: " + 5  # This will cause an error!
# Instead, use paste():
result3 <- paste("Number:", 5)
print(result3)
```

## Practical Examples

### Example 1: Survey Data Analysis

```{r}
#| label: survey-example

# Survey responses
respondent_id <- 1:5
age <- c(25, 30, 28, 35, 29)
income <- c(45000, 52000, 48000, 65000, 51000)
satisfied <- c(TRUE, TRUE, FALSE, TRUE, TRUE)
feedback <- c("Great!", "Good service", "Could be better", "Excellent", "Very happy")

# Analysis
avg_age <- mean(age)
avg_income <- mean(income)
satisfaction_rate <- mean(satisfied)
num_responses <- length(respondent_id)

# Create summary
summary_text <- paste("Survey Results:",
                     "\nAverage age:", round(avg_age, 1),
                     "\nAverage income: $", format(avg_income, big.mark = ","),
                     "\nSatisfaction rate:", round(satisfaction_rate * 100, 1), "%",
                     "\nTotal responses:", num_responses)

cat(summary_text)
```

### Example 2: Data Quality Checks

```{r}
#| label: data-quality

# Simulated data with quality issues
temperatures <- c(72, 75, NA, 80, 999, -50, 77)

# Quality checks
valid_range <- temperatures >= 0 & temperatures <= 120
has_missing <- is.na(temperatures)
suspicious_values <- temperatures > 100 | temperatures < 0

# Results
cat("Valid temperatures:", sum(valid_range, na.rm = TRUE), "\n")
cat("Missing values:", sum(has_missing), "\n")
cat("Suspicious values:", sum(suspicious_values, na.rm = TRUE), "\n")

# Clean the data
clean_temperatures <- temperatures[valid_range & !has_missing]
print(clean_temperatures)
```

## Common Pitfalls and Best Practices

### 1. Integer vs Double Confusion

```{r}
#| label: integer-confusion

# This might surprise you!
x <- 5
y <- 5L

identical(x, y)  # FALSE! Different types
x == y          # TRUE (values are equal)

# Best practice: be explicit about integer types when needed
```

### 2. Character to Numeric Conversion

```{r}
#| label: character-numeric-confusion

# Common mistake
numbers_as_text <- c("1", "2", "3")
# numbers_as_text + 1  # This would error!

# Correct approach
numbers <- as.numeric(numbers_as_text)
numbers + 1  # Now this works
```

### 3. Logical Arithmetic

```{r}
#| label: logical-arithmetic-tips

# Useful for counting
test_scores <- c(85, 92, 78, 96, 75, 88)
passing_scores <- test_scores >= 80

# Count passing scores
num_passing <- sum(passing_scores)

# Percentage passing
pct_passing <- mean(passing_scores) * 100

cat("Students passing:", num_passing, "\n")
cat("Percentage passing:", round(pct_passing, 1), "%\n")
```

## Exercises

### Exercise 1: Type Exploration
Create variables of each data type and explore their properties:

```{r}
#| label: exercise-1-demo
#| eval: false

# Create one variable of each type
my_integer <- ___
my_double <- ___
my_character <- ___
my_logical <- ___

# Check their types using typeof()
# Convert between types using as.* functions
# Try some arithmetic operations
```

### Exercise 2: Real-World Data Types
Given this data about employees, determine the appropriate data type for each variable:

- Employee ID numbers
- Employee names
- Salaries
- Whether they work remotely
- Years of experience
- Department codes

### Exercise 3: Type Coercion Challenge
What will be the result and type of each expression?

```{r}
#| label: exercise-3-demo
#| eval: false

TRUE + 2
"5" + 3
as.logical(0)
as.logical(-1)
as.integer(3.9)
paste(TRUE, 5)
```

## Summary

Understanding R's data types is crucial for effective programming:

- **Numeric types**: integers (with `L`) and doubles (default for numbers)
- **Character type**: text strings in quotes
- **Logical type**: TRUE/FALSE values
- **Special values**: NA (missing), NULL (absent), Inf (infinity)

Key points to remember:

1. **Everything in R has a type**
2. **R can convert between types** (sometimes automatically)
3. **Check types when debugging** using `typeof()` and `class()`
4. **Be explicit about integers** using the `L` suffix
5. **Logical values are powerful** for filtering and counting

Next, we'll explore how to combine these basic types into vectors, R's fundamental data structure!