---
title: "Module 2: R Fundamentals"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
---

## Welcome to R Programming! 🚀

Now that you have R and RStudio set up, it's time to dive into the fundamentals of R programming. This module will introduce you to the core concepts that form the foundation of data analysis in R.

## What You'll Learn

In this module, we'll explore the essential building blocks of R programming:

### 1. Data Types and Objects
- Understanding R's fundamental data types: integers, doubles, characters, and logicals
- Working with different object types and their properties
- Type conversion and checking data types

### 2. Vectors: R's Building Blocks
- Creating and manipulating vectors
- Element-wise operations
- Vector indexing and subsetting
- Working with named vectors

### 3. Data Structures
- Lists: heterogeneous data containers
- Data frames: the backbone of data analysis
- Matrices and arrays for multidimensional data

### 4. Control Structures
- Conditional statements (if/else)
- Loops (for, while)
- Vectorized operations vs. loops

### 5. Functions
- Using built-in functions
- Creating your own functions
- Function arguments and return values
- Scope and environments

### 6. Working with Files
- Reading and writing data files
- File paths and project organization
- Basic data import/export

## Learning Objectives

By the end of this module, you will be able to:

- ✅ Understand and work with R's fundamental data types
- ✅ Create, manipulate, and subset vectors effectively
- ✅ Build and work with lists and data frames
- ✅ Write conditional statements and loops
- ✅ Create and use custom functions
- ✅ Import and export data files
- ✅ Organize your R projects effectively

## Why These Concepts Matter

Understanding R fundamentals is crucial because:

1. **Foundation for Everything**: Every data analysis task builds on these concepts
2. **Efficient Problem Solving**: Knowing the right data structure saves time and effort
3. **Debugging Skills**: Understanding how R works helps you fix problems
4. **Code Quality**: Well-structured code is easier to read, maintain, and share

## Module Overview

### The R Philosophy

R is designed around a few key principles:

- **Everything is an object**: In R, everything from numbers to functions is an object
- **Vectorization**: Operations work on entire vectors, not just single values
- **Functional programming**: Functions are first-class objects that can be passed around
- **Statistical focus**: Built with statistical analysis and data science in mind

### Getting Started

Let's begin with a simple example that demonstrates several key concepts:

```{r}
#| label: intro-example

# Create a vector of temperatures
temperatures <- c(72, 75, 68, 80, 77, 73, 71)

# Calculate basic statistics
avg_temp <- mean(temperatures)
max_temp <- max(temperatures)
min_temp <- min(temperatures)

# Create a summary data frame
temp_summary <- data.frame(
  statistic = c("Average", "Maximum", "Minimum"),
  value = c(avg_temp, max_temp, min_temp)
)

print(temp_summary)
```

This simple example demonstrates:
- Creating vectors with `c()`
- Using built-in functions like `mean()`, `max()`, `min()`
- Creating data frames
- Combining different data types

## Module Structure

This module is organized into six main sections:

1. **[Data Types and Objects](data-types.qmd)**: Learn about R's fundamental data types
2. **[Vectors](vectors.qmd)**: Master R's most important data structure
3. **[Lists and Data Frames](data-structures.qmd)**: Work with complex data containers
4. **[Control Structures](control-structures.qmd)**: Add logic and iteration to your code
5. **[Functions](functions.qmd)**: Create reusable code and understand scope
6. **[Working with Files](working-with-files.qmd)**: Import, export, and organize data

## Prerequisites

Before starting this module, make sure you have:

- R and RStudio installed (covered in Module 1)
- Basic familiarity with the RStudio interface
- Understanding of how to create and run R scripts

## Practice Philosophy

Throughout this module, we emphasize:

- **Hands-on learning**: Every concept includes practical examples
- **Real-world applications**: Examples use realistic data scenarios
- **Progressive complexity**: Start simple, build to more complex concepts
- **Common patterns**: Learn the most frequently used approaches

## Getting Help

As you work through this module, remember these resources:

- **R Help System**: Use `?function_name` to get help on any function
- **RStudio Help**: Use the Help tab in RStudio
- **Error Messages**: Read them carefully - they often contain helpful information
- **Practice**: The more you code, the more intuitive R becomes

## A Quick Motivation

Here's why learning R fundamentals is worth the effort:

```{r}
#| label: motivation-example

# Generate some sample data
set.seed(123)  # For reproducible results
sales_data <- data.frame(
  month = month.name[1:12],
  sales = round(rnorm(12, mean = 50000, sd = 10000)),
  expenses = round(rnorm(12, mean = 30000, sd = 5000))
)

# Calculate profit and growth
sales_data$profit <- sales_data$sales - sales_data$expenses
sales_data$profit_margin <- round(sales_data$profit / sales_data$sales * 100, 1)

# Find best performing months
best_month <- sales_data[which.max(sales_data$profit), "month"]
worst_month <- sales_data[which.min(sales_data$profit), "month"]

cat("Best month:", best_month, "\n")
cat("Worst month:", worst_month, "\n")
cat("Average profit margin:", mean(sales_data$profit_margin), "%\n")
```

In just a few lines of R code, we've:
- Created a dataset
- Performed calculations
- Added new variables
- Found extremes
- Calculated summaries

This is the power of R fundamentals in action!

## Ready to Begin?

Let's start with the foundation: [Data Types and Objects](data-types.qmd)

## Summary

R fundamentals form the bedrock of all data analysis in R. While these concepts might seem abstract at first, they quickly become second nature with practice. Remember:

- **Start small**: Master basic concepts before moving to complex ones
- **Practice regularly**: Coding is learned by doing
- **Don't memorize**: Focus on understanding patterns and principles
- **Use resources**: R has excellent built-in help and documentation

The investment you make in learning these fundamentals will pay dividends throughout your R journey. Every advanced technique builds on these basic concepts, so take time to really understand them.

Happy coding! 🎯