---
title: "Vectors: R's Building Blocks"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to Vectors

Vectors are the most fundamental data structure in R. In fact, even single values in R are vectors with one element! Understanding vectors is crucial because virtually everything in R is built upon them.

A **vector** is a sequence of data elements of the same type. Think of it as a container that holds multiple values in a specific order.

## Creating Vectors

### The `c()` Function

The most common way to create vectors is using the `c()` function (which stands for "combine" or "concatenate"):

```{r}
#| label: vector-creation

# Numeric vectors
numbers <- c(1, 2, 3, 4, 5)
temperatures <- c(72.5, 75.2, 68.9, 80.1, 77.3)

# Character vectors
names <- c("Alice", "Bob", "Charlie", "Diana")
colors <- c("red", "green", "blue", "yellow")

# Logical vectors
answers <- c(TRUE, FALSE, TRUE, TRUE, FALSE)

# Print the vectors
print(numbers)
print(names)
print(answers)
```

### Vector Properties

Every vector has important properties:

```{r}
#| label: vector-properties

scores <- c(85, 92, 78, 96, 88)

# Length: number of elements
length(scores)

# Type: what kind of data
typeof(scores)

# Class: object class
class(scores)

# Structure: comprehensive overview
str(scores)
```

### Creating Sequences

R provides several ways to create vectors with patterns:

```{r}
#| label: vector-sequences

# Simple sequences
seq1 <- 1:10          # 1, 2, 3, ..., 10
seq2 <- 10:1          # 10, 9, 8, ..., 1

# Using seq() function
seq3 <- seq(from = 0, to = 100, by = 10)    # 0, 10, 20, ..., 100
seq4 <- seq(0, 1, length.out = 11)         # 11 equally spaced numbers

# Repeated values
rep1 <- rep(5, times = 8)                  # 5, 5, 5, 5, 5, 5, 5, 5
rep2 <- rep(c(1, 2, 3), times = 3)        # 1, 2, 3, 1, 2, 3, 1, 2, 3
rep3 <- rep(c(1, 2, 3), each = 3)         # 1, 1, 1, 2, 2, 2, 3, 3, 3

print(seq3)
print(rep2)
print(rep3)
```

## Vector Indexing and Subsetting

### Accessing Elements by Position

```{r}
#| label: vector-indexing

fruits <- c("apple", "banana", "cherry", "date", "elderberry")

# Single element (note: R uses 1-based indexing!)
fruits[1]     # First element
fruits[3]     # Third element
fruits[5]     # Last element

# Multiple elements
fruits[c(1, 3, 5)]    # Elements 1, 3, and 5
fruits[1:3]           # Elements 1 through 3
fruits[c(2, 4)]       # Elements 2 and 4
```

### Negative Indexing

Use negative indices to exclude elements:

```{r}
#| label: negative-indexing

numbers <- c(10, 20, 30, 40, 50)

# Exclude specific elements
numbers[-1]           # All except the first
numbers[-c(1, 5)]     # All except first and last
numbers[-(2:4)]       # All except elements 2 through 4

print(numbers[-c(1, 5)])
```

### Logical Indexing

Use logical vectors to subset based on conditions:

```{r}
#| label: logical-indexing

ages <- c(23, 35, 28, 42, 19, 31, 27)

# Find elements meeting a condition
ages > 30             # Logical vector
ages[ages > 30]       # Elements where condition is TRUE

# Multiple conditions
ages[ages >= 25 & ages <= 35]     # Ages between 25 and 35
ages[ages < 25 | ages > 40]       # Ages less than 25 OR greater than 40

# Store logical vector for reuse
adults <- ages >= 18
ages[adults]
```

### Subsetting with Names

Vectors can have named elements:

```{r}
#| label: named-vectors

# Create named vector
student_grades <- c(alice = 92, bob = 87, charlie = 95, diana = 89)
print(student_grades)

# Access by name
student_grades["alice"]
student_grades[c("alice", "charlie")]

# Get names
names(student_grades)

# Add names to existing vector
scores <- c(85, 90, 78, 92)
names(scores) <- c("Math", "Science", "English", "History")
print(scores)
```

## Element-wise Operations

One of R's greatest strengths is **vectorization** - operations work on entire vectors automatically:

### Arithmetic Operations

```{r}
#| label: vector-arithmetic

# Create vectors
a <- c(2, 4, 6, 8, 10)
b <- c(1, 2, 3, 4, 5)

# Element-wise arithmetic
a + b     # Add corresponding elements
a - b     # Subtract corresponding elements
a * b     # Multiply corresponding elements
a / b     # Divide corresponding elements
a ^ b     # Raise a to the power of b

# Operations with single values (recycling)
a + 10    # Add 10 to each element
a * 2     # Multiply each element by 2
a / 2     # Divide each element by 2

print(a + b)
print(a * 2)
```

### Comparison Operations

```{r}
#| label: vector-comparisons

scores <- c(85, 92, 78, 96, 88, 74, 91)

# Comparisons return logical vectors
high_scores <- scores > 90
passing_scores <- scores >= 80
failing_scores <- scores < 70

print(high_scores)
print(passing_scores)

# Count TRUE values
sum(high_scores)      # How many scored above 90?
sum(passing_scores)   # How many passed?

# Percentage calculations
mean(high_scores) * 100    # Percentage with high scores
```

### Logical Operations

```{r}
#| label: vector-logical

x <- c(TRUE, FALSE, TRUE, FALSE, TRUE)
y <- c(FALSE, FALSE, TRUE, TRUE, TRUE)

# Element-wise logical operations
x & y     # AND operation
x | y     # OR operation
!x        # NOT operation

print(x & y)
print(x | y)
```

## Vector Functions

R provides many built-in functions that work with vectors:

### Mathematical Functions

```{r}
#| label: vector-math-functions

values <- c(1, 4, 9, 16, 25)

# Basic functions
sum(values)           # Sum of all elements
mean(values)          # Average
median(values)        # Median
min(values)           # Minimum value
max(values)           # Maximum value
range(values)         # Min and max
var(values)           # Variance
sd(values)            # Standard deviation

# Element-wise mathematical functions
sqrt(values)          # Square root of each element
log(values)           # Natural logarithm
round(sqrt(values), 2) # Round to 2 decimal places

print(sqrt(values))
```

### Statistical Functions

```{r}
#| label: vector-stats

# Generate some sample data
set.seed(123)
sample_data <- round(rnorm(20, mean = 75, sd = 10))

# Comprehensive statistics
summary(sample_data)

# Quantiles
quantile(sample_data)
quantile(sample_data, probs = c(0.25, 0.5, 0.75, 0.95))

# Ranking and ordering
sort(sample_data)                    # Sort in ascending order
sort(sample_data, decreasing = TRUE) # Sort in descending order
order(sample_data)                   # Indices that would sort the vector
rank(sample_data)                    # Ranks of each element
```

### Finding and Counting

```{r}
#| label: vector-finding

grades <- c(85, 92, 78, 96, 88, 74, 91, 89)

# Find specific values
which(grades > 90)           # Positions of elements > 90
which.max(grades)            # Position of maximum value
which.min(grades)            # Position of minimum value

# Check for presence
85 %in% grades               # Is 85 in the vector?
c(85, 100) %in% grades       # Which of these are in the vector?

# Count specific values
sum(grades == 85)            # How many times does 85 appear?
sum(grades > 90)             # How many scores above 90?

# Unique values
duplicated_values <- c(1, 2, 2, 3, 3, 3, 4)
unique(duplicated_values)    # Get unique values
duplicated(duplicated_values) # Which are duplicates?
```

## Modifying Vectors

### Adding Elements

```{r}
#| label: vector-modification

# Start with a vector
original <- c(1, 2, 3)

# Add elements to the end
extended <- c(original, 4, 5)
print(extended)

# Add elements to the beginning
prepended <- c(0, original)
print(prepended)

# Insert elements in the middle
# (This requires more complex indexing)
middle_insert <- c(original[1:2], 2.5, original[3])
print(middle_insert)
```

### Replacing Elements

```{r}
#| label: vector-replacement

scores <- c(85, 92, 78, 96, 88)

# Replace specific positions
scores[3] <- 82          # Replace 3rd element
scores[c(1, 5)] <- c(87, 90)  # Replace 1st and 5th elements

print(scores)

# Conditional replacement
ages <- c(23, 35, 28, 42, 19, 31, 27)
ages[ages < 25] <- 25    # Set minimum age to 25
print(ages)
```

### Removing Elements

```{r}
#| label: vector-removal

numbers <- c(10, 20, 30, 40, 50)

# Remove by position
shortened <- numbers[-3]              # Remove 3rd element
multiple_removed <- numbers[-c(1, 5)] # Remove 1st and 5th

print(shortened)
print(multiple_removed)

# Remove by condition
grades <- c(85, 92, 78, 96, 88, 74, 91)
passing_only <- grades[grades >= 80]  # Keep only passing grades
print(passing_only)
```

## Working with Missing Values

### Creating and Detecting Missing Values

```{r}
#| label: missing-values

# Vector with missing values
incomplete_data <- c(1, 2, NA, 4, 5, NA, 7)

# Detect missing values
is.na(incomplete_data)
which(is.na(incomplete_data))     # Positions of NA values
sum(is.na(incomplete_data))       # Count of NA values

# Complete cases (non-missing)
complete.cases(incomplete_data)
incomplete_data[complete.cases(incomplete_data)]
```

### Handling Missing Values in Calculations

```{r}
#| label: na-calculations

data_with_na <- c(10, 15, NA, 20, 25, NA, 30)

# Many functions return NA if any element is NA
mean(data_with_na)           # Returns NA
sum(data_with_na)            # Returns NA

# Use na.rm = TRUE to exclude NA values
mean(data_with_na, na.rm = TRUE)
sum(data_with_na, na.rm = TRUE)
sd(data_with_na, na.rm = TRUE)

# Functions that handle NA by default
length(data_with_na)         # Counts NA values too
length(na.omit(data_with_na)) # Length after removing NA
```

## Vector Recycling

When vectors of different lengths are used together, R "recycles" the shorter vector:

```{r}
#| label: vector-recycling

# Vectors of different lengths
long_vector <- c(1, 2, 3, 4, 5, 6)
short_vector <- c(10, 20)

# The short vector gets recycled
result <- long_vector + short_vector
print(result)  # c(11, 22, 13, 24, 15, 26)

# Recycling with single values
add_five <- long_vector + 5  # 5 is recycled to match length
print(add_five)

# Warning when lengths don't divide evenly
uneven_example <- c(1, 2, 3, 4, 5) + c(10, 20, 30)  # Warning!
```

## Practical Examples

### Example 1: Grade Analysis

```{r}
#| label: grade-analysis

# Student grades for a class
student_names <- c("Alice", "Bob", "Charlie", "Diana", "Eve", "Frank", "Grace")
midterm_scores <- c(85, 78, 92, 88, 79, 94, 87)
final_scores <- c(88, 82, 89, 91, 83, 96, 90)

# Calculate overall grades (60% final, 40% midterm)
overall_grades <- 0.4 * midterm_scores + 0.6 * final_scores

# Assign letter grades
letter_grades <- ifelse(overall_grades >= 90, "A",
                       ifelse(overall_grades >= 80, "B",
                             ifelse(overall_grades >= 70, "C",
                                   ifelse(overall_grades >= 60, "D", "F"))))

# Create a summary
grade_summary <- data.frame(
  Student = student_names,
  Midterm = midterm_scores,
  Final = final_scores,
  Overall = round(overall_grades, 1),
  Grade = letter_grades
)

print(grade_summary)

# Class statistics
cat("Class average:", round(mean(overall_grades), 1), "\n")
cat("Students with A:", sum(letter_grades == "A"), "\n")
cat("Passing rate:", round(mean(overall_grades >= 60) * 100, 1), "%\n")
```

### Example 2: Temperature Conversion

```{r}
#| label: temperature-conversion

# Daily temperatures in Fahrenheit
fahrenheit_temps <- c(68, 72, 75, 71, 69, 74, 78, 76, 73, 70)

# Convert to Celsius
celsius_temps <- (fahrenheit_temps - 32) * 5/9

# Categorize temperatures
temp_categories <- ifelse(celsius_temps < 15, "Cold",
                         ifelse(celsius_temps < 25, "Mild", "Warm"))

# Summary
temp_summary <- data.frame(
  Day = 1:10,
  Fahrenheit = fahrenheit_temps,
  Celsius = round(celsius_temps, 1),
  Category = temp_categories
)

print(temp_summary)

# Find extreme days
hottest_day <- which.max(celsius_temps)
coldest_day <- which.min(celsius_temps)

cat("Hottest day:", hottest_day, "with", round(celsius_temps[hottest_day], 1), "°C\n")
cat("Coldest day:", coldest_day, "with", round(celsius_temps[coldest_day], 1), "°C\n")
```

## Common Mistakes and Best Practices

### 1. Remember 1-based Indexing

```{r}
#| label: indexing-reminder

my_vector <- c("a", "b", "c", "d", "e")

# R uses 1-based indexing (not 0-based like many languages)
my_vector[1]  # First element (not my_vector[0])
my_vector[5]  # Last element
```

### 2. Vector Type Consistency

```{r}
#| label: type-consistency

# Vectors can only hold one type of data
mixed_attempt <- c(1, "two", 3, "four")
print(mixed_attempt)  # Everything becomes character!
typeof(mixed_attempt)

# Use lists for mixed types (covered in next section)
```

### 3. NA Propagation

```{r}
#| label: na-propagation

# One NA can affect entire calculations
values_with_na <- c(1, 2, NA, 4, 5)
mean(values_with_na)           # Returns NA
mean(values_with_na, na.rm = TRUE)  # Proper way to handle
```

## Exercises

### Exercise 1: Vector Creation and Manipulation
1. Create a vector of the first 20 even numbers
2. Create a vector with your name repeated 5 times
3. Create a vector of 15 random numbers between 1 and 100

### Exercise 2: Data Analysis Practice
Given these test scores: `scores <- c(78, 85, 92, 88, 79, 95, 87, 83, 90, 86)`

1. Calculate the mean, median, and standard deviation
2. Find how many scores are above average
3. Identify the positions of scores above 90
4. Replace any score below 80 with 80

### Exercise 3: Real-world Application
You have monthly sales data: `sales <- c(12000, 15000, 13500, 16000, 14200, 17500)`

1. Calculate the total yearly sales
2. Find the month with highest sales
3. Calculate the percentage increase from the first month to the last month
4. Identify months where sales exceeded $15,000

## Summary

Vectors are fundamental to R programming:

- **Creation**: Use `c()`, sequences (`1:10`, `seq()`), and repetition (`rep()`)
- **Indexing**: Access elements by position `[1]`, multiple positions `[c(1,3,5)]`, or conditions `[x > 5]`
- **Operations**: Vectorized arithmetic and comparisons work element-wise
- **Functions**: Many built-in functions work naturally with vectors
- **Modification**: Add, replace, or remove elements as needed

Key principles:
- Vectors hold elements of the same type
- R uses 1-based indexing
- Operations are vectorized by default
- Missing values (NA) propagate through calculations

Understanding vectors is essential because they form the foundation for more complex data structures like data frames and lists, which we'll explore next!