---
title: "Working with Files in R"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to File Operations

Working with files is a fundamental part of data analysis. In this section, you'll learn how to:

- **Read data** from various file formats
- **Write data** to files for sharing and storage
- **Navigate file systems** and organize projects
- **Handle different data formats** (CSV, Excel, text files)
- **Work with file paths** across different operating systems

R provides powerful tools for file operations that integrate seamlessly with data analysis workflows.

## Understanding File Paths

### Absolute vs Relative Paths

```{r}
#| label: file-paths
#| eval: false

# Absolute paths (full path from root)
# Windows: "C:/Users/YourName/Documents/data.csv"
# Mac/Linux: "/home/username/documents/data.csv"

# Relative paths (relative to current working directory)
# "data.csv"           - file in current directory
# "data/sales.csv"     - file in 'data' subdirectory
# "../data.csv"        - file in parent directory
```

### Working Directory Operations

```{r}
#| label: working-directory

# Check current working directory
current_dir <- getwd()
print(current_dir)

# List files in current directory
files_here <- list.files()
print(files_here)

# List files with full paths
files_full <- list.files(full.names = TRUE)
head(files_full)

# List files with specific pattern
csv_files <- list.files(pattern = "\\.csv$")
print(csv_files)

# Check if file exists
file.exists("data.csv")
file.exists("nonexistent.csv")
```

### Creating Directories and File Paths

```{r}
#| label: directory-operations

# Create directory (if it doesn't exist)
if (!dir.exists("output")) {
  dir.create("output")
  cat("Created 'output' directory\n")
}

# Create nested directories
if (!dir.exists("data/raw")) {
  dir.create("data/raw", recursive = TRUE)
  cat("Created 'data/raw' directory structure\n")
}

# Build file paths safely (works across operating systems)
data_path <- file.path("data", "raw", "sales_data.csv")
print(data_path)

# Get file information
if (file.exists(data_path)) {
  file_info <- file.info(data_path)
  print(file_info)
}
```

## Reading Data Files

### Reading CSV Files

CSV (Comma-Separated Values) files are the most common format for data exchange:

```{r}
#| label: reading-csv

# Create sample CSV data for demonstration
sample_data <- data.frame(
  name = c("Alice", "Bob", "Charlie", "Diana"),
  age = c(25, 30, 35, 28),
  salary = c(50000, 60000, 55000, 52000),
  department = c("Sales", "IT", "Marketing", "HR")
)

# Write sample data to demonstrate reading
write.csv(sample_data, "sample_employees.csv", row.names = FALSE)

# Read CSV file
employees <- read.csv("sample_employees.csv")
print(employees)

# Check structure
str(employees)

# Read CSV with custom settings
employees_custom <- read.csv("sample_employees.csv",
                            stringsAsFactors = FALSE,  # Keep strings as characters
                            header = TRUE,             # First row contains column names
                            sep = ",")                 # Comma separator
print(employees_custom)
```

### Handling Different Separators

```{r}
#| label: different-separators

# Create tab-separated file
write.table(sample_data, "sample_employees.tsv",
           sep = "\t", row.names = FALSE, quote = FALSE)

# Read tab-separated file
employees_tsv <- read.table("sample_employees.tsv",
                           header = TRUE, sep = "\t",
                           stringsAsFactors = FALSE)
print(employees_tsv)

# Create semicolon-separated file (common in European locales)
write.table(sample_data, "sample_employees_semi.csv",
           sep = ";", row.names = FALSE, quote = FALSE)

# Read semicolon-separated file
employees_semi <- read.table("sample_employees_semi.csv",
                            header = TRUE, sep = ";",
                            stringsAsFactors = FALSE)
print(employees_semi)
```

### Reading Text Files

```{r}
#| label: reading-text

# Create a simple text file
text_content <- c(
  "This is line 1",
  "This is line 2",
  "This is line 3 with more text",
  "Final line"
)

writeLines(text_content, "sample_text.txt")

# Read entire file as character vector
file_lines <- readLines("sample_text.txt")
print(file_lines)

# Read specific number of lines
first_two_lines <- readLines("sample_text.txt", n = 2)
print(first_two_lines)

# Read file as single string
file_content <- paste(readLines("sample_text.txt"), collapse = "\n")
cat(file_content)
```

### Reading Excel Files

```{r}
#| label: reading-excel
#| eval: false

# Note: This requires the readxl package
# install.packages("readxl")
library(readxl)

# Read Excel file (first sheet by default)
excel_data <- read_excel("data.xlsx")

# Read specific sheet
excel_data_sheet2 <- read_excel("data.xlsx", sheet = "Sheet2")

# Read specific sheet by number
excel_data_sheet1 <- read_excel("data.xlsx", sheet = 1)

# Read with specific range
excel_range <- read_excel("data.xlsx", range = "A1:D10")

# Skip rows and specify column types
excel_custom <- read_excel("data.xlsx",
                          skip = 2,  # Skip first 2 rows
                          col_types = c("text", "numeric", "date"))
```

## Writing Data Files

### Writing CSV Files

```{r}
#| label: writing-csv

# Create analysis results
analysis_results <- data.frame(
  metric = c("Total Sales", "Average Sale", "Max Sale", "Min Sale"),
  value = c(125000, 2500, 15000, 250),
  period = rep("Q1 2024", 4)
)

# Write to CSV
write.csv(analysis_results, "analysis_results.csv", row.names = FALSE)

# Write with custom settings
write.csv(analysis_results, "analysis_results_custom.csv",
         row.names = FALSE,    # Don't include row numbers
         quote = FALSE,        # Don't quote text fields
         na = "")             # How to represent missing values

# Verify the file was written
if (file.exists("analysis_results.csv")) {
  cat("File written successfully!\n")
  cat("File size:", file.info("analysis_results.csv")$size, "bytes\n")
}
```

### Writing Other Formats

```{r}
#| label: writing-other-formats

# Write tab-separated file
write.table(analysis_results, "analysis_results.tsv",
           sep = "\t", row.names = FALSE, quote = FALSE)

# Write space-separated file
write.table(analysis_results, "analysis_results.txt",
           sep = " ", row.names = FALSE, quote = FALSE)

# Write with custom separator
write.table(analysis_results, "analysis_results_pipe.txt",
           sep = "|", row.names = FALSE, quote = FALSE)

# Write R data file (preserves data types perfectly)
save(analysis_results, file = "analysis_results.RData")

# Write individual object to RDS file (more efficient)
saveRDS(analysis_results, "analysis_results.rds")
```

### Writing Text Files

```{r}
#| label: writing-text

# Create a report
report_lines <- c(
  "Data Analysis Report",
  "====================",
  "",
  paste("Generated on:", Sys.Date()),
  paste("Total records analyzed:", nrow(sample_data)),
  "",
  "Summary Statistics:",
  paste("Average age:", round(mean(sample_data$age), 1)),
  paste("Average salary: $", format(mean(sample_data$salary), big.mark = ",")),
  "",
  "End of report"
)

# Write report to text file
writeLines(report_lines, "analysis_report.txt")

# Write using cat() for more control
cat("Custom Report\n",
    "=============\n\n",
    "This report was generated using cat().\n",
    "Current time:", format(Sys.time()), "\n",
    file = "custom_report.txt")

# Append to existing file
cat("\nAdditional information added later.\n",
    file = "custom_report.txt", append = TRUE)
```

## Project Organization

### Creating a Standard Project Structure

```{r}
#| label: project-structure

# Function to create standard project directories
create_project_structure <- function(project_name) {

  # Create main project directory
  if (!dir.exists(project_name)) {
    dir.create(project_name)
  }

  # Create subdirectories
  subdirs <- c("data/raw", "data/processed", "scripts", "output", "docs")

  for (subdir in subdirs) {
    full_path <- file.path(project_name, subdir)
    if (!dir.exists(full_path)) {
      dir.create(full_path, recursive = TRUE)
      cat("Created:", full_path, "\n")
    }
  }

  # Create README file
  readme_content <- c(
    paste("#", project_name),
    "",
    "## Project Structure",
    "- `data/raw/` - Original, unmodified data",
    "- `data/processed/` - Cleaned and processed data",
    "- `scripts/` - R scripts for analysis",
    "- `output/` - Results, plots, reports",
    "- `docs/` - Documentation",
    "",
    paste("Created:", Sys.Date())
  )

  writeLines(readme_content, file.path(project_name, "README.md"))
  cat("Created README.md\n")

  cat("Project structure created successfully!\n")
}

# Create a sample project
create_project_structure("my_analysis_project")

# List the created structure
if (dir.exists("my_analysis_project")) {
  cat("\nProject structure:\n")
  files <- list.files("my_analysis_project", recursive = TRUE, include.dirs = TRUE)
  for (file in files) {
    cat("  ", file, "\n")
  }
}
```

### File Naming Conventions

```{r}
#| label: file-naming

# Good file naming practices
create_analysis_files <- function() {

  # Use ISO date format for chronological ordering
  today <- format(Sys.Date(), "%Y%m%d")

  # Descriptive names with underscores
  good_names <- c(
    paste0(today, "_sales_analysis.R"),
    paste0(today, "_customer_data_cleaning.R"),
    "01_data_import.R",
    "02_data_cleaning.R",
    "03_exploratory_analysis.R",
    "04_modeling.R",
    "05_report_generation.R"
  )

  # Create empty files to demonstrate naming
  for (name in good_names) {
    file_path <- file.path("scripts", name)
    if (!file.exists(file_path)) {
      # Create directory if it doesn't exist
      if (!dir.exists("scripts")) {
        dir.create("scripts")
      }
      # Create empty file
      file.create(file_path)
      cat("Created:", name, "\n")
    }
  }

  cat("\nFile naming best practices:\n")
  cat("✓ Use dates in YYYYMMDD format\n")
  cat("✓ Use descriptive names\n")
  cat("✓ Use underscores instead of spaces\n")
  cat("✓ Use numbers for sequential scripts\n")
  cat("✓ Avoid special characters\n")
}

create_analysis_files()
```

## Advanced File Operations

### Reading Large Files Efficiently

```{r}
#| label: large-files

# Function to read large files in chunks
read_large_csv_chunked <- function(filename, chunk_size = 1000) {

  # First, count total rows
  total_rows <- length(readLines(filename)) - 1  # Subtract header
  cat("Total rows to process:", total_rows, "\n")

  # Read file in chunks
  all_data <- list()
  connection <- file(filename, "r")

  # Read header first
  header <- readLines(connection, n = 1)

  chunk_num <- 1
  while (TRUE) {
    # Read chunk
    lines <- readLines(connection, n = chunk_size)

    if (length(lines) == 0) break  # End of file

    # Convert to data frame
    chunk_text <- c(header, lines)
    chunk_data <- read.csv(text = paste(chunk_text, collapse = "\n"))

    all_data[[chunk_num]] <- chunk_data
    cat("Processed chunk", chunk_num, "with", nrow(chunk_data), "rows\n")

    chunk_num <- chunk_num + 1
  }

  close(connection)

  # Combine all chunks
  final_data <- do.call(rbind, all_data)
  return(final_data)
}

# For demonstration, create a larger file
large_sample <- data.frame(
  id = 1:500,
  name = paste("Person", 1:500),
  value = runif(500, 1, 100),
  category = sample(c("A", "B", "C"), 500, replace = TRUE)
)

write.csv(large_sample, "large_sample.csv", row.names = FALSE)

# Read it back in chunks
# chunked_data <- read_large_csv_chunked("large_sample.csv", chunk_size = 100)
```

### Working with Compressed Files

```{r}
#| label: compressed-files

# Create sample data
compression_demo <- data.frame(
  x = 1:1000,
  y = rnorm(1000),
  text = paste("Text entry", 1:1000)
)

# Write regular CSV
write.csv(compression_demo, "uncompressed.csv", row.names = FALSE)

# Write compressed CSV (gzip)
gz_connection <- gzfile("compressed.csv.gz", "w")
write.csv(compression_demo, gz_connection, row.names = FALSE)
close(gz_connection)

# Read compressed file
compressed_data <- read.csv("compressed.csv.gz")

# Compare file sizes
uncompressed_size <- file.info("uncompressed.csv")$size
compressed_size <- file.info("compressed.csv.gz")$size

cat("Uncompressed size:", uncompressed_size, "bytes\n")
cat("Compressed size:", compressed_size, "bytes\n")
cat("Compression ratio:", round(compressed_size / uncompressed_size * 100, 1), "%\n")
```

### File Manipulation

```{r}
#| label: file-manipulation

# Copy files
if (file.exists("sample_employees.csv")) {
  file.copy("sample_employees.csv", "backup_employees.csv")
  cat("File copied successfully\n")
}

# Rename files
if (file.exists("backup_employees.csv")) {
  file.rename("backup_employees.csv", "employees_backup.csv")
  cat("File renamed successfully\n")
}

# Get file information
file_info <- file.info("sample_employees.csv")
cat("File size:", file_info$size, "bytes\n")
cat("Created:", file_info$ctime, "\n")
cat("Modified:", file_info$mtime, "\n")

# Remove files (be careful!)
temp_files <- c("temp1.txt", "temp2.txt")
for (temp_file in temp_files) {
  cat("Creating temporary file:", temp_file, "\n")
  writeLines("Temporary content", temp_file)
}

# List temporary files
existing_temp <- temp_files[file.exists(temp_files)]
cat("Temporary files found:", paste(existing_temp, collapse = ", "), "\n")

# Remove temporary files
for (temp_file in existing_temp) {
  file.remove(temp_file)
  cat("Removed:", temp_file, "\n")
}
```

## Data Import/Export Best Practices

### Robust Data Reading Function

```{r}
#| label: robust-reading

# Robust function for reading CSV files
read_csv_safe <- function(filename, ...) {

  # Check if file exists
  if (!file.exists(filename)) {
    stop("File does not exist: ", filename)
  }

  # Get file information
  file_info <- file.info(filename)
  cat("Reading file:", filename, "\n")
  cat("File size:", file_info$size, "bytes\n")

  # Try to read the file
  tryCatch({
    data <- read.csv(filename, stringsAsFactors = FALSE, ...)

    cat("Successfully read", nrow(data), "rows and", ncol(data), "columns\n")

    # Basic data quality checks
    if (nrow(data) == 0) {
      warning("File contains no data rows")
    }

    # Check for missing values
    missing_counts <- sapply(data, function(x) sum(is.na(x)))
    if (any(missing_counts > 0)) {
      cat("Missing values found:\n")
      for (col in names(missing_counts)[missing_counts > 0]) {
        cat("  ", col, ":", missing_counts[col], "\n")
      }
    }

    return(data)

  }, error = function(e) {
    stop("Error reading file: ", e$message)
  })
}

# Test the robust reading function
safe_data <- read_csv_safe("sample_employees.csv")
str(safe_data)
```

### Data Export with Metadata

```{r}
#| label: export-with-metadata

# Function to export data with metadata
export_with_metadata <- function(data, filename, description = "") {

  # Create metadata
  metadata <- list(
    filename = filename,
    description = description,
    export_date = Sys.time(),
    rows = nrow(data),
    columns = ncol(data),
    column_names = names(data),
    column_types = sapply(data, class),
    r_version = R.version.string
  )

  # Write data file
  write.csv(data, filename, row.names = FALSE)

  # Write metadata file
  metadata_filename <- sub("\\.csv$", "_metadata.txt", filename)

  metadata_text <- c(
    "Data Export Metadata",
    "===================",
    "",
    paste("File:", metadata$filename),
    paste("Description:", metadata$description),
    paste("Export Date:", metadata$export_date),
    paste("R Version:", metadata$r_version),
    "",
    "Data Summary:",
    paste("  Rows:", metadata$rows),
    paste("  Columns:", metadata$columns),
    "",
    "Column Information:",
    paste("  Names:", paste(metadata$column_names, collapse = ", ")),
    "",
    "Column Types:"
  )

  for (i in seq_along(metadata$column_types)) {
    metadata_text <- c(metadata_text,
                      paste("  ", metadata$column_names[i], ":", metadata$column_types[i]))
  }

  writeLines(metadata_text, metadata_filename)

  cat("Data exported to:", filename, "\n")
  cat("Metadata exported to:", metadata_filename, "\n")

  return(invisible(metadata))
}

# Test the export function
export_with_metadata(sample_data, "employees_with_metadata.csv",
                    "Sample employee data for training purposes")
```

## Practical Examples

### Example 1: Data Pipeline

```{r}
#| label: data-pipeline

# Complete data processing pipeline
process_sales_data <- function(input_file, output_dir = "output") {

  cat("=== Sales Data Processing Pipeline ===\n")

  # Create output directory
  if (!dir.exists(output_dir)) {
    dir.create(output_dir)
  }

  # Step 1: Read raw data
  cat("Step 1: Reading raw data...\n")
  tryCatch({
    raw_data <- read.csv(input_file, stringsAsFactors = FALSE)
    cat("  ✓ Read", nrow(raw_data), "records\n")
  }, error = function(e) {
    stop("Failed to read input file: ", e$message)
  })

  # Step 2: Data cleaning
  cat("Step 2: Cleaning data...\n")

  # Remove duplicates
  initial_rows <- nrow(raw_data)
  clean_data <- unique(raw_data)
  duplicates_removed <- initial_rows - nrow(clean_data)
  if (duplicates_removed > 0) {
    cat("  ✓ Removed", duplicates_removed, "duplicate records\n")
  }

  # Handle missing values (example: remove rows with missing sales)
  if ("sales" %in% names(clean_data)) {
    clean_data <- clean_data[!is.na(clean_data$sales), ]
    cat("  ✓ Removed records with missing sales values\n")
  }

  # Step 3: Add calculated fields
  cat("Step 3: Adding calculated fields...\n")
  if ("sales" %in% names(clean_data)) {
    clean_data$sales_category <- ifelse(clean_data$sales > mean(clean_data$sales),
                                       "High", "Low")
    cat("  ✓ Added sales category field\n")
  }

  # Step 4: Generate summary
  cat("Step 4: Generating summary...\n")
  summary_stats <- list(
    total_records = nrow(clean_data),
    date_processed = Sys.time()
  )

  if ("sales" %in% names(clean_data)) {
    summary_stats$total_sales <- sum(clean_data$sales)
    summary_stats$avg_sales <- mean(clean_data$sales)
    summary_stats$max_sales <- max(clean_data$sales)
    summary_stats$min_sales <- min(clean_data$sales)
  }

  # Step 5: Export results
  cat("Step 5: Exporting results...\n")

  # Export cleaned data
  output_file <- file.path(output_dir, "cleaned_sales_data.csv")
  write.csv(clean_data, output_file, row.names = FALSE)
  cat("  ✓ Exported cleaned data to", output_file, "\n")

  # Export summary
  summary_file <- file.path(output_dir, "processing_summary.txt")
  summary_text <- c(
    "Sales Data Processing Summary",
    "============================",
    "",
    paste("Processing Date:", summary_stats$date_processed),
    paste("Total Records:", summary_stats$total_records),
    ""
  )

  if ("sales" %in% names(clean_data)) {
    summary_text <- c(summary_text,
                     paste("Total Sales: $", format(summary_stats$total_sales, big.mark = ",")),
                     paste("Average Sales: $", round(summary_stats$avg_sales, 2)),
                     paste("Max Sales: $", format(summary_stats$max_sales, big.mark = ",")),
                     paste("Min Sales: $", format(summary_stats$min_sales, big.mark = ",")))
  }

  writeLines(summary_text, summary_file)
  cat("  ✓ Exported summary to", summary_file, "\n")

  cat("=== Pipeline completed successfully! ===\n")
  return(clean_data)
}

# Create sample sales data for testing
sample_sales <- data.frame(
  date = seq(as.Date("2024-01-01"), by = "day", length.out = 100),
  sales = round(runif(100, 1000, 5000), 2),
  region = sample(c("North", "South", "East", "West"), 100, replace = TRUE)
)

# Add some duplicates and missing values for demonstration
sample_sales <- rbind(sample_sales, sample_sales[1:5, ])  # Add duplicates
sample_sales$sales[c(10, 25, 50)] <- NA  # Add missing values

write.csv(sample_sales, "sample_sales_raw.csv", row.names = FALSE)

# Run the pipeline
processed_data <- process_sales_data("sample_sales_raw.csv")
```

### Example 2: Automated Reporting

```{r}
#| label: automated-reporting

# Function to generate automated reports
generate_data_report <- function(data, output_file = "data_report.txt") {

  # Start report
  report_lines <- c(
    "AUTOMATED DATA REPORT",
    paste(rep("=", 40), collapse = ""),
    "",
    paste("Generated:", format(Sys.time(), "%Y-%m-%d %H:%M:%S")),
    paste("R Version:", R.version.string),
    ""
  )

  # Data overview
  report_lines <- c(report_lines,
    "DATA OVERVIEW",
    paste(rep("-", 20), collapse = ""),
    paste("Rows:", nrow(data)),
    paste("Columns:", ncol(data)),
    paste("Column Names:", paste(names(data), collapse = ", ")),
    ""
  )

  # Column analysis
  report_lines <- c(report_lines, "COLUMN ANALYSIS", paste(rep("-", 20), collapse = ""))

  for (col_name in names(data)) {
    col_data <- data[[col_name]]

    report_lines <- c(report_lines, paste("Column:", col_name))
    report_lines <- c(report_lines, paste("  Type:", class(col_data)[1]))
    report_lines <- c(report_lines, paste("  Missing:", sum(is.na(col_data))))

    if (is.numeric(col_data)) {
      report_lines <- c(report_lines,
        paste("  Min:", round(min(col_data, na.rm = TRUE), 3)),
        paste("  Max:", round(max(col_data, na.rm = TRUE), 3)),
        paste("  Mean:", round(mean(col_data, na.rm = TRUE), 3)),
        paste("  Std Dev:", round(sd(col_data, na.rm = TRUE), 3))
      )
    } else if (is.character(col_data) || is.factor(col_data)) {
      unique_vals <- length(unique(col_data[!is.na(col_data)]))
      report_lines <- c(report_lines, paste("  Unique Values:", unique_vals))

      if (unique_vals <= 10) {
        top_values <- names(sort(table(col_data), decreasing = TRUE))[1:min(5, unique_vals)]
        report_lines <- c(report_lines, paste("  Top Values:", paste(top_values, collapse = ", ")))
      }
    }

    report_lines <- c(report_lines, "")
  }

  # Write report
  writeLines(report_lines, output_file)
  cat("Report generated:", output_file, "\n")

  # Return report content for display
  return(report_lines)
}

# Generate report for our sample data
report_content <- generate_data_report(sample_data, "employee_data_report.txt")

# Display first part of report
cat(paste(report_content[1:20], collapse = "\n"))
```

## Common Pitfalls and Solutions

### 1. Encoding Issues

```{r}
#| label: encoding-issues

# Handle different text encodings
read_with_encoding <- function(filename, encoding = "UTF-8") {

  # Try different encodings if the first one fails
  encodings_to_try <- c(encoding, "UTF-8", "latin1", "CP1252")

  for (enc in encodings_to_try) {
    tryCatch({
      cat("Trying encoding:", enc, "\n")
      data <- read.csv(filename, fileEncoding = enc, stringsAsFactors = FALSE)
      cat("Success with encoding:", enc, "\n")
      return(data)
    }, error = function(e) {
      cat("Failed with", enc, ":", e$message, "\n")
    })
  }

  stop("Could not read file with any of the attempted encodings")
}
```

### 2. Large File Handling

```{r}
#| label: large-file-tips

# Tips for handling large files
large_file_tips <- function() {
  cat("Tips for working with large files:\n")
  cat("1. Use read.csv with nrows parameter to test structure:\n")
  cat("   sample <- read.csv('large_file.csv', nrows = 100)\n\n")

  cat("2. Specify column types to save memory:\n")
  cat("   col_classes <- c('character', 'numeric', 'Date')\n")
  cat("   data <- read.csv('file.csv', colClasses = col_classes)\n\n")

  cat("3. Consider using data.table or readr for better performance:\n")
  cat("   library(data.table)\n")
  cat("   data <- fread('large_file.csv')\n\n")

  cat("4. Process in chunks if memory is limited\n")
  cat("5. Use compression for storage (gzip, bzip2)\n")
  cat("6. Consider binary formats like RDS for R-specific data\n")
}

large_file_tips()
```

## Exercises

### Exercise 1: File Organization
Create a function that:
1. Takes a directory path as input
2. Creates a standard data science project structure
3. Adds a README file with project description
4. Creates sample R scripts with appropriate names

### Exercise 2: Data Import Pipeline
Write a robust data import function that:
1. Checks if the file exists
2. Determines the file type automatically
3. Reads the data with appropriate functions
4. Performs basic data quality checks
5. Returns a list with data and quality report

### Exercise 3: Batch Processing
Create a function that:
1. Processes multiple CSV files in a directory
2. Combines them into a single dataset
3. Handles files with different structures gracefully
4. Exports the combined data with a processing log

### Exercise 4: Report Generation
Build an automated reporting system that:
1. Takes a dataset as input
2. Generates summary statistics
3. Creates visualizations (if plotting libraries available)
4. Exports everything to a formatted text report
5. Includes metadata about the analysis

## Summary

Working with files effectively is crucial for data analysis workflows:

### Key Concepts:
- **File paths**: Understanding absolute vs relative paths and cross-platform compatibility
- **Reading data**: CSV, text, Excel, and other formats with appropriate functions
- **Writing data**: Exporting results in various formats with proper encoding
- **Project organization**: Creating consistent, logical file structures
- **Error handling**: Robust file operations with proper validation

### Best Practices:
- **Use consistent naming conventions** for files and directories
- **Validate inputs** before processing files
- **Handle errors gracefully** with informative messages
- **Document data sources** and processing steps
- **Use compression** for large files to save space
- **Create backups** of important data

### Common Functions:
- **Reading**: `read.csv()`, `readLines()`, `read.table()`
- **Writing**: `write.csv()`, `writeLines()`, `save()`, `saveRDS()`
- **File operations**: `file.exists()`, `list.files()`, `file.copy()`, `file.remove()`
- **Directory operations**: `getwd()`, `setwd()`, `dir.create()`, `file.path()`

### Advanced Features:
- **Chunk processing** for large files
- **Compression support** with gzip connections
- **Encoding handling** for international text
- **Metadata preservation** for reproducible analyses

Proper file handling skills will serve you well throughout your data analysis career. They ensure your work is reproducible, shareable, and maintainable!

This completes our exploration of R fundamentals. You now have the foundation to tackle real-world data analysis projects!