---
title: "R as a Calculator"
---

## Introduction

One of the best ways to start learning R is to use it as a sophisticated calculator. R excels at mathematical operations and provides many built-in functions that go far beyond what a standard calculator can do. In this lesson, we'll explore R's mathematical capabilities and start building your confidence with the R console.

## Learning Objectives

By the end of this lesson, you should be able to:

- translate a mathematical expression into valid R syntax;
- store a result with `<-` and inspect the stored object;
- apply one calculation to every element of a vector; and
- distinguish an ordinary value from `NA`, `NaN`, and `Inf`.

These outcomes matter beyond arithmetic: the same syntax drives financial
models, transformations of data-frame columns, and statistical calculations.

## Getting Started

Open RStudio and look for the **Console** pane (usually in the bottom-left). You'll see a prompt that looks like this:

```bash
R version 4.5.1 (2025-06-13) -- "Great Square Root"
Copyright (C) 2025 The R Foundation for Statistical Computing
Platform: x86_64-pc-linux-gnu

R is free software and comes with ABSOLUTELY NO WARRANTY.
You are welcome to redistribute it under certain conditions.
Type 'license()' or 'licence()' for distribution details.

R is a collaborative project with many contributors.
Type 'contributors()' for more information and
'citation()' on how to cite R or R packages in publications.

Type 'demo()' for some demos, 'help()' for on-line help, or
'help.start()' for an HTML browser interface to help.
Type 'q()' to quit R.

> 
```

This prompt is where you can type R commands and see immediate results.

## Basic Arithmetic Operations

Let's start with fundamental arithmetic operations:

### Addition and Subtraction

```{r}
# Addition
5 + 3

# Subtraction
10 - 4

# You can chain operations
15 + 7 - 3
```

### Multiplication and Division

```{r}
# Multiplication
6 * 7

# Division
20 / 4

# Combined operations
3 * 4 + 2
```

### Exponentiation

```{r}
# Using ^ operator
2^3

# Using ** operator (alternative)
2**3

# Square root (special case)
sqrt(16)

# nth root using exponentiation
8^(1/3)  # cube root of 8
```

### Modular Arithmetic

```{r}
# Modulo (remainder after division)
17 %% 5

# Integer division (quotient without remainder)
17 %/% 5
```

## Order of Operations

R follows standard mathematical order of operations (PEMDAS):

```{r}
# Without parentheses
2 + 3 * 4

# With parentheses to change order
(2 + 3) * 4

# Complex expression
2^3 + 4 * (5 - 2) / 6
```

## Mathematical Functions

R includes many built-in mathematical functions:

### Basic Functions

```{r}
# Absolute value
abs(-5)

# Rounding functions
round(3.14159, 2)    # Round to 2 decimal places
ceiling(3.2)         # Round up
floor(3.9)           # Round down

# Sign function
sign(-42)            # Returns -1, 0, or 1
```

### Logarithmic and Exponential Functions

```{r}
# Natural logarithm
log(10)

# Logarithm base 10
log10(100)

# Logarithm with custom base
log(8, base = 2)

# Exponential function
exp(1)               # e^1

# Power of 10
10^2
```

### Trigonometric Functions

```{r}
# Basic trigonometric functions (in radians)
sin(pi/2)
cos(0)
tan(pi/4)

# Inverse functions
asin(1)
acos(1)
atan(1)

# Convert degrees to radians
degrees_to_radians <- function(degrees) degrees * pi / 180
sin(degrees_to_radians(90))
```

## Special Numbers and Constants

R has several built-in constants:

```{r}
# Pi
pi

# Euler's number
exp(1)

# Infinity
Inf
1/0                  # Also gives Inf

# Not a Number
NaN
0/0                  # Results in NaN

# Not Available (missing value)
NA
```

## Working with Vectors

One of R's superpowers is vectorized operations:

### Creating Vectors

```{r}
# Using c() function to combine values
numbers <- c(1, 2, 3, 4, 5)
numbers

# Creating sequences
1:10                 # Numbers 1 through 10
seq(0, 20, by = 2)   # Even numbers from 0 to 20
```

### Vector Arithmetic

```{r}
# Operations on entire vectors
x <- c(1, 2, 3, 4)
y <- c(5, 6, 7, 8)

# Element-wise operations
x + y
x * y
x^2

# Operations with single numbers (recycling)
x + 10
x * 2
```

R evaluates these expressions element by element. In `x + y`, the first value
of `x` is paired with the first value of `y`, and so on. A length-one value such
as `10` is reused for every element. This special case of **recycling** is
useful; recycling two longer vectors of incompatible lengths is usually a bug
and produces a warning.

### Useful Vector Functions

```{r}
# Create some data
scores <- c(85, 92, 78, 96, 88, 75, 89)

# Summary statistics
sum(scores)          # Total
mean(scores)         # Average
median(scores)       # Middle value
max(scores)          # Maximum
min(scores)          # Minimum
range(scores)        # Min and max
length(scores)       # Number of elements

# More advanced statistics
var(scores)          # Variance
sd(scores)           # Standard deviation
```

## Logical Operations

R can also perform logical operations:

### Comparison Operators

```{r}
# Equal to
5 == 5

# Not equal to
5 != 3

# Greater than / Less than
10 > 5
3 < 7

# Greater than or equal to / Less than or equal to
5 >= 5
4 <= 3
```

### Logical Operators

```{r}
# AND operator
TRUE & TRUE
(5 > 3) & (2 < 4)

# OR operator
TRUE | FALSE
(5 > 10) | (2 < 4)

# NOT operator
!TRUE
!(5 > 10)
```

## Practical Examples

Let's work through some practical calculation examples:

### Example 1: Compound Interest

```{r}
# Calculate compound interest
# Formula: A = P(1 + r/n)^(nt)
principal <- 1000    # Initial amount
rate <- 0.05         # 5% annual interest rate
n <- 12              # Compounded monthly
t <- 5               # 5 years

final_amount <- principal * (1 + rate/n)^(n*t)
final_amount

# Interest earned
interest_earned <- final_amount - principal
interest_earned
```

### Example 2: Body Mass Index (BMI)

```{r}
# BMI calculation: weight (kg) / height (m)^2
weight_kg <- 70
height_m <- 1.75

bmi <- weight_kg / height_m^2
round(bmi, 1)

# BMI for multiple people
weights <- c(65, 70, 80, 55)
heights <- c(1.70, 1.75, 1.80, 1.60)
bmis <- weights / heights^2
round(bmis, 1)
```

### Example 3: Temperature Conversion

```{r}
# Celsius to Fahrenheit: F = (C * 9/5) + 32
celsius_temps <- c(0, 20, 30, 37, 100)
fahrenheit_temps <- (celsius_temps * 9/5) + 32
fahrenheit_temps

# Fahrenheit to Celsius: C = (F - 32) * 5/9
fahrenheit_example <- c(32, 68, 86, 98.6, 212)
celsius_converted <- (fahrenheit_example - 32) * 5/9
celsius_converted
```

## Comments in R

Notice the use of `#` for comments in the examples above. Comments are crucial for explaining your code:

```{r}
# This is a comment - R ignores this line
2 + 2    # You can also add comments at the end of lines

# Comments can span multiple lines
# Use them to explain what your calculations do
# Future you (and others) will thank you!
```

## Tips for Using R as a Calculator

### Use the Up Arrow

- Press the **up arrow** key in the console to recall previous commands
- This saves time when you want to modify a calculation slightly

### Scientific Notation

```{r}
# R uses scientific notation for very large or small numbers
1000000
0.0001

# You can control the display
options(scipen = 999)  # Avoid scientific notation
1000000

options(scipen = 0)    # Reset to default
```

### Assignment vs. Printing

```{r}
# This just calculates and displays the result
5 + 3

# This calculates and stores the result in a variable
result <- 5 + 3

# To see the stored value, type the variable name
result

# Or use parentheses to assign and display
(result2 <- 10 * 2)
```

Assignment and printing are separate actions. `<-` binds a name to a value but
normally prints nothing; entering the name asks R to display the value. Choose
meaningful names such as `monthly_revenue`, not names such as `x1`, once an
object represents a business concept.

## Common Mistakes to Avoid

### Decimal Comparisons

Computers store most decimal fractions approximately. Consequently, this
mathematically natural comparison is not a reliable test:

```{r}
0.1 + 0.2 == 0.3
isTRUE(all.equal(0.1 + 0.2, 0.3))
```

Use `all.equal()` (wrapped in `isTRUE()` when a single logical result is
required) for computed decimal values.

### Case Sensitivity

```{r}
#| eval: false
# R is case-sensitive
PI        # This will give an error
pi        # This works
```

### Missing Parentheses

```{r}
#| eval: false
# Incorrect
mean c(1, 2, 3)    # Missing parentheses - ERROR

# Correct
mean(c(1, 2, 3))
```

### Division by Zero

```{r}
# Be careful with division
5 / 0    # Returns Inf
0 / 0    # Returns NaN
```

## Practice Exercises

Try these calculations on your own:

1. Calculate the area of a circle with radius 7 (use π)
2. Find the hypotenuse of a right triangle with sides 3 and 4
3. Calculate the average of these test scores: 88, 92, 76, 95, 89, 82
4. Convert 25°C to Fahrenheit
5. Calculate 15% tip on a $47.50 restaurant bill

::: {.callout-note}
## Solutions

1. `pi * 7^2` = 153.94
2. `sqrt(3^2 + 4^2)` = 5
3. `mean(c(88, 92, 76, 95, 89, 82))` = 87
4. `(25 * 9/5) + 32` = 77°F
5. `47.50 * 0.15` = $7.125
:::

## Summary

In this lesson, you've learned to use R as a powerful calculator that can:

- Perform basic and advanced mathematical operations
- Work with vectors for batch calculations
- Use built-in mathematical functions
- Handle logical operations
- Apply mathematical concepts to real-world problems

## Next Steps

Now that you're comfortable with R's calculation capabilities, let's explore:

- [RStudio Interface](rstudio-interface.qmd) - Understanding your development environment
- [R Scripts and Projects](r-scripts-projects.qmd) - Organizing your R work

::: {.callout-tip}
## Pro Tip

The more you practice using R for calculations, the more natural it becomes. Try replacing your regular calculator with R for daily calculations – you'll be surprised how quickly you improve!
:::
