---
title: "R Scripts and Projects"
---

## Introduction

As you progress from simple calculations to complex analyses, organizing your R work becomes crucial. R scripts and RStudio Projects are the foundation of reproducible, organized data analysis. In this lesson, you'll learn how to structure your R code professionally and manage your analytical projects effectively.

Think of R scripts as recipes and Projects as well-organized kitchens – both are essential for creating consistent, reproducible results.

## Understanding R Scripts

### What is an R Script?

An R script is a plain text file with a `.R` extension that contains R commands. Unlike typing commands directly in the console, scripts allow you to:

- **Save your work** for future use
- **Document your analysis** with comments
- **Share your methods** with others
- **Reproduce results** exactly
- **Debug and modify** code systematically

### Creating Your First Script

1. **File > New File > R Script** (or `Ctrl+Shift+N`)
2. Save immediately: **File > Save** (or `Ctrl+S`)
3. Give it a meaningful name: `my_first_analysis.R`

### Script Structure Best Practices

Here's a well-structured R script template:

```{r}
#| eval: false
# ============================================================================
# Project: Introduction to R Analysis
# Script: my_first_analysis.R
# Author: Your Name
# Date: 2024-09-21
# Last Modified: 2024-09-21
#
# Purpose: Demonstrate best practices for R script organization
#
# Data: Built-in R datasets
# Output: Summary statistics and basic plots
# ============================================================================

# SETUP ======================================================================

# Clear workspace (optional, be careful!)
# rm(list = ls())

# Load required libraries
library(tidyverse)  # Data manipulation and visualization
library(here)       # File path management

# Set working directory (if not using projects)
# setwd("~/Documents/R-Analysis")

# Source additional scripts if needed
# source("helper_functions.R")

# CONSTANTS AND CONFIGURATION ================================================

# Define constants
SIGNIFICANCE_LEVEL <- 0.05
DEFAULT_PLOT_WIDTH <- 8
DEFAULT_PLOT_HEIGHT <- 6

# Set plot theme
theme_set(theme_minimal())

# DATA IMPORT ================================================================

# Load built-in dataset for this example
data("mtcars")

# In real projects, you might load data like this:
# my_data <- read_csv(here("data", "raw", "dataset.csv"))

# DATA EXPLORATION ===========================================================

# Quick overview
glimpse(mtcars)
summary(mtcars)

# Check for missing values
sum(is.na(mtcars))

# DATA ANALYSIS ==============================================================

# Calculate summary statistics
mpg_stats <- mtcars %>%
  summarise(
    mean_mpg = mean(mpg),
    median_mpg = median(mpg),
    sd_mpg = sd(mpg),
    min_mpg = min(mpg),
    max_mpg = max(mpg)
  )

print(mpg_stats)

# Create visualizations
mpg_histogram <- ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 10, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Distribution of Miles Per Gallon",
    x = "Miles Per Gallon",
    y = "Frequency"
  )

print(mpg_histogram)

# RESULTS EXPORT =============================================================

# Save plot
ggsave(
  filename = here("output", "mpg_distribution.png"),
  plot = mpg_histogram,
  width = DEFAULT_PLOT_WIDTH,
  height = DEFAULT_PLOT_HEIGHT,
  dpi = 300
)

# Save results
write_csv(mpg_stats, here("output", "mpg_summary_stats.csv"))

# SESSION INFO ===============================================================

# Document session for reproducibility
sessionInfo()

# End of script
```

### Key Script Elements

#### 1. Header Comments
Always start with a comprehensive header:
- Project name and purpose
- Author and date information
- Brief description of what the script does
- Data sources and outputs

#### 2. Setup Section
- Load required libraries
- Set global options
- Define constants
- Source helper functions

#### 3. Organized Sections
Use clear section headers:
```{r}
#| eval: false
# SECTION NAME ================================================================
# or
# Section Name ----
```

#### 4. Meaningful Comments
```{r}
#| eval: false
# Good comments explain WHY, not just WHAT
mpg_threshold <- 20  # Fuel efficiency benchmark for classification

# Bad comment (obvious)
mpg_threshold <- 20  # Set mpg_threshold to 20
```

## Working with Scripts

### Running Script Code

**Option 1: Line by Line**
- Place cursor on line
- Press `Ctrl+Enter` (Windows/Linux) or `Cmd+Enter` (Mac)

**Option 2: Selected Code**
- Highlight code block
- Press `Ctrl+Enter`

**Option 3: Entire Script**
- Press `Ctrl+Shift+Enter` or click "Source"

**Option 4: From Console**
```{r}
#| eval: false
source("my_script.R")
```

### Script Navigation

**Code Sections**
Create collapsible sections with `----` or `====`:
```{r}
#| eval: false
# Data Import ----
# (code here)

# Data Cleaning ====
# (code here)
```

**Go to Function/Variable**
- `Ctrl+.` opens "Go to File/Function" dialog
- Type name to jump quickly

**Outline View**
- Click the outline button in Source pane
- See all functions and sections

## Introduction to RStudio Projects

### Why Use Projects?

RStudio Projects solve common organizational problems:

**Without Projects:**
```{r}
#| eval: false
# Absolute paths (bad!)
setwd("/Users/john/Documents/my_analysis")
data <- read.csv("/Users/john/Documents/my_analysis/data/file.csv")

# Problems:
# - Not portable between computers
# - Difficult to share
# - Hard to organize multiple analyses
```

**With Projects:**
```{r}
#| eval: false
# Relative paths (good!)
data <- read.csv("data/file.csv")
# or even better:
data <- read_csv(here("data", "file.csv"))
```

### Project Benefits

1. **Automatic working directory** - no more `setwd()`
2. **Organized file structure** - everything in one place
3. **Portable** - works on any computer
4. **Version control ready** - easy Git integration
5. **Workspace isolation** - separate environments for different projects

### Creating a New Project

**Method 1: From RStudio**
1. **File > New Project**
2. Choose project type:
   - **New Directory**: Start fresh
   - **Existing Directory**: Use existing folder
   - **Version Control**: Clone from Git

**Method 2: New Directory Walkthrough**
1. Select "New Directory"
2. Choose "New Project"
3. Directory name: `my-r-analysis`
4. Choose parent directory
5. Optional: Initialize Git repository
6. Click "Create Project"

### Recommended Project Structure

```
my-r-analysis/
├── my-r-analysis.Rproj    # Project file
├── README.md              # Project description
├── data/                  # Data folder
│   ├── raw/              # Original, unmodified data
│   └── processed/        # Cleaned, processed data
├── scripts/              # R scripts
│   ├── 01_data_import.R
│   ├── 02_data_cleaning.R
│   └── 03_analysis.R
├── output/               # Results folder
│   ├── figures/          # Plots and visualizations
│   └── tables/           # Summary tables
├── docs/                 # Documentation
└── renv/                 # Package management (optional)
```

### Setting Up Project Structure

Create this structure using R:

```{r}
#| eval: false
# Create project folders
dir.create("data")
dir.create("data/raw")
dir.create("data/processed")
dir.create("scripts")
dir.create("output")
dir.create("output/figures")
dir.create("output/tables")
dir.create("docs")

# Create README file
writeLines(
  c("# My R Analysis Project",
    "",
    "## Description",
    "Brief description of your project",
    "",
    "## Structure",
    "- `data/`: Data files",
    "- `scripts/`: R scripts",
    "- `output/`: Results and figures",
    "- `docs/`: Documentation"),
  "README.md"
)
```

## File Management Best Practices

### Naming Conventions

**Files and Folders:**
- Use lowercase with hyphens: `data-cleaning.R`
- Or use underscores: `data_cleaning.R`
- Be descriptive: `monthly_sales_analysis.R`
- Use prefixes for ordering: `01_import.R`, `02_clean.R`

**Variables and Functions:**
```{r}
#| eval: false
# Good naming
customer_ages <- c(25, 30, 45, 22)
calculate_total_revenue <- function(price, quantity) { ... }

# Avoid
x <- c(25, 30, 45, 22)
fun1 <- function(a, b) { ... }
```

### Working with Paths

**Use the `here` Package:**
```{r}
#| eval: false
# Install if needed
install.packages("here")
library(here)

# Benefits of here()
data_path <- here("data", "raw", "sales.csv")
# Works on Windows: data/raw/sales.csv
# Works on Mac/Linux: data/raw/sales.csv

# Read data
sales_data <- read_csv(here("data", "raw", "sales.csv"))

# Save output
write_csv(results, here("output", "tables", "summary.csv"))
```

### Data Import Best Practices

```{r}
#| eval: false
# Good data import workflow
import_sales_data <- function() {
  # Define file path
  file_path <- here("data", "raw", "sales_2024.csv")

  # Check if file exists
  if (!file.exists(file_path)) {
    stop("Sales data file not found: ", file_path)
  }

  # Import with explicit column types
  sales_data <- read_csv(
    file_path,
    col_types = cols(
      date = col_date(format = "%Y-%m-%d"),
      customer_id = col_character(),
      amount = col_double(),
      product = col_character()
    )
  )

  # Basic validation
  if (nrow(sales_data) == 0) {
    warning("Sales data is empty")
  }

  return(sales_data)
}

# Use the function
sales_data <- import_sales_data()
```

## Reproducible Analysis Workflow

### Script Dependencies

**Master Script Approach:**
```{r}
#| eval: false
# main_analysis.R
# ============================================================================
# Master Script: Sales Analysis Pipeline
# ============================================================================

# Setup
source(here("scripts", "00_setup.R"))

# Data pipeline
source(here("scripts", "01_data_import.R"))
source(here("scripts", "02_data_cleaning.R"))
source(here("scripts", "03_data_analysis.R"))
source(here("scripts", "04_generate_report.R"))

# Session info
sessionInfo()
```

### Package Management

**Using `renv` for Reproducibility:**
```{r}
#| eval: false
# Install renv
install.packages("renv")

# Initialize project environment
renv::init()

# Install packages
install.packages(c("tidyverse", "here", "lubridate"))

# Take snapshot of current packages
renv::snapshot()

# Restore environment (on another computer)
renv::restore()
```

## Common Project Pitfalls

### Pitfall 1: Hardcoded Paths
```{r}
#| eval: false
# Bad
setwd("C:/Users/John/Documents/analysis")
data <- read.csv("C:/Users/John/Documents/analysis/data.csv")

# Good
data <- read_csv(here("data", "data.csv"))
```

### Pitfall 2: Not Using Version Control
- Initialize Git repository when creating project
- Use meaningful commit messages
- Push to GitHub/GitLab for backup

### Pitfall 3: Mixing Data and Scripts
```{r}
#| eval: false
# Bad structure
project/
├── analysis.R
├── data.csv
├── plot.R
├── results.csv

# Good structure
project/
├── scripts/
│   ├── analysis.R
│   └── plot.R
├── data/
│   └── data.csv
└── output/
    └── results.csv
```

### Pitfall 4: No Documentation
Always include:
- README.md with project description
- Comments in scripts explaining complex logic
- Data dictionaries for datasets

## Practical Exercise

Let's create a complete project:

1. **Create New Project**
   - Name: "iris-analysis"
   - Choose appropriate location

2. **Set Up Structure**
   ```{r}
   #| eval: false
   # Run this in your new project
   dir.create("data")
   dir.create("scripts")
   dir.create("output")
   ```

3. **Create Analysis Script**
   ```{r}
   #| eval: false
   # scripts/iris_analysis.R

   # ========================================================================
   # Iris Dataset Analysis
   # Author: Your Name
   # Date: Today's Date
   # ========================================================================

   # Setup
   library(tidyverse)
   library(here)

   # Load data (built-in dataset)
   data("iris")

   # Quick exploration
   glimpse(iris)
   summary(iris)

   # Analysis
   species_summary <- iris %>%
     group_by(Species) %>%
     summarise(
       mean_sepal_length = mean(Sepal.Length),
       mean_petal_length = mean(Petal.Length),
       count = n()
     )

   print(species_summary)

   # Visualization
   scatter_plot <- ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
     geom_point(size = 3) +
     labs(
       title = "Iris: Sepal Length vs Petal Length",
       x = "Sepal Length (cm)",
       y = "Petal Length (cm)"
     ) +
     theme_minimal()

   print(scatter_plot)

   # Save results
   write_csv(species_summary, here("output", "species_summary.csv"))
   ggsave(here("output", "iris_scatter.png"), scatter_plot, width = 8, height = 6)
   ```

4. **Run and Test**
   - Run script section by section
   - Check that files are created in output folder
   - Verify paths work correctly

## Summary

You now understand how to organize R work professionally:

### R Scripts
- **Structure** scripts with clear headers and sections
- **Comment** code to explain your reasoning
- **Use** meaningful variable and file names
- **Organize** code logically from setup to results

### RStudio Projects
- **Create** projects for each analysis
- **Use** consistent folder structures
- **Leverage** relative paths with `here()`
- **Document** your work with README files

### Best Practices
- **Plan** your analysis structure before coding
- **Test** scripts section by section
- **Save** intermediate results
- **Document** your session info for reproducibility

## Next Steps

With Module 1 complete, you're ready to dive deeper into R:

- **Module 2**: [R Language Fundamentals](../module-2/index.qmd) - Objects, data types, and control structures
- **Module 3**: [Introduction to Tidyverse](../module-3/index.qmd) - Modern R data manipulation

::: {.callout-tip}
## Pro Tips for Success

1. **Start every analysis** with a new project
2. **Use the same folder structure** across projects
3. **Write scripts as if someone else** will read them
4. **Save your work frequently** and use version control
5. **Test code in small chunks** before running entire scripts
:::

Congratulations! You've completed Module 1 and have a solid foundation for R programming. The habits you develop now will serve you well throughout your data science journey.