---
title: "Working with Dates and Times using lubridate"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to Date and Time Data

Working with dates and times is crucial for time series analysis, scheduling, and understanding temporal patterns in data. The `lubridate` package makes date manipulation intuitive and powerful, providing functions that closely match how we naturally think about dates and times.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(lubridate)

# Key lubridate functions we'll explore
cat("Key lubridate functions:\n")
cat("- Parsing: ymd(), mdy(), dmy(), ymd_hms()\n")
cat("- Extracting: year(), month(), day(), hour(), minute()\n")
cat("- Arithmetic: +, -, years(), months(), days(), hours()\n")
cat("- Rounding: floor_date(), ceiling_date(), round_date()\n")
cat("- Intervals: interval(), duration(), period()\n")
cat("- Time zones: with_tz(), force_tz()\n")
```

## Parsing Dates from Text

### Basic Date Parsing

```{r}
#| label: date-parsing-basic

# Different date formats commonly found in data
date_strings <- c(
  "2024-01-15",           # ISO format
  "01/15/2024",           # US format
  "15/01/2024",           # European format
  "January 15, 2024",     # Written format
  "15-Jan-2024",          # Mixed format
  "2024-01-15 14:30:00"   # With time
)

# Use appropriate parsing functions based on order
iso_dates <- ymd("2024-01-15")           # Year-Month-Day
us_dates <- mdy("01/15/2024")            # Month-Day-Year
euro_dates <- dmy("15/01/2024")          # Day-Month-Year
written_dates <- mdy("January 15, 2024") # Month-Day-Year (text month)

cat("Parsed dates (all represent the same date):\n")
print(iso_dates)
print(us_dates)
print(euro_dates)
print(written_dates)

# Verify they're all the same
identical(iso_dates, us_dates, euro_dates, written_dates)
```

### Parsing with Times

```{r}
#| label: datetime-parsing

# Different datetime formats
datetime_strings <- c(
  "2024-01-15 14:30:00",
  "01/15/2024 2:30 PM",
  "15-01-2024 14:30",
  "2024-01-15T14:30:00Z"
)

# Parse datetimes
ymd_hms("2024-01-15 14:30:00")
mdy_hms("01/15/2024 2:30:00 PM")
dmy_hm("15-01-2024 14:30")

# Handle different separators
ymd_hms("2024/01/15 14:30:00")
ymd_hms("2024.01.15 14.30.00")

# Parse times only
hms("14:30:00")
hm("14:30")
ms("30:45")  # Minutes:Seconds
```

### Handling Messy Date Data

```{r}
#| label: messy-dates

# Real-world messy date data
messy_dates <- c(
  "2024-01-15",
  "01/16/2024",
  "17-Jan-2024",
  "2024-1-18",          # Single digit month
  "01/19/24",           # Two-digit year
  "20th January 2024",  # Ordinal day
  "invalid date",
  NA,
  "2024/01/21"
)

# Parse the whole vector by listing the formats we expect
parsed_dates <- parse_date_time(
  messy_dates,
  orders = c("ymd", "mdy", "dmy", "dBy"),
  quiet = TRUE
) %>%
  as_date()

cat("Original messy dates:\n")
print(messy_dates)
cat("\nParsed dates:\n")
print(parsed_dates)

# Count parsing failures
cat("\nParsing success rate:", mean(!is.na(parsed_dates)), "\n")
```

`parse_date_time()` is vectorised, so it handles the complete column in one
call. The strings in `orders` describe the component order: `ymd` means
year-month-day, while `dBy` allows a day, a full month name, and a year. Invalid
text and missing input become `NA`; always count those failures rather than
silently assuming every value parsed correctly.

## Extracting Date Components

### Basic Date Components

```{r}
#| label: date-components

# Sample dates
sample_dates <- ymd(c("2024-01-15", "2024-06-30", "2024-12-25"))
sample_datetimes <- ymd_hms(c("2024-01-15 09:30:00", "2024-06-30 15:45:30", "2024-12-25 20:15:45"))

# Extract date components
year(sample_dates)
month(sample_dates)
month(sample_dates, label = TRUE)        # Month names
month(sample_dates, label = TRUE, abbr = FALSE)  # Full month names
day(sample_dates)
mday(sample_dates)      # Day of month (same as day())
yday(sample_dates)      # Day of year
wday(sample_dates)      # Day of week (numeric)
wday(sample_dates, label = TRUE)         # Day of week names
wday(sample_dates, label = TRUE, abbr = FALSE)   # Full day names

# Week information
week(sample_dates)      # Week of year
isoweek(sample_dates)   # ISO week (Monday as week start)
```

### Time Components

```{r}
#| label: time-components

# Extract time components
hour(sample_datetimes)
minute(sample_datetimes)
second(sample_datetimes)

# AM/PM
am(sample_datetimes)
pm(sample_datetimes)

# Extract multiple components at once
datetime_components <- tibble(
  datetime = sample_datetimes,
  year = year(datetime),
  month = month(datetime, label = TRUE),
  day = day(datetime),
  weekday = wday(datetime, label = TRUE),
  hour = hour(datetime),
  minute = minute(datetime),
  am_pm = if_else(am(datetime), "AM", "PM")
)

print(datetime_components)
```

### Creating Date Summaries

```{r}
#| label: date-summaries

# Create sample transaction data
set.seed(123)
transactions <- tibble(
  transaction_id = 1:100,
  date = sample(seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "day"), 100),
  amount = round(runif(100, 10, 500), 2)
)

# Add date components for analysis
transactions <- transactions %>%
  mutate(
    year = year(date),
    month = month(date, label = TRUE),
    quarter = quarter(date),
    weekday = wday(date, label = TRUE),
    is_weekend = wday(date) %in% c(1, 7),  # Sunday = 1, Saturday = 7
    week_of_year = week(date)
  )

# Analyze patterns
cat("Sales by day of week:\n")
transactions %>%
  group_by(weekday) %>%
  summarise(
    total_sales = sum(amount),
    avg_sale = round(mean(amount), 2),
    transaction_count = n(),
    .groups = "drop"
  ) %>%
  print()

cat("\nWeekend vs Weekday sales:\n")
transactions %>%
  group_by(is_weekend) %>%
  summarise(
    total_sales = sum(amount),
    avg_sale = round(mean(amount), 2),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(is_weekend, "Weekend", "Weekday")) %>%
  select(day_type, everything(), -is_weekend) %>%
  print()
```

## Date Arithmetic and Manipulation

### Basic Date Arithmetic

```{r}
#| label: date-arithmetic

# Start with a date
start_date <- ymd("2024-01-15")

# Add and subtract time periods
start_date + days(30)          # 30 days later
start_date + months(3)         # 3 months later
start_date + years(1)          # 1 year later

# Combine periods
start_date + years(1) + months(6) + days(15)

# Subtract periods
start_date - days(10)
start_date - months(2)

# Date sequences
seq(from = start_date, to = start_date + months(6), by = "month")
seq(from = start_date, length.out = 10, by = "2 weeks")

# Business days (excluding weekends)
business_days <- seq(from = start_date, to = start_date + days(20), by = "day")
business_days[!wday(business_days) %in% c(1, 7)]  # Remove weekends
```

### Period vs Duration vs Interval

```{r}
#| label: periods-durations-intervals

# Periods: Human-friendly units (months, years can vary)
period_1_month <- period(1, "month")
period_30_days <- period(30, "days")

# Durations: Exact elapsed time stored in seconds
duration_30_days <- ddays(30)

# Compare period vs duration
leap_year_date <- ymd("2024-01-31")  # 2024 is a leap year
cat("Starting date:", as.character(leap_year_date), "\n")
cat("Plus 1 month (period):", as.character(leap_year_date + months(1)), "\n")
cat("Plus 30 days (duration):", as.character(leap_year_date + duration_30_days), "\n")

# Intervals: Specific time spans between two dates
start_date <- ymd("2024-01-01")
end_date <- ymd("2024-12-31")
year_interval <- interval(start_date, end_date)

cat("Interval:", as.character(year_interval), "\n")
cat("Duration in days:", as.numeric(year_interval, "days"), "\n")
cat("Duration in months:", as.numeric(year_interval, "months"), "\n")

# Check if dates fall within interval
test_dates <- ymd(c("2023-12-31", "2024-06-15", "2025-01-01"))
test_dates %within% year_interval
```

A period such as `months(1)` follows the calendar; 31 January plus one month is
not a valid calendar date and therefore returns `NA`. The month-safe operators
`%m+%` and `%m-%` can roll such calculations back to the final valid day. A
duration such as `ddays(30)` instead adds exactly 30 × 24 hours, which is a
different business rule.

### Date Rounding and Truncation

```{r}
#| label: date-rounding

# Sample datetime data
sample_datetime <- ymd_hms("2024-03-15 14:37:23")

# Round to different units
floor_date(sample_datetime, "month")     # Beginning of month
floor_date(sample_datetime, "week")      # Beginning of week (Sunday)
floor_date(sample_datetime, "day")       # Beginning of day
floor_date(sample_datetime, "hour")      # Beginning of hour

ceiling_date(sample_datetime, "month")   # End of month
ceiling_date(sample_datetime, "week")    # End of week
ceiling_date(sample_datetime, "day")     # End of day

round_date(sample_datetime, "hour")      # Nearest hour
round_date(sample_datetime, "15 minutes") # Nearest 15 minutes

# Useful for grouping data
sales_data <- tibble(
  timestamp = ymd_hms("2024-01-15 09:00:00") + minutes(seq(0, 480, 30)),  # Every 30 min for 8 hours
  sales = round(runif(17, 100, 500), 2)
)

# Group by hour
hourly_sales <- sales_data %>%
  mutate(hour = floor_date(timestamp, "hour")) %>%
  group_by(hour) %>%
  summarise(total_sales = sum(sales), .groups = "drop")

print(hourly_sales)
```

## Time Zones and International Dates

### Working with Time Zones

```{r}
#| label: time-zones

# Create datetime in different time zones
utc_time <- ymd_hms("2024-01-15 12:00:00", tz = "UTC")
ny_time <- ymd_hms("2024-01-15 12:00:00", tz = "America/New_York")
london_time <- ymd_hms("2024-01-15 12:00:00", tz = "Europe/London")
tokyo_time <- ymd_hms("2024-01-15 12:00:00", tz = "Asia/Tokyo")

cat("Same wall clock time in different zones:\n")
cat("UTC:", as.character(utc_time), "\n")
cat("New York:", as.character(ny_time), "\n")
cat("London:", as.character(london_time), "\n")
cat("Tokyo:", as.character(tokyo_time), "\n")

# Convert between time zones
cat("\nSame moment in different zones:\n")
cat("UTC:", as.character(utc_time), "\n")
cat("In New York:", as.character(with_tz(utc_time, "America/New_York")), "\n")
cat("In London:", as.character(with_tz(utc_time, "Europe/London")), "\n")
cat("In Tokyo:", as.character(with_tz(utc_time, "Asia/Tokyo")), "\n")

# Force timezone (changes the timezone label without converting)
cat("\nForcing timezone (wall clock time stays same):\n")
utc_forced <- force_tz(ymd_hms("2024-01-15 12:00:00"), "UTC")
ny_forced <- force_tz(ymd_hms("2024-01-15 12:00:00"), "America/New_York")
cat("UTC forced:", as.character(utc_forced), "\n")
cat("NY forced:", as.character(ny_forced), "\n")
cat("Difference:", as.numeric(difftime(ny_forced, utc_forced, units = "hours")), "hours\n")
```

### Handling Daylight Saving Time

```{r}
#| label: daylight-saving

# Daylight saving time transitions in 2024
# Spring forward: March 10, 2024 (2:00 AM becomes 3:00 AM)
# Fall back: November 3, 2024 (2:00 AM becomes 1:00 AM)

# Create times around DST transition
spring_transition <- ymd_hms("2024-03-10 01:30:00", tz = "America/New_York")
spring_after <- spring_transition + hours(1)

fall_transition <- ymd_hms("2024-11-03 01:30:00", tz = "America/New_York")
fall_after <- fall_transition + hours(1)

cat("Spring DST transition (spring forward):\n")
cat("Before:", as.character(spring_transition), "\n")
cat("After +1 hour:", as.character(spring_after), "\n")

cat("\nFall DST transition (fall back):\n")
cat("Before:", as.character(fall_transition), "\n")
cat("After +1 hour:", as.character(fall_after), "\n")

# Working with DST-aware periods
cat("\nDST-aware periods:\n")
dst_start <- ymd_hms("2024-03-09 12:00:00", tz = "America/New_York")
dst_plus_24h <- dst_start + hours(24)  # Exactly 24 hours
dst_plus_1d <- dst_start + days(1)     # 1 calendar day (23 hours due to DST)

cat("Start:", as.character(dst_start), "\n")
cat("Plus 24 hours:", as.character(dst_plus_24h), "\n")
cat("Plus 1 day:", as.character(dst_plus_1d), "\n")
```

## Real-World Date Applications

### Example 1: Business Analytics

```{r}
#| label: business-analytics

# Create realistic business data
set.seed(123)
business_data <- tibble(
  date = sample(seq(ymd("2023-01-01"), ymd("2024-12-31"), by = "day"), 500),
  revenue = round(rnorm(500, 1000, 300), 2),
  customers = round(rnorm(500, 50, 15)),
  product_category = sample(c("Electronics", "Clothing", "Home", "Sports"), 500, replace = TRUE)
) %>%
  arrange(date)

# Add comprehensive date features
business_data <- business_data %>%
  mutate(
    year = year(date),
    month = month(date, label = TRUE),
    quarter = paste0("Q", quarter(date)),
    weekday = wday(date, label = TRUE),
    is_weekend = wday(date) %in% c(1, 7),
    week_of_year = week(date),
    month_year = floor_date(date, "month"),

    # Business calendar features
    is_holiday_season = month(date) %in% c(11, 12),  # Nov-Dec
    is_summer = month(date) %in% c(6, 7, 8),         # Jun-Aug
    days_since_start = as.numeric(difftime(date, min(date), units = "days")),

    # Seasonal indicators
    season = case_when(
      month(date) %in% c(12, 1, 2) ~ "Winter",
      month(date) %in% c(3, 4, 5) ~ "Spring",
      month(date) %in% c(6, 7, 8) ~ "Summer",
      month(date) %in% c(9, 10, 11) ~ "Fall"
    )
  )

# Analyze seasonal patterns
seasonal_analysis <- business_data %>%
  group_by(season, year) %>%
  summarise(
    avg_daily_revenue = round(mean(revenue), 2),
    avg_daily_customers = round(mean(customers), 1),
    total_days = n(),
    .groups = "drop"
  )

cat("Seasonal business patterns:\n")
print(seasonal_analysis)

# Monthly trends
monthly_trends <- business_data %>%
  group_by(month_year) %>%
  summarise(
    total_revenue = sum(revenue),
    total_customers = sum(customers),
    avg_order_value = round(total_revenue / total_customers, 2),
    .groups = "drop"
  ) %>%
  mutate(
    revenue_growth = round((total_revenue / lag(total_revenue) - 1) * 100, 1),
    customer_growth = round((total_customers / lag(total_customers) - 1) * 100, 1)
  )

cat("\nRecent monthly trends:\n")
monthly_trends %>%
  tail(6) %>%
  print()
```

### Example 2: Employee Scheduling

```{r}
#| label: employee-scheduling

# Create employee shift data
shift_schedule <- tibble(
  employee_id = rep(1:10, each = 30),
  shift_date = rep(seq(ymd("2024-01-01"), by = "day", length.out = 30), 10),
  shift_start = sample(c("06:00", "14:00", "22:00"), 300, replace = TRUE),
  hours_worked = sample(c(8, 10, 12), 300, replace = TRUE)
) %>%
  mutate(
    shift_start_time = ymd_hm(paste(shift_date, shift_start)),
    shift_end_time = shift_start_time + hours(hours_worked),

    # Shift classification
    shift_type = case_when(
      hour(shift_start_time) < 10 ~ "Morning",
      hour(shift_start_time) < 18 ~ "Afternoon",
      TRUE ~ "Night"
    ),

    # Date features
    weekday = wday(shift_date, label = TRUE),
    is_weekend = wday(shift_date) %in% c(1, 7),
    week_of_year = week(shift_date)
  )

# Analyze shift patterns
shift_analysis <- shift_schedule %>%
  group_by(shift_type, weekday) %>%
  summarise(
    total_shifts = n(),
    total_hours = sum(hours_worked),
    avg_hours_per_shift = round(mean(hours_worked), 1),
    .groups = "drop"
  )

cat("Shift distribution by type and day:\n")
print(shift_analysis)

# Identify potential overtime issues
overtime_analysis <- shift_schedule %>%
  group_by(employee_id, week = floor_date(shift_date, "week")) %>%
  summarise(
    weekly_hours = sum(hours_worked),
    shifts_worked = n(),
    weekend_shifts = sum(is_weekend),
    .groups = "drop"
  ) %>%
  mutate(
    potential_overtime = weekly_hours > 40,
    excessive_hours = weekly_hours > 50
  )

cat("\nEmployees with potential overtime (>40 hours/week):\n")
overtime_analysis %>%
  filter(potential_overtime) %>%
  arrange(desc(weekly_hours)) %>%
  head(10) %>%
  print()
```

### Example 3: Event Planning and Deadlines

```{r}
#| label: event-planning

# Create project timeline
project_events <- tibble(
  event_name = c("Project Kickoff", "Requirements Complete", "Design Review",
                "Development Start", "Alpha Release", "Beta Release",
                "User Testing", "Final Review", "Launch"),
  planned_date = ymd(c("2024-01-15", "2024-02-01", "2024-02-15",
                      "2024-03-01", "2024-04-15", "2024-05-15",
                      "2024-06-01", "2024-06-15", "2024-07-01")),
  actual_date = ymd(c("2024-01-15", "2024-02-03", "2024-02-18",
                     "2024-03-05", "2024-04-20", "2024-05-22",
                     "2024-06-05", NA, NA))  # Future events not yet complete
)

# Calculate delays and time to completion
project_analysis <- project_events %>%
  mutate(
    # Days between planned and actual
    delay_days = as.numeric(difftime(actual_date, planned_date, units = "days")),

    # Time since project start
    days_from_start_planned = as.numeric(difftime(planned_date, first(planned_date), units = "days")),
    days_from_start_actual = as.numeric(difftime(actual_date, first(planned_date), units = "days")),

    # Status
    status = case_when(
      is.na(actual_date) ~ "Pending",
      delay_days > 0 ~ "Delayed",
      delay_days == 0 ~ "On Time",
      delay_days < 0 ~ "Early"
    ),

    # Days until planned completion (for pending items)
    days_until_planned = ifelse(is.na(actual_date),
                               as.numeric(difftime(planned_date, Sys.Date(), units = "days")),
                               NA)
  )

cat("Project timeline analysis:\n")
project_analysis %>%
  select(event_name, planned_date, actual_date, delay_days, status, days_until_planned) %>%
  print()

# Calculate project health metrics
project_health <- project_analysis %>%
  filter(!is.na(actual_date)) %>%
  summarise(
    completed_events = n(),
    avg_delay = round(mean(delay_days, na.rm = TRUE), 1),
    total_delay = sum(pmax(delay_days, 0), na.rm = TRUE),
    on_time_rate = round(mean(delay_days <= 0, na.rm = TRUE) * 100, 1)
  )

cat("\nProject health metrics:\n")
print(project_health)

# Predict remaining timeline
remaining_events <- project_events %>%
  filter(is.na(actual_date))

if (nrow(remaining_events) > 0) {
  avg_delay <- project_health$avg_delay

  predicted_completion <- remaining_events %>%
    mutate(
      predicted_date = planned_date + days(ceiling(avg_delay)),
      days_until_predicted = as.numeric(difftime(predicted_date, Sys.Date(), units = "days"))
    )

  cat("\nPredicted completion dates (based on average delay):\n")
  predicted_completion %>%
    select(event_name, planned_date, predicted_date, days_until_predicted) %>%
    print()
}
```

## Advanced Date Techniques

### Rolling Date Windows

```{r}
#| label: rolling-windows

# Create daily sales data
daily_sales <- tibble(
  date = seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "day"),
  sales = round(rnorm(366, 1000, 200) + sin(seq_along(date) * 2 * pi / 365) * 100, 2)  # Seasonal pattern
)

# Calculate rolling averages
daily_sales <- daily_sales %>%
  mutate(
    # Rolling 7-day average
    sales_7d_avg = zoo::rollmean(sales, 7, fill = NA, align = "right"),

    # Rolling 30-day average
    sales_30d_avg = zoo::rollmean(sales, 30, fill = NA, align = "right"),

    # Year-over-year comparison (simulated)
    sales_yoy = sales * runif(n(), 0.9, 1.2),  # Simulate YoY growth
    yoy_change = round((sales / sales_yoy - 1) * 100, 1),

    # Month-to-date and quarter-to-date
    month_start = floor_date(date, "month"),
    quarter_start = floor_date(date, "quarter")
  )

# Show recent rolling averages
cat("Recent sales with rolling averages:\n")
daily_sales %>%
  filter(date >= ymd("2024-12-20")) %>%
  select(date, sales, sales_7d_avg, sales_30d_avg) %>%
  print()
```

### Business Day Calculations

```{r}
#| label: business-days

# Function to check if date is a business day
is_business_day <- function(date) {
  # Exclude weekends
  weekday <- wday(date)
  !weekday %in% c(1, 7)  # Not Sunday or Saturday
}

# Function to add business days
add_business_days <- function(start_date, days_to_add) {
  current_date <- start_date
  days_added <- 0

  while (days_added < days_to_add) {
    current_date <- current_date + days(1)
    if (is_business_day(current_date)) {
      days_added <- days_added + 1
    }
  }
  return(current_date)
}

# Calculate business days between dates
count_business_days <- function(start_date, end_date) {
  if (start_date > end_date) return(0)

  date_seq <- seq(from = start_date + days(1), to = end_date, by = "day")
  sum(is_business_day(date_seq))
}

# Test business day functions
start_date <- ymd("2024-01-15")  # Monday
cat("Start date:", format(start_date, "%A, %B %d, %Y"), "\n")

# Add 5 business days
end_date <- add_business_days(start_date, 5)
cat("5 business days later:", format(end_date, "%A, %B %d, %Y"), "\n")

# Count business days in January 2024
jan_start <- ymd("2024-01-01")
jan_end <- ymd("2024-01-31")
business_days_jan <- count_business_days(jan_start, jan_end)
cat("Business days in January 2024:", business_days_jan, "\n")

# Create business day sequence
business_day_seq <- seq(from = ymd("2024-01-01"), to = ymd("2024-01-31"), by = "day")
business_days_only <- business_day_seq[is_business_day(business_day_seq)]

cat("First 10 business days of 2024:\n")
print(head(business_days_only, 10))
```

## Best Practices and Common Pitfalls

### Date Parsing Best Practices

```{r}
#| label: date-best-practices

# Best Practice 1: Always specify format when possible
# Good
explicit_dates <- ymd("2024-01-15")

# Less reliable (could be misinterpreted)
ambiguous_dates <- as.Date("01/02/2024", format = "%m/%d/%Y")  # Is this Jan 2 or Feb 1?

# Best Practice 2: Handle parsing failures gracefully
messy_input <- c("2024-01-15", "invalid", "2024/01/16", NA)

safe_parse <- function(date_strings) {
  results <- ymd(date_strings, quiet = TRUE)

  # Report parsing issues
  failed <- is.na(results) & !is.na(date_strings)
  if (any(failed)) {
    cat("Failed to parse:", sum(failed), "dates\n")
    cat("Failed values:", paste(date_strings[failed], collapse = ", "), "\n")
  }

  return(results)
}

parsed_safely <- safe_parse(messy_input)

# Best Practice 3: Validate date ranges
validate_date_range <- function(dates, min_date = ymd("1900-01-01"), max_date = Sys.Date()) {
  valid <- !is.na(dates) & dates >= min_date & dates <= max_date

  if (!all(valid, na.rm = TRUE)) {
    invalid_dates <- dates[!valid & !is.na(dates)]
    cat("Found", length(invalid_dates), "dates outside valid range\n")
    cat("Invalid dates:", paste(as.character(invalid_dates), collapse = ", "), "\n")
  }

  return(valid)
}

# Test validation
test_dates <- ymd(c("1850-01-01", "2024-01-15", "2050-01-01"))
validation_results <- validate_date_range(test_dates)
```

### Common Mistakes

```{r}
#| label: date-mistakes

# Mistake 1: Confusing time zones
# Problem: Creating datetime without specifying timezone
ambiguous_time <- ymd_hms("2024-01-15 12:00:00")  # What timezone?
cat("Ambiguous timezone:", tz(ambiguous_time), "\n")

# Better: Always specify timezone when it matters
explicit_time <- ymd_hms("2024-01-15 12:00:00", tz = "America/New_York")
cat("Explicit timezone:", tz(explicit_time), "\n")

# Mistake 2: Not handling leap years
# Problem: Assuming February always has 28 days
leap_year_check <- function(year) {
  feb_29 <- ymd(paste(year, "02", "29", sep = "-"))
  !is.na(feb_29)
}

cat("Is 2024 a leap year?", leap_year_check(2024), "\n")
cat("Is 2023 a leap year?", leap_year_check(2023), "\n")

# Mistake 3: Ignoring DST in calculations
# Problem: Assuming all days have 24 hours
dst_spring <- ymd("2024-03-10", tz = "America/New_York")  # Spring forward
hours_in_day <- as.numeric(difftime(dst_spring + days(1), dst_spring, units = "hours"))
cat("Hours in DST transition day:", hours_in_day, "\n")

# Mistake 4: Not validating date arithmetic
# Problem: Invalid dates from arithmetic
jan_31 <- ymd("2024-01-31")
invalid_result <- jan_31 + months(1)  # What's January 31 + 1 month?
cat("Jan 31 + 1 month:", as.character(invalid_result), "\n")

# Better: Use %m+% for month arithmetic that handles this
valid_result <- jan_31 %m+% months(1)
cat("Jan 31 %m+% 1 month:", as.character(valid_result), "\n")
```

## Exercises

### Exercise 1: Sales Data Analysis
Given daily sales data with various date formats:
1. Parse dates from mixed formats
2. Extract seasonal and trend components
3. Calculate rolling averages and growth rates
4. Identify business day vs weekend patterns

### Exercise 2: Employee Time Tracking
Create a time tracking system that:
1. Handles different time zones for global employees
2. Calculates overtime based on business rules
3. Accounts for holidays and vacation days
4. Generates payroll reports by pay period

### Exercise 3: Project Timeline Management
Build a project management system that:
1. Tracks milestones and deadlines
2. Calculates critical path and delays
3. Handles business day scheduling
4. Predicts completion dates based on current progress

### Exercise 4: Event Scheduling
Design an event scheduling application that:
1. Handles recurring events (daily, weekly, monthly)
2. Manages time zone conversions for global events
3. Avoids scheduling conflicts
4. Sends reminders based on time until event

## Summary

The `lubridate` package makes date and time manipulation intuitive and powerful:

### Key Functions:
- **Parsing**: `ymd()`, `mdy()`, `dmy()`, `ymd_hms()`
- **Extracting**: `year()`, `month()`, `day()`, `hour()`, `minute()`
- **Arithmetic**: `+`, `-`, `years()`, `months()`, `days()`
- **Rounding**: `floor_date()`, `ceiling_date()`, `round_date()`
- **Time zones**: `with_tz()`, `force_tz()`, `tz()`

### Best Practices:
- **Specify time zones explicitly** when they matter
- **Handle parsing failures gracefully** with error checking
- **Validate date ranges** to catch data entry errors
- **Use appropriate period types** (period vs duration)
- **Account for leap years and DST** in calculations

### Common Applications:
- **Time series analysis**: Extracting seasonal patterns
- **Business analytics**: Calculating rolling metrics
- **Scheduling**: Managing appointments and deadlines
- **International applications**: Handling multiple time zones

### Remember:
- Dates in R are stored as numbers (days since 1970-01-01)
- Time zones can be tricky - always be explicit
- Business day calculations need custom logic
- DST transitions affect duration calculations
- Use `%m+%` for safer month arithmetic

Date and time manipulation is essential for many data analysis tasks. With lubridate, you can handle even complex temporal data scenarios with confidence!

This completes our exploration of data types in the tidyverse. You now have the tools to work effectively with strings, factors, and dates in your data analysis projects!
