---
title: "Module 4: Data Types in Tidyverse"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
---

## Welcome to Data Types in Tidyverse! 🔤📅

Now that you understand the basics of tidyverse data manipulation, it's time to dive deep into working with specific data types. Real-world data comes in many forms - text, categories, dates, and times - and the tidyverse provides specialized packages to handle each type effectively.

## What You'll Learn

In this module, we'll explore how to work with three fundamental data types using tidyverse packages:

### 1. Strings with stringr
- Pattern matching and text manipulation
- Regular expressions for complex text processing
- Cleaning and standardizing text data
- Extracting information from unstructured text

### 2. Factors with forcats
- Creating and managing categorical data
- Reordering factor levels for better visualization
- Collapsing and recoding categories
- Handling missing levels and unexpected values

### 3. Dates and Times with lubridate
- Parsing dates from various formats
- Extracting date components (year, month, day)
- Date arithmetic and time intervals
- Working with time zones and periods

## Learning Objectives

By the end of this module, you will be able to:

- ✅ Manipulate and clean text data using stringr functions
- ✅ Use regular expressions for pattern matching
- ✅ Create and manage factor variables with forcats
- ✅ Reorder and recode categorical data effectively
- ✅ Parse, manipulate, and format dates with lubridate
- ✅ Perform date arithmetic and handle time zones
- ✅ Apply these skills to real-world data cleaning tasks

## Why These Data Types Matter

Understanding how to work with different data types is crucial because:

1. **Real-world messiness**: Data rarely comes in perfect numeric form
2. **Text processing**: Much information is stored as text that needs cleaning
3. **Categorical insights**: Factors help organize and analyze groupings
4. **Time series analysis**: Dates enable temporal analysis and trends
5. **Data quality**: Proper type handling prevents analysis errors

## The Tidyverse Approach

The tidyverse provides specialized packages for each data type:

```{r}
#| label: tidyverse-packages
#| message: false

# Load the core tidyverse (includes readr, dplyr, ggplot2, etc.)
library(tidyverse)

# Specialized packages for data types
library(stringr)   # String manipulation
library(forcats)   # Factor handling
library(lubridate) # Date and time manipulation

# Check versions
cat("Tidyverse approach to data types:\n")
cat("- stringr version:", as.character(packageVersion("stringr")), "\n")
cat("- forcats version:", as.character(packageVersion("forcats")), "\n")
cat("- lubridate version:", as.character(packageVersion("lubridate")), "\n")
```

## Module Overview

### The Philosophy: Type-Specific Tools

Each tidyverse package follows consistent design principles:

```{r}
#| label: design-principles

# Example of consistent function naming
sample_text <- c("Hello World", "data science", "R Programming")
sample_factors <- factor(c("High", "Medium", "Low", "High"))
sample_dates <- c("2024-01-15", "2024-03-20", "2024-12-01")

# stringr functions start with str_
str_to_upper(sample_text)
str_length(sample_text)

# forcats functions start with fct_
fct_count(sample_factors)
fct_relevel(sample_factors, "Low", "Medium", "High")

# lubridate has intuitive parsing functions
ymd(sample_dates)
mdy("03/15/2024")
```

This consistency makes the tidyverse easier to learn and remember!

## Real-World Data Challenges

Let's look at a typical messy dataset that demonstrates why we need specialized data type tools:

```{r}
#| label: messy-data-example

# Simulate messy real-world data
messy_survey <- tibble(
  id = 1:6,
  name = c("John DOE", "jane smith", "Bob-Johnson", "Mary O'Connor", "李小明", "José García"),
  satisfaction = c("Very Satisfied", "satisfied", "NEUTRAL", "Dissatisfied", "very satisfied", "Satisfied"),
  date_submitted = c("2024-01-15", "01/16/2024", "2024-1-17", "16-Jan-2024", "2024/01/18", "Jan 19, 2024"),
  comments = c("Great service!", "good stuff", "It's okay...", "Could be better!", "非常好！", "¡Excelente!"),
  score = c("4.5", "4", "3", "2", "5", "4")
)

print(messy_survey)

cat("\nChallenges in this data:\n")
cat("- Names: inconsistent capitalization and formats\n")
cat("- Satisfaction: inconsistent factor levels\n")
cat("- Dates: multiple formats that need parsing\n")
cat("- Comments: mixed languages and punctuation\n")
cat("- Scores: stored as text instead of numbers\n")
```

By the end of this module, you'll know how to clean and standardize all of these issues!

## Module Structure

This module is organized into three main sections:

1. **[Working with Strings (stringr)](strings-stringr.qmd)**: Master text manipulation and pattern matching
2. **[Managing Factors (forcats)](factors-forcats.qmd)**: Handle categorical data effectively
3. **[Dates and Times (lubridate)](dates-lubridate.qmd)**: Parse and manipulate temporal data

Each section includes:
- Core functions and concepts
- Regular expressions and advanced techniques
- Real-world examples and case studies
- Best practices and common pitfalls
- Hands-on exercises

## Prerequisites

Before starting this module, make sure you have:

- Completed Module 2 (R Fundamentals) and Module 3 (Tidyverse Introduction)
- Understanding of tibbles, pipes, and basic dplyr operations
- R and tidyverse packages installed and loaded

## Quick Motivation Example

Here's a preview of what you'll be able to do:

```{r}
#| label: motivation-preview

# Transform the messy survey data
cleaned_survey <- messy_survey %>%
  # Clean names with stringr
  mutate(
    name_clean = str_to_title(str_replace_all(name, "[^A-Za-z\\s]", " ")),
    name_clean = str_squish(name_clean)
  ) %>%
  # Standardize satisfaction with forcats
  mutate(
    satisfaction_clean = str_to_lower(satisfaction),
    satisfaction_clean = case_when(
      str_detect(satisfaction_clean, "very satisfied") ~ "Very Satisfied",
      str_detect(satisfaction_clean, "satisfied") ~ "Satisfied",
      str_detect(satisfaction_clean, "neutral") ~ "Neutral",
      str_detect(satisfaction_clean, "dissatisfied") ~ "Dissatisfied",
      TRUE ~ satisfaction_clean
    ),
    satisfaction_factor = factor(satisfaction_clean,
                                levels = c("Very Satisfied", "Satisfied", "Neutral", "Dissatisfied"))
  ) %>%
  # Parse dates with lubridate
  mutate(
    date_clean = case_when(
      str_detect(date_submitted, "^\\d{4}-\\d{1,2}-\\d{1,2}$") ~ ymd(date_submitted),
      str_detect(date_submitted, "^\\d{1,2}/\\d{1,2}/\\d{4}$") ~ mdy(date_submitted),
      str_detect(date_submitted, "^\\d{4}/\\d{1,2}/\\d{1,2}$") ~ ymd(date_submitted),
      str_detect(date_submitted, "\\d{1,2}-[A-Za-z]{3}-\\d{4}") ~ dmy(date_submitted),
      str_detect(date_submitted, "[A-Za-z]{3}\\s+\\d{1,2},\\s+\\d{4}") ~ mdy(date_submitted),
      TRUE ~ as.Date(NA)
    )
  ) %>%
  # Clean scores
  mutate(score_numeric = as.numeric(score)) %>%
  # Select clean columns
  select(id, name_clean, satisfaction_factor, date_clean, score_numeric, comments)

print(cleaned_survey)

# Quick analysis with clean data
cat("\nWith clean data, analysis becomes straightforward:\n")
cleaned_survey %>%
  group_by(satisfaction_factor) %>%
  summarise(
    count = n(),
    avg_score = round(mean(score_numeric, na.rm = TRUE), 2),
    date_range = paste(min(date_clean, na.rm = TRUE), "to", max(date_clean, na.rm = TRUE))
  ) %>%
  print()
```

Amazing! In just a few lines of code, we've cleaned names, standardized categories, parsed dates, and converted scores to numbers. This is the power of tidyverse data type packages!

## Getting Started

Ready to master data types in the tidyverse? Let's begin with string manipulation using the powerful stringr package: **[Working with Strings (stringr)](strings-stringr.qmd)**

## Summary

Working with different data types is a fundamental skill in data analysis. The tidyverse provides three excellent packages that make text, categorical, and date data much easier to work with:

- **stringr**: Consistent, powerful string manipulation
- **forcats**: Intuitive factor handling and reordering
- **lubridate**: Natural date and time parsing

These tools will transform how you clean and prepare real-world data for analysis. Let's dive in! 🚀