---
title: "Data Transformation with dplyr Basics"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to dplyr

dplyr is the tidyverse's grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges. Think of dplyr verbs as the fundamental building blocks for data transformation.

```{r}
#| label: setup
library(tidyverse)

# We'll use the built-in mtcars dataset for examples
data(mtcars)
cars <- as_tibble(mtcars, rownames = "model")
glimpse(cars)
```

## The Five Key Verbs

dplyr provides five key verbs for data manipulation:

1. **`select()`** - Choose columns
2. **`filter()`** - Choose rows
3. **`mutate()`** - Create or modify columns
4. **`arrange()`** - Reorder rows
5. **`summarize()`** - Reduce multiple values to a single summary

## select(): Choose Columns

### Basic Column Selection

```{r}
#| label: select-basics

# Select specific columns by name
cars %>%
  select(model, mpg, cyl, hp) %>%
  head()

# Select columns by position
cars %>%
  select(1:4) %>%
  head()

# Select all columns except some
cars %>%
  select(-am, -gear, -carb) %>%
  head()
```

### Selection Helpers

dplyr provides helpful functions for column selection:

```{r}
#| label: select-helpers

# Select columns starting with "d"
cars %>%
  select(model, starts_with("d")) %>%
  head()

# Select columns ending with "p"
cars %>%
  select(model, ends_with("p")) %>%
  head()

# Select columns containing "a"
cars %>%
  select(model, contains("a")) %>%
  head()

# Select columns matching a pattern
cars %>%
  select(model, matches("^[md]")) %>%
  head()
```

### Renaming While Selecting

```{r}
#| label: select-rename

# Rename columns while selecting
cars %>%
  select(
    car = model,
    miles_per_gallon = mpg,
    cylinders = cyl,
    horsepower = hp
  ) %>%
  head()

# Or use rename() to keep all columns
cars %>%
  rename(
    miles_per_gallon = mpg,
    cylinders = cyl
  ) %>%
  select(model, miles_per_gallon, cylinders) %>%
  head()
```

### Reordering Columns

```{r}
#| label: select-reorder

# Move specific columns to the front
cars %>%
  select(model, mpg, hp, everything()) %>%
  head()

# Use relocate() for more control
cars %>%
  relocate(hp, .before = mpg) %>%
  select(model, hp, mpg, cyl) %>%
  head()
```

## filter(): Choose Rows

### Basic Filtering

```{r}
#| label: filter-basics

# Filter for cars with mpg > 20
cars %>%
  filter(mpg > 20) %>%
  select(model, mpg, cyl)

# Multiple conditions (AND)
cars %>%
  filter(mpg > 20, cyl == 4) %>%
  select(model, mpg, cyl)

# Multiple conditions (OR)
cars %>%
  filter(mpg > 30 | hp > 200) %>%
  select(model, mpg, hp)
```

### Complex Filtering

```{r}
#| label: filter-complex

# Using %in% for multiple values
cars %>%
  filter(cyl %in% c(4, 6)) %>%
  select(model, cyl, mpg) %>%
  head()

# Combining conditions
cars %>%
  filter(
    (mpg > 20 & cyl == 4) | (hp > 200 & cyl == 8)
  ) %>%
  select(model, mpg, cyl, hp)

# Filter with between()
cars %>%
  filter(between(mpg, 15, 25)) %>%
  select(model, mpg) %>%
  head()
```

### Filtering with String Matching

```{r}
#| label: filter-strings

# Filter models containing "Merc"
cars %>%
  filter(str_detect(model, "Merc")) %>%
  select(model, mpg)

# Case-insensitive matching
cars %>%
  filter(str_detect(model, "(?i)MAZDA")) %>%
  select(model, mpg)
```

### Filtering Missing Values

```{r}
#| label: filter-na

# Create data with NAs
cars_with_na <- cars %>%
  mutate(mpg = if_else(row_number() %in% c(2, 5, 8), NA_real_, mpg))

# Remove rows with NA in mpg
cars_with_na %>%
  filter(!is.na(mpg)) %>%
  select(model, mpg) %>%
  head()

# Keep only complete cases
cars_with_na %>%
  filter(complete.cases(.)) %>%
  nrow()
```

## mutate(): Create or Modify Columns

### Creating New Columns

```{r}
#| label: mutate-create

# Create new columns based on existing ones
cars %>%
  mutate(
    mpg_per_cyl = mpg / cyl,
    hp_per_cyl = hp / cyl,
    performance_ratio = hp / wt,
    is_efficient = mpg > 20
  ) %>%
  select(model, mpg_per_cyl, hp_per_cyl, performance_ratio, is_efficient) %>%
  head()
```

### Modifying Existing Columns

```{r}
#| label: mutate-modify

# Modify existing columns
cars %>%
  mutate(
    mpg = round(mpg, 0),
    wt = wt * 1000,  # Convert to pounds
    model = str_to_upper(model)
  ) %>%
  select(model, mpg, wt) %>%
  head()
```

### Conditional Mutations

```{r}
#| label: mutate-conditional

# Using if_else()
cars %>%
  mutate(
    efficiency = if_else(mpg > 20, "Efficient", "Not Efficient"),
    size = if_else(wt < 3, "Light", "Heavy")
  ) %>%
  select(model, mpg, wt, efficiency, size) %>%
  head()

# Using case_when() for multiple conditions
cars %>%
  mutate(
    performance = case_when(
      hp < 100 ~ "Low Power",
      hp < 150 ~ "Medium Power",
      hp < 200 ~ "High Power",
      TRUE ~ "Very High Power"
    ),
    efficiency = case_when(
      mpg > 30 ~ "Excellent",
      mpg > 25 ~ "Good",
      mpg > 20 ~ "Fair",
      TRUE ~ "Poor"
    )
  ) %>%
  select(model, hp, mpg, performance, efficiency) %>%
  head(10)
```

### Window Functions in mutate()

```{r}
#| label: mutate-window

# Add ranking and cumulative statistics
cars %>%
  arrange(desc(mpg)) %>%
  mutate(
    mpg_rank = row_number(),
    mpg_dense_rank = dense_rank(mpg),
    mpg_percent_rank = percent_rank(mpg),
    cumulative_avg_mpg = cummean(mpg)
  ) %>%
  select(model, mpg, mpg_rank, cumulative_avg_mpg) %>%
  head(10)
```

## arrange(): Reorder Rows

### Basic Sorting

```{r}
#| label: arrange-basics

# Sort by mpg (ascending)
cars %>%
  arrange(mpg) %>%
  select(model, mpg) %>%
  head()

# Sort by mpg (descending)
cars %>%
  arrange(desc(mpg)) %>%
  select(model, mpg) %>%
  head()
```

### Multiple Column Sorting

```{r}
#| label: arrange-multiple

# Sort by multiple columns
cars %>%
  arrange(cyl, desc(mpg)) %>%
  select(model, cyl, mpg) %>%
  head(10)

# Complex sorting with calculated values
cars %>%
  mutate(efficiency_score = mpg / wt) %>%
  arrange(desc(efficiency_score)) %>%
  select(model, mpg, wt, efficiency_score) %>%
  head()
```

## summarize(): Reduce to Single Values

### Basic Summarization

```{r}
#| label: summarize-basics

# Calculate summary statistics
cars %>%
  summarize(
    avg_mpg = mean(mpg),
    median_mpg = median(mpg),
    sd_mpg = sd(mpg),
    min_mpg = min(mpg),
    max_mpg = max(mpg),
    n_cars = n()
  )

# Multiple summary statistics
cars %>%
  summarize(
    across(c(mpg, hp, wt),
           list(mean = mean,
                median = median,
                sd = sd),
           .names = "{.col}_{.fn}")
  )
```

### Counting and Proportions

```{r}
#| label: summarize-count

# Count unique values
cars %>%
  summarize(
    n_total = n(),
    n_efficient = sum(mpg > 20),
    prop_efficient = mean(mpg > 20),
    n_unique_cyl = n_distinct(cyl)
  )

# Using count() shortcut
cars %>%
  count(cyl, sort = TRUE)

# Count with weights
cars %>%
  count(cyl, wt = hp, name = "total_hp")
```

## group_by(): The Power Multiplier

### Grouped Operations

```{r}
#| label: group-by-basics

# Group by cylinder and calculate summaries
cars %>%
  group_by(cyl) %>%
  summarize(
    n = n(),
    avg_mpg = mean(mpg),
    avg_hp = mean(hp),
    avg_wt = mean(wt),
    .groups = "drop"
  )

# Multiple grouping variables
cars %>%
  mutate(transmission = if_else(am == 1, "Manual", "Automatic")) %>%
  group_by(cyl, transmission) %>%
  summarize(
    n = n(),
    avg_mpg = mean(mpg),
    .groups = "drop"
  ) %>%
  arrange(cyl, transmission)
```

### Grouped Mutations

```{r}
#| label: group-mutate

# Add group-level statistics to each row
cars %>%
  group_by(cyl) %>%
  mutate(
    avg_mpg_for_cyl = mean(mpg),
    mpg_diff_from_group = mpg - avg_mpg_for_cyl,
    mpg_rank_in_group = rank(desc(mpg))
  ) %>%
  select(model, cyl, mpg, avg_mpg_for_cyl, mpg_diff_from_group, mpg_rank_in_group) %>%
  arrange(cyl, mpg_rank_in_group) %>%
  head(10)
```

### Grouped Filtering

```{r}
#| label: group-filter

# Keep only the most efficient car in each cylinder group
cars %>%
  group_by(cyl) %>%
  filter(mpg == max(mpg)) %>%
  select(model, cyl, mpg) %>%
  arrange(cyl)

# Keep groups meeting certain criteria
cars %>%
  group_by(cyl) %>%
  filter(mean(mpg) > 20) %>%
  ungroup() %>%
  count(cyl)
```

## Combining Multiple Operations

### Complex Data Pipelines

```{r}
#| label: complex-pipeline

# Comprehensive analysis pipeline
analysis <- cars %>%
  # Add calculated columns
  mutate(
    efficiency = mpg / wt,
    performance = hp / wt,
    transmission = if_else(am == 1, "Manual", "Automatic")
  ) %>%
  # Filter for relevant cases
  filter(complete.cases(.)) %>%
  # Group for analysis
  group_by(cyl, transmission) %>%
  # Calculate summaries
  summarize(
    n_cars = n(),
    avg_mpg = mean(mpg),
    avg_efficiency = mean(efficiency),
    avg_performance = mean(performance),
    best_mpg = max(mpg),
    .groups = "drop"
  ) %>%
  # Add overall rankings
  mutate(
    efficiency_rank = dense_rank(desc(avg_efficiency)),
    performance_rank = dense_rank(desc(avg_performance))
  ) %>%
  # Sort by efficiency
  arrange(efficiency_rank)

analysis
```

### Real-World Example: Sales Analysis

```{r}
#| label: sales-example

# Create sample sales data
set.seed(123)
sales <- tibble(
  date = sample(seq.Date(from = as.Date("2024-01-01"),
                        to = as.Date("2024-03-31"),
                        by = "day"), 500, replace = TRUE),
  product = sample(c("Widget A", "Widget B", "Widget C", "Widget D"),
                  500, replace = TRUE),
  region = sample(c("North", "South", "East", "West"),
                 500, replace = TRUE),
  quantity = sample(1:20, 500, replace = TRUE),
  price = sample(c(9.99, 14.99, 19.99, 24.99), 500, replace = TRUE),
  discount = sample(c(0, 0.1, 0.15, 0.2), 500, replace = TRUE,
                   prob = c(0.6, 0.2, 0.15, 0.05))
)

# Complex analysis
sales_analysis <- sales %>%
  # Calculate revenue
  mutate(
    revenue = quantity * price * (1 - discount),
    month = format(date, "%Y-%m"),
    has_discount = discount > 0
  ) %>%
  # Monthly product performance by region
  group_by(month, product, region) %>%
  summarize(
    n_transactions = n(),
    total_quantity = sum(quantity),
    total_revenue = sum(revenue),
    avg_discount = mean(discount),
    pct_with_discount = mean(has_discount),
    .groups = "drop"
  ) %>%
  # Add product-level rankings within each month
  group_by(month) %>%
  mutate(
    revenue_rank = dense_rank(desc(total_revenue)),
    quantity_rank = dense_rank(desc(total_quantity))
  ) %>%
  ungroup() %>%
  # Focus on top performers
  filter(revenue_rank <= 3) %>%
  arrange(month, revenue_rank)

head(sales_analysis, 10)
```

## Common Patterns and Best Practices

### Pattern 1: Filter-Select-Arrange

```{r}
#| label: pattern-fsa

# Common pattern for data exploration
cars %>%
  filter(mpg > 20) %>%
  select(model, mpg, cyl, hp) %>%
  arrange(desc(mpg))
```

### Pattern 2: Group-Summarize-Arrange

```{r}
#| label: pattern-gsa

# Common pattern for aggregation
cars %>%
  group_by(cyl) %>%
  summarize(
    n = n(),
    avg_mpg = mean(mpg),
    avg_hp = mean(hp),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_mpg))
```

### Pattern 3: Mutate-Filter-Select

```{r}
#| label: pattern-mfs

# Common pattern for feature engineering
cars %>%
  mutate(
    efficiency_ratio = mpg / wt,
    is_efficient = efficiency_ratio > 7
  ) %>%
  filter(is_efficient) %>%
  select(model, mpg, wt, efficiency_ratio)
```

## Advanced Tips

### 1. Using across() for Multiple Columns

```{r}
#| label: across-examples

# Apply same operation to multiple columns
cars %>%
  mutate(across(c(mpg, hp, wt), ~round(.x, 0))) %>%
  select(model, mpg, hp, wt) %>%
  head()

# Summarize multiple columns
cars %>%
  group_by(cyl) %>%
  summarize(across(c(mpg, hp, wt),
                  list(mean = mean, sd = sd),
                  .names = "{.col}_{.fn}")) %>%
  round(2)
```

### 2. Using slice() Functions

```{r}
#| label: slice-functions

# Get top N rows per group
cars %>%
  group_by(cyl) %>%
  slice_max(mpg, n = 2) %>%
  select(model, cyl, mpg)

# Get random sample per group
set.seed(123)
cars %>%
  group_by(cyl) %>%
  slice_sample(n = 2) %>%
  select(model, cyl, mpg) %>%
  arrange(cyl)
```

### 3. Scoped Variants

```{r}
#| label: scoped-variants

# Select columns by condition
cars %>%
  select(model, where(is.numeric)) %>%
  head(3)

# Summarize all numeric columns
cars %>%
  summarize(across(where(is.numeric), mean)) %>%
  round(2)
```

## Exercises

### Exercise 1: Basic Verbs
Using the `iris` dataset:
1. Select only the Petal columns and Species
2. Filter for Petal.Length > 4
3. Create a new column for Petal.Area (Length × Width)
4. Arrange by Petal.Area descending
5. Show only the top 10 rows

### Exercise 2: Grouped Operations
Using `mtcars`:
1. Group by number of gears
2. Calculate average mpg and hp for each group
3. Add a column showing the difference from overall average mpg
4. Keep only groups where average mpg > 20

### Exercise 3: Complex Pipeline
Create a pipeline that:
1. Filters cars with 4 or 6 cylinders
2. Creates an efficiency score (mpg × 1000 / (hp × wt))
3. Groups by cylinder count
4. Finds the most and least efficient car in each group
5. Presents the results in a clean summary table

## Summary

dplyr provides a powerful, intuitive grammar for data manipulation:

- **`select()`**: Choose your columns
- **`filter()`**: Choose your rows
- **`mutate()`**: Create new variables
- **`arrange()`**: Order your data
- **`summarize()`**: Calculate summaries
- **`group_by()`**: Split-apply-combine operations

These verbs can be combined in endless ways to solve virtually any data manipulation challenge. The key is to think of data transformation as a series of simple steps, each accomplishing one specific task.

Next, we'll explore [data reshaping with tidyr](tidyr-principles.qmd) to complement these transformation skills!