---
title: "Module 3: Introduction to the Tidyverse"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
---

## Welcome to the Tidyverse! 🌟

The tidyverse is a collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures. The tidyverse makes data manipulation, exploration, and visualization faster and more intuitive.

## What is the Tidyverse?

The tidyverse is an opinionated collection of R packages that work in harmony because they share common data representations and API design. The packages are designed to work together naturally, making data analysis workflows more efficient and readable.

### Core Tidyverse Packages

The tidyverse includes several core packages that you'll use in almost every analysis:

- **ggplot2**: Create elegant data visualizations using the grammar of graphics
- **dplyr**: A grammar of data manipulation, providing a consistent set of verbs
- **tidyr**: Tidy messy data and reshape data structures
- **readr**: Fast and friendly reading of rectangular data
- **purrr**: Functional programming tools for working with functions and vectors
- **tibble**: Modern re-imagining of the data frame
- **stringr**: Cohesive set of functions for working with strings
- **forcats**: Tools for working with categorical variables (factors)

## Installing and Loading the Tidyverse

```{r}
#| label: install-tidyverse
#| eval: false

# Install the tidyverse (only need to do this once)
install.packages("tidyverse")

# Load the tidyverse
library(tidyverse)
```

When you load the tidyverse, you'll see which packages are attached and any conflicts with other packages:

```{r}
#| label: load-tidyverse
#| message: true

library(tidyverse)
```

## The Tidyverse Philosophy

### 1. Tidy Data Principles

The tidyverse is built around the concept of tidy data, which has three key characteristics:

1. Each variable forms a column
2. Each observation forms a row
3. Each type of observational unit forms a table

### 2. Consistent Grammar

The tidyverse provides a consistent grammar for data manipulation. Functions are named using verbs that describe what they do:

- `select()`: choose columns
- `filter()`: choose rows
- `mutate()`: create new columns
- `summarize()`: reduce multiple values to a single summary
- `arrange()`: reorder rows

### 3. The Pipe Operator

The pipe operator (`%>%` or `|>`) is central to tidyverse workflows, allowing you to chain operations together in a readable, left-to-right fashion.

## Module Overview

In this module, we'll explore the fundamental packages and concepts of the tidyverse:

### Topics Covered

1. **The Pipe Operator**: Learn to chain operations for cleaner, more readable code
2. **Tibbles**: Modern data frames with improved printing and subsetting
3. **Data Import with readr**: Efficiently read rectangular data from files
4. **Data Transformation with dplyr**: Master the five key verbs for data manipulation
5. **Data Tidying with tidyr**: Reshape and organize messy data

### Learning Objectives

By the end of this module, you will be able to:

- ✅ Understand the tidyverse philosophy and ecosystem
- ✅ Use the pipe operator to create readable data pipelines
- ✅ Import data from various file formats using readr
- ✅ Perform basic data manipulations with dplyr
- ✅ Reshape data between wide and long formats with tidyr
- ✅ Work with tibbles as an enhanced alternative to data frames

## Quick Example: The Power of the Tidyverse

Let's see a quick example that demonstrates the elegance of tidyverse code:

```{r}
#| label: tidyverse-example

# Create some sample data
sales_data <- tibble(
  date = seq.Date(from = as.Date("2024-01-01"),
                  to = as.Date("2024-01-10"),
                  by = "day"),
  product = rep(c("A", "B"), 5),
  units = sample(10:50, 10),
  price = rep(c(9.99, 14.99), 5)
)

# Analyze the data using tidyverse functions
sales_summary <- sales_data %>%
  mutate(revenue = units * price) %>%
  group_by(product) %>%
  summarize(
    total_units = sum(units),
    total_revenue = sum(revenue),
    avg_daily_revenue = mean(revenue),
    .groups = "drop"
  ) %>%
  arrange(desc(total_revenue))

print(sales_summary)
```

This example shows how tidyverse functions work together to:
1. Create new variables with `mutate()`
2. Group data by categories with `group_by()`
3. Calculate summaries with `summarize()`
4. Sort results with `arrange()`
5. Chain it all together with the pipe `%>%`

## Base R vs. Tidyverse

While base R is powerful and important to understand, the tidyverse often provides more intuitive solutions:

```{r}
#| label: base-vs-tidyverse

# Base R approach
base_result <- aggregate(
  sales_data$units,
  by = list(product = sales_data$product),
  FUN = sum
)
names(base_result)[2] <- "total_units"

# Tidyverse approach
tidy_result <- sales_data %>%
  group_by(product) %>%
  summarize(total_units = sum(units))

# Both give the same result, but tidyverse is more readable
print(base_result)
print(tidy_result)
```

## Getting Help

The tidyverse has excellent documentation and resources:

- **Official website**: [tidyverse.org](https://www.tidyverse.org)
- **R for Data Science book**: Free online book by Hadley Wickham
- **RStudio Cheat Sheets**: Visual guides for each package
- **Package documentation**: Use `?function_name` or `help(package = "package_name")`

## What's Next?

In the following sections, we'll dive deep into each component:

- [The Pipe Operator](pipe-operator.qmd): Master the art of chaining operations
- [Working with Tibbles](tibbles.qmd): Modern data frames for the 21st century
- [Importing Data with readr](readr-import.qmd): Get your data into R efficiently
- [Data Manipulation with dplyr](dplyr-basics.qmd): Transform your data with elegance
- [Tidying Data with tidyr](tidyr-principles.qmd): Reshape data for analysis

## Practice Exercises

### Exercise 1: Install and Explore
1. Install the tidyverse if you haven't already
2. Load the tidyverse and examine which packages are attached
3. Check for any conflicts with `tidyverse_conflicts()`

### Exercise 2: First Pipeline
Create a simple pipeline that:
1. Creates a tibble with student grades
2. Calculates the average grade per subject
3. Arranges the results from highest to lowest average

### Exercise 3: Compare Approaches
Take a simple data manipulation task and implement it in both base R and tidyverse. Which do you find more readable?

## Summary

The tidyverse represents a modern, coherent approach to data analysis in R. Its consistent design principles, readable syntax, and powerful tools make it an essential part of any R programmer's toolkit. As we progress through this module, you'll gain hands-on experience with each of the core packages and learn to leverage their combined power for efficient data analysis.

Remember: the tidyverse is not just about individual functions, but about a philosophy of data analysis that emphasizes clarity, consistency, and reproducibility. Welcome to a more elegant way of working with data! 🎉