---
title: "The Pipe Operator: Making R Code Readable"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to the Pipe

The pipe operator is one of the most transformative features in modern R programming. It allows you to chain operations together, creating readable, left-to-right workflows that mirror how we think about data transformations.

## Two Pipes: `%>%` and `|>`

R now has two pipe operators:

1. **`%>%`** - The magrittr pipe (comes with tidyverse)
2. **`|>`** - The native pipe (built into R 4.1+)

Both accomplish the same goal: passing the result of one function as the first argument to the next function.

```{r}
#| label: setup
library(tidyverse)
```

## How the Pipe Works

### The Basic Concept

The pipe takes the output from the left side and passes it as the first argument to the function on the right side.

```{r}
#| label: basic-pipe

# Without pipe - nested functions (hard to read)
result1 <- round(sqrt(sum(c(1, 4, 9, 16, 25))), 2)

# With pipe - sequential operations (easy to read)
result2 <- c(1, 4, 9, 16, 25) %>%
  sum() %>%
  sqrt() %>%
  round(2)

# Both give the same result
print(result1)
print(result2)
```

### Reading Pipe Chains

Read the pipe as "then":
- Take this data
- **then** do this
- **then** do that
- **then** do another thing

## Why Use the Pipe?

### 1. Improved Readability

Compare these three approaches to the same problem:

```{r}
#| label: readability-comparison

# Create sample data
numbers <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

# Approach 1: Nested functions (hard to read)
nested_result <- mean(sqrt(abs(numbers[numbers > 3])))

# Approach 2: Intermediate variables (verbose)
filtered <- numbers[numbers > 3]
absolute <- abs(filtered)
square_root <- sqrt(absolute)
intermediate_result <- mean(square_root)

# Approach 3: Pipe (clear and concise)
pipe_result <- numbers %>%
  .[. > 3] %>%
  abs() %>%
  sqrt() %>%
  mean()

# All give the same result
print(c(nested_result, intermediate_result, pipe_result))
```

### 2. Natural Workflow

The pipe mirrors how we think about data analysis:

```{r}
#| label: natural-workflow

# Create a dataset
students <- tibble(
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  math = c(85, 92, 78, 95, 88),
  science = c(90, 88, 82, 92, 85),
  english = c(88, 85, 90, 87, 92)
)

# Natural thought process:
# "Take the students data,
#  calculate the average score,
#  filter for high performers,
#  arrange by average score"

result <- students %>%
  mutate(avg_score = (math + science + english) / 3) %>%
  filter(avg_score >= 85) %>%
  arrange(desc(avg_score))

print(result)
```

## Pipe with Data Manipulation

### Using Pipes with dplyr

The pipe works beautifully with dplyr verbs:

```{r}
#| label: pipe-with-dplyr

# Create sample sales data
sales <- tibble(
  date = rep(seq.Date(from = as.Date("2024-01-01"),
                      to = as.Date("2024-01-05"),
                      by = "day"), each = 3),
  store = rep(c("North", "South", "East"), 5),
  sales = round(runif(15, 1000, 5000)),
  returns = round(runif(15, 0, 200))
)

# Complex data pipeline
summary <- sales %>%
  mutate(net_sales = sales - returns) %>%
  group_by(store) %>%
  summarize(
    total_sales = sum(sales),
    total_returns = sum(returns),
    net_revenue = sum(net_sales),
    avg_daily_sales = mean(sales),
    .groups = "drop"
  ) %>%
  mutate(return_rate = total_returns / total_sales * 100) %>%
  arrange(desc(net_revenue))

print(summary)
```

### Piping into Visualization

You can pipe directly into ggplot2:

```{r}
#| label: pipe-to-plot
#| fig-width: 8
#| fig-height: 5

sales %>%
  group_by(date) %>%
  summarize(daily_total = sum(sales), .groups = "drop") %>%
  ggplot(aes(x = date, y = daily_total)) +
  geom_line(color = "steelblue", size = 1.5) +
  geom_point(color = "steelblue", size = 3) +
  theme_minimal() +
  labs(title = "Daily Sales Trend",
       x = "Date",
       y = "Total Sales")
```

## Advanced Pipe Techniques

### 1. Using the Placeholder

The dot (`.`) represents the piped data:

```{r}
#| label: pipe-placeholder

# When you need to use the data in a non-first argument position
numbers <- 1:10

# Using . to specify where the piped data goes
result <- numbers %>%
  lm(formula = . ~ seq_along(.), data = data.frame(. = .)) %>%
  summary()

# More practical example
mtcars %>%
  lm(mpg ~ wt, data = .) %>%
  summary() %>%
  .$r.squared
```

### 2. Piping with Anonymous Functions

For R 4.1+, you can use the new anonymous function syntax:

```{r}
#| label: pipe-anonymous

# Using anonymous functions in pipes
1:10 %>%
  {\(x) x * 2}() %>%
  {\(x) x + 10}() %>%
  mean()

# More practical example with data frame
mtcars %>%
  {\(df) df[df$mpg > 20, ]}() %>%
  nrow()
```

### 3. Side Effects with `%T>%`

The tee pipe (`%T>%`) passes the left-hand side forward while also performing a side effect:

```{r}
#| label: tee-pipe
#| fig-width: 8
#| fig-height: 5

library(magrittr)  # For %T>%

# Create data, plot it, AND continue processing
result <- mtcars %>%
  filter(cyl == 4) %T>%
  {print(paste("Processing", nrow(.), "cars with 4 cylinders"))} %>%
  select(mpg, wt) %T>%
  plot() %>%
  summarize(
    avg_mpg = mean(mpg),
    avg_weight = mean(wt)
  )

print(result)
```

## Common Pipe Patterns

### Pattern 1: Read, Clean, Transform, Visualize

```{r}
#| label: pattern-read-clean
#| eval: false

# Common data analysis workflow
"data.csv" %>%
  read_csv() %>%
  filter(!is.na(important_column)) %>%
  mutate(new_variable = calculation) %>%
  group_by(category) %>%
  summarize(metric = mean(value)) %>%
  ggplot(aes(x = category, y = metric)) +
  geom_col()
```

### Pattern 2: Multiple Transformations

```{r}
#| label: pattern-multiple

# Create example data
transactions <- tibble(
  customer_id = sample(1:100, 500, replace = TRUE),
  amount = round(runif(500, 10, 500), 2),
  category = sample(c("Food", "Electronics", "Clothing", "Other"),
                   500, replace = TRUE),
  date = sample(seq.Date(from = as.Date("2024-01-01"),
                        to = as.Date("2024-03-31"),
                        by = "day"),
               500, replace = TRUE)
)

# Complex transformation pipeline
customer_summary <- transactions %>%
  mutate(month = format(date, "%Y-%m")) %>%
  group_by(customer_id, month, category) %>%
  summarize(
    total_spent = sum(amount),
    n_transactions = n(),
    .groups = "drop"
  ) %>%
  group_by(customer_id) %>%
  mutate(
    pct_of_customer_total = total_spent / sum(total_spent) * 100
  ) %>%
  filter(pct_of_customer_total > 25) %>%
  arrange(customer_id, desc(pct_of_customer_total))

head(customer_summary, 10)
```

## Pipe Best Practices

### 1. Keep Chains Reasonable

```{r}
#| label: reasonable-chains

# Good: Clear, focused pipeline
good_pipeline <- mtcars %>%
  filter(cyl %in% c(4, 6)) %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg))

# Consider breaking very long chains
step1 <- mtcars %>%
  filter(cyl %in% c(4, 6)) %>%
  mutate(efficiency = mpg / wt)

step2 <- step1 %>%
  group_by(cyl) %>%
  summarize(
    avg_mpg = mean(mpg),
    avg_efficiency = mean(efficiency)
  )

final_result <- step2 %>%
  mutate(category = ifelse(avg_mpg > 25, "High", "Medium"))
```

### 2. Use Meaningful Variable Names

```{r}
#| label: meaningful-names

# Bad: Generic names
df <- mtcars %>% filter(mpg > 20)
df2 <- df %>% select(mpg, wt)
result <- df2 %>% summarize(mean(mpg))

# Good: Descriptive names
fuel_efficient_cars <- mtcars %>%
  filter(mpg > 20)

mpg_weight_data <- fuel_efficient_cars %>%
  select(mpg, wt)

average_mpg <- mpg_weight_data %>%
  summarize(mean_mpg = mean(mpg))
```

### 3. Format for Readability

```{r}
#| label: format-readability
#| eval: false

# Bad: Everything on one line
result <- data %>% filter(x > 5) %>% mutate(y = x * 2) %>% group_by(category) %>% summarize(mean = mean(y))

# Good: One verb per line, aligned
result <- data %>%
  filter(x > 5) %>%
  mutate(y = x * 2) %>%
  group_by(category) %>%
  summarize(mean = mean(y))

# Good: With comments for complex operations
result <- data %>%
  # Remove outliers
  filter(x > 5 & x < 100) %>%
  # Create derived variable
  mutate(y = x * 2) %>%
  # Calculate summaries by group
  group_by(category) %>%
  summarize(mean = mean(y))
```

## Common Pitfalls and Solutions

### Pitfall 1: Forgetting Grouping

```{r}
#| label: pitfall-grouping

# Problem: Forgetting that data is still grouped
problem <- mtcars %>%
  group_by(cyl) %>%
  filter(mpg > mean(mpg))  # This filters within groups!

# Solution: Explicitly ungroup when needed
solution <- mtcars %>%
  group_by(cyl) %>%
  filter(mpg > mean(mpg)) %>%
  ungroup()  # Now subsequent operations work on all data
```

### Pitfall 2: Order of Operations

```{r}
#| label: pitfall-order
#| error: true

# This will error - can't use a column before creating it
# mtcars %>%
#   filter(efficiency > 5) %>%
#   mutate(efficiency = mpg / wt)

# Correct order
mtcars %>%
  mutate(efficiency = mpg / wt) %>%
  filter(efficiency > 5) %>%
  head(3)
```

## Exercises

### Exercise 1: Basic Piping
Convert this nested function call to a pipe chain:
```{r}
#| eval: false
round(mean(sqrt(abs(c(-4, -9, -16, -25)))), 2)
```

### Exercise 2: Data Pipeline
Using the built-in `iris` dataset:
1. Filter for flowers with Sepal.Length > 5
2. Calculate the ratio of Petal.Length to Petal.Width
3. Group by Species
4. Find the mean ratio for each species
5. Arrange in descending order

### Exercise 3: Complex Pipeline
Create a pipeline that:
1. Generates 100 random numbers from a normal distribution
2. Keeps only positive values
3. Squares them
4. Takes the top 10 values
5. Calculates their mean

## Summary

The pipe operator fundamentally changes how we write R code:

- **Readability**: Code reads left-to-right, top-to-bottom
- **Debugging**: Easy to run partial pipelines to check intermediate results
- **Modularity**: Each step does one thing
- **Maintainability**: Easy to add, remove, or modify steps

As you continue with the tidyverse, the pipe will become second nature. It's not just a convenience—it's a different way of thinking about data transformation that makes your code more expressive and your intentions clearer.

Next, we'll explore [tibbles](tibbles.qmd), the tidyverse's modern take on data frames!