---
title: "Data Import with readr"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to readr

The readr package provides a fast and friendly way to read rectangular data (like csv, tsv, and fixed width files) into R. It's part of the core tidyverse and automatically returns tibbles instead of data frames.

```{r}
#| label: setup
library(tidyverse)
```

## Why Use readr?

### Advantages Over Base R

1. **Speed**: ~10x faster than base R functions
2. **Tibbles**: Returns tibbles by default
3. **Consistency**: Consistent naming and behavior
4. **Progress bars**: Shows progress for large files
5. **Better defaults**: More sensible type guessing
6. **Encoding support**: Better handling of character encodings

```{r}
#| label: comparison-example
#| eval: false

# Base R
df_base <- read.csv("data.csv")         # Returns data.frame
                                        # Strings become factors (pre R 4.0)
                                        # Slower for large files

# readr
df_readr <- read_csv("data.csv")        # Returns tibble
                                        # Strings stay as characters
                                        # Much faster
                                        # Shows column specifications
```

## Reading CSV Files

### Basic CSV Import

Let's create a sample CSV file to work with:

```{r}
#| label: create-sample-csv

# Create a sample CSV file
sample_data <- tibble(
  id = 1:5,
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  age = c(25, 30, 35, 28, 32),
  salary = c(50000, 60000, 75000, 55000, 65000),
  hire_date = c("2020-01-15", "2019-06-01", "2018-03-20",
                "2021-02-10", "2019-11-30")
)

# Write to CSV
write_csv(sample_data, "employees.csv")

# Read it back
employees <- read_csv("employees.csv")
employees
```

### Understanding Column Specifications

readr automatically detects column types:

```{r}
#| label: column-specs

# readr shows what it detected
employees <- read_csv("employees.csv", show_col_types = TRUE)

# Get the specification
spec(employees)

# You can also explicitly set column types
employees_typed <- read_csv(
  "employees.csv",
  col_types = cols(
    id = col_integer(),
    name = col_character(),
    age = col_integer(),
    salary = col_double(),
    hire_date = col_date(format = "%Y-%m-%d")
  )
)

employees_typed
```

## Column Type Specifications

### Available Column Types

```{r}
#| label: column-types-demo

# Create a file with various data types
diverse_data <- tibble(
  integers = c(1, 2, 3),
  doubles = c(1.5, 2.7, 3.9),
  logicals = c(TRUE, FALSE, TRUE),
  characters = c("apple", "banana", "cherry"),
  dates = as.Date(c("2024-01-01", "2024-02-01", "2024-03-01")),
  times = hms::hms(hours = c(9, 14, 18), minutes = c(30, 15, 45)),
  datetimes = as.POSIXct(c("2024-01-01 09:30:00",
                           "2024-02-01 14:15:00",
                           "2024-03-01 18:45:00"))
)

write_csv(diverse_data, "diverse_data.csv")

# Read with automatic detection
auto_read <- read_csv("diverse_data.csv")
glimpse(auto_read)
```

### Compact Column Specification

Use a compact string specification:

```{r}
#| label: compact-spec

# Use single letters for common types
# i = integer, d = double, c = character, l = logical
# D = date, T = datetime, t = time

data_compact <- read_csv(
  "diverse_data.csv",
  col_types = "iddcDtT"  # One letter per column
)

glimpse(data_compact)
```

## Handling Common Data Issues

### Missing Values

```{r}
#| label: missing-values

# Create data with missing values
missing_data <- "id,value,category
1,10,A
2,NA,B
3,20,NA
4,,C
5,30,D"

# Default: NA and empty strings are treated as missing
df_default <- read_csv(missing_data)
df_default

# Specify custom NA values
df_custom <- read_csv(
  missing_data,
  na = c("", "NA", "N/A", "missing")
)
df_custom
```

### Parsing Problems

```{r}
#| label: parsing-problems

# Data with parsing issues
problematic_data <- "number,text
1,apple
two,banana
3,cherry
4.5,date"

# Read and check for problems
df_problems <- read_csv(problematic_data)

# See what went wrong
problems(df_problems)

# Fix by specifying correct types
df_fixed <- read_csv(
  problematic_data,
  col_types = cols(
    number = col_character(),  # Read as character to preserve all data
    text = col_character()
  )
)
df_fixed
```

### Skip Rows and Comments

```{r}
#| label: skip-rows

# Data with header information
messy_file <- "# Company Sales Report
# Generated: 2024-01-01
# Department: Analytics

id,product,sales
1,Widget A,1000
2,Widget B,1500
# Mid-year update
3,Widget C,2000"

# Skip lines and ignore comments
clean_data <- read_csv(
  messy_file,
  skip = 3,        # Skip first 3 lines
  comment = "#"    # Ignore lines starting with #
)
clean_data
```

## Reading Other File Formats

### TSV Files (Tab-Separated Values)

```{r}
#| label: read-tsv

# Create a TSV file
tsv_data <- tibble(
  gene = c("BRCA1", "TP53", "EGFR"),
  chromosome = c(17, 17, 7),
  gene_function = c("DNA repair", "Tumor suppressor", "Growth factor receptor")
)

write_tsv(tsv_data, "genes.tsv")

# Read TSV
genes <- read_tsv("genes.tsv")
genes
```

### Delimited Files

```{r}
#| label: read-delim

# Create a pipe-delimited file
pipe_data <- "name|age|city
Alice|25|New York
Bob|30|Los Angeles
Charlie|35|Chicago"

write_file(pipe_data, "pipe_data.txt")

# Read with custom delimiter
df_pipe <- read_delim("pipe_data.txt", delim = "|")
df_pipe

# Semicolon-delimited (common in European data)
semi_data <- "product;price;quantity
Apple;1,50;100
Banana;0,75;150
Orange;2,00;80"

write_file(semi_data, "semi_data.csv")

# Read with semicolon delimiter and European decimal mark
df_semi <- read_delim(
  "semi_data.csv",
  delim = ";",
  locale = locale(decimal_mark = ",")
)
df_semi
```

### Fixed Width Files

```{r}
#| label: fixed-width

# Create a fixed width file
fwf_data <- "001John    Smith   25  Engineer
002Jane    Doe     30  Manager
003Bob     Johnson 28  Analyst"

write_file(fwf_data, "fixed_width.txt")

# Read fixed width file with column positions
df_fwf <- read_fwf(
  "fixed_width.txt",
  col_positions = fwf_widths(
    widths = c(3, 8, 8, 4, 10),
    col_names = c("id", "first", "last", "age", "job")
  )
)
df_fwf
```

## Working with Large Files

### Reading in Chunks

```{r}
#| label: chunked-reading

# Create a larger file
large_data <- tibble(
  id = 1:10000,
  value = rnorm(10000),
  category = sample(LETTERS[1:5], 10000, replace = TRUE)
)

write_csv(large_data, "large_file.csv")

# Read in chunks for processing
process_chunk <- function(chunk, pos) {
  # Process each chunk
  summarize(chunk,
            chunk = pos,
            n = n(),
            mean_value = mean(value))
}

# Process file in chunks of 2000 rows
chunk_results <- read_csv_chunked(
  "large_file.csv",
  chunk_size = 2000,
  callback = DataFrameCallback$new(process_chunk)
)

chunk_results
```

### Lazy Reading

```{r}
#| label: lazy-reading
#| eval: false

# For very large files, use lazy reading
# This doesn't load the entire file into memory

library(vroom)  # Alternative package for very large files

# Lazy read - data is only read when needed
lazy_df <- vroom("very_large_file.csv")

# Operations are performed lazily
result <- lazy_df %>%
  filter(category == "A") %>%
  group_by(year) %>%
  summarize(total = sum(value))
```

## Advanced Import Options

### Locale Settings

```{r}
#| label: locale-settings

# European format data
euro_data <- "date;amount;percentage
01/02/2024;1.234,56;45,5%
15/03/2024;2.345,67;32,1%
30/04/2024;3.456,78;28,9%"

write_file(euro_data, "euro_data.csv")

# Read with European locale
euro_df <- read_delim(
  "euro_data.csv",
  delim = ";",
  locale = locale(
    date_format = "%d/%m/%Y",
    decimal_mark = ",",
    grouping_mark = "."
  )
)
euro_df
```

### Column Selection

```{r}
#| label: column-selection

# Create data with many columns
wide_data <- tibble(
  id = 1:5,
  name = c("A", "B", "C", "D", "E"),
  col1 = rnorm(5),
  col2 = rnorm(5),
  col3 = rnorm(5),
  col4 = rnorm(5),
  col5 = rnorm(5)
)

write_csv(wide_data, "wide_data.csv")

# Read only specific columns
selected <- read_csv(
  "wide_data.csv",
  col_select = c(id, name, col1, col3)
)
selected

# Use tidyselect helpers
selected2 <- read_csv(
  "wide_data.csv",
  col_select = c(id, name, starts_with("col"))
)
head(selected2, 2)
```

## Writing Data

### Writing CSV Files

```{r}
#| label: write-csv

# Create sample data
output_data <- tibble(
  id = 1:5,
  value = rnorm(5),
  date = Sys.Date() + 0:4,
  timestamp = Sys.time() + 0:4 * 3600
)

# Write CSV (no row names, readable format)
write_csv(output_data, "output.csv")

# Write with European format
write_csv2(output_data, "output_euro.csv")  # Uses semicolon delimiter

# Write TSV
write_tsv(output_data, "output.tsv")

# Write with custom NA representation
data_with_na <- tibble(
  x = c(1, NA, 3),
  y = c("a", "b", NA)
)

write_csv(data_with_na, "with_na.csv", na = "missing")
```

### Appending to Files

```{r}
#| label: append-files

# Initial data
initial <- tibble(id = 1:3, value = c(10, 20, 30))
write_csv(initial, "append_test.csv")

# New data to append
new_data <- tibble(id = 4:6, value = c(40, 50, 60))

# Append to existing file
write_csv(new_data, "append_test.csv", append = TRUE)

# Read the combined file
read_csv("append_test.csv")
```

## Best Practices

### 1. Always Inspect Your Data

```{r}
#| label: inspect-data

# After reading, always check:
data <- read_csv("employees.csv")

# Structure
glimpse(data)

# First few rows
head(data)

# Summary statistics
summary(data)

# Check for parsing problems
problems(data)
```

### 2. Be Explicit About Types

```{r}
#| label: explicit-types
#| eval: false

# Good: Explicit column types
data <- read_csv(
  "file.csv",
  col_types = cols(
    id = col_integer(),
    amount = col_double(),
    date = col_date(format = "%Y-%m-%d"),
    category = col_factor(levels = c("A", "B", "C"))
  )
)

# This prevents surprises and makes code more robust
```

### 3. Handle Encodings Properly

```{r}
#| label: encoding
#| eval: false

# For files with special characters
data <- read_csv(
  "international_data.csv",
  locale = locale(encoding = "UTF-8")
)

# Common encodings: UTF-8, Latin1, Windows-1252
```

## Common Patterns

### Pattern 1: Import and Clean

```{r}
#| label: import-clean-pattern

# Common workflow: import, clean, transform
clean_data <- read_csv("employees.csv") %>%
  # Remove any rows with missing values
  drop_na() %>%
  # Parse dates properly
  mutate(
    hire_date = as.Date(hire_date),
    years_employed = as.numeric(Sys.Date() - hire_date) / 365.25
  ) %>%
  # Filter and arrange
  filter(age >= 25) %>%
  arrange(desc(salary))

clean_data
```

### Pattern 2: Multiple File Import

```{r}
#| label: multiple-files

# Create multiple files
for(i in 1:3) {
  tibble(
    file_id = i,
    value = rnorm(5),
    category = sample(c("A", "B"), 5, replace = TRUE)
  ) %>%
    write_csv(paste0("data_", i, ".csv"))
}

# Read all files at once
all_files <- list.files(pattern = "data_\\d+\\.csv", full.names = TRUE)

combined_data <- all_files %>%
  map_dfr(read_csv, show_col_types = FALSE) %>%
  arrange(file_id)

combined_data
```

## Exercises

### Exercise 1: Basic Import
1. Create a CSV file with student grades
2. Read it with appropriate column types
3. Check for any parsing problems

### Exercise 2: Messy Data
Create a CSV file with:
- Comment lines starting with #
- Empty rows
- Mixed date formats
- Custom NA values
Then read it correctly using readr options.

### Exercise 3: Large File Simulation
1. Generate a CSV with 100,000 rows
2. Read only specific columns
3. Process it in chunks to calculate group summaries

### Exercise 4: International Data
Create a CSV file with European formatting (semicolon delimiter, comma decimal) and read it properly.

## Cleanup

```{r}
#| label: cleanup
#| echo: false

# Remove created files
files_to_remove <- c(
  "employees.csv", "diverse_data.csv", "genes.tsv",
  "pipe_data.txt", "semi_data.csv", "fixed_width.txt",
  "large_file.csv", "euro_data.csv", "wide_data.csv",
  "output.csv", "output_euro.csv", "output.tsv",
  "with_na.csv", "append_test.csv",
  paste0("data_", 1:3, ".csv")
)

for(file in files_to_remove) {
  if(file.exists(file)) {
    file.remove(file)
  }
}
```

## Summary

The readr package provides a modern, fast, and user-friendly approach to data import:

- **Consistent API**: All read functions work similarly
- **Automatic type detection**: Smart guessing with explicit overrides
- **Better defaults**: Returns tibbles, preserves strings
- **Helpful feedback**: Shows column specs and parsing problems
- **Performance**: Much faster than base R for large files
- **Flexibility**: Handles various formats and encodings

Mastering readr is essential for efficient data import in the tidyverse ecosystem. Combined with dplyr and tidyr, it forms the foundation of most data analysis workflows.

Next, we'll explore [data transformation with dplyr](dplyr-basics.qmd)!