---
title: "Tibbles: Modern Data Frames"
author: "IND218"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## What are Tibbles?

Tibbles are a modern reimagining of data frames, designed to work seamlessly within the tidyverse. While they behave like data frames in most ways, tibbles have several improvements that make them safer and more user-friendly for data analysis.

```{r}
#| label: setup
library(tidyverse)
```

## Creating Tibbles

### From Scratch with `tibble()`

The most common way to create a tibble is with the `tibble()` function:

```{r}
#| label: create-tibble

# Create a tibble
my_tibble <- tibble(
  id = 1:5,
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  score = c(85.5, 92.3, 78.9, 95.1, 88.7),
  passed = score >= 80
)

my_tibble
```

### Converting Data Frames to Tibbles

You can convert existing data frames to tibbles:

```{r}
#| label: convert-to-tibble

# Traditional data frame
df <- data.frame(
  x = 1:3,
  y = c("a", "b", "c")
)

# Convert to tibble
tbl <- as_tibble(df)

# Compare the printing
print(df)
print(tbl)
```

### Using `tribble()` for Row-wise Creation

`tribble()` allows you to create tibbles row-by-row, which is great for small datasets:

```{r}
#| label: tribble

# Create a tibble row by row
grades <- tribble(
  ~student,  ~math, ~science, ~english,
  #---------|------|---------|---------|
  "Alice",      85,       90,       88,
  "Bob",        92,       88,       85,
  "Charlie",    78,       82,       90,
  "Diana",      95,       92,       87
)

grades
```

## Key Differences from Data Frames

### 1. Enhanced Printing

Tibbles have a more informative print method:

```{r}
#| label: enhanced-printing

# Create a larger tibble
large_tibble <- tibble(
  id = 1:1000,
  value = rnorm(1000),
  category = sample(letters[1:5], 1000, replace = TRUE),
  timestamp = Sys.time() + 1:1000
)

# Tibble shows first 10 rows and all columns that fit on screen
large_tibble

# Data frame would print all 1000 rows!
# df <- as.data.frame(large_tibble)
# df  # Don't run this - it would print everything!
```

### 2. Column Types are Preserved

Tibbles don't automatically convert strings to factors:

```{r}
#| label: preserve-types

# Data frame converts strings to factors (in older R versions)
df <- data.frame(
  text = c("apple", "banana", "cherry"),
  value = 1:3
)

# Tibble preserves strings as character vectors
tbl <- tibble(
  text = c("apple", "banana", "cherry"),
  value = 1:3
)

# Check the types
str(df)
str(tbl)
```

### 3. Stricter Subsetting

Tibbles are more strict about subsetting:

```{r}
#| label: strict-subsetting
#| error: true

# Create a tibble
tbl <- tibble(x = 1:3, y = 4:6)

# Single bracket always returns a tibble
tbl[1]       # Returns a tibble with one column
tbl["x"]     # Returns a tibble with column x

# Double bracket and $ extract the column
tbl[[1]]     # Returns the vector
tbl$x        # Returns the vector
tbl[["x"]]   # Returns the vector

# Tibbles don't do partial matching
# tbl$xy     # This would error - no partial matching!
```

## Working with Tibbles

### Adding and Modifying Columns

```{r}
#| label: modify-columns

# Start with a basic tibble
students <- tibble(
  name = c("Alice", "Bob", "Charlie"),
  midterm = c(85, 78, 92),
  final = c(88, 82, 90)
)

# Add columns with mutate
students <- students %>%
  mutate(
    average = (midterm + final) / 2,
    grade = case_when(
      average >= 90 ~ "A",
      average >= 80 ~ "B",
      average >= 70 ~ "C",
      TRUE ~ "D"
    )
  )

students
```

### Accessing Tibble Information

```{r}
#| label: tibble-info

# Create a sample tibble
data <- tibble(
  id = 1:100,
  group = sample(LETTERS[1:4], 100, replace = TRUE),
  value = rnorm(100, mean = 50, sd = 10)
)

# Get dimensions
nrow(data)
ncol(data)
dim(data)

# Get column names
names(data)
colnames(data)

# Check if it's a tibble
is_tibble(data)
is.data.frame(data)  # TRUE - tibbles are also data frames

# Get a glimpse of the structure
glimpse(data)
```

## Advanced Tibble Features

### List Columns

Tibbles can contain list columns, allowing complex data structures:

```{r}
#| label: list-columns

# Create a tibble with list columns
complex_data <- tibble(
  id = 1:3,
  name = c("Experiment A", "Experiment B", "Experiment C"),
  measurements = list(
    c(1.2, 1.5, 1.3),
    c(2.1, 2.3, 2.2, 2.4),
    c(3.1, 3.0)
  ),
  metadata = list(
    list(date = "2024-01-01", operator = "Alice"),
    list(date = "2024-01-02", operator = "Bob"),
    list(date = "2024-01-03", operator = "Charlie")
  )
)

complex_data

# Access list column elements
complex_data$measurements[[1]]
complex_data$metadata[[2]]$operator
```

### Nested Tibbles

You can nest tibbles within tibbles:

```{r}
#| label: nested-tibbles

# Create nested data
sales_data <- tibble(
  region = c("North", "South", "East", "West"),
  data = list(
    tibble(month = 1:3, sales = c(100, 120, 110)),
    tibble(month = 1:3, sales = c(80, 85, 90)),
    tibble(month = 1:3, sales = c(95, 100, 105)),
    tibble(month = 1:3, sales = c(110, 115, 125))
  )
)

sales_data

# Access nested tibble
sales_data$data[[1]]

# Work with nested data using purrr
sales_data %>%
  mutate(total_sales = map_dbl(data, ~sum(.x$sales)))
```

### Tibble Validation

Tibbles perform validation to ensure data integrity:

```{r}
#| label: validation
#| error: true

# Recycling rules are stricter
# This works - length 1 recycles
tibble(x = 1:4, y = 1)

# This errors - inconsistent lengths
# tibble(x = 1:4, y = 1:3)  # Error!

# Column names must be unique
# tibble(x = 1, x = 2)  # Error!

# All columns must have the same length (or length 1)
tibble(
  x = 1:3,
  y = 1,      # Length 1 - OK, will be recycled
  z = 4:6     # Length 3 - OK
)
```

## Tibbles in Data Analysis Workflows

### Example: Customer Analysis Pipeline

```{r}
#| label: analysis-workflow

# Create customer data
customers <- tibble(
  customer_id = 1:20,
  age = sample(18:65, 20, replace = TRUE),
  region = sample(c("North", "South", "East", "West"), 20, replace = TRUE),
  purchases = sample(0:50, 20, replace = TRUE),
  member_since = sample(2018:2024, 20, replace = TRUE)
)

# Analysis pipeline
customer_summary <- customers %>%
  mutate(
    years_member = 2024 - member_since,
    age_group = case_when(
      age < 30 ~ "Young",
      age < 50 ~ "Middle",
      TRUE ~ "Senior"
    )
  ) %>%
  group_by(region, age_group) %>%
  summarize(
    n_customers = n(),
    avg_purchases = mean(purchases),
    total_purchases = sum(purchases),
    avg_tenure = mean(years_member),
    .groups = "drop"
  ) %>%
  arrange(desc(total_purchases))

customer_summary
```

### Combining Multiple Tibbles

```{r}
#| label: combining-tibbles

# Product information
products <- tibble(
  product_id = 1:5,
  product_name = c("Widget A", "Widget B", "Gadget C", "Gadget D", "Tool E"),
  price = c(19.99, 29.99, 49.99, 39.99, 24.99)
)

# Sales records
sales <- tibble(
  sale_id = 1:10,
  product_id = sample(1:5, 10, replace = TRUE),
  quantity = sample(1:5, 10, replace = TRUE),
  date = seq.Date(from = as.Date("2024-01-01"),
                  length.out = 10,
                  by = "day")
)

# Join tibbles
sales_details <- sales %>%
  left_join(products, by = "product_id") %>%
  mutate(revenue = quantity * price) %>%
  select(sale_id, date, product_name, quantity, price, revenue)

sales_details
```

## Best Practices with Tibbles

### 1. Use Consistent Naming

```{r}
#| label: naming-practices

# Good: Consistent, descriptive names
good_tibble <- tibble(
  customer_id = 1:3,
  customer_name = c("Alice", "Bob", "Charlie"),
  customer_age = c(25, 30, 35),
  purchase_amount = c(100, 200, 150)
)

# Better: Even more consistent with snake_case
better_tibble <- tibble(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  amount_purchased = c(100, 200, 150)
)
```

### 2. Document Complex Structures

```{r}
#| label: document-structures

# Document what each column represents
experiment_data <- tibble(
  # Unique identifier for each experiment
  exp_id = 1:5,

  # Timestamp when experiment started (UTC)
  start_time = Sys.time() + 0:4 * 3600,

  # List of measurements taken during experiment
  measurements = list(
    rnorm(10), rnorm(12), rnorm(8), rnorm(15), rnorm(10)
  ),

  # Nested metadata about experimental conditions
  conditions = list(
    list(temp = 20, pressure = 1.0),
    list(temp = 25, pressure = 1.0),
    list(temp = 20, pressure = 1.5),
    list(temp = 25, pressure = 1.5),
    list(temp = 22.5, pressure = 1.25)
  )
)

glimpse(experiment_data)
```

### 3. Prefer Tibbles for New Code

```{r}
#| label: prefer-tibbles

# When reading data, get tibbles directly
# read_csv() returns a tibble
# read.csv() returns a data frame

# When creating data, use tibble() not data.frame()
modern_approach <- tibble(
  x = 1:3,
  y = c("a", "b", "c")
)

# When converting existing data frames
legacy_df <- data.frame(x = 1:3, y = 4:6)
modernized <- as_tibble(legacy_df)
```

## Common Operations Reference

```{r}
#| label: operations-reference

# Create a reference tibble
ref <- tibble(
  id = 1:5,
  value = c(10, 20, 30, 40, 50),
  category = c("A", "B", "A", "B", "A")
)

# Selection operations
ref %>% select(id, value)              # Select columns
ref %>% select(-category)              # Drop columns
ref %>% filter(value > 20)             # Filter rows
ref %>% slice(1:3)                     # Select rows by position

# Modification operations
ref %>% mutate(double = value * 2)     # Add column
ref %>% rename(val = value)            # Rename column
ref %>% arrange(desc(value))           # Sort rows

# Aggregation operations
ref %>%
  group_by(category) %>%
  summarize(
    mean_value = mean(value),
    total = sum(value),
    count = n()
  )
```

## Exercises

### Exercise 1: Create and Explore
Create a tibble with information about 5 books (title, author, year, pages). Print it and use `glimpse()` to examine its structure.

### Exercise 2: List Columns
Create a tibble where each row represents a student and includes:
- Name (character)
- Grades (list of numeric scores)
- Subjects (list of course names)

### Exercise 3: Tibble vs Data Frame
Create the same dataset as both a data frame and a tibble. Compare:
- How they print
- How they handle column access
- What happens with partial name matching

### Exercise 4: Complex Analysis
Using the built-in `diamonds` dataset (convert to tibble first):
1. Group by cut and color
2. Calculate average price and carat
3. Create a list column with all clarity values per group
4. Find which combination has the highest average price

## Summary

Tibbles are the foundation of tidy data analysis:

- **Safer defaults**: No automatic type conversion, no partial matching
- **Better printing**: Shows only what fits, includes type information
- **Modern features**: List columns, nested data structures
- **Consistent behavior**: Predictable subsetting and modification
- **Tidyverse integration**: Works seamlessly with all tidyverse functions

As you work more with the tidyverse, tibbles will become your default data structure. They eliminate many of the surprises and inconsistencies of traditional data frames while adding powerful new capabilities.

Next, we'll learn about [importing data with readr](readr-import.qmd) to get your data into tibble format!