---
title: "Data Tidying Principles with tidyr"
author: "IND215"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
---

## Introduction to tidyr

tidyr helps you create **tidy data**, where:

1. Every column is a variable
2. Every row is an observation
3. Every cell is a single value

Most real-world data violates these principles. tidyr provides tools to reshape data into tidy format, making it easier to work with dplyr and ggplot2.

```{r}
#| label: setup
library(tidyverse)

# Set seed for reproducibility
set.seed(123)
```

## What is Tidy Data?

### The Tidy Data Principles

Let's see the difference between messy and tidy data:

```{r}
#| label: tidy-principles

# Messy data: Multiple variables in column names
messy_data <- tibble(
  country = c("USA", "UK", "Japan"),
  `2020_population` = c(331, 67, 126),
  `2020_gdp` = c(21433, 2708, 5065),
  `2021_population` = c(332, 68, 125),
  `2021_gdp` = c(22996, 3131, 4941)
)

print(messy_data)

# Tidy version: Each variable in its own column
tidy_data <- tibble(
  country = rep(c("USA", "UK", "Japan"), each = 2),
  year = rep(c(2020, 2021), 3),
  population = c(331, 332, 67, 68, 126, 125),
  gdp = c(21433, 22996, 2708, 3131, 5065, 4941)
)

print(tidy_data)
```

## Pivoting: Reshaping Data

### pivot_longer(): From Wide to Long

`pivot_longer()` makes datasets longer by increasing rows and decreasing columns:

```{r}
#| label: pivot-longer-basic

# Wide format data
wide_sales <- tibble(
  product = c("Widget A", "Widget B", "Widget C"),
  jan = c(100, 150, 120),
  feb = c(110, 140, 130),
  mar = c(120, 160, 140),
  apr = c(130, 155, 135)
)

print(wide_sales)

# Convert to long format
long_sales <- wide_sales %>%
  pivot_longer(
    cols = jan:apr,           # Columns to pivot
    names_to = "month",        # Name for new column from column names
    values_to = "sales"        # Name for new column from values
  )

print(long_sales)
```

### Advanced pivot_longer()

```{r}
#| label: pivot-longer-advanced

# Data with multiple variables in column names
complex_wide <- tibble(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie"),
  math_2022 = c(85, 78, 92),
  math_2023 = c(88, 82, 90),
  science_2022 = c(90, 85, 88),
  science_2023 = c(92, 87, 91)
)

print(complex_wide)

# Pivot with name separation
complex_long <- complex_wide %>%
  pivot_longer(
    cols = -c(id, name),                    # Everything except id and name
    names_to = c("subject", "year"),        # Split column names into two variables
    names_sep = "_",                        # Separator in column names
    values_to = "score"
  ) %>%
  mutate(year = as.integer(year))           # Convert year to integer

print(complex_long)
```

### pivot_wider(): From Long to Wide

`pivot_wider()` makes datasets wider by increasing columns and decreasing rows:

```{r}
#| label: pivot-wider-basic

# Start with long format
long_data <- tibble(
  date = rep(c("2024-01-01", "2024-01-02"), each = 3),
  metric = rep(c("views", "clicks", "conversions"), 2),
  value = c(1000, 50, 5, 1200, 65, 8)
)

print(long_data)

# Convert to wide format
wide_data <- long_data %>%
  pivot_wider(
    names_from = metric,       # Column to get new column names from
    values_from = value        # Column to get values from
  )

print(wide_data)
```

### Advanced pivot_wider()

```{r}
#| label: pivot-wider-advanced

# Data with multiple value columns
multi_long <- tibble(
  year = rep(c(2022, 2023), each = 6),
  quarter = rep(c("Q1", "Q2"), 6),
  region = rep(c("North", "South", "East"), 4),
  sales = round(runif(12, 100, 500)),
  costs = round(runif(12, 50, 200))
)

print(multi_long)

# Pivot with multiple value columns
multi_wide <- multi_long %>%
  pivot_wider(
    names_from = c(quarter, region),    # Multiple columns for names
    names_sep = "_",                    # Separator for new column names
    values_from = c(sales, costs)       # Multiple value columns
  )

print(multi_wide)
```

## Separating and Uniting Columns

### separate(): Split One Column into Multiple

```{r}
#| label: separate

# Data with combined information
combined_data <- tibble(
  employee = c("John_Smith", "Jane_Doe", "Bob_Johnson"),
  dept_role = c("Sales_Manager", "IT_Developer", "HR_Director"),
  date_time = c("2024-01-15_09:30", "2024-01-15_10:00", "2024-01-15_14:30")
)

print(combined_data)

# Separate into multiple columns
separated_data <- combined_data %>%
  separate(employee, into = c("first", "last"), sep = "_") %>%
  separate(dept_role, into = c("department", "role"), sep = "_") %>%
  separate(date_time, into = c("date", "time"), sep = "_")

print(separated_data)
```

### separate_rows(): Split and Expand Rows

```{r}
#| label: separate-rows

# Data with multiple values in cells
multi_value_data <- tibble(
  product = c("Widget A", "Widget B", "Widget C"),
  colors = c("red,blue", "green,yellow,orange", "black,white"),
  sizes = c("S,M,L", "M,L", "S,M,L,XL")
)

print(multi_value_data)

# Separate into multiple rows
expanded_data <- multi_value_data %>%
  separate_rows(colors, sep = ",") %>%
  separate_rows(sizes, sep = ",")

print(expanded_data)
```

### unite(): Combine Multiple Columns

```{r}
#| label: unite

# Data with separate columns
separate_data <- tibble(
  year = c(2024, 2024, 2024),
  month = c(1, 2, 3),
  day = c(15, 20, 25),
  city = c("New York", "Los Angeles", "Chicago"),
  state = c("NY", "CA", "IL")
)

print(separate_data)

# Unite columns
united_data <- separate_data %>%
  unite(date, year, month, day, sep = "-") %>%
  unite(location, city, state, sep = ", ")

print(united_data)
```

## Handling Missing Values

### complete(): Fill in Missing Combinations

```{r}
#| label: complete

# Data with missing combinations
incomplete_data <- tibble(
  year = c(2022, 2022, 2023),
  quarter = c("Q1", "Q3", "Q2"),
  sales = c(100, 150, 120)
)

print(incomplete_data)

# Complete all combinations
complete_data <- incomplete_data %>%
  complete(year, quarter, fill = list(sales = 0))

print(complete_data)

# Complete with sequences
sales_data <- tibble(
  date = as.Date(c("2024-01-01", "2024-01-03", "2024-01-07")),
  sales = c(100, 150, 200)
)

complete_sales <- sales_data %>%
  complete(
    date = seq.Date(min(date), max(date), by = "day"),
    fill = list(sales = 0)
  )

print(complete_sales)
```

### drop_na(): Remove Rows with Missing Values

```{r}
#| label: drop-na

# Data with missing values
data_with_na <- tibble(
  id = 1:6,
  x = c(1, 2, NA, 4, NA, 6),
  y = c(NA, 2, 3, 4, 5, 6),
  z = c(1, 2, 3, NA, 5, 6)
)

print(data_with_na)

# Drop rows with any NA
data_with_na %>% drop_na()

# Drop rows with NA in specific columns
data_with_na %>% drop_na(x)

# Drop rows with NA in multiple specific columns
data_with_na %>% drop_na(x, y)
```

### fill(): Fill Missing Values

```{r}
#| label: fill

# Time series with gaps
time_series <- tibble(
  date = seq.Date(from = as.Date("2024-01-01"),
                  to = as.Date("2024-01-10"),
                  by = "day"),
  value = c(10, NA, NA, 15, NA, 20, NA, NA, 25, 30)
)

print(time_series)

# Fill down (carry forward)
time_series %>%
  fill(value, .direction = "down")

# Fill up (carry backward)
time_series %>%
  fill(value, .direction = "up")

# Fill both directions
time_series %>%
  fill(value, .direction = "downup")
```

### replace_na(): Replace NA with Specific Values

```{r}
#| label: replace-na

# Data with NAs
na_data <- tibble(
  product = c("A", "B", "C", "D"),
  price = c(10, NA, 15, 20),
  discount = c(0.1, 0.2, NA, NA),
  category = c("Food", NA, "Electronics", NA)
)

print(na_data)

# Replace NAs with specific values
na_data %>%
  replace_na(list(
    price = 0,
    discount = 0,
    category = "Other"
  ))
```

## Nesting and Unnesting

### nest(): Create List-Columns

```{r}
#| label: nest

# Grouped data
student_scores <- tibble(
  student = rep(c("Alice", "Bob", "Charlie"), each = 4),
  subject = rep(c("Math", "Science", "English", "History"), 3),
  score = c(85, 90, 88, 92,   # Alice
           78, 85, 82, 88,    # Bob
           92, 88, 90, 95)    # Charlie
)

print(student_scores)

# Nest by student
nested_scores <- student_scores %>%
  nest(data = c(subject, score))

print(nested_scores)

# Access nested data
nested_scores$data[[1]]
```

### unnest(): Expand List-Columns

```{r}
#| label: unnest

# Start with nested data
nested_data <- tibble(
  group = c("A", "B"),
  data = list(
    tibble(x = 1:3, y = 4:6),
    tibble(x = 7:9, y = 10:12)
  )
)

print(nested_data)

# Unnest the data
unnested_data <- nested_data %>%
  unnest(data)

print(unnested_data)
```

## Real-World Examples

### Example 1: Survey Data Cleaning

```{r}
#| label: survey-example

# Messy survey data
survey_raw <- tibble(
  respondent_id = 1:5,
  q1_2023 = c(5, 4, 3, 5, 4),
  q2_2023 = c(4, 4, 3, 4, 5),
  q1_2024 = c(5, 5, 4, 5, 4),
  q2_2024 = c(5, 4, 4, 5, 5),
  demographics = c("25-34,Male,Urban",
                  "35-44,Female,Rural",
                  "18-24,Male,Urban",
                  "45-54,Female,Suburban",
                  "25-34,Male,Rural")
)

# Clean and tidy the data
survey_clean <- survey_raw %>%
  # Separate demographics
  separate(demographics, into = c("age_group", "gender", "location"),
          sep = ",") %>%
  # Pivot questions longer
  pivot_longer(
    cols = starts_with("q"),
    names_to = c("question", "year"),
    names_sep = "_",
    values_to = "rating"
  ) %>%
  # Convert year to integer
  mutate(year = as.integer(year))

print(survey_clean)

# Analyze by demographics
survey_summary <- survey_clean %>%
  group_by(year, gender, question) %>%
  summarize(
    avg_rating = mean(rating),
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = question,
    values_from = avg_rating
  )

print(survey_summary)
```

### Example 2: Financial Data Transformation

```{r}
#| label: financial-example

# Quarterly financial data in wide format
financial_wide <- tibble(
  company = c("TechCorp", "RetailCo", "FinanceInc"),
  revenue_q1_2023 = c(100, 80, 120),
  revenue_q2_2023 = c(110, 85, 125),
  revenue_q3_2023 = c(120, 90, 130),
  revenue_q4_2023 = c(130, 95, 140),
  costs_q1_2023 = c(60, 50, 70),
  costs_q2_2023 = c(65, 52, 72),
  costs_q3_2023 = c(68, 54, 75),
  costs_q4_2023 = c(70, 56, 78)
)

# Transform to tidy format
financial_tidy <- financial_wide %>%
  pivot_longer(
    cols = -company,
    names_to = c("metric", "quarter", "year"),
    names_sep = "_",
    values_to = "value"
  ) %>%
  mutate(
    year = as.integer(year),
    quarter = str_to_upper(quarter)
  ) %>%
  pivot_wider(
    names_from = metric,
    values_from = value
  ) %>%
  mutate(
    profit = revenue - costs,
    margin = profit / revenue * 100
  )

print(financial_tidy)

# Quarterly summary
quarterly_summary <- financial_tidy %>%
  group_by(quarter) %>%
  summarize(
    avg_revenue = mean(revenue),
    avg_costs = mean(costs),
    avg_margin = mean(margin),
    .groups = "drop"
  ) %>%
  arrange(quarter)

print(quarterly_summary)
```

### Example 3: Time Series Data Preparation

```{r}
#| label: timeseries-example

# Irregular time series data
irregular_ts <- tibble(
  date = as.Date(c("2024-01-01", "2024-01-03", "2024-01-04",
                  "2024-01-08", "2024-01-10")),
  sensor_a = c(10, 12, NA, 15, 18),
  sensor_b = c(20, NA, 23, 24, 25),
  sensor_c = c(30, 32, 33, NA, 36)
)

# Complete and clean the time series
regular_ts <- irregular_ts %>%
  # Complete all dates
  complete(
    date = seq.Date(min(date), max(date), by = "day")
  ) %>%
  # Fill missing values
  fill(everything(), .direction = "downup") %>%
  # Pivot to long format for analysis
  pivot_longer(
    cols = starts_with("sensor"),
    names_to = "sensor",
    values_to = "reading"
  ) %>%
  # Add derived features
  mutate(
    day_of_week = weekdays(date),
    is_weekend = day_of_week %in% c("Saturday", "Sunday")
  )

print(regular_ts)

# Daily summary across sensors
daily_summary <- regular_ts %>%
  group_by(date, is_weekend) %>%
  summarize(
    mean_reading = mean(reading, na.rm = TRUE),
    sd_reading = sd(reading, na.rm = TRUE),
    .groups = "drop"
  )

print(daily_summary)
```

## Best Practices

### 1. Start with Tidy Data in Mind

```{r}
#| label: best-practice-tidy
#| eval: false

# Design your data collection to be tidy from the start
# Good: Each observation is a row
good_design <- tibble(
  date = rep(dates, each = n_products),
  product = rep(products, n_dates),
  metric = value
)

# Avoid: Wide format with dates/products as columns
avoid_design <- tibble(
  product = products,
  `2024-01-01` = values1,
  `2024-01-02` = values2
)
```

### 2. Use Meaningful Column Names

```{r}
#| label: best-practice-names

# Good: Clear, descriptive names
good_names <- tibble(
  measurement_date = Sys.Date(),
  temperature_celsius = 25,
  humidity_percent = 60
)

# Better: Consistent naming convention
better_names <- tibble(
  date_measured = Sys.Date(),
  temp_c = 25,
  humid_pct = 60
)
```

### 3. Document Your Transformations

```{r}
#| label: best-practice-document

# Create sample data for documentation example
raw_data <- tibble(
  product_code = c("A-2023", "B-2023", "C-2024"),
  jan = c(100, NA, 150),
  feb = c(110, 120, NA),
  mar = c(120, 130, 170)
)

# Document each step
documented_pipeline <- raw_data %>%
  # Step 1: Convert wide sales data to long format
  pivot_longer(
    cols = jan:mar,
    names_to = "month",
    values_to = "sales"
  ) %>%
  # Step 2: Separate year from product code
  separate(
    product_code,
    into = c("category", "year"),
    sep = "-"
  ) %>%
  # Step 3: Fill missing sales with 0
  replace_na(list(sales = 0))

print(documented_pipeline)
```

## Common Patterns

### Pattern 1: Wide to Long to Summary

```{r}
#| label: pattern-wls

# Common analysis pattern
wide_data <- tibble(
  group = c("A", "B"),
  day1 = c(10, 20),
  day2 = c(15, 22),
  day3 = c(12, 25)
)

summary <- wide_data %>%
  pivot_longer(cols = starts_with("day"),
              names_to = "day",
              values_to = "value") %>%
  group_by(group) %>%
  summarize(
    mean = mean(value),
    sd = sd(value),
    .groups = "drop"
  )

print(summary)
```

### Pattern 2: Separate, Clean, Unite

```{r}
#| label: pattern-scu

# Clean messy identifiers
messy_ids <- tibble(
  raw_id = c("USA_2024_Q1", "CAN_2024_Q2", "MEX_2024_Q1"),
  value = c(100, 150, 120)
)

clean_ids <- messy_ids %>%
  separate(raw_id, into = c("country", "year", "quarter"), sep = "_") %>%
  mutate(
    year = as.integer(year),
    quarter = factor(quarter, levels = c("Q1", "Q2", "Q3", "Q4"))
  ) %>%
  unite(period, year, quarter, sep = "-", remove = FALSE)

print(clean_ids)
```

## Exercises

### Exercise 1: Basic Pivoting
Create a dataset of monthly temperatures for 3 cities across 4 months. Start in wide format, then:
1. Pivot to long format
2. Calculate average temperature per city
3. Pivot back to wide with cities as columns

### Exercise 2: Data Cleaning Pipeline
Given messy survey data with:
- Combined name fields ("FirstName LastName")
- Multiple responses in single cells ("option1;option2;option3")
- Missing value combinations

Clean the data to create a tidy dataset ready for analysis.

### Exercise 3: Complex Reshaping
Create a sales dataset with:
- Multiple products
- Multiple regions
- Quarterly data for 2 years

Practice pivoting between different representations for different types of analysis.

## Summary

tidyr provides essential tools for data tidying:

- **`pivot_longer()`/`pivot_wider()`**: Reshape between wide and long formats
- **`separate()`/`unite()`**: Split and combine columns
- **`complete()`**: Fill in missing combinations
- **`drop_na()`/`fill()`/`replace_na()`**: Handle missing values
- **`nest()`/`unnest()`**: Create and expand list-columns

These functions, combined with dplyr, form the foundation of data manipulation in the tidyverse. Tidy data makes subsequent analysis much easier and more intuitive.

The key insight: spend time getting your data tidy, and the rest of your analysis will be much simpler!